### 2014-10-15 AdamRebuttal.txt
stuff

Reviewer A:

We want to emphasize why the result about sequential consistency is important.
Here we use "sequential consistency" as a name for the LTS in Figure 3, which
actually completely specifies the observable behavior of a full multiprocessor
system.  Thus, by proving a relationship with that LTS, we are proving what is
perhaps the most natural version of "full functional correctness" in this
domain.

Regarding store atomicity and weak memory models: we define store atomicity as
a property of main memory only, which is quite complex in its own right, with
hierarchies of caches, etc.  There are additional cache-like elements attached
to the individual processors, called by names like "store buffer".  If we
consider these extra buffers as part of the memory, then you're certainly right
that weak memory models don't guarantee atomic stores. However, weak-memory
systems can be built on top of the same store-atomic main memory interface we
give in the paper; it's only the processor-local store buffers that introduce
weaker behavior.

We agree, though, that we should include less speculation on adaptation to weak
memory; and thanks for the bug reports on our discussion of related work.

Reviewer B:

We agree that the underlying LTS proof techniques are not new.  We view our
contribution as more of demonstrating a new application in an area that badly
needs techniques at this level of modularity.

For both Reviewers B and C:

Sorry for not explaining more of our rationale for considering these LTSes as
practical hardware descriptions.  The cache-coherence algorithm, for instance,
is directly transliterated out of [fill in self-contained paper references
here].  We would like to have an automatic translation from Bluespec code, but
we think the two abstraction levels are close enough that it is reasonable to
draw conclusions from our manual translations.

Reviewer C:

Our main theorem uses termination as an arbitrary property that is sufficient
to make the theorem interesting but still easily stated. That's for the same
reason that many problems in computability theory are equivalent to the halting
problem in suitable senses.

Reviewer D:

We apologize for not getting across the sense of the main theorem. Our
optimized architecture actually does fulfill the SC spec, without any side
conditions on programs!  The explanation of why is not trivial, and the proof
carried out throughout the paper justifies the claim.

We also view simplicity of the proof outline (at the level given in the paper)
as a feature, not a bug, of the approach. The actual proof in Coq is much
longer and more detailed (about 10,000 lines).

### 2014-10-15 ArvindRebuttalLatest.txt
stuff

We are thankful to the reviewers for their valuable and in-depth comments. We
hope the following clarifications would make them more enthusiastic about
advocating its publication in POPL.

1. Store-atomicity is ubiquitous: Even though systems implement memory models
weaker than sequential consistency (SC), real multiprocessor systems can be
decomposed into multiple parts, one of which is an atomic-memory. In the
simplest form, atomic memory is just a monolithic multiported global memory
(GM), but more commonly, it is implemented using a hierarchy of coherent
caches. Weak memory model behaviors can be understood in terms of local storage
(LS) between each processor and GM. If the processor memory interface is drawn
such that LS is part of GM, then GM will not behave like an atomic store. By
making LS part of the processor model, we assert that all the nuances of the
known weak memory models can be modeled as part of the processor description.
We address the two examples cited by Reviewer A at the end. The proof of
correctness of cache coherence is completely orthogonal to the semantics of the
processor side and hence the memory model issues.

2. Choice of SC: The processor description in Figure 7 implements SC. This
description will have to be modified for weaker memory models, but the
correctness criteria for ROB in Figure 6 will remain the same. For example, one
may not even issue a verification load or issue it sooner. We acknowledge that
more work needs to be done to formally model weaker memory models, including a
treatment of fences.

3. SC performance issues: In the current architecture literature, it's hotly
debated whether weaker memory models can provide better performance than
properly implemented SC. Our SC processor in Figure 7 has high performance
because it is not restricted to sustain only a fixed number of (speculative)
loads in flight. Furthermore, instructions dependent upon these loads can be
processed out of order. A weaker memory model cannot improve upon this
important aspect. The main restriction in our model is that a processor can
issue only one store or one verification load which may be improvable by a
weaker model.

4. LTS for describing and synthesizing hardware:  A Bluespec program is a
transition system and experience has shown that Bluespec descriptions can be
both concise and generate efficient hardware. The cache coherence protocol of
Figure 10 was initially implemented and synthesized into efficient hardware
from Bluespec (Khan et al ISPASS-2012, see also Dave et al Memocode 2006). A
Bluespec program can be transliterated into an LTS in Coq; currently, this is
done manually.

The processor LTS in Figure 7 has an abstract specification of the ROB and
branch predictor with a clearly stated correctness criteria. To derive the
actual hardware, the LTSes for these components have to be specified in much
greater detail. Our ISA is general enough to encompass any CISC ISA including
x86. However the complexity of the detailed LTSes for ROB etc will depend upon
the specific ISA.

5. Reusability of our Cache Coherence proof: MOSI, MOESI, Firefly etc. are very
similar to MSI protocol, though their transition systems will be different. The
way our proof is organized, we will have to prove Lemmas 5 and 6, and
Invariants 2, 3, and 4 for each of these protocols; the rest of the proof will
remain unchanged. In the current version, these lemmas and invariants account
for x% of the proof. Also all these lemmas deal with less than 4 states and
thus can be proven easily using model checking.

------- 
Examples For Reviewer A

For explaining Reviewer A's IRIW behavior, consider a system containing several
in-order cores, each core connected to what we call as ''non-coherent cache''
NCC. All NCCs are connected to a global atomic memory. An NCC can read any
location from memory and store it, and the core can use the data from NCC, if
present. But stores from the core always happen to the global memory, and the
NCC can discard the location that it has, anytime. So, in the example,
initially, NCC of core 4 reads in the location x. Now the sequence of
operations is
Core 1: Wx1 (into global memory)
Core 2: Rx (gets 1 from global memory)
        Ry (gets 0 from global memory, initial value)
Core 3: Wy1 (into global memory)
Core 4: Ry (gets 1 from global memory)
        Rx (gets 0 from NCC)

Similarly, the second example about store buffers can be explained using a
store buffer instead of the NCC, into which stores are written, and loads can
read from. The sequence for this to explain the behavior is:
Core 1: Wx1 (into store buffer, not to memory)
        Ry (read from memory 0)
Core 2: Wy1 (into store buffer, not to memory)
        Rx (read from memory 0)

This global memory is, in reality, implemented using cache coherent protocols.
The separation of coherent caches implementing atomic memory semantics and the
rest of the system is an important contribution of this paper; a version of
this was presented in Arvind (Memory Model = Instruction Reordering + Store
Atomicity, ISCA 2006), and this paper gives a formal proof of this separation,
while extending this idea to speculating systems.

### 2014-10-15 rebuttal.txt
stuff

Store-atomicity vs weak memory-models:
We define store-atomicity as a property of just the memory-subsystem, which is
quite complex in its own right, with hierarchies of caches.  There are
additional cache-like elements attached to individual processors, called by
names like "store-buffer".  If we consider these extra buffers as part of the
memory, then Reviewer-A is correct in saying the weak memory-models don't
guarantee atomic stores. However, weak-memory systems are built on top of the
same store-atomic main memory interface defined in LTS of Figure-5; it's only
the processor-local buffers that introduce weaker behavior (we address
Reviewer-A's examples at the end). Proving that the cache-LTS (Figure-10)
preserves store-atomicity (Figure-5) is a major contribution of this paper. We
agree, however, that extension to weaker memory-models needs work.

Choice of SC:
We define "sequential-consistency" as the LTS in Figure-3, which completely
specifies the observable behavior of a full multiprocessor system, including
termination. Thus, by proving a relationship with that LTS, we are proving what
is perhaps the most natural version of "full functional correctness" in this
domain.

Main result (Theorem-10):
We apologize for not getting across the sense of the main theorem. A system
with our speculative processors and store-atomic memory actually does fulfill
the SC spec, without any side conditions on programs or hidden assumptions! We
think the proof is nontrivial. We view simplicity of our proof outline as a
feature, not a bug. The actual proof in Coq is about 10000 lines.

Proof reusability:
MOSI, MOESI, etc. are very similar to MSI protocol. The way our proof is
organized, we will have to just prove Lemmas 5, 6, and Invariants 2, 3, 4 for
each of these protocols. These lemmas deal with less than 4 states and thus can
also be proven easily using model-checking.

LTS specifications for hardware:
Sorry for not explaining our rationale for considering LTSes as practical
hardware descriptions.  The cache-LTS (Figure-10), for instance, is a direct
transliteration from Bluespec-code [Khan]. See also [Dave].  We plan on
automating the translation from Bluespec-code, but we think the two
abstractions are close enough that it is reasonable to draw conclusions from
manual translations.

Processor Specification:
The processor-LTS (Figure-7) is intentionally abstract to capture a superset of
microarchitectures; eg, we do not give the actual transitions for a reorder
buffer implementation but instead give its correctness criteria (Figure-8).
Our ISA is abstract enough to include CISC-instructions and if desired can be
extended with atomic-memory-instructions and memory-barriers.

SC performance:
In the architecture literature, it has been shown that an aggressive
speculative implementation of SC (like ours) may perform as well as weaker
memory-models ([Falsafi], [Hill]). These arguments are orthogonal to
this paper's technical contributions.

Termination-property:
Our main theorem uses termination as an arbitrary property which makes the
theorem interesting but still easily stated. That's for the same reason that
many problems in computability-theory are equivalent to halting-problem in
suitable senses.

-----------------------

Reviewer-C
1. Please note that the buffers between processor and memory-subsystems --
states "ins" and "outs" in Figures 4, 5, 7, 10 -- are multi-capacity buffers.

Reviewer-A
1. Thanks for the bug reports on our related work discussion.

2. Examples:
IRIW behavior:
Consider a system containing several in-order cores, each core connected to
''non-coherent cache'' NCC. All NCCs are connected to a global atomic sotore.
An NCC can read any location from memory and keep it, and the core can load the
location from NCC, if present. But stores from the core always happen to the
global store, and the NCC can discard the location that it has, anytime.

At the beginning, NCC of core 4 reads in the location for x. The following
sequence gives the required behavior:
Core 1: Wx1 (into global memory)
Core 2: Rx (gets 1 from global memory)
        Ry (gets initial value 0 from global memory)
Core 3: Wy1 (into global memory)
Core 4: Ry (gets 1 from global memory)
        Rx (gets 0 from NCC)

Store buffer behavior:
Instead of NCC, each core has a store-buffer into which stores are written, and
loads can read from. The following sequence gives the required behavior:
Core 1: Wx1 (into store buffer, not to memory)
        Ry (gets initial value 0 from global memory)
Core 2: Wy1 (into store buffer, not to memory)
        Rx (gets initial value 0 from global memory)
--------------------------------------------------

Citations in the rebuttal:

[Falsafi] Guiady, C.; Falsafi, B.; Vijaykumar, T. N., "Is SC+ILP=RC?," Computer
Architecture, 1999. Proceedings of the 26th International Symposium on , vol.,
no., pp.162,171, 1999 doi: 10.1109/ISCA.1999.765948 URL:
http://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=765948&isnumber=16584

[Hill] Mark D. Hill. 1998. Multiprocessors Should Support Simple
Memory-Consistency Models. Computer 31, 8 (August 1998), 28-34.
DOI=10.1109/2.707614 http://dx.doi.org/10.1109/2.707614

[Khan] Asif Khan, Muralidaran Vijayaraghavan, Silas Boyd-Wickizer, and Arvind.
2012. Fast and cycle-accurate modeling of a multicore processor. In Proceedings
of the 2012 IEEE International Symposium on Performance Analysis of Systems &
Software (ISPASS '12). IEEE Computer Society, Washington, DC, USA, 178-187.
DOI=10.1109/ISPASS.2012.6189224 http://dx.doi.org/10.1109/ISPASS.2012.6189224

[Dave] N. Dave, M. C. Ng, and Arvind. 2005. Automatic synthesis of
cache-coherence protocol processors using Bluespec. In Proceedings of the 2nd
ACM/IEEE International Conference on Formal Methods and Models for Co-Design
(MEMOCODE '05). IEEE Computer Society, Washington, DC, USA, 25-34.
DOI=10.1109/MEMCOD.2005.1487887 http://dx.doi.org/10.1109/MEMCOD.2005.1487887

### 2014-10-15 Abstract.tex
stuff

We provide a modular proof of a high-performance implementation of Lamport's
sequential consistency in a system with distributed memory. Unlike
Lamport's reference operational model, a high-performance implementation
processes many memory accesses concurrently. Processors employ
\emph{speculative execution} to generate many concurrent loads to pipeline the
memory system effectively. We define what it means for a speculative processor
to be correct independently of the memory system. We then define a memory
system with caches and show that our cache-coherence protocol preserves
\emph{store atomicity} (independently of the processor), a property that all
reasonable memories have in common.  We offer the first formal proof that when
an atomic memory is connected to $n$ speculative processors, the system as a
whole is sequentially consistent. The proof is composed modularly from proofs
of the individual components. Additionally, the cache-coherent memory we
present in this paper is the first fully verified system that supports
arbitrary cache hierarchies. We describe all our components using labeled
transition systems and machine-verify all our proofs using the Coq proof
assistant.

### 2014-10-15 Acknowledgements.tex
stuff

%\subsection*{Acknowledgments}
{\footnotesize\noindent\textbf{Acknowledgments:} The parts of this work done
at MIT was supported by Quanta Research agreement dated 4/1/05 and
Angstrom Project under Grant No. HR0011-10-9-0009. The research at SRI
was supported by Defense Advanced Research Projects Agency (DARPA) and
Air Force Research Laboratory (AFRL), under contract FA8750-11-C-0249
and National Science Foundation under Grant No. CCF-1217498. The
views, opinions, and/or findings contained in this report are those of
the authors and should not be interpreted as representing the official
views or policies, either expressed or implied, of the Quanta
Research, the Defense Advanced Research Projects Agency, the United
States Air Force or the Department of Defense. We would like to thank
Richard Uhler for helping focus this paper in early drafts.% }
%
%{ We would like to thank Richard Uhler for helping focus this paper in
%  early drafts.  This work was supported by the Defense Advanced
%  Research Projects Agency (DARPA) and the United States Air Force,
%  under Contract No. FA8750-11-C-0249. Any opinions, findings and
%  conclusions or recommendations expressed in this material are those
%  of the authors and do not necessarily reflect the views of DARPA or
%  the United States Air Force.  }

### 2014-10-15 Architect.tex
stuff

\section{Architectural Implications}\label{sec:implications}

We have demonstrated a high-performance implementation of SC with
a distributed shared-memory system and speculative processors. However, here
are some of the concerns that an architect of a high-performance multi-processor
may have about our design:

\begin{enumerate}
\item Splitting each load into a speculative load and a verification
load increases the number of requests to the L1 cache. While verification loads
are extremely likely to hit in a cache and therefore have low latency,
nevertheless they fundamentally increase cache pressure on L1. Can we
eliminate some of this verification traffic?

\item The correctness of our model with respect to sequential consistency
crucially relies on the fact that a commit operation for only one load or store
can be active per processor at a time. Can we pipeline commit requests, \ie{}
have several of them issued at a time?

\item In some designs, a processor does not receive an explicit acknowledgement
for a store operation. Intuitively the store acknowledgment means that a store
has been placed in a global order of all stores being issued to the memory, and
this condition can potentially be determined even before the data is stored in
any cache. Can this optimization be incorporated in our framework?
\end{enumerate}

A trick architects use to reduce the need for verification loads is to use the
invalidation signal received by L1 (\eg{} in the MIPS R10000 processor~\cite{yeager1996mips}) 
to invalidate speculative loads for that address in the reorder buffer. 
The reason this works is that such a signal is an
indicator of the fact that another processor has changed the value of that
address before the load has been committed. This scheme is simultaneously
incomplete, in that it does not alleviate the need for verification loads in the
context of address speculation; and conservative, in that it invalidates loads
whose values may not have changed in spite of a write into the corresponding
cache line. Incorporating this solution will require a very small change in our
processor interface, \viz{} that of including invalidation signals from L1
caches.

The solutions for the remaining two issues require making some extra
assumptions about the memory interface. These assumptions can be enforced by
having an extra module (Load Store Buffer, LSB) in front of the memory we have
presented. The processor can issue a commit store request to the LSB and
assume that it receives the acknowledgment straightaway, continuing to issue
further commit requests to the LSB. A verification load request in the LSB
still has to receive a response from L1. It will be sent to L1 only after all
previous commit store requests in the LSB have been sent to L1. The formal
proof of why this scheme works is beyond the scope of this paper.

A more aggressive solution to maintaining the strict load-store orderings
defined by SC is to change the memory model itself. Indeed, many processors use
a more relaxed memory model; for instance, Intel x86 and Sparc use
TSO~\cite{intel64and,weaver1994sparc}, ARM and PowerPC both use their own
relaxed memory models~\cite{grisenthwaite2009arm,power2009version}.
Two things need to be noted about all relaxed memory models. First, every SC
behavior is
always a legal behavior in a relaxed model, making our speculating processor
correct in such a setting. Second, the ISA must be augmented with explicit
memory fence instructions that prevent reorderings of memory instructions
across fences, if the application demands preserving the order. Adding explicit memory fences is generally slower than implicitly
having the restrictions within the memory model, but the belief is that
explicit fences are only needed sparingly in practice and on the whole will
give better performance than the equivalent SC system.
%\murali{Starting from "two things", the rest seems out of place}

In order to use our methodology for systems with relaxed memory models, we
would have to provide an LTS description for each model. One can potentially
adjust our processor model accordingly (for
instance, more aggressive issue of even verification loads) and add the LSB described
earlier. The details are not obvious, however. Alglave
\etal{}~\cite{alglave2012formal,alglave2014herding} provide a framework for 
axiomatic specification of memory models.  Axioms regarding relaxed memory
models have also been proposed through analysis of empirically observed behavior
of certain ``litmus tests''~\cite{mador2012axiomatic,sarkar2012synchronising}.
There has also been work in providing operational semantics that describe the
observed behavior of real hardware~\cite{sarkar2011understanding}, and proving
the equivalence between the axiomatic specification of TSO and operational
semantics of a simple in-order processor with a store buffer and instantaneous
global memory~\cite{x86tsocacm10}. However, others have suggested
simplifying memory models, in other words to implement SC in real
hardware~\cite{hill1998multiprocessors}.

### 2014-10-15 AtomicStore.tex
stuff

\section{Enabling Practical Memory Subsystems: Atomic Store}

The behavior of a memory system realizing different memory models or
interfacing with different levels speculation from the processors would be
dramatically different. For instance, a memory system realizing Total Store
Order allows significantly the more memory requests to be in flight than in a
sequential consistent system. As such one might expect that there should be
relatively little sharing between designs.

%% At a high-level, the physical interface of memory is fairly consistent
%% regardless of the memory model or instruction execution scheme. The processor
%% makes requests (loads and stores) and gets responses back from the
%% memory.

Memory subsystems are designed to allow parallel operation

The most natural interface for a memory subsystem is to expect the interface to

Memory needs caches, highly pipelined. All worries happen at a single address. No ordering guarantees between two different addresses. This is a natural representation because.

Clearly this model is not SC. However, it is highly atuned to high-performance results.

### 2014-10-15 Conclusion.tex
stuff

\section{Conclusions and Future Work}\label{sec:conclusion}

In this paper we developed a modular proof structure for distributed
shared-memory hardware systems, corresponding naturally to the modularization seen in
hardware implementation. This makes it feasible for verification of the
system to be carried out throughout the implementation process without
disrupting the architect's design process.

While we provide a clean interface for an SC system, we are working on
encompassing relaxed memory models commonly used in a modern processors.

Another direction that bears investigation is extending our results to
cover finer-grained architectural details.
However, the behaviors of microarchitectural structures,
\ie{} register renaming structures \etc{}, vary significantly across
implementations and can be quite complex. It is not clear if a
practical modularization can be found that strikes a balance between flexibility and
simplicity.

Finally, to make full use of our results for implementations, hardware
descriptions must be translated to and from our labeled transition system
representation. As we implied in the discussion of our cache-coherence system,
Bluespec SystemVerilog's rule-based representation is very close to LTSes and
in some cases can be transliterated directly.  There has been work to generate
circuits from Bluespec-like descriptions in Coq \cite{Braibant2013Fesi} and we
are pursuing further automatization.

### 2014-10-15 Implementation.tex
stuff

\section{How to Build and Sequentially System using Atomic Store}

We can have a simple system using a lamport-like processor which will nopipelining of memory. Trivially SC with atomic store. This is obviously correct, but from a point of view of practice, we must pipeline this. We need to refine this model into a

(How to use a non-monolithic memory).

\section{How to build a monolithic processor}

A speculative Lamport system consists of a monolithic memory and other loads and stores. This is how this system works. Need to discuss

* need definitions of spec processor. (Must be complete and sound)

A full realistic implementation:

### 2014-10-15 Introduction.tex
stuff

\section{Introduction}
\label{sec:Introduction}

A modern high-performance, cache-coherent, distributed-memory hardware system is
inherently complex. Such a system implements memory models like sequential
consistency (SC)~\cite{lamport1979make}, Total Store Order (TSO), and
Relaxed Memory Order (RMO)~\cite{weaver1994sparc}, each
of which characterizes how loads and stores may be issued by a processor and
how the shared-memory component responds to these requests.  These systems by
their nature are highly concurrent and nondeterministic.

Hardware designers recognize the challenge of implementing such systems, to such an extent that they have
already become some of the most serious real-world adopters of formal methods.
Hardware verification is dominated by model-checking, where abstraction
techniques are used to reduce designs to finite state spaces, which can be
explored exhaustively.  There are limits to the construction of sound
abstractions, so verifications of, e.g., cache-coherence protocols have almost
always treated systems with concrete topologies, involving particular finite
numbers of caches and processors. For instance, explicit-state model checking tools
like Murphi~\cite{murphi} or TLC~\cite{tlc} are only able to handle small and
simplified systems, \eg{} a single-level cache hierarchy with three CPUs, two
addresses, and one bit of data. Symbolic model checking techniques have fared
slightly better; McMillan \etal{} have verified a 2-level MSI protocol based on
the Gigamax distributed multiprocessor using SMV~\cite{gigamax}. Optimizations
on these techniques (\eg{} partial order reduction~\cite{part}, symmetry
reduction~\cite{sym1, sym2}, compositional reasoning~\cite{comp}, \etc{}) scale
the approach further, but they are still unable to handle realistic
hierarchical protocols. We might hope to achieve higher
assurance and understanding of design ideas by verifying \emph{infinite families}
of hardware designs, which resist reduction to finite-state models.  A
potential related benefit is reducing the computational resources required
to formally verify a new design in the family, applying a more general theorem without
much new analysis.

A related benefit of a deductive approach is \emph{modularity}.  It seems to
be rare for hardware practitioners to verify full designs that include
distinct functional components, like processors and memory systems.  One
reason is the algorithmic overhead of exploring such large state spaces.
If we can instead give localized correctness theorems for different
subsystems, then we can compose these theorems in a black-box way to
produce full-system theorems.

The traditional notion of modularity in hardware is closely tied to
\emph{Finite State Machines} (FSMs), more specificially synchronous FSMs. The biggest
problem with using FSM abstraction is that it restricts significantly the kinds of
refinements that can be performed on a module. For example, if we change the
timing of an adder so that it takes 2 clock cycles instead of 1, the whole
system is likely to break. Even though big modules in hardware are often
designed with asynchronous handshaking protocols, current
hardware verification methodologies do not deal with timing refinements.
In this paper, we use the formalism of \emph{labeled transition systems} that
is well-known in process-calculus circles.  It turns out to be exactly the
right way of capturing contracts on interaction between hardware components,
enabling both modular refinement and verification.

%Proofs in this style are quite common in the software world.  The function
%preconditions and postconditions of Hoare logic provide a canonical
%example, and there have been many innovations in modular proof since Hoare's
%original work.  Somehow the hardware verification world has taken less
%advantage of modularity.  In this paper, we propose a new program in
%hardware verification, based on modular proofs using the formalism of
%\emph{labelled transition systems} that is well-known in process-calculus
%circles.  It turns out to be exactly the right way of capturing contracts on
%interaction between hardware components.

Computer architects have intuitions about the contracts of classic
subsystems.  However, these intuitions have not been formalized
in enough detail to enable rigorous proofs.  In carrying out that exercise,
we have also learned some things about architectural design that should
be of interest outside a formal-methods context.

Concretely, we provide a modularization of distributed shared-memory systems
that makes minimal assumptions about networks, provides an abstraction for
memory that allows high-performance implementation of any reasonable memory
model, and allows a full range of realistic performance optimizations in
processors without violating modularity. We want our modular system to permit
\emph{black-box refinement}, that is, nothing beyond the interface contract can
be assumed about other modules.

In each instance we describe a reference model and an implementation model for
a module using labeled transition systems and formally prove that the
implementation is correct (sound) with respect to our reference. For example,
our reference model for memory is monolithic, where each load or store request
generates a (conceptually) instantaneous response, but where requests may be processed
in any order. Such a model is intuitively \emph{store atomic} in the sense that
the effect of a store is visible to all subsequent loads. We then describe a
realistic cache-coherence protocol comprised of a hierarchy of caches, and
formally prove that this system can never produce
a behavior that is not allowed by the reference model. Similarly, we define a
reference model for a processor that executes one instruction at a time and
an implementation model that exploits speculation to increase parallelism,
and we show that the speculative processor implements the reference processor,
in a suitable sense.

In the end, we prove a theorem about an infinite family of hardware systems,
each system having an arbitrary number of speculative processors connected to
a memory with an arbitrary hierarchy of caches.  All our proofs are mechanized
in the Coq proof assistant, and they are modular in the sense of allowing
further optimization to either processor or memory without needing to touch
the proof of the other.  We characterize overall system behavior with a very
simple transition system formalizing \emph{sequential consistency}, and we
conjecture that the methodology generalizes to other popular memory models.

\begin{figure}
\centering
\includegraphics[scale=.43]{zoom}
\caption{Overall system}
\label{zoom}
\end{figure}

Figure \ref{zoom} sketches the sort of hardware system that we have verified.
We show the top-level system design at top right, and we zoom in on boxes standing for the
memory system and a single processor.

Memory is composed of a \emph{hierarchy of caches}, where each cache node (labeled
like ``L1,'' ``L2,'' etc.) communicates only with its neighbors in the graph.
Each processor has a dedicated L1 cache, from which we do our best to satisfy
all memory requests, to avoid the latency of a round-trip with main memory.
With an L1 cache miss, we may still find a hit in a parent cache below the
level of main memory, realizing a smaller but still significant speedup.
We have verified a \emph{directory-based} protocol for coordinating an arbitrary
tree of caches, where each node stores a conservative approximation of its
children's states.

A processor is decomposed into several components.  We have the normal
architectural state, such as values of registers.  Our proofs are generic over
\emph{a family of instruction set architectures}, with parameters for opcode sets and
functions for executing opcodes and decoding them from memory.  Other key
components are a \emph{branch predictor}, which guesses at the control-flow path
that a processor will follow, to facilitate speculation; and a
\emph{reorder buffer (ROB)}, which decides which instructions along that path to
try executing ahead of schedule.  Our proofs apply to an arbitrary branch predictor,
and they work for any reorder buffer satisfying a simple semantic condition.

Figure~\ref{proofs} gives the overall proof structure that we employ to verify
the system of Figure~\ref{zoom}. As we will explain later, $P_\text{so}$
represents the speculative out-of-order processor, $M_c$ the cache memory,
$M_m$ the simple memory, $P_\text{ref}$ a simple decoupled processor, and
SC the simple reference model having multiple processors executing each
instruction atomically thus implementing sequential consistency.  Notation
$A^n$ is for $n$ copies of system $A$ running in parallel.
$\sqsubseteq$ is a refinement operator, capturing a suitable notion of when
one system implements another.  We go into more detail on the symbols from
the figure in the rest of the paper.

Our current framework establishes theorems of the form ``if system $A$ has a run
with some particular observable behavior, then system $B$ also has a run with
the same behavior.''  In this sense, we say that $A$ correctly implements $B$.
Other important properties, such as \emph{deadlock freedom} for $A$ (which
might get stuck without producing any useful behavior), we leave for future
work.

\begin{figure}
\includegraphics[scale=.45]{proofs}
\caption{Overall proof structure}
\label{proofs}
\end{figure}

The main contributions of this paper are:

\begin{itemize}

\item The \emph{first machine-checked proof} of an implementation of sequential
  consistency involving speculative processors and cache-coherent distributed
  memory.

\item The \emph{first machine-checked proof} of an invalidation-based
  cache-coherence protocol for distributed memories comprised of arbitrary
  cache hierarchies. We believe such a memory is also the correct abstraction for
  implementing relaxed memory models.

\item A general \emph{modular verification methodology} applicable to hardware
  algorithms, based on \emph{labeled transition systems}.  We are able to verify
  the several components of Figure~\ref{zoom} against general formal interfaces,
  enabling mixing and matching of different realizations of each interface,
  without doing any new proofs that peek beneath abstraction boundaries.

\item We use semantics to \emph{elucidate folklore ideas in computer architecture}.
  For instance, we formalize a variety of processor speculation patterns with
  the idea of speculative loads followed by verification loads, and we capture
  a key requirement on memory systems as refinement of a simple transition
  system.

\end{itemize}

As a side effect, we %debunk a common misconception about the semantics of the
have shown a (rather surprising) architectural result about the semantics of
the interface between processor and memory system. One can establish a
sequentially consistent system even if the buffers between the memory system
and the processor do not preserve message order. We prove as much by
establishing simulation of sequential consistency without relying on an
ordering assumption.

\paragraph{Paper Organization:} In Section~\ref{sec:lts} we introduce
our flavor of the labeled transition systems formalism,
including a definition of trace refinement. In
Section~\ref{sec:sc} we give a simple formal model of sequential consistency,
which we decompose further for decoupled shared-memory systems in Section~\ref{sec:ref}.
The following sections refine the two main subcomponents from Figure~\ref{zoom}.
Section~\ref{sec:ooo} discusses definition and verification of a
speculative processor model, and Sections~\ref{sec:cc} and~\ref{sec:ccproof}
respectively define and prove our hierarchical cache-coherence protocol.
Finally, in Section~\ref{sec:finalresult} we show
the whole-system modular proof of our complex system, followed by
discussion of implications in Section~\ref{sec:implications}
and further discussion of related work and other conclusions in
Sections~\ref{relatedWork} and \ref{sec:conclusion}.

The Coq source code of our proofs is included as the non-anonymous
supplement to this paper.

%% Modularity has long been understood as a key property for effective design and
%% verification of complex systems, \eg\ distributed memory systems. For the
%% designer, it allows a separation of concerns increasing robustness by allowing
%% the behavior encapsulated by a modular boundary to be realized by multiple
%% implementation any of which may be safely dropped into the system. It also
%% allows a greater measure of parallelization of design improving development
%% time. Similarly verification is simplified as modular interface agreements
%% provide a set of natural lemmas to verify. This leads us towards a
%% decomposition of the whole system verification task into smaller more
%% manageable chunks.

%% Given its complexity, it is a practical necessity for distributed shared memory
%% systems to be built a strong idea of modularity. The standard modularization
%% (see Figure~\ref{fig:highleveldsm}) separates the ISA-level processors nodes
%% which are responsible for executing threads of computation, and a unified
%% shared memory subsystem which is responsible for ensuring that memory requests
%% it receives are handled correctly, \ie{} in line with the memory model which
%% dictates which store values a load may observe in the system. These processors
%% may then be refined for performance, while the monolithic memory module is
%% improved by introducing layers of caching and a coherence protocol to guarantee
%% memory consistency.

%% This high-level approaches works reasonably well for implementation; separate
%% design groups tackle can each task in isolation. However, when we consider
%% verification, the modularity enforced during the implementation has little
%% benefit and in practice we must resort to full system verification. The core of
%% this problem lies in the fact that the simple high-level decomposition of
%% memory and processors does not fully take into account the space of speculation
%% techniques used in the processor.

%% All practical processor instruction set architectures (ISAs) are fairly
%% sequential; the result of executing an instruction may affect which next
%% instruction is executed. As a result for performance some sort of prediction is
%% used to allow instructions to be speculatively started and executed before the
%% previous instructions have completed and the values need to execute are known.
%% Additional bookkeeping is used to determine when our guess is wrong and
%% false-path instructions are undone leaving no affect on the architectural
%% state. As precise exception handling is a requirement these speculative
%% instructions are committed, \ie{} expose their effects of execution to the
%% whole system, in sequence.

%% Though most forms of speculation (\eg{} branch prediction, instruction prefetch
%% from pipelining, value prediction) are often considered as being orthogonal to
%% memory issues, speculation changes the sequence of memory requests the
%% processor sends to the memory. This may be merely handling requests out of
%% order, \eg{} issuing a load request before a logically earlier one has issued
%% (a common occurrence due to data dependencies in out-of-order processors). We
%% may also allow incorrect memory operations to be issued, \eg{} executing a load
%% from a false path. It is worth noting that while such load operations are
%% possible to speculate, we cannot speculatively issues store operations to the
%% memory as they become exposed to the whole system and speculation mechanisms
%% are isolated to a single processor. As a consequence of this, our memory does
%% not know what order requests it is sent should happen. Thus the memory in
%% isolation cannot directly realize the memory model and in actually implements a
%% much weaker model leaving the full enforcement of the memory model to the
%% processor which can enforce this by restricting when it issues requests.

%% The goal of this work is to introduce an interface enabling true modular
%% implementation and verification of shared memory systems. That is, we should be
%% able to describe a modular decomposition where we may correct replace either
%% the processor nodes, the memory with locally verified alternatives without
%% changing the correctness of our whole system proof of correctness or
%% restricting our ability to describe shared memory systems. As part of this we
%% will describe the processor interface in a way which allows the full range of speculation

%% We do so by showing the construction and initial refinements of a sequentially
%% consistent shared memory system. We formally defining a minimal cache coherent
%% speculation-friendly memory interface (\emph{Store Atomicity}), and use this as
%% the interface for our memory subsystem. Using this we specify the notion of
%% correct for a processor and prove that our system is correct assuming the
%% correct implementation of these two objects.  We then construct both a naive
%% reference processor and memory subsystem and provide proofs of correctness in
%% Coq. We then refine both components into more realistic implementations
%% including taking our memory subsystem to hardware-level description of a
%% arbitrary-level hierarchical cache coherence implementation and verify these
%% module. We show that the cross products of possible implementations are correct
%% and their proof automatically from the modularly.

%%  LocalWords:  Modularity modularity

### 2014-10-15 RelatedWork.tex
stuff

\section{Related Work}
\label{relatedWork}

Formal verification has been used to verify multiprocessor
systems, especially with model checking
\cite{burch1994automatic,mcmillan1998verification}.

More recently, Intel's execution cluster (\ie{} the ALU and its
interface) has been formally verified using model checking
\cite{kaivola2009replacing}.

Theorem provers have also been used to verify microprocessors, \eg{} HOL has been
used to verify an academic microprocessor AVI-1 \cite{windley1995formal}.
Term-rewriting based techniques have been used to specify microprocessors, with
the goal of making them more amenable to verification \cite{shen1999using}.

%\murali{Next paragraph is all I know about this stuff, should probably be
%removed}
%\adam{What are you referring to with ``it'' in your comment?  I don't understand what this paragraph refers to.}
%It has been verified using model checking \cite{qadeer2003verifying},
%with a fixed number of processors and memory locations.  Model-checking has also
%been used to verify communication fabrics \cite{chatterjee2012automatic}.

Theorem provers have also been used in verifying cache coherence protocols.
The correctness of the FLASH cache coherence protocol~\cite{flash} for a
single-level hierarchy was proven with PVS~\cite{park} using a mapping, called
``aggregation of transactions,'' from multiple localized transactions to a
single global transaction, and finally using model checking to prove the correctness of
the global transactions model. However, they do not provide a methodology to
prove the aggregation of transactions for other protocols, which is where, we
believe, the difficulty lies; in contrast, our modular proof relies on generic
invariants, making it more adaptable to different protocols.

Cachet~\cite{StoyShenArvind:Proofs}, another single-level cache coherence
protocol, was proven to be implementable in the CRF memory
model~\cite{Shen:CRF}. Shen \etal{} provided a paper-and-pencil proof showing
Cachet does not violate the CRF memory model, while Stoy \etal{} implemented
portions of the paper-and-pencil proof using PVS. However, neither proof
establishes that the protocol is store atomic. Rather, they show that Cachet
obeys the CRF memory model, which is not restricted to store-atomic behavior.

%There has been a lot of work in providing frameworks for axiomatic
%specifications of memory models \cite{alglave2012formal,alglave2014herding}
%and in providing axiomatic memory model specifications for real systems
%through analyis of empirically observed behavior of certain ``litmus tests''
%\cite{mador2012axiomatic,sarkar2012synchronising}. There has also been some
%work in providing operational semantics that describes the observed behavior of
%real hardware \cite{sarkar2011understanding}. Finally, there has been work
%formally relating the axiomatic specification of TSO to operational semantics
%of real x86 and Sparc machines \cite{x86tsocacm10}.
%
%The plethora of work in the area of memory models indeed suggests that this is
%a complex problem, exacerbated by the hardware architect's desire to increase
%hardware performance at the cost of complicating the memory models further. There
%are two aspects where our work can help in tackling this complex problem: (1) we
%are proposing a modular approach for verifying systems in which the components,
%namely the processors and the memory, can be verified against the memory model
%contract independent of each other, and (2) we are also proposing a design
%in which a system obeying sequential consistency is made into a high
%performance machine through the use of speculative loads followed by
%verification. The only downside is waiting for store acknowledgements, but this
%wait does not preclude the machine from performing further computations for
%future instructions without actually committing them. We are hoping to develop
%both these ideas -- of extending our proof approach for more complex memory
%models and designing real hardware which is sequentially consistent but has
%high performance.
%
%
%BELOW IS UNEDITTED FROM OLD PAPER

%These are lifted as is from that guy. Need to figure out how to restate them, and make it more concise.
%
%For verification of high level specifications, modern industrial practice
%consists of modeling small instances of the protocols, e.g., three CPUs
%handling two addresses and one bit of data, in terms of interleaving atomic
%steps, in guard/action languages such as Murphi [50] or TLA+ [84], and
%exploring the reachable states through explicit state enumeration.
%Symbolic methods, e.g., BDD [31] or SAT [129, 142], are yet to succeed for the
%verification of cache coherence protocols.
%Monolithic formal verification methods – methods that treat the protocol as a whole – have
%been used fairly routinely for verifying cache coherence protocols from the early 1990s
%[10 (2 nodes, murphi)]. However, these monolithic techniques will not be able to handle the very
%large state space of hierarchical protocols. Compositional verification techniques are essential. 
%
%For coherence protocols which only consider one level (or nonhierarchical protocols),
%various techniques have been proposed for the verification. These techniques basically
%include explicit state enumeration [133] and symbolic model checking [102]
%Opti-
%mizations on these techniques include partial order reduction [24], symmetry reduction
%[25, 72], compositional reasoning [96], predicate abstraction [22, 49, 82], etc. There is
%also a large body of work on parameterized verification [42, 53, 54, 100, 117]. However,
%none of them have shown the capability to verify hierarchical coherence protocols with
%realistic complexity.
%
%McMillan et al. [97, 102] modeled a 2-level MSI coherence protocol based on the
%Gigamax distributed multiprocessor [102]. In the protocol, bus snooping is used in both
%levels. SMV [97] was used to verify this protocol with two clusters each having six
%processors, for both safety and liveness properties. In terms of the complexity, many
%realistic details are abstracted away from the protocol so that the protocol does not have
%the typical complexity that a hierarchical coherence protocol has.
%
%=======
%Theorem provers have also been used in verifying cache coherence protocols,
%\eg{} FLASH~\cite{park} and Cachet~\cite{Shen:CRF, StoyShenArvind:Proofs} have
%been proved using the PVS theorem prover. Both works prove correctness of a
%single-level hierarchy while our work proves it for a multi-level hierarchy.
%The work by Park \etal{} on FLASH proves a mapping, called ``aggregation of
%transactions'', from multiple localized transactions to a single global
%transaction with PVS, and uses model-checking to prove the correctness of the
%global transactions model. However, they do not provide a methodology to prove
%the aggregation of transactions for other protocols, which we believe to be very
%difficult; in contrast, our modular proof relies on generic invariants making it
%more naturally adaptable to different protocols.
%>>>>>>> 96e9880a190d9acb6201b10e2b105d18c72f56aa

### 2014-10-15 SequentialConsistency.tex
stuff

\section{Defining Sequential Consistency}

To dicsuss the modularization of our system we must first fully define what we
mean by sequential consistency. Informally, sequential consistency states that
the memory operations from all processors in the system are evaluated
atomically with regards to each other by the memory and are evaluated in order
locally but with no ordering requirements between operations from different
processors.

Formally Sequential Consistency can be described as:

\xxx{SC Definition HERE}.

\xxx{TOTAL ORDERING CONDITION}

Total ordering is not generally stated.

### 2014-10-15 TransitionsUseful.tex
stuff

\section{Labeled Transition Systems}\label{sec:lts}

We make extensive use of the general theory of labeled transition
systems, a semantics approach especially relevant to communicating
concurrent systems.  As we are formalizing processors for
Turing-complete machine languages, it is challenging to prove that a
system preserves almost any aspect of processor behavior from a model
like SC.  To focus our theorems, we pick the time-honored property of
\emph{termination}.  An optimized system should terminate or diverge
iff the reference system could also terminate or diverge,
respectively.  Our basic definitions of transition systems build in
special treatment of halting, so that we need not mention it
explicitly in most contexts to come.

\begin{defn}
A \textbf{labeled transition system (LTS)} is a ternary relation, over
$\mathcal S^H \times \mathcal L^\epsilon \times \mathcal S^H$, for some sets
$\mathcal S$ of states and $\mathcal L$ of labels. We usually do not mention
these sets explicitly, as they tend to be clear from context. We write
$X^\epsilon$ for lifting of a set $X$ to have an extra ``empty'' element
$\epsilon$ (like an \texttt{option} type in ML). We write $X^H$ for lifting of
a set $X$ to have an extra ``halt'' element $H$. We also implicitly consider
each LTS to be associated with an \emph{initial} state in $\mathcal S$.
\end{defn}

For LTS $A$, we write $\io{A}{s}{s'}{\ell}$ as shorthand for $(s, \ell, s') \in
A$, and we write $A_0$ for $A$'s initial state. The intuition is that $A$ is
one process within a concurrent system. The label $\ell$ from set $\mathcal L$
of labels is produced when $A$ participates in some IO exchange with another
process; otherwise it is an empty or ``silent'' label $\epsilon$. As a
shorthand, we sometimes omit labels for $\epsilon$ steps.

\subsection{Basic Constructions on LTSes}

From an LTS representing single-step system evolution, we often want to build
an LTS capturing arbitrary-length evolutions.

\begin{defn}
The \textbf{transitive-reflexive closure} of $A$, written $A^{*}$, is a derived
LTS.  Where $A$'s states and labels are $\mathcal S$ and $\mathcal L$, the
states of $A^{*}$ are $\mathcal S$, and the labels are $\mathcal L^{*}$, or
\emph{sequences} of labels from the original system.  $A^{*}$ steps from $s$ to
$s'$ when there exist zero or more transitions in $A$ that move from $s$ to
$s'$.  The label of this transition is the \emph{concatenation} of all labels
generated in $A$, where the empty or ``silent'' label $\epsilon$ is treated as
an identity element for concatenation.
\end{defn}

We also want to compose $n$ copies of an LTS together, with no explicit
communication between them.  We apply this construction later to lift a
single-CPU system to a multi-CPU system.

\begin{defn}
The \textbf{$n$-repetition} of $A$, written $\smulti{n}{A}$, is a derived LTS.
Where $A$'s states and labels are $\mathcal S$ and $\mathcal L$, the states of
$\smulti{n}{A}$ are $\mathcal S^n$, and the labels are $[1, n] \times \mathcal
L$, or pairs that \emph{tag} labels with which component system generated them.
These labels are generated only when the component system generates a label.
The whole system halts whenever one of the components halts.  We define the
transition relation with the following inference rules. We treat $n$-tuples as
functions keyed on numeric positions, and we write $\theta[i := v]$ for
updating $n$-tuple $\theta$ to overwrite the $i$th position with value $v$, and we treat any label $(i, \epsilon)$ as $\epsilon$.
$$\inference[i]
{\theta(i) = a & \io{$A$}{a}{a'}{\ell} & a' \neq H}
{\io{$\smulti{n}{A}$}{\theta}{\theta[i:=a']}{i,\ell}}
\quad \inference[H]
{\io{$A$}{\theta(i)}{H}{\ell}}
{\io{$\smulti{n}{A}$}{\theta}{H}{i,\ell}}
$$
\end{defn}

Eventually, we need processes to be able to communicate with each
other, which we formalize via the $+$ composition operator. It
connects same-label transitions in the two systems, treating the label
as a cooperative communication event that may now be hidden from the
outside world, as an $\epsilon$ label.

\begin{defn}\label{plus}
Where $A$ and $B$ are two LTSes sharing labels set $\mathcal L$,
and with state sets $\mathcal S_A$ and
$\mathcal S_B$ respectively, the \textbf{communicating composition} $A + B$ is
a new LTS with states $\mathcal S_A \times \mathcal S_B$ and an empty label set,
defined as follows:
\vspace{-.2in}\begin{center}
$$\inference[$A$]
{\io{$A$}{a}{a'}{} & a' \neq H}
{\io{$A + B$}{a, b}{a', b}{}}
\quad \inference[$B$]
{\io{$B$}{b}{b'}{} & b' \neq H}
{\io{$A + B$}{a, b}{a, b'}{}}$$
$$\inference[H$_A$]
{\io{$A$}{a}{H}{}}
{\io{$A + B$}{a, b}{H}{}}
\quad \inference[H$_B$]
{\io{$B$}{b}{H}{}}
{\io{$A + B$}{a, b}{H}{}}$$
$$\inference[Join]
{\io{$A$}{a}{a'}{\ell} & \io{$B$}{b}{b'}{\ell} & a', b' \neq H}
{\io{$A + B$}{a, b}{a', b'}{}}$$
\end{center}
\end{defn}

\subsection{Refinement Between LTSes}

We need a notion of when one LTS ``implements'' another.  Intuitively,
transition labels and halting are all that the outside world
can observe. Two systems that produce identical labels and termination behavior
under all circumstances can be considered as safe substitutes for
one another. We will only need an asymmetrical notion of compatibility:

\begin{defn}\label{refines}
For some label domain $\mathcal L$, let $f : \mathcal L \to \mathcal L^\epsilon$ 
be a function that is able to replace labels with alternative
labels, or erase them altogether. Let LTSes $A$ and $B$ have the same label
sets $\mathcal L$. We say that
\textbf{$A$ trace-refines $B$ w.r.t. $f$}, or $A \sqsubseteq_f B$, if:
\begin{multline*}
\forall s_A, \eta. \; \io{$A^{*}$}{A_0}{s_A}{\eta} \Rightarrow \exists s_B. \;
\io{$B^{*}$}{B_0}{s_B}{f(\eta)} \; \wedge \\
(s_A = H \Leftrightarrow s_B = H)
\end{multline*}
Each label in the trace is replaced by the mapping of $f$ on it, and
labels mapped to $\epsilon$ by $f$ are dropped. $f$ is overloaded
to denote that multi-label version when applied to $\eta$.
\end{defn}

As a shorthand, we write $A \sqsubseteq B$ for $A \sqsubseteq_\mathsf{id} B$,
for $\mathsf{id}$ an identity function, forcing traces in the two systems to
match exactly. Under this notion of identical traces, we say that $A$
is sound w.r.t. $B$.

\subsection{A Few Useful Lemmas}

The reflexivity and transitivity properties of refinement are crucial to help us
structure modular proofs.

\begin{theorem}\label{transitive}
$\sqsubseteq$ is reflexive and transitive.
\end{theorem}

Another crucial proof step is to lift a single-process result to the $n$-way
replication of that process.

\begin{theorem}\label{liftn}
If $A \sqsubseteq_f B$, then $\smulti{n}{A} \sqsubseteq_{f^n} \smulti{n}{B}$, where
$f^n$ is $f$ lifted appropriately to deal with indices ($f^n(i, \ell) =
(i, \ell')$ when $f(\ell) = \ell'$, and $f^n(i, \ell) = \epsilon$ when
$f(\ell) = \epsilon$).
\end{theorem}

We will also need a result applying to communicating compositions.

\begin{theorem}\label{liftplus}
If $A \sqsubseteq_f A'$ and $B \sqsubseteq_f B'$, then $A + B \sqsubseteq A' +
B'$. In other words, if systems $A$ and $B$ are individually simulated by $A'$ and
$B'$ on identical alteration of the traces, then the composed
system $A+B$ will be sound with respect to $A'+B'$.
\end{theorem}

\begin{figure}
\centering
\begin{boxedminipage}[c]{.48\textwidth}
%\centering
\inference
[Halt]
{\theta(i) = (s, \pc) & \dec(s,\pc) = H}
{\io{SC}{\theta,m}{H}{}}

\inference
[NonMem]
{\theta(i) = (s, \pc) & \dec(s,\pc) = (\nm, x) \\ \exec(s, \pc, (\nm, x)) = (\pc', s')}
{\io{SC}{\theta,m}{\theta[i \coloneqq (s', \pc')], m}{}}

\inference
[Load]
{\theta(i) = (s,\pc) & \dec(s,\pc) = (\ld, x, a) \\ \exec(s, \pc, (\ld, x, m(a))) = (\pc', s')}
{\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m}{}}

\inference
[Store]
{\theta(i) = (s,\pc) & \dec(s,\pc) = (\st, a, v) \\ \exec(s, \pc, (\st)) = (\pc', s')}
{\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m[a\coloneqq v]}{}}
\end{boxedminipage}
\caption{LTS for sequential consistency with $n$ simple processors}
\label{Ref}
\end{figure}

\section{Specifying Sequential Consistency}\label{sec:sc}

Our final theorem in this paper establishes that a particular complex hardware
system implements sequential consistency (SC) properly.  We state the theorem
in terms of the trace refinement relation $\sqsubseteq$ developed in the prior
section.  Therefore, we need to define an LTS embodying SC.  The simpler this
system, the better.  We do not need to worry about its performance properties,
since we will prove that an optimized system remains faithful to it.

Figure~\ref{Ref} defines an $n$-processor, sequentially consistent system as an
LTS.  The definition is parametrized over details of an \emph{instruction set
architecture} (ISA).  In particular, the ISA gives us some domains of
architectural states $s$ (e.g., register files) and of program counters $\pc$.
A function $\dec(s, \pc)$ figures out which instruction $\pc$ references in the
current state, returning the instruction's ``decoded'' form.  A companion
function $\exec(s, \pc, \mathit{dec})$ actually executes the instruction,
returning the next program counter $\pc'$ and the new state $s'$.

The legal instruction forms, which are outputs of $\dec$, are $(\nm, x)$, for
an operation not accessing memory; $(\ld, x, a)$, for a memory load from
address $a$; $(\st, a, v)$, for a memory store of value $v$ to address $a$; and
$H$, for a ``halt'' instruction that moves the LTS to state $H$. The parameter
$x$ above represents the rest of the instruction, including the
opcode, registers, constants, \etc{}

The legal inputs to $\exec$ encode both a decoded instruction and any relevant
responses from the memory system.  These inputs are $(\nm, x)$ and $\st$, which
need no extra input from the memory; and $(\ld, x, v)$, where $v$ gives the
contents of the requested memory cell.

We define the initial state of SC as $(\theta_0, m_0)$, where $m_0$ is
some initial memory fixed throughout our development, mapping every
address to value $v_0$; and $\theta_0$ maps every processor ID to
$(s_0, \pc_0)$, using architecture-specific default values $s_0$ and
$\pc_0$.

%\medskip

This LTS encodes Lamport's notion of SC, where processors take turns executing
nondeterministically in a simple interleaving.  We need know very few details
of the ISA to define the SC model.  Our final, optimized system is parametrized
over an ISA in the same way.  In the course of the rest of this paper, we will
define an optimized system $O$ and prove $O \sqsubseteq \text{SC}$.

However, we are not interested merely in verifying the particular system $O$.
Instead, we want to structure our proof \emph{modularly}, so that different
common hardware components can be verified separately, and, in a black-box
manner, we may assemble a correctness proof for any composition of these
components.  For instance, we want to verify CPUs and memory systems
separately, picking any realization of each to produce a verified system, with
minimal new proof effort. As a more concrete example, memory systems should be
verifiable independently of which weak memory model processors assume; we want
to reuse memory proofs for different kinds of processors.

To support this style of modular decomposition, we need to introduce a few
intermediate systems.

\begin{figure}
\centering
\begin{boxedminipage}[c]{.48\textwidth}
\inference
[Halt]
{\dec(s,\pc) = H}
{\io{P$_\text{ref}$}{s,\pc,\bot}{H}{}}

\inference
[NM]
{\dec(s,\pc) = (\nm, x) & \exec(s, \pc, (\nm, x)) = (\pc', s')}
{\ioe{P$_\text{ref}$}{s,\pc,\bot}{s', \pc', \bot}}

\inference
[LdReq]
{\dec(s,\pc) = (\ld, x, a)}
{\ior{P$_\text{ref}$}{s, \pc, \bot}
{s, \pc, \top}{\ld, a}}

\inference
[StReq]
{\dec(s,\pc) = (\st, a, v)}
{\ior{P$_\text{ref}$}{s, \pc, \bot}
{s, \pc, \top}{\st, a, v}}

\inference
[LdRep]
{\dec(s,\pc) = (\ld, x, a) & \exec(s, \pc, (\ld, x, v)) = (\pc', s')}
{\iol{P$_\text{ref}$}{s, \pc, \top}
{s', \pc', \bot}{\ld, v}}

\inference
[StRep]
{\dec(s,\pc) = (\st, a, v) & \exec(s, \pc, (\st)) = (\pc', s')}
{\iol{P$_\text{ref}$}{s, \pc, \top}
{s', \pc', \bot}{\st}}
\end{boxedminipage}

\caption{LTS for a simple decoupled processor (P$_{\text{ref}}$)}
\label{Pref$}
\end{figure}

\begin{figure*}
\centering
\begin{boxedminipage}[c]{.48\textwidth}
\inference
[Insert]
{}
{\iorn{$M_m$}{\inp, \outp, m}{\inp[i \coloneqq \{q\} \uplus \inp(i)], \outp, m}{i}{q}}

\inference
[Remove]
{\outp(i) = \mathit{rs} \uplus \{r\}}
{\ioln{$M_m$}{\inp, \outp, m}{\inp, \outp[i \coloneqq \mathit{rs}], m}{i}{r}}

\inference
[SpLoad]
{\inp(i) = \mathit{qs} \uplus \{\spc\ld,t,a\}}
{\io{$M_m$}{\inp, \outp, m}{i, \inp[i \coloneqq \mathit{qs}], \outp[i \coloneqq
(\spc\ld, t, m(a)) \uplus \outp(i)], m}{}}

\inference
[Load]
{\inp(i) = \mathit{qs} \uplus \{\ld,a\}}
{\io{$M_m$}{\inp, \outp, m}{i, \inp[i \coloneqq \mathit{qs}], \outp[i \coloneqq
(\ld, m(a)) \uplus \outp(i)], m}{}}

\inference
[Store]
{\inp(i) = \mathit{qs} \uplus \{\st,a,v\}}
{\io{$M_m$}{\inp, \outp, m}{i, \inp[i \coloneqq \mathit{qs}], \outp[i \coloneqq
(\st) \uplus \outp(i)], m[a\coloneqq v]}{}}%i, (\st, a,v)}}

\end{boxedminipage}
\caption{LTS for a simple memory}
\label{$M_m$}
\end{figure*}

\section{Respecifying Sequential Consistency with Communication}\label{sec:ref}

Realistic hardware systems do not implement the monolithic SC model of
Figure~\ref{Ref} directly.  Instead, there is usually a split between
processors and memory. Here we formalize that split using LTSes that 
compose to produce a system refining the simple SC model.

Figure~\ref{Pref$} defines an LTS for a simple \emph{decoupled} processor
(P$_\text{ref}$).  That is, memory does not appear within a processor's state, but
instead, to load from or store to an address, we send \emph{requests} to the
memory system and receive its \emph{responses}.  Both kinds of messages are
encoded as labels, $\text{ToM}$ for requests to memory and $\text{ToP}$ for
responses from memory back to the processor.

A state of P$_{\text{ref}}$ is a triple $(s, \pc, \wait)$, giving the current architectural
state $s$ and program counter $\pc$, as well as a Boolean flag $\wait$
indicating whether the processor is blocked waiting for a response from the
memory system.

As in the SC model, we change the state of the processor to $H$ whenever
$\dec$ returns $H$.  Note that there is some subtlety in the design of
this semantics regarding halting; we plan ahead to be sure we will be able to prove the
appropriate simulation facts.  Those proofs would be harder if, \eg{}
a processor could halt while it is still waiting for a memory response.

As initial state for system $P_\text{ref}$, we use $(s_0, \pc_0, \bot)$.

%\medskip

Figure~\ref{$M_m$} defines a simple memory, meant to be composed with P$_{\text{ref}}$
processors.  The input-output interface of this memory $M_m$ is largely what
one would expect from the P$_{\text{ref}}$ definition.  However, in anticipation of the
more complex processor we will define later, we add requests and responses for
\emph{speculative loads}.  These are load operations that a processor may
attempt, before it knows that the program will actually need the result.  Many
speculative loads from a single processor may be in flight at once, improving
performance via parallelism.  A request to memory like $(\spc\ld, t, a)$ asks
for the value of memory cell $a$, associating a \emph{tag} $t$ that the
processor can use to match responses to requests.  Those responses take the
form $(\spc\ld, t, v)$, giving the value $v$ of the requested memory address.

A memory state is a triple $(\inp, \outp, m)$, giving not just the
memory $m$ itself, but also queues $\inp$ and $\outp$ for receiving
processor requests and batching up responses to processors,
respectively.  We define the initial state of the $M_m$ LTS as
$(\emptyset, \emptyset, m_0)$, with empty queues.

Note that we have used $\uplus$ for inserting (or removing) from the buffers
$\inp$ and $\outp$. The intent is to convey that $\inp$ and $\outp$ are treated
as unordered sets. We will later use $\coloncolon$ for inserting (or removing)
from FIFO buffers. Intuitively, the unordered requests/responses set suffices
even for requests/responses for the same address because the processor always
waits for a response before sending the next request.
%\medskip

Now we can compose these LTSes to produce an implementation of SC.  Rule SpLoad
in Figure~\ref{$M_m$} will be irrelevant in this particular composition, since
P$_{\text{ref}}$ never does speculation.

For a system of $n$ processors, our decoupled SC implementation is
$\text{P$^n_{\text{ref}}$} + M_m$.

\begin{figure*}[t]
\begin{displaymath}
\upd((s, \pc, \_), \mathit{rs}) =
\left\{
\begin{array}{ll}
s', \pc' &: (\ld, v) \in \mathit{rs} \wedge \dec(pc, s) = (\ld, x, a) \wedge
\exec(pc, s, (\ld, x, v)) = (\pc', s'))\\
s', \pc' &: (\st) \in \mathit{rs} \wedge \dec(pc, s) = (\st, a, v) \wedge
\exec(pc, s, (\st)) = (\pc', s')\\
s, \pc &: \text{otherwise if } \dec(pc,s) \neq H\\
H &: \text{otherwise}
\end{array}
\right.
\end{displaymath}
\caption{Mapping of states from $(\text{P$^n_\text{ref}$} + M_m)$ to SC}
\label{smap}
\end{figure*}

\begin{theorem}
$\text{P$^n_{\text{ref}}$} + M_m \sqsubseteq SC$\label{scthm}
\end{theorem}
\begin{proof}
\vspace{-.1cm}
By induction on traces of the decoupled system.  We need to choose an
abstraction function $f$ from states of the complex system to states of
the simple system.  This function must be inductive in the
appropriate sense: a step from $s$ to $s'$ on the left of the
simulation relation must be matched by sequences of steps on the right
from $f(s)$ to $f(s')$.

Figure~\ref{smap} defines a suitable function $\upd$ applying to the
state of a single processor, and our abstraction function is the
natural lifting to $n$ processors.
The single-processor version $\upd$ maps
the decoupled state of a processor to an SC state, when also given the
set $rs$ of memory responses that have not been processed yet. If such a
response exists, then the updated $s', \pc'$ is the result of
receiving the response and performing the state updates on the current $s,
\pc$ pair.  We can also prove that multiple responses will not be present in
the $rs$ set at any point, intuitively because the processor goes into a wait
state as soon as it sends a request, and the single response, if present, will
correspond exactly to the instruction currently waiting for the response.
Therefore, the apparently ambiguous definition of $\upd$ is actually
well-founded.  The only interesting case in the proof arises when the memory adds a
response to $rs$; one has to show that the SC system can lead to the same
(projected) state.
\end{proof}

\section{Speculative Out-Of-Order Processor}\label{sec:ooo}

The previous section divided a multiprocessor system into two rather
simple components, processors and memories.  We are now ready to begin
developing more complex, optimized versions of each piece, starting
with processors.

We implement a \emph{speculative} processor, which may create many
simultaneous outstanding requests to the memory, as an optimization to
increase parallelism.  Our processor proof is in some sense generic
over correct speculation strategies.  We parametrize over two key
components of a processor design: a \emph{branch predictor}, which
makes guesses about processor-local control flow in advance of
%receiving enough memory responses to resolve
resolving conditional jumps; and a
\emph{reorder buffer}, which decides what speculative instructions (like memory loads)
are worth issuing at which moments (in effect \emph{reordering} later
instructions to happen before earlier instructions have finished).

The branch predictor is the simpler of the two components to model.
We indicate branch-predictor state with metavariable $\ppc$.
The operations on such state are $\curPpc(\ppc)$, to extract the
current program-counter prediction; $\nextPpc(\ppc)$, to ask the
predictor to step forward by one instruction; and $\setBp(\ppc, \pc)$,
to reset prediction to begin at a known-accurate position $\pc$.  It
turns out that we need to impose no explicit correctness criterion on
branch predictors.  The processor uses predictions only as hints, and
it always resets the predictor using $\setBp$ after detecting an
inaccurate hint.

The interface and formal contract of a reorder buffer are more
involved.  We write $\rob$ as a metavariable for reorder buffer
state.  The associated operations are:
\begin{itemize}
\item $\phi$, the state of an empty buffer.
\item $\add(\pc, \rob)$, which appends the program instruction at location $\pc$ to the list of instructions that the buffer is allowed to consider executing.
\item $\compute(\rob)$, which models a
step of computation inside the buffer, returning both an updated state
and an optional speculative load to issue. For instance, it invokes the $\dec$
and $\exec$ functions (as defined for SC) internally to obtain the next program
counter, state, \etc{} (but the actual states are not updated).
\item $\upd\ld(\rob,t,v)$, which informs the buffer that the memory
has returned result value $v$ for the speculative load with tag $t$.
\item $\commit(\rob)$, which returns the next instruction in serial
program order, if we have accumulated enough memory responses to execute it
accurately, or returns
$\epsilon$ otherwise.  When $\commit$ returns an instruction, it also
returns the associated program counter plus the next program counter
we would advance to afterward.
 Furthermore, the instruction is
extended with any relevant response from memory (used only for load
instructions, obtained through $\upd\ld$) and with the new architectural state (\eg{} register
file) after execution.
\item $\retire(\rob)$, which informs the buffer that its $\commit$
instruction was executed successfully, so it is time to move on to the
next instruction.
\end{itemize}

We postpone characterization of correct reorder buffers until after we
present the LTS for speculating processors.

\begin{figure*}[t]
\centering
\begin{boxedminipage}[c]{.95\textwidth}
\inference[Fetch]
{}
{\ioe{P$_\text{so}$}{s,\pc,\wait,\rob,\ppc}{s,\pc,\wait,\add(\curPpc(\ppc),\rob),\nextPpc(\ppc)}}

\inference[Compute]
{\compute(\rob) = (\rob',\epsilon)}
{\ioe{\text{P$_\text{so}$}}{s,\pc,\wait,\rob,\ppc}{s,\pc,\wait,\rob',\ppc}}

\inference[SpecLdReq]
{\compute(\rob) = (\rob',(\spc\ld,t,a))}
{\ior{\text{P$_\text{so}$}}{s,\pc,\wait,\rob,\ppc}{s,\pc,\wait,\rob',\ppc}
{\spc\ld, t, a}}

\inference[SpecLdRep]
{}
{\iol{\text{P$_\text{so}$}}{s,\pc,\wait,\rob,\ppc}
{s,\pc,\wait,\upd\ld(\rob, t, v),\ppc
}{\spc\ld,t,v}}

\inference[Abort]
{\commit(\rob) = (\pc',\_,\_) & \pc' \neq \pc}
{\ioe{\text{P$_\text{so}$}}{s,\pc,\wait,\rob,\ppc}{s,\pc,\wait,\phi,\setBp(\ppc,\pc)}}

\inference[Halt]
{\commit(\rob) = H}
{\ioe{\text{P$_\text{so}$}}{s,\pc,\bot,\rob,\ppc}{H}}

\inference[Nm]
{\commit(\rob) = (\pc,\pc',(\nm,s'))}
{\ioe{\text{P$_\text{so}$}}{s,\pc,\bot,\rob,\ppc}{s',\pc',\bot,\retire(\rob),\ppc}}

\inference[StReq]
{\commit(\rob) = (\pc,\pc',(\st,a,v,s'))}
{\ior{\text{P$_\text{so}$}}{s,\pc,\bot,\rob,\ppc}{s,\pc,\top,\rob,\ppc}
{\st, a, v}}

\inference[LdReq]
{\commit(\rob) = (\pc,\pc',(\ld,x,a,v,s'))}
{\ior{\text{P$_\text{so}$}}{s,\pc,\bot,\rob,\ppc}{s,\pc,\top,\rob,\ppc}{\ld, a}}
\inference[StRep]
{\commit(\rob) = (\pc,\pc',(\st,a,v,s'))}
{\iol{\text{P$_\text{so}$}}{s,\pc,\top,\rob,\ppc}{s',\pc',\bot,\retire(\rob),\ppc}{\st}}

\inference[LdRepGood]
{\commit(\rob) = (\pc,\pc',(\ld,x,a,v,s'))}
{\iol{\text{P$_\text{so}$}}{s,\pc,\top,\rob,\ppc}{s',\pc',\bot,\retire(\rob),\ppc}{\ld,v}}

\inference[LdRepBad]
{\commit(\rob) = (\pc,\pc',(\ld,x,a,v',s')) & v' \neq v & \exec(s,\pc,(\ld,x,v)) = (\pc'', s'')}
{\iol{\text{P$_\text{so}$}}{s,\pc,\top,\rob,\ppc}{s'',\pc'',\bot,\phi,\setBp(\ppc,\pc)}{\ld, v}}

\end{boxedminipage}

\caption{Speculating, out-of-order issue processor}
\label{Ooo}
\end{figure*}

Figure~\ref{Ooo} defines that LTS,
P$_{\texttt{so}}$. This processor may issue arbitrary speculative
loads, but it only ever \emph{commits} the next instruction in serial
program order. 
We will see the processor issue two kinds of loads, a speculative
load ($\spc\ld$) and a commit or real load ($\ld$).  To maintain SC, every speculative
load must have a matching verification load later on, and we maintain
the illusion that the program only depends on the results of
verification loads, which, along with stores, \emph{must be issued in
serial program order}.

When committing a
previously issued speculative load instruction, the associated speculative
memory load response is verified against the new commit load response. If the
resulting values do not match, the processor terminates all past uncommitted
speculation, by emptying the reorder buffer and resetting the
next predicted program counter in the branch predictor to the correct next value. This step mimics
the mechanism implemented in most modern-day processors, operating on the
conservative assumption that the processor may have performed arbitrary
incorrect speculation after receiving the inaccurate response from
memory.  (Note that here ``inaccurate'' does not mean ``wrong at
the time the response was received,'' but rather ``wrong at commit
time'' or ``correct before, but not anymore.'')

A full processor state is $(s, \pc, \wait, \rob, \ppc)$, comprising
architectural state, the program counter, a Boolean flag indicating
whether the processor is waiting for a memory response about an
instruction being committed, and the reorder buffer and branch
predictor states. Its initial state is given by $(s_0, \pc_0, \bot, \phi, \ppc_0)$.
The interface of this processor with memory (\ie{}
communication labels with $\toM, \toP$) is identical
to that of the reference processor, but now we can exercise a memory
system's transitions for speculative loads.

%The processor state is represented in the form $(s,\pc,\wait,\rob,\ppc)$, where
%the extra states compared to the decoupled processor of Figure~\ref{Pref$}. The
%additional states $\rob$ and $\ppc$ correspond to the reorder buffer holding
%the speculative operations and the speculative program counter indicating the
%next instruction to execute speculatively.

\begin{figure*}[t]
\begin{inv}
If $P_\text{so}$ reaches a state $(s, \pc, \wait, \rob, \ppc)$, \ie{}

$\io{P$_\text{so}^{*}$}{s_0,\pc_0,\bot,\phi,\ppc_0}{s,\pc,\wait,\rob,\ppc}{\eta} \Rightarrow$
\begin{displaymath}
\left\{
\begin{array}{ll}
\commit(\rob) = (\pc, \pc', (\nm, s')) &\Rightarrow
\exists x. \; \dec(s,\pc) = (\nm, x) \wedge \exec(s,\pc, (\nm,x)) =
(\pc', s')\\
\commit(\rob) = (\pc, \pc', (\ld, x, a, v, s')) & \Rightarrow
\dec(s,\pc) = (\ld, x,a) \wedge \exec(s,\pc, (\ld,x,v)) = (\pc', s')\\
\commit(\rob) = (\pc, \pc', (\st, a, v, s')) &\Rightarrow
\dec(s,\pc) = (\st, a,v) \wedge \exec(s,\pc, (\st)) =
(\pc', s')\\
\commit(\rob) = H & \Rightarrow
\dec(s, \pc) = H\\
\end{array}
\right.\end{displaymath}
\label{rob}
\end{inv}
\vspace{-.5cm}
\caption{Correctness of reorder buffer}
\label{robfig}
\end{figure*}

Finally, we impose a general correctness condition on reorder
buffers.  They must maintain Invariant~\ref{rob} in Figure~\ref{robfig}.
Intuitively, whenever the buffer claims in a $\commit$
output that a particular instruction is next to execute, causing
certain state changes, that instruction must really be next in line according
to how the program runs in the SC system, and running it must really cause
those state changes.

When this condition holds, we may conclude the correctness theorem for
out-of-order processors.  We use a trace-transformation function
$\noSpec$ that drops all speculative-load requests and responses.
See Definition~\ref{refines} for a review of how we use such
functions in framing trace refinement.  Intuitively, we prove that any
behavior by the speculating processor can be matched by the simple
processor, with speculative messages erased.

\begin{theorem}
\label{ocorrect}
$\text{P$_\text{so}$} \sqsubseteq_\noSpec$ P$_\text{ref}$
\end{theorem}
\begin{proof}
By induction on P$_\text{so}$ traces, using a proper abstraction
function (which drops the speculative messages and the $\rob$ and $\ppc$ states) to relate the two systems.  Invariant~\ref{rob} is crucial to
relate the reorder buffer's behavior with the simple in-order
execution of P$_\text{ref}$.
\end{proof}

\begin{corollary}
\label{ges}
$\text{P$^n_\text{so}$} \sqsubseteq_{\noSpec^n}$ P$_\text{ref}^n$
\end{corollary}
\begin{proof}
Direct consequence of Theorems~\ref{ocorrect} and~\ref{liftn} (where the
latter is one of our preliminary results about $n$-repetition).
\end{proof}

\section{Cache-Based Memory System}\label{sec:cc}

We now turn our attention toward a more efficient implementation of
memory.  It is physically infeasible to connect many processors
directly to a single memory with low enough latency to handle most
memory requests quickly enough.  Instead, each processor will have its
own \emph{L1 cache} of recently used memory cells.  Whenever possible, we
want to service memory requests from processors' own caches, only
contacting memory for addresses not found in the caches.  To further
avoid contacting main memory, we create additional higher-level
caches, connecting each L1 cache to an L2 cache, potentially
connecting L2 caches to L3, etc., bottoming out in main memory at the
root of our tree. The zoomed-in portion of the memory in Figure~\ref{zoom}
diagrams the basic layout.

Now we have concurrent interaction of many processors with many caches
with main memory, and the relationship with the original $M_m$ system
is far from direct.  However, this intricate concurrent execution is
crucial to hiding the latency of main-memory access.
Figure~\ref{cache} formalizes as an LTS $M_c$ the algorithm we implemented,
for providing the memory abstraction on top of a cache hierarchy.
We have what is called an \emph{invalidating directory-based hierarchical
cache-coherence protocol}.  It is a transliteration of a realistic
algorithm implemented in the Bluespec hardware description
language~\cite{Bluespec:TFRG}.

We now try to explain aspects of this transition system at an
intuitive level.  Thanks to the overall modular proof structure in
this paper, it is not essential to understand the details of the cache
system, to understand any other part of our development.  Therefore,
some readers may want to skip this explanation upon first reading,
skipping to the next section for a key theorem relating $M_c$ and $M_m$.

We describe a state of the system using 8 fields: $\dt$, $\ch$, $\s$, $\dst$,
$\wt$, $\dwt$, $\inp$, $\outp$. The $\inp$ and $\outp$ sets are the interfaces
to the processors and are exactly the same as in the simple memory LTS
(Figure~\ref{$M_m$}). The rest are summarized in Figure~\ref{fields}.
We use $\parent(c,p)$ to denote that $p$ is the parent of $c$ and $\leaf(c)$ to 
denote that $c$ is a leaf or L1 cache.

\begin{figure}
\centering
\begin{tabular}{|l|p{5.5cm}|}
\hline
$\dt(c,a)$ & Data in cache $c$ for address $a$\\
$\s(c,a)$ & Coherence state of cache $c$ for address $a$\\
$\dst(p,c,a)$ & Directory state of cache $p$ for child $c$ and address $a$\\
$\wt(c,a)$ & Gives cache $c$'s desired coherence state to upgrade to (for address $a$), or $\epsilon$ if none\\
$\dwt(p,c,a)$ & Gives cache $p$'s desired coherence state for its child $c$ to downgrade to (for address $a$), or $\epsilon$ if none\\
$\ch$ & Channels between nodes in the system\\
\hline
\end{tabular}
\caption{Fields in a cache-based memory system}
\label{fields}
\end{figure}

Fields $\dt, \s$ are maps over pairs of cache names and memory addresses.
The $\dt$ map assigns each pair a memory-cell value, and the $\s$ map
assigns each pair a coherence state.  A coherence state is $M, S$ or
$I$, broadly representing permissions to modify, read, or do nothing
with an address, respectively, the decreasing permissions denoted by $M > S > I$. This interpretation of permissions is
accurate at the leaf caches connected to the processors (\ie{} L1 caches),
but the interpretation is subtly different at the higher level caches.

Mapping $\wt$ goes from cache name and address to either $\epsilon$ or a
coherence state.  It denotes what permission the cache is waiting for,
if any.  That is, a cache may have decided to \emph{upgrade} its
coherence state for some address to a more permissive value, but the
cache is waiting for acknowledgment from another cache before
finalizing the upgrade.

Since we have a directory-based protocol, the $\s$ and
$\wt$ maps have mirror-image counterparts $\dst$ and $\dwt$.
The former maps triples of parent name, child name, and address to
coherence state. It represents the parent's notion of the
coherence state of the child for that address. We later prove that this notion
is always conservative, \ie{} if the parent assumes that a child does not have
a particular permission, then it is guaranteed in this system that the child
will not have that permission.  The $\dwt$ field also represents a mapping from parent
name, child name, and address to either $\epsilon$ or coherence state. It
denotes whether the parent is waiting for any downgrade response from the
corresponding child, and if so, the coherence state that the child must
downgrade to as well.

All communication channels in the system are represented by field $\ch$.
The channels are asymmetric as
follows. Channels going from parent to children (indexed by the names of the
parent and child) maintain FIFO order of messages between the same parent-child
pair. They carry both request messages (asking the child to downgrade) and
response messages (acknowledging that the child can upgrade). Channels going
from children to their parents are further split into request and response
channels, thus indexed by tuples (child name, parent name, request or response).
These channels need not maintain the FIFO order for messages between the same
child-parent pair (notice the use of $\coloncolon$ to add and remove elements
from the parent to child channel, as opposed to $\uplus$ for the 
child to parent channel). The reason is that only one request or one response is
sent for a particular address between a child-parent pair -- this invariant can
be proved for the LTS in  Figure \ref{cache}. The order between a request and
response message sent from the same child to its parent is lost, but our
protocol ensures that nothing bad happens as a result.  Notice rules
ParentRecvRsp and ParentRecvReq in Figure \ref{cache}, each of which looks at the
previous coherence state of the incoming message and compares it with its own
directory state to decide whether to process the message.

%\medskip

Here is an intuition on how the transitions work in the common case.  A cache
can decide, spontaneously, to upgrade its coherence state, in which case it
sends an upgrade request to its parent. The parent then makes a local decision
on whether to send a response to the requesting child or not, based on its
directory approximation and its own coherence state $\s$. If $\s$ is lower than
the requested upgrade, then it cannot handle the request, and instead must
decide to upgrade $\s$.  Once the parent's $\s$ is not lower than the requested
upgrade, it makes sure that the rest of its children are ``compatible'' with
the requested upgrade (given by $\compat$ definition below).  If not, the
parent must send requests to the incompatible children to downgrade. Finally, when
the $\s$'s upgrade and children's downgrade responses are all received, the
original request can be responded to. The transitions for sending requests are
allowed to happen spontaneously without any trigger. An L1 cache can process a
load request or a store request only if it is in at least the \Sh{} state or
\Mo{} state, respectively, for the requested address; otherwise it must
spontaneously request to upgrade.

\begin{defn}
\begin{displaymath}
\compat(p, c, x, a) = \left\{
\begin{array}{ll}
\forall c' \neq c. \; \dst(p,c',a) = I &: x = M\\
\forall c' \neq c. \; \dst(p,c',a) \le S &: x = S\\
\end{array}
\right.
\end{displaymath}
\end{defn}

The reason for having separate request and response channels for the same
child-parent pair is to avoid deadlocks. Throughout the protocol, we maintain
the invariant
that a downgrade request from the parent always gets priority, \ie{} if a child
is waiting for an upgrade response, it must still try to satisfy a downgrade
request from the parent. If both the parent and the child are waiting
for their respective responses, there will be a deadlock. This tie cannot
be broken arbitrarily: if the parent concedes and starts processing a child's
request, then it may have to send downgrade requests to other children to make
the states compatible, but the other children may have sent upgrade requests
and remain waiting, leading to a deadlock. The only way to break the
vicious waiting cycle is for the children to concede and respond when they get
downgrade requests. Of course, they have to downgrade the states of their own
children first, and this process goes on recursively until the downgrade requests
reach the leaf caches, which can respond immediately.

A complication arises because a cache can voluntarily decide to downgrade its
state.  This transition is used to model invalidation of cache lines to make
room for a different location (we permit a cache to downgrade from $M$ to $S$
instead of only to $I$).
As a result, the parent's $\dst$ and the
corresponding $\s$ of the child may go out of sync, leading to the parent
requesting a child to downgrade when it already has. To handle this situation,
the child has to drop the downgrade request when it has already downgraded to
the required state (Rule DropReq in Figure \ref{cache}), to avoid deadlocks
by not dequeuing the request.

%Dropping the request
%removes the entry from the buffer, allowing the child to process further messages
%from the parent -- otherwise there would have been a deadlock.

\begin{figure*}
\centering

\begin{boxedminipage}[c]{.95\textwidth}

\textbf{Processor/Memory Interface}

\inference
[Insert]
{}
{\iorn{$M_c$}{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}{\dt, \ch,
    \s, \dst, \wt, \dwt, \inp[i \coloneqq \{q\} \uplus \inp(i)],
    \outp}{i}{q}}
\vspace{.1in}
\inference
[Remove]
{\outp(i) = \mathit{rs} \uplus \{r\}}
{\ioln{$M_c$}{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp[i \coloneqq \mathit{rs}]}{i}{r}}
\vspace{.1in}
\inference[SpecLoad]
{\inp(c) = \{(\spc\ld, t, a)\} \uplus \rs & \s(c,a) \geq S}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
{
\dt, \ch, \s, \dst, \wt, \dwt, \inp[c \coloneqq \rs],\\
\outp[c \coloneqq \outp(c) \uplus \{(\spc\ld, t, \dt(c, a))\}]
}}
\vspace{.1in}
\inference[Load]
{\inp(c) = \{(\ld, a)\} \uplus \rs & \s(c,a) \geq S}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
{\dt, \ch, \s, \dst, \wt, \dwt, \inp[c \coloneqq \rs], \\
\outp[c \coloneqq \outp(c) \uplus \{(\ld, \dt(c, a))\}]}}
\vspace{.1in}
\inference[Store]
{\inp(c) = \{(\st, a, v)\} \uplus \rs & \s(c,a) \geq M}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
{\dt[(c, a) \coloneqq v], \ch, \s, \dst, \wt, \dwt, \inp[c \coloneqq \rs],
  \\
\outp[c \coloneqq \outp(c) \uplus \{(\st)\}]}}

\textbf{Child Upgrade}

\inference[ChildSendReq]
{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \\ \dwt, \inp, \outp}
{\dt, \ch[ (c, p, \req) \coloneqq (a, \s(c, a), x) \uplus \ch(c,p,\req)], \s, \dst,\\
\wt[(c,a)\coloneqq x], \dwt, \inp, \outp}}
\vspace{.1in}
\inference[ParentRecvReq]
{\parent(c,p) & \ch(c, p, \req) = \{(a, y, x)\} \uplus \rs & \s(p, a) \geq x
\\ \compat(p, c, x, a) & \dwt(p,c,a) = \epsilon & \dst(p,c,a) \le y}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst,\\ \wt, \dwt, \inp, \outp}
{\dt, \ch[(c,p,\req) \coloneqq \rs][(p, c) \coloneqq (\resp, (a, \dst(p,c,a), x,\\ \;\;\;\;
\ite{\dst(p,c,a) = I}{\dt(p,a)}{\_})) \coloncolon \ch(p,c)], \\ \s, \dst[(p,c, a)\coloneqq x],
\wt, \dwt, \inp, \outp}}
\vspace{.1in}
\inference[ChildRecvRsp]
{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\resp, (a, y, x, v))}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \\ \dwt, \inp, \outp}
{\dt[(c,a) \coloneqq \ite{y = I}{v}{\dt(c,a)}], \ch[(p,c) \coloneqq \rs], \s[(c,a) \coloneqq x], \dst,\\
\wt[(c,a)\coloneqq \ite{\wt(c,a) \le x}{\epsilon}{\wt(c,a)}], \dwt, \inp, \outp}}

\textbf{Parent Downgrade}

\inference[ParentSendReq]
{\parent(c,p) & \dst(p,c,a) > x & \dwt(p,c,a) = \epsilon}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \\ \dwt, \inp, \outp}
{\dt, \ch[(p, c)\coloneqq (\req, (a, \dst(p, c, a), x)) \coloncolon \ch(p,c)], \s, \dst,\\
\wt, \dwt[(p, c,a)\coloneqq x], \inp, \outp}}
\vspace{.1in}
\inference[ChildRecvReq]
{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\req, (a, y, x)) & (\forall i. \; \parent(i, c) \Rightarrow \dst(c, i, a) \le x)
& \s(c,a) > x}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst,\\ \wt, \dwt, \inp, \outp}
{\dt, \ch[(p,c) \coloneqq \rs][(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
\ite{\dst(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
\wt, \dwt, \inp, \outp}}
\vspace{.1in}
\inference[ParentRecvRsp]
{\parent(c,p) & \ch(c,p,\resp) = \{(a, y, x, v)\} \uplus \rs & \dst(p,c,a) = y}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \\ \dwt, \inp, \outp}
{\dt[(p,a) \coloneqq \ite{y = M}{v}{\dt(p,a)}], \ch[(c,p,\resp) \coloneqq \rs], \s, \dst[(p,c,a) \coloneqq x],\\
\wt, \dwt[(p,c,a)\coloneqq \ite{\dwt(p,c,a) \ge x}{\epsilon}{\dwt(p,c,a)}], \inp, \outp}}

\textbf{Voluntary downgrade for replacement}

\inference[VolResp]
{\parent(c,p) & (\forall i. \; \parent(i, c) \Rightarrow \dst(c, i, a) \le x)
& \s(c,a) > x}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst,\\ \wt, \dwt, \inp, \outp}
{\dt, \ch[(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
\ite{\s(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
\wt, \dwt, \inp, \outp}}

\textbf{Dropping request because of voluntary downgrade}

\inference[DropReq]
{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\req, (a, y, x)) & 
 \s(c,a) \le x}
{\ioxe{$M_c$}
{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
{\dt, \ch[(p,c) \coloneqq \rs], \s, \dst,
\wt, \dwt, \inp, \outp}}
\end{boxedminipage}

\caption{LTS for cache-coherent shared-memory system}
\label{cache}
\end{figure*}

\section{Proof that the MSI Protocol Preserves Store Atomicity}\label{sec:ccproof}
\label{safety}

We must prove the following theorem, \ie{} the cache-based system is sound with
respect to the simple memory.
\begin{theorem}
\label{ccorrect}
$M_c \sqsubseteq M_m$
\end{theorem}

The key property we need for this proof is what we define as
``store atomicity'' below. Throughout this section, we say
\emph{time} to denote the number of transitions that occurred before reaching
the specified state.

\begin{defn}
Store atomicity is defined as the following property:
For a load (or speculative load) request $\toM(\ld, a)$ (or $\toM(\spc\ld, t, a)$)
from any processor, the response sent at time $T$
$\toP(\ld, v)$ (or $\toP(\spc\ld, t, v)$) should be such that
\begin{enumerate}
\item $v = v_0$ (the initial value of any memory address) and no store
  request $\toM(\st, a, v')$ has been processed at any time $T'$ such
  that $T' < T$ or
\item There is a store request $\toM(\st, a, v)$ that was processed at time $T_q$ such that
$T_q < T$ and no other store
request $\toM(\st, a, v')$ was processed at any time $T'$ such that $T_q < T' < T$.
\end{enumerate}
\label{sa}
\end{defn}

We can easily show that any system obeying store atomicity
is sound with respect to $M_m$. Intuitively, the store atomicity
property precisely describes the property of a memory, \viz{} that if a store
happens, then that location gets updated such that any subsequent load gets
back the updated value.

The proof that $M_c$ obeys store atomicity is involved enough to deserve a
section's worth of explanation, which we now turn to.

We first reduce the proof of store atomicity to several lemmas. We then
present a key insight that we discovered for proving these lemmas in the form
of three fundamental invariants. We believe that other invalidation-based
protocols with inclusive caches (\ie{} those cache hierarchies where a parent
cache contains all the locations that any of its children contains) like MESI,
MOSI, and MOESI also satisfy our invariants and lemmas, making their verification
easier by reusing most of our MSI proof; we omit the discussion of extension of
this proof to other protocols for space reasons.

%\adamc{...and maybe we don't, for space reasons. \texttt{:-)}}

%We use the terms write to mean handling a store request, \ie{} executing the
%StoreReq transition and read to mean handling a load request, \ie{} executing
%the LoadReq transition.

The lemmas and invariants presented below hold after any number of transitions
starting from the initial state.

While we have proofs in Coq for each lemma, our discussion will be mainly
informal for brevity. All the proofs can be found in our supplementary material.
%\adamc{or just remind of the source code URL given earlier}

%\subsection{Invariants}
%\label{firstLevel}
Intuitively our protocol is correct because it is guaranteed that any cache that attempts to
handle a load request has the ``latest'' value.

\begin{lemma}
\textit{latestValue}:
At any time $T$, a cache $c$ that is \clean{} (defined below) for an address
$a$ will have the most up-to-date value for that address, \ie{}
\begin{enumerate}
\item $\dt(c,a) = v_0$ and no store request $\toM(\st, a, v)$ has been processed at
any time $T'$ such that $T' < T$ or
\item There is a store request $\toM(\st, a, v)$ that was processed at time $T_q$ such that
$T_q < T \wedge \dt(c,a) = v$ and no other store
request $\toM(\st, a, v')$ was processed at any time $T'$ such that $T_q < T' < T$.
\end{enumerate}
\label{latestValue}
\end{lemma}

\vspace{-.15in}

\begin{defn}
\textit{clean}: A cache is said to be \textit{clean} for an address $a$ if and only if
the state of the cache for that address is at least \Sh{} and the directory 
state for any of its children is at most \Sh.
\label{clean}
\end{defn}

It is pretty intuitive to prove the store atomicity property given
Lemma~\ref{latestValue}. A leaf cache can respond to a request only when it
has the required permission, at which point it is guaranteed to have the latest
value, because it is \textit{clean} (as it has no children).

Lemma~\ref{latestValue} requires further decomposition into more
lemmas.

\begin{lemma}
\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
that neither is an ancestor of the other in the cache-tree hierarchy (where
the ancestor relation is the reflexive-transitive closure of the parent relation),
for each address $a$, $\s(c_1,a)$ and $\s(c_2,a)$ are
\textit{compatible} (defined below).
\label{nonAncestorCompatible}
\end{lemma}
\begin{defn}
\textit{compatible}: Values of states $x_1$ and $x_2$ are \textit{compatible}
iff whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In.
\label{compatible}
\end{defn}

It is easy to see that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also
compatible if $x_1$ and $x_2$ are compatible.

\begin{lemma}
\textit{noTransitWrite}: Whenever data for an address $a$ is in transit (\ie{}
$\forall T. \; T_s \le T \le T_r$ where $T_s$ is the time of sending the data and
$T_r$ the time of receiving the data), no cache can process a store request for
$a$, and the data must be sent from a \textit{clean} cache.
\label{noTransitWrite}
\end{lemma}

\begin{lemma}
\textit{conservative}: The state of an address in a cache is never greater than
the corresponding entry in its parent's directory.
\label{conservative}
\end{lemma}

\begin{lemma}
\textit{localCompatibility1}:
The state of a cache is never less than its directory entry for any of
its children.
\label{localCompatibility1}
\end{lemma}
\begin{lemma}
\textit{localCompatibility2}: Let $c_1$ and $c_2$ be distinct caches with the
same parent $p$. Then, $\s(c_1, a)$ and $\s(c_2, a)$ are
\textit{compatible}.
\label{localCompatibility2}
\end{lemma}

Lemmas~\ref{conservative} and~\ref{localCompatibility1} imply a
$\ge$ relation between the states of a parent and its child for any address.
Since the ancestor relation is the reflexive-transitive closure of the parent
relation, the $\ge$ relation is preserved between the states of two caches $p$
and $c$ where $p$ is the ancestor of $c$.

Any two caches $c_1, c_2$ neither of which is an ancestor of the other have a
lowest common ancestor $x$, which has
two distinct children $x_1$ and $x_2$ such that they are the ancestors of $c_1$
and $c_2$ respectively. For any address $a$, $\s(x_1, a)$ and
$\s(x_2, a)$ are compatible by Lemma~\ref{localCompatibility2}. From
the previous paragraph about the relation between the states of a cache and its
ancestor, it follows that $c_1$ and $c_2$ are also compatible, proving
Lemma~\ref{nonAncestorCompatible}.

Using Lemmas~\ref{nonAncestorCompatible} and~\ref{noTransitWrite}, one can
prove Lemma~\ref{latestValue}. The proof is by induction on time $T$; we assume
that any \clean{} (node, address) pair has the latest value until time
$T$. Consider a (node, addr) pair at time $T+1$.
If the pair was \clean{} at $T$, then no cache,
save itself (because of Lemma~\ref{nonAncestorCompatible}), can process any
store request for that address, thus continuing to be \clean{} and having the latest value.
And if the pair was not \clean{} at time $T$, it will receive data from a cache
that is \clean{} for the (node, address) pair, and no store
request can be processed when the data is in transit (by
Lemma~\ref{noTransitWrite}). Therefore, by the induction hypothesis the cache will
continue to have the latest value whenever it is \clean{} for an address at time $T+1$.

\subsection{Key Insight for Proving Lemmas}

We present what we believe to be the key insight for proving any
cache-coherence protocol.  In fact, the request/response message semantics for
the MSI protocol was designed to satisfy the following three invariants.

\begin{inv}
\textit{stateChange}:
\begin{enumerate}
\item A cache downgrades its state for an address iff it sends a response for that address.
\item A parent upgrades the directory entry for an address for one of its
children iff it sends a response for that child and address.
\item The state or directory entry changes only on sending or
receiving a response $(\resp, (a, y, x, d))$, to $x$.
\end{enumerate}
\label{stateChange}
\end{inv}

\vspace{-.1in}

Invariant~\ref{stateChange} ensures that a response message acts as a sync
between a parent's directory entry for a child and the state of the child, as
both the sender and the receiver change their respective directory entry or
state to the same $x$ in a response $(\resp, (a, y, x, d))$

\begin{inv}
\textit{noCross}: A response from a parent to its child for an address cannot
be in transit while another response from the same child to the parent for the
same address is in transit.
\label{noCross}
\end{inv}

\begin{inv}
\textit{respFIFO}: Responses are received in the order in which they are sent,
for a source-destination pair, for the same address.
\label{respFIFO}
\end{inv}

\begin{figure}
\centering
\includegraphics[scale=.45]{resps}
\caption{An example showing a sequence of responses between a child and its parent}
\label{resps}
\end{figure}

Figure~\ref{resps} shows an example of the effect of
Invariants~\ref{stateChange}, \ref{noCross} and \ref{respFIFO}. The
directory entry of the parent or the state of the child changes on sending and
receiving a response, and the new state is shown at the points of sending or
receiving responses.  Both the cache and its parent follow the same sequence of
state and directory entry changes, but the receiver always lags slightly behind
the sender.  The receiver cannot send any response before receiving because of
Invariant~\ref{noCross}, thereby eliminating any scenarios where the lag
complicates reasoning.

Using these three invariants (Invariants~\ref{stateChange}, \ref{noCross} and
\ref{respFIFO}), it is easy to see that Lemma~\ref{conservative} holds.
Intuitively, going back to Figure~\ref{resps}, we can see that at all times,
either there is no response in flight, only responses from the parent are in
flight, or only responses from the child are in flight. A response from child
downgrades the state, and a response from parent upgrades the directory. Since
the other side simply lags behind, Lemma~\ref{conservative} holds.
Since Invariants~\ref{stateChange}, \ref{noCross} and \ref{respFIFO} deal with
just one cache's state for a single address and the corresponding directory
entry, these properties can also potentially be verified using model-checking.
Thus, if some other protocol is verified to obey Invariants~\ref{stateChange},
\ref{noCross} and \ref{respFIFO} using model-checking, then
Lemma~\ref{conservative} is already automatically verified for that
protocol.

Analyzing a parent-child pair will suffice in proving
Lemmas~\ref{localCompatibility1} and \ref{localCompatibility2}.  Proof of
Lemma~\ref{noTransitWrite} is similar to the proof of
Lemma~\ref{nonAncestorCompatible}.

%Figure \ref{something} gives the structure of the proof for Theorem \ref{ccorrect}
%\begin{figure*}
%\begin{prooftree}
%\AxiomC{stateChange}
%\AxiomC{noCross}
%\AxiomC{respFifo}
%\TrinaryInfC{conservative}
%\AxiomC{localCompatibility1}
%\AxiomC{localCompatibility2}
%\TrinaryInfC{nonAncestorCompatible}
%\AxiomC{noTransitWrite}
%\BinaryInfC{latestValue}
%\AxiomC{$\ldots$ (trivial)}
%\BinaryInfC{StoreAtomicity}
%\UnaryInfC{$M_c \sqsubseteq M_m$}
%\end{prooftree}
%\caption{Proving $M_c \sqsubseteq M_m$}
%\label{something}
%\end{figure*}

\section{The Final Result}\label{sec:finalresult}

With our two main results about optimized processors and memories, it
is now straightforward to complete the correctness proof of the
composed optimized system.

First, we need to know that, whenever the simple memory can generate
some trace of messages, it could also generate the same trace with all
speculative messages removed.  We need this property to justify the
introduction of speculation, during our final series of refinements
from the optimized system to SC.

\begin{theorem}
\label{mcorrect}
$M_m \sqsubseteq_{\noSpec^n} M_m$
\end{theorem}
\begin{proof}
By induction on traces, with an identity abstraction function.
\end{proof}

That theorem turns out to be the crucial ingredient to justify placing
a speculative processor in-context with simple memory.

\begin{theorem}
\label{ocorrect1}
$\text{P$^n_\text{so}$} + M_m \sqsubseteq \text{P$^n_\text{ref}$} + M_m$
\end{theorem}
\begin{proof}
Follows from Theorem \ref{liftplus} (one of our preliminary results
about $+$), Corollary \ref{ges}, and Theorem \ref{mcorrect}.
\end{proof}

The last theorem kept the memory the same while refining the
processor.  The next one does the opposite, switching out memory.

\begin{theorem}
\label{ocorrect2}
$\text{P$^n_\text{so}$} + M_c \sqsubseteq \text{P$^n_\text{so}$} + M_m$
\end{theorem}
\begin{proof}
Follows from Theorems \ref{ccorrect} and \ref{liftplus} plus reflexivity of $\sqsubseteq$ (Theorem \ref{transitive}).
\end{proof}

\begin{theorem}
\label{ofull}
$\text{P$^n_\text{so}$} + M_c \sqsubseteq SC$
\end{theorem}
\begin{proof}
We twice apply $\sqsubseteq$ transitivity (Theorem \ref{transitive}) to
connect Theorems \ref{ocorrect2}, \ref{ocorrect1}, and \ref{scthm}.
\end{proof}

%Figure \ref{overall} gives a summary the overall proof structure that we have
%employed to prove the correctness of the full system.
%
%\begin{figure*}
%\begin{prooftree}
%\AxiomC{$M_c \sqsubseteq M_m$}
%\AxiomC{$\sqsubseteq$ is reflexive}
%\AxiomC{$\begin{array}{l} \forall f, A, A', B, B'. A \sqsubseteq_f A' \wedge B \sqsubseteq_f B' \\\;\;\;\;\Rightarrow A+B \sqsubseteq A'+B'(\text{Theorem 3})\end{array}$}
%\TrinaryInfC{P$_\text{so}^n+M_c \sqsubseteq $ P$_\text{so}^n+M_m$}
%\end{prooftree}
%\begin{prooftree}
%\AxiomC{P$_\text{so} \sqsubseteq_\noSpec $ P$_\text{ref}$}
%\AxiomC{$\begin{array}{l}\forall f, A, B, n. A \sqsubseteq_f B \Rightarrow \\ \;\;\;\;\;\;\;A^n \sqsubseteq_{f^n} B^n(\text{Theorem 2})\end{array}$}
%\BinaryInfC{P$_\text{so}^n \sqsubseteq_\noSpec $ P$_\text{ref}^n$}
%\AxiomC{$M_m \sqsubseteq_\noSpec M_m$}
%\AxiomC{$\begin{array}{l} \forall f, A, A', B, B'. A \sqsubseteq_f A' \wedge B \sqsubseteq_f B' \\\;\;\;\;\Rightarrow A+B \sqsubseteq A'+B'(\text{Theorem 3})\end{array}$}
%\TrinaryInfC{P$_\text{so}^n+M_m \sqsubseteq $ P$_\text{ref}^n+M_m$}
%\end{prooftree}
%\begin{prooftree}
%\AxiomC{P$_\text{so}^n+M_c \sqsubseteq $ P$_\text{so}^n+M_m$}
%\AxiomC{P$_\text{so}^n+M_m \sqsubseteq $ P$_\text{ref}^n+M_m$}
%\AxiomC{P$_\text{ref}^n + M_m \sqsubseteq SC$}
%\AxiomC{$\sqsubseteq $ is transitive}
%\QuaternaryInfC{P$_\text{so}^n + M_c \sqsubseteq SC$}
%\end{prooftree}
%\caption{Overall proof structure}
%\label{overall}
%\end{figure*}

### 2014-10-15 TransitionsUsefulOld.tex
stuff

--- 2014-10-15 TransitionsUseful.tex
@@
 
-We first reduce the proof of store atomicity to several lemmas. We then
-present a key insight that we discovered for proving these lemmas in the form
-of three fundamental invariants. We believe that other invalidation-based
+We first reduce the proof of store atomicity to several invariants. We then
+present a key insight that we discovered for proving these invariants in the form
+of three more fundamental invariants. We believe that other invalidation-based
 protocols with inclusive caches (\ie{} those cache hierarchies where a parent
 cache contains all the locations that any of its children contains) like MESI,
-MOSI, and MOESI also satisfy our invariants and lemmas, making their verification
+MOSI, and MOESI also satisfy all of our invariants, making their verification
 easier by reusing most of our MSI proof; we omit the discussion of extension of
@@
 
-The lemmas and invariants presented below hold after any number of transitions
+The invariants presented below hold after any number of transitions
 starting from the initial state.
 
-While we have proofs in Coq for each lemma, our discussion will be mainly
+While we have proofs in Coq for each invariant, our discussion will be mainly
 informal for brevity. All the proofs can be found in our supplementary material.
@@
 
-%\subsection{Invariants}
-%\label{firstLevel}
+\subsection{Invariants}
+\label{firstLevel}
 Intuitively our protocol is correct because it is guaranteed that any cache that attempts to
@@
 
-\begin{lemma}
+\begin{inv}
 \textit{latestValue}:
@@
 \label{latestValue}
-\end{lemma}
+\end{inv}
 
@@
 It is pretty intuitive to prove the store atomicity property given
-Lemma~\ref{latestValue}. A leaf cache can respond to a request only when it
+Invariant~\ref{latestValue}. A leaf cache can respond to a request only when it
 has the required permission, at which point it is guaranteed to have the latest
@@
 
-Lemma~\ref{latestValue} requires further decomposition into more
-lemmas.
-
-\begin{lemma}
+Invariant~\ref{latestValue} requires further decomposition into more
+invariants.
+
+\begin{inv}
 \textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
@@
 \label{nonAncestorCompatible}
-\end{lemma}
+\end{inv}
 \begin{defn}
@@
 
-\begin{lemma}
+\begin{inv}
 \textit{noTransitWrite}: Whenever data for an address $a$ is in transit (\ie{}
@@
 \label{noTransitWrite}
-\end{lemma}
-
-\begin{lemma}
+\end{inv}
+
+\begin{inv}
 \textit{conservative}: The state of an address in a cache is never greater than
@@
 \label{conservative}
-\end{lemma}
-
-\begin{lemma}
+\end{inv}
+
+\begin{inv}
 \textit{localCompatibility1}:
@@
 \label{localCompatibility1}
-\end{lemma}
-\begin{lemma}
+\end{inv}
+\begin{inv}
 \textit{localCompatibility2}: Let $c_1$ and $c_2$ be distinct caches with the
@@
 \label{localCompatibility2}
-\end{lemma}
-
-Lemmas~\ref{conservative} and~\ref{localCompatibility1} imply a
+\end{inv}
+
+Invariants~\ref{conservative} and~\ref{localCompatibility1} imply a
 $\ge$ relation between the states of a parent and its child for any address.
@@
 and $c_2$ respectively. For any address $a$, $\s(x_1, a)$ and
-$\s(x_2, a)$ are compatible by Lemma~\ref{localCompatibility2}. From
+$\s(x_2, a)$ are compatible by Invariant~\ref{localCompatibility2}. From
 the previous paragraph about the relation between the states of a cache and its
 ancestor, it follows that $c_1$ and $c_2$ are also compatible, proving
-Lemma~\ref{nonAncestorCompatible}.
-
-Using Lemmas~\ref{nonAncestorCompatible} and~\ref{noTransitWrite}, one can
-prove Lemma~\ref{latestValue}. The proof is by induction on time $T$; we assume
+Invariant~\ref{nonAncestorCompatible}.
+
+Using Invariants~\ref{nonAncestorCompatible} and~\ref{noTransitWrite}, one can
+prove Invariant~\ref{latestValue}. The proof is by induction on time $T$; we assume
 that any \clean{} (node, address) pair has the latest value until time
@@
 If the pair was \clean{} at $T$, then no cache,
-save itself (because of Lemma~\ref{nonAncestorCompatible}), can process any
+save itself (because of Invariant~\ref{nonAncestorCompatible}), can process any
 store request for that address, thus continuing to be \clean{} and having the latest value.
@@
 request can be processed when the data is in transit (by
-Lemma~\ref{noTransitWrite}). Therefore, by the induction hypothesis the cache will
+Invariant~\ref{noTransitWrite}). Therefore, by the induction hypothesis the cache will
 continue to have the latest value whenever it is \clean{} for an address at time $T+1$.
 
-\subsection{Key Insight for Proving Lemmas}
+\subsection{Key Insight for Proving Invariants}
 
@@
 Using these three invariants (Invariants~\ref{stateChange}, \ref{noCross} and
-\ref{respFIFO}), it is easy to see that Lemma~\ref{conservative} holds.
+\ref{respFIFO}), it is easy to see that Invariant~\ref{conservative} holds.
 Intuitively, going back to Figure~\ref{resps}, we can see that at all times,
@@
 downgrades the state, and a response from parent upgrades the directory. Since
-the other side simply lags behind, Lemma~\ref{conservative} holds.
+the other side simply lags behind, Invariant~\ref{conservative} holds.
 Since Invariants~\ref{stateChange}, \ref{noCross} and \ref{respFIFO} deal with
@@
 \ref{noCross} and \ref{respFIFO} using model-checking, then
-Lemma~\ref{conservative} is already automatically verified for that
+Invariant~\ref{conservative} is already automatically verified for that
 protocol.
@@
 Analyzing a parent-child pair will suffice in proving
-Lemmas~\ref{localCompatibility1} and \ref{localCompatibility2}.  Proof of
-Lemma~\ref{noTransitWrite} is similar to the proof of
-Lemma~\ref{nonAncestorCompatible}.
+Invariants~\ref{localCompatibility1} and \ref{localCompatibility2}.  Proof of
+Invariant~\ref{noTransitWrite} is similar to the proof of
+Invariant~\ref{nonAncestorCompatible}.
 

### 2014-10-15 Abstract.tex
all the changes

@@
-We provide a modular proof of a high-performance implementation of Lamport's
-sequential consistency in a system with distributed memory. Unlike
-Lamport's reference operational model, a high-performance implementation
-processes many memory accesses concurrently. Processors employ
-\emph{speculative execution} to generate many concurrent loads to pipeline the
-memory system effectively. We define what it means for a speculative processor
-to be correct independently of the memory system. We then define a memory
-system with caches and show that our cache-coherence protocol preserves
-\emph{store atomicity} (independently of the processor), a property that all
-reasonable memories have in common.  We offer the first formal proof that when
-an atomic memory is connected to $n$ speculative processors, the system as a
-whole is sequentially consistent. The proof is composed modularly from proofs
-of the individual components. Additionally, the cache-coherent memory we
-present in this paper is the first fully verified system that supports
-arbitrary cache hierarchies. We describe all our components using labeled
-transition systems and machine-verify all our proofs using the Coq proof
+We provide a framework for modular verification of multi-processor hardware
+systems with respect to memory model specification. In this framework, we make
+use of a key observation that eventhough different memory models have different
+overall system behaviors, they all share a coherent cache hierarchy. We prove
+that this coherent cache hierarchy implements the \emph{store atomicity}
+property, which makes it behave like an atomic instantaneous memory with
+respect to the rest of the system. We further modularize the proof by defining
+what is meant for a processor to be correct independently of the memory system.
+As a concrete instance of our framework, we walk through a modular proof of
+high-performance multi-processor implementation of Lamport's Sequential
+Consistency model. The processors employ \emph{speculative execution} to
+generate many concurrent loads to the memory sub-system, while obeying our
+correctness criteria. We offer the first formal proof that when an atomic
+memory is connected to $n$ speculative processors, the system as a whole is
+sequentially consistent.  Additionally, the cache-coherent memory we present in
+this paper is the first fully verified system that supports arbitrary cache
+hierarchies. Our framework involves describing all our components using labeled
+transition systems and we machine-verify all our proofs using the Coq proof
 assistant.

### 2014-10-15 Introduction.tex
all the changes

@@
 
-Hardware designers recognize the challenge of implementing such systems, to such an extent that they have
-already become some of the most serious real-world adopters of formal methods.
-Hardware verification is dominated by model-checking, where abstraction
-techniques are used to reduce designs to finite state spaces, which can be
-explored exhaustively.  There are limits to the construction of sound
-abstractions, so verifications of, e.g., cache-coherence protocols have almost
-always treated systems with concrete topologies, involving particular finite
-numbers of caches and processors. For instance, explicit-state model checking tools
-like Murphi~\cite{murphi} or TLC~\cite{tlc} are only able to handle small and
+The goal of this work is to provide a framework for full verification of such
+complex hardware systems. Hardware designers already recognize the challenge of
+implementing such systems, to such an extent that they have already become some
+of the most serious real-world adopters of formal methods.  Hardware
+verification is dominated by model-checking, where abstraction techniques are
+used to reduce designs to finite state spaces, which can be explored
+exhaustively.  There are limits to the construction of sound abstractions, so
+verifications of, e.g., cache-coherence protocols have almost always treated
+systems with concrete topologies, involving particular finite numbers of caches
+and processors. For instance, explicit-state model checking tools like
+Murphi~\cite{murphi} or TLC~\cite{tlc} are only able to handle small and
 simplified systems, \eg{} a single-level cache hierarchy with three CPUs, two
@@
 the approach further, but they are still unable to handle realistic
-hierarchical protocols. We might hope to achieve higher
-assurance and understanding of design ideas by verifying \emph{infinite families}
-of hardware designs, which resist reduction to finite-state models.  A
-potential related benefit is reducing the computational resources required
-to formally verify a new design in the family, applying a more general theorem without
-much new analysis.
+hierarchical protocols. We might hope to achieve higher assurance and
+understanding of design ideas by verifying \emph{infinite families} of hardware
+designs, which resist reduction to finite-state models.  A potential related
+benefit is reducing the computational resources required to formally verify a
+new design in the family, applying a more general theorem without much new
+analysis.
 
-A related benefit of a deductive approach is \emph{modularity}.  It seems to
-be rare for hardware practitioners to verify full designs that include
-distinct functional components, like processors and memory systems.  One
-reason is the algorithmic overhead of exploring such large state spaces.
-If we can instead give localized correctness theorems for different
-subsystems, then we can compose these theorems in a black-box way to
+Modularity has long been understood as a key property for effective design and
+verification of such complex systems. For the designer, it allows a separation
+of concerns increasing robustness by allowing the behavior encapsulated by a
+modular boundary to be realized by multiple implementation any of which may be
+safely dropped into the system. It also allows a greater measure of
+parallelization of design improving development time. Similarly verification is
+simplified as modular interface agreements provide a set of natural lemmas to
+verify. This leads us towards a decomposition of the whole system verification
+task into lemmas of sub-systems, which can be composed in a black-box manner to
 produce full-system theorems.
 
-The traditional notion of modularity in hardware is closely tied to
-\emph{Finite State Machines} (FSMs), more specificially synchronous FSMs. The biggest
-problem with using FSM abstraction is that it restricts significantly the kinds of
-refinements that can be performed on a module. For example, if we change the
-timing of an adder so that it takes 2 clock cycles instead of 1, the whole
-system is likely to break. Even though big modules in hardware are often
-designed with asynchronous handshaking protocols, current
-hardware verification methodologies do not deal with timing refinements.
-In this paper, we use the formalism of \emph{labeled transition systems} that
-is well-known in process-calculus circles.  It turns out to be exactly the
-right way of capturing contracts on interaction between hardware components,
-enabling both modular refinement and verification.
+This works introduces an interface enabling true modular implementation and
+verification of shared memory systems. We want our modular system to permit
+\emph{black-box refinement}, that is, nothing beyond the interface contract can
+be assumed about other modules.  That is, we should be able to describe a
+modular decomposition where we may replace either the processor nodes, the
+memory with locally verified alternatives without changing the correctness of
+our whole system.  To this end, we identify a key property called \emph{store
+atomicity} shared by the memory component of every memory model. This property
+effectively restricts the behavior of the memory sub-system to that of an
+\emph{atomic instantaneous memory}. We prove that this property is in fact what
+is guaranteed using the cache coherence protocols in a multi-processor system
+containing a hierarchy of caches, enabling replacement of the cache hierarchy
+with an atomic instantaneous memory. Similarly, on the processor side, we
+define what is meant to be a correct single processor. These invariants about
+the processor and the memory sub-system remains constant across memory models.
+We show a concrete instance of this framework by proving that a highly
+speculative multi-processor system connected to a coherent cache hierarchy
+implements the sequential consistency memory model.
 
-%Proofs in this style are quite common in the software world.  The function
-%preconditions and postconditions of Hoare logic provide a canonical
-%example, and there have been many innovations in modular proof since Hoare's
-%original work.  Somehow the hardware verification world has taken less
-%advantage of modularity.  In this paper, we propose a new program in
-%hardware verification, based on modular proofs using the formalism of
-%\emph{labelled transition systems} that is well-known in process-calculus
-%circles.  It turns out to be exactly the right way of capturing contracts on
-%interaction between hardware components.
-
-Computer architects have intuitions about the contracts of classic
-subsystems.  However, these intuitions have not been formalized
-in enough detail to enable rigorous proofs.  In carrying out that exercise,
-we have also learned some things about architectural design that should
-be of interest outside a formal-methods context.
-
-Concretely, we provide a modularization of distributed shared-memory systems
-that makes minimal assumptions about networks, provides an abstraction for
-memory that allows high-performance implementation of any reasonable memory
-model, and allows a full range of realistic performance optimizations in
-processors without violating modularity. We want our modular system to permit
-\emph{black-box refinement}, that is, nothing beyond the interface contract can
-be assumed about other modules.
-
-In each instance we describe a reference model and an implementation model for
-a module using labeled transition systems and formally prove that the
-implementation is correct (sound) with respect to our reference. For example,
-our reference model for memory is monolithic, where each load or store request
-generates a (conceptually) instantaneous response, but where requests may be processed
-in any order. Such a model is intuitively \emph{store atomic} in the sense that
-the effect of a store is visible to all subsequent loads. We then describe a
-realistic cache-coherence protocol comprised of a hierarchy of caches, and
-formally prove that this system can never produce
-a behavior that is not allowed by the reference model. Similarly, we define a
-reference model for a processor that executes one instruction at a time and
-an implementation model that exploits speculation to increase parallelism,
-and we show that the speculative processor implements the reference processor,
-in a suitable sense.
-
-In the end, we prove a theorem about an infinite family of hardware systems,
-each system having an arbitrary number of speculative processors connected to
-a memory with an arbitrary hierarchy of caches.  All our proofs are mechanized
-in the Coq proof assistant, and they are modular in the sense of allowing
-further optimization to either processor or memory without needing to touch
-the proof of the other.  We characterize overall system behavior with a very
-simple transition system formalizing \emph{sequential consistency}, and we
-conjecture that the methodology generalizes to other popular memory models.
-
-\begin{figure}
-\centering
-\includegraphics[scale=.43]{zoom}
-\caption{Overall system}
-\label{zoom}
-\end{figure}
-
-Figure \ref{zoom} sketches the sort of hardware system that we have verified.
-We show the top-level system design at top right, and we zoom in on boxes standing for the
-memory system and a single processor.
-
-Memory is composed of a \emph{hierarchy of caches}, where each cache node (labeled
-like ``L1,'' ``L2,'' etc.) communicates only with its neighbors in the graph.
-Each processor has a dedicated L1 cache, from which we do our best to satisfy
-all memory requests, to avoid the latency of a round-trip with main memory.
-With an L1 cache miss, we may still find a hit in a parent cache below the
-level of main memory, realizing a smaller but still significant speedup.
-We have verified a \emph{directory-based} protocol for coordinating an arbitrary
-tree of caches, where each node stores a conservative approximation of its
-children's states.
-
-A processor is decomposed into several components.  We have the normal
-architectural state, such as values of registers.  Our proofs are generic over
-\emph{a family of instruction set architectures}, with parameters for opcode sets and
-functions for executing opcodes and decoding them from memory.  Other key
-components are a \emph{branch predictor}, which guesses at the control-flow path
-that a processor will follow, to facilitate speculation; and a
-\emph{reorder buffer (ROB)}, which decides which instructions along that path to
-try executing ahead of schedule.  Our proofs apply to an arbitrary branch predictor,
-and they work for any reorder buffer satisfying a simple semantic condition.
-
-Figure~\ref{proofs} gives the overall proof structure that we employ to verify
-the system of Figure~\ref{zoom}. As we will explain later, $P_\text{so}$
-represents the speculative out-of-order processor, $M_c$ the cache memory,
-$M_m$ the simple memory, $P_\text{ref}$ a simple decoupled processor, and
-SC the simple reference model having multiple processors executing each
-instruction atomically thus implementing sequential consistency.  Notation
-$A^n$ is for $n$ copies of system $A$ running in parallel.
-$\sqsubseteq$ is a refinement operator, capturing a suitable notion of when
-one system implements another.  We go into more detail on the symbols from
-the figure in the rest of the paper.
-
-Our current framework establishes theorems of the form ``if system $A$ has a run
-with some particular observable behavior, then system $B$ also has a run with
-the same behavior.''  In this sense, we say that $A$ correctly implements $B$.
-Other important properties, such as \emph{deadlock freedom} for $A$ (which
-might get stuck without producing any useful behavior), we leave for future
-work.
-
-\begin{figure}
-\includegraphics[scale=.45]{proofs}
-\caption{Overall proof structure}
-\label{proofs}
-\end{figure}
+Our proof methodology involves describing a reference model of a component and
+an implementation model for that component using labeled transition systems
+(LTS) and formally proving that the implementation is correct (sound) with
+respect to our reference. Using LTS for describing hardware has already been
+studied \cite{HoeArvind:TRSSynthesis1, Hoe:TCAD}, and there are compilers converting an LTS into efficient
+hardware \cite{BSV:LangRef, Bluespec:TFRG}.
+In fact, a cache coherence protocol implementation in Bluespec was discussed in \cite{DNA:CoherenceImplementation}.
+In fact, the cache coherence protocol that we
+study in this work is obtained from a direct transliteration of Bluespec code
+used to implement a multi-processor PowerPC system \cite{Khan:PowerPc}. The
+resulting implementation from that Bluespec code is pretty efficient; an 8-core
+system can run 55 million instructions overall per second on the BEE FPGA
+board, with a 2-level cache hierarchy. All our proofs are mechanized in the Coq
+proof assistant, and are included as the non-anonymous supplement to this
+paper.
 
@@
 \begin{itemize} 
+\item Identification of an interface between the processor and the memory
+components of a multi-processor system that enables modular implementation and
+verification.
+
+\item A general \emph{modular verification methodology} applicable to hardware
+  algorithms, based on \emph{labeled transition systems}.  We are able to verify
+  the several components of Figure~\ref{zoom} against general formal interfaces,
+  enabling mixing and matching of different realizations of each interface,
+  without doing any new proofs that peek beneath abstraction boundaries.
 
@@
   implementing relaxed memory models.
-
-\item A general \emph{modular verification methodology} applicable to hardware
-  algorithms, based on \emph{labeled transition systems}.  We are able to verify
-  the several components of Figure~\ref{zoom} against general formal interfaces,
-  enabling mixing and matching of different realizations of each interface,
-  without doing any new proofs that peek beneath abstraction boundaries.
-
-\item We use semantics to \emph{elucidate folklore ideas in computer architecture}.
-  For instance, we formalize a variety of processor speculation patterns with
-  the idea of speculative loads followed by verification loads, and we capture
-  a key requirement on memory systems as refinement of a simple transition
-  system.
-
 \end{itemize}
 
-As a side effect, we %debunk a common misconception about the semantics of the
-have shown a (rather surprising) architectural result about the semantics of
-the interface between processor and memory system. One can establish a
-sequentially consistent system even if the buffers between the memory system
-and the processor do not preserve message order. We prove as much by
-establishing simulation of sequential consistency without relying on an
-ordering assumption.
-
-\paragraph{Paper Organization:} In Section~\ref{sec:lts} we introduce
-our flavor of the labeled transition systems formalism,
-including a definition of trace refinement. In
-Section~\ref{sec:sc} we give a simple formal model of sequential consistency,
-which we decompose further for decoupled shared-memory systems in Section~\ref{sec:ref}.
-The following sections refine the two main subcomponents from Figure~\ref{zoom}.
-Section~\ref{sec:ooo} discusses definition and verification of a
-speculative processor model, and Sections~\ref{sec:cc} and~\ref{sec:ccproof}
-respectively define and prove our hierarchical cache-coherence protocol.
-Finally, in Section~\ref{sec:finalresult} we show
-the whole-system modular proof of our complex system, followed by
-discussion of implications in Section~\ref{sec:implications}
-and further discussion of related work and other conclusions in
+\paragraph{Paper Organization:} In Section~\ref{sec:lts} we introduce our
+flavor of the labeled transition systems formalism, including a definition of
+trace refinement. In Section~\ref{sec:store-atomicity}, we show a generic
+decomposition of any multi-processor system independent of the memory model
+that it implements, and discuss the store-atomicity property of the memory
+sub-component. In Section~\ref{sec:sc} we
+give a simple formal model of sequential consistency.  The
+following sections refine the two main subcomponents from Figure~\ref{zoom}.
+Section~\ref{sec:ooo} discusses definition and verification of a speculative
+processor model, and Sections~\ref{sec:cc} and~\ref{sec:ccproof} respectively
+define and prove our hierarchical cache-coherence protocol.  Finally, in
+Section~\ref{sec:finalresult} we show the whole-system modular proof of our
+complex system and a discussion of related work and other conclusions in
 Sections~\ref{relatedWork} and \ref{sec:conclusion}.
 
-The Coq source code of our proofs is included as the non-anonymous
-supplement to this paper.
+%% The goal of this work is to introduce an interface enabling true modular
+%% implementation and verification of shared memory systems. That is, we should be
+%% able to describe a modular decomposition where we may correct replace either
+%% the processor nodes, the memory with locally verified alternatives without
+%% changing the correctness of our whole system proof of correctness or
+%% restricting our ability to describe shared memory systems. As part of this we
+%% will describe the processor interface in a way which allows the full range of speculation
+
+%% We do so by showing the construction and initial refinements of a sequentially
+%% consistent shared memory system. We formally defining a minimal cache coherent
+%% speculation-friendly memory interface (\emph{Store Atomicity}), and use this as
+%% the interface for our memory subsystem. Using this we specify the notion of
+%% correct for a processor and prove that our system is correct assuming the
+%% correct implementation of these two objects.  We then construct both a naive
+%% reference processor and memory subsystem and provide proofs of correctness in
+%% Coq. We then refine both components into more realistic implementations
+%% including taking our memory subsystem to hardware-level description of a
+%% arbitrary-level hierarchical cache coherence implementation and verify these
+%% module. We show that the cross products of possible implementations are correct
+%% and their proof automatically from the modularly.
+
+%The traditional notion of modularity in hardware is closely tied to
+%\emph{Finite State Machines} (FSMs), more specificially synchronous FSMs. The biggest
+%problem with using FSM abstraction is that it restricts significantly the kinds of
+%refinements that can be performed on a module. For example, if we change the
+%timing of an adder so that it takes 2 clock cycles instead of 1, the whole
+%system is likely to break. Even though big modules in hardware are often
+%designed with asynchronous handshaking protocols, current
+%hardware verification methodologies do not deal with timing refinements.
+%In this paper, we use the formalism of \emph{labeled transition systems} that
+%is well-known in process-calculus circles.  It turns out to be exactly the
+%right way of capturing contracts on interaction between hardware components,
+%enabling both modular refinement and verification.
+
+%\begin{figure}
+%\centering
+%\includegraphics[scale=.43]{zoom}
+%\caption{Overall system}
+%\label{zoom}
+%\end{figure}
+%
+%Figure \ref{zoom} sketches the sort of hardware system that we have verified.
+%We show the top-level system design at top right, and we zoom in on boxes standing for the
+%memory system and a single processor.
+%
+%Memory is composed of a \emph{hierarchy of caches}, where each cache node (labeled
+%like ``L1,'' ``L2,'' etc.) communicates only with its neighbors in the graph.
+%Each processor has a dedicated L1 cache, from which we do our best to satisfy
+%all memory requests, to avoid the latency of a round-trip with main memory.
+%With an L1 cache miss, we may still find a hit in a parent cache below the
+%level of main memory, realizing a smaller but still significant speedup.
+%We have verified a \emph{directory-based} protocol for coordinating an arbitrary
+%tree of caches, where each node stores a conservative approximation of its
+%children's states.
+%
+%A processor is decomposed into several components.  We have the normal
+%architectural state, such as values of registers.  Our proofs are generic over
+%\emph{a family of instruction set architectures}, with parameters for opcode sets and
+%functions for executing opcodes and decoding them from memory.  Other key
+%components are a \emph{branch predictor}, which guesses at the control-flow path
+%that a processor will follow, to facilitate speculation; and a
+%\emph{reorder buffer (ROB)}, which decides which instructions along that path to
+%try executing ahead of schedule.  Our proofs apply to an arbitrary branch predictor,
+%and they work for any reorder buffer satisfying a simple semantic condition.
+%
+%Figure~\ref{proofs} gives the overall proof structure that we employ to verify
+%the system of Figure~\ref{zoom}. As we will explain later, $P_\text{so}$
+%represents the speculative out-of-order processor, $M_c$ the cache memory,
+%$M_m$ the simple memory, $P_\text{ref}$ a simple decoupled processor, and
+%SC the simple reference model having multiple processors executing each
+%instruction atomically thus implementing sequential consistency.  Notation
+%$A^n$ is for $n$ copies of system $A$ running in parallel.
+%$\sqsubseteq$ is a refinement operator, capturing a suitable notion of when
+%one system implements another.  We go into more detail on the symbols from
+%the figure in the rest of the paper.
+%
+%Our current framework establishes theorems of the form ``if system $A$ has a run
+%with some particular observable behavior, then system $B$ also has a run with
+%the same behavior.''  In this sense, we say that $A$ correctly implements $B$.
+%Other important properties, such as \emph{deadlock freedom} for $A$ (which
+%might get stuck without producing any useful behavior), we leave for future
+%work.
+%
+%\begin{figure}
+%\includegraphics[scale=.45]{proofs}
+%\caption{Overall proof structure}
+%\label{proofs}
+%\end{figure}
 

### 2014-10-15 RelatedWork.tex
all the changes

@@
 
-%There has been a lot of work in providing frameworks for axiomatic
-%specifications of memory models \cite{alglave2012formal,alglave2014herding}
-%and in providing axiomatic memory model specifications for real systems
-%through analyis of empirically observed behavior of certain ``litmus tests''
-%\cite{mador2012axiomatic,sarkar2012synchronising}. There has also been some
-%work in providing operational semantics that describes the observed behavior of
-%real hardware \cite{sarkar2011understanding}. Finally, there has been work
-%formally relating the axiomatic specification of TSO to operational semantics
-%of real x86 and Sparc machines \cite{x86tsocacm10}.
-%
+\xxx{TODO: Correct This}
+
+There has been a lot of work in providing frameworks for axiomatic
+specifications of memory models \cite{alglave2012formal,alglave2014herding}
+and in providing axiomatic memory model specifications for real systems
+through analyis of empirically observed behavior of certain ``litmus tests''
+\cite{mador2012axiomatic,sarkar2012synchronising}. There has also been some
+work in providing operational semantics that describes the observed behavior of
+real hardware \cite{sarkar2011understanding}. Finally, there has been work
+formally relating the axiomatic specification of TSO to operational semantics
+of real x86 and Sparc machines \cite{x86tsocacm10}.
+
 %The plethora of work in the area of memory models indeed suggests that this is
@@
 %
-%
+
 %BELOW IS UNEDITTED FROM OLD PAPER

### 2014-10-15 TransitionsUseful.tex
all the changes

@@
 functions keyed on numeric positions, and we write $\theta[i := v]$ for
-updating $n$-tuple $\theta$ to overwrite the $i$th position with value $v$, and we treat any label $(i, \epsilon)$ as $\epsilon$.
+updating $n$-tuple $\theta$ to overwrite the $i$th position with value $v$, and
+we treat any label $(i, \epsilon)$ as $\epsilon$.
 $$\inference[i]
@@
 
+\section{Decomposing a multi-processor system}
+\label{sec:store-atomicity}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.43]{zoom}
+\caption{Components of a multi-processors system}
+\label{zoom}
+\end{figure}
+
+Any multi-processor component can be divided into 3 components as shown in
+Figure \ref{zoom}. 
+We show the top-level system design at top right, and we zoom in on boxes standing for the
+memory system and a single processor ($P_i$).
+The processor component $P_i$ can be implemented in a
+variety of ways, from a simple in-order issue, to a complex out-of-order issue
+core, with aggressive speculation. The memory component is normally implemented
+using a hierarchy of caches, in order to increase the performance of the
+overall system, since the latency of accessing a memory directly is very slow
+compared to accessing a much smaller cache. The component between each
+processor and the global memory sub-system contains local buffers, $B_i$, specific to
+the processor $P_i$ they are connected to.
+
+\begin{figure}
+\centering
+\begin{boxedminipage}[c]{.48\textwidth}
+\inference
+[Load]
+{}
+{\io{$M_m$}{m}{m}{(\ld\req,a),(\ld\resp,m(a))}}
+
+\inference
+[Store]
+{}
+{\io{$M_m$}{m}{m[a\coloneqq v]}{(\st, a, v)}}
+
+\end{boxedminipage}
+\caption{LTS for a simple memory}
+\label{$M_m$}
+\end{figure}
+
+Independent of the memory models, the memory component always implements the
+\emph{store atomicity} property. This property can be understood using the
+behavior defined by the operational semantics of Figure \ref{$M_m$}. The
+different behaviors permitted by different memory models arise merely because
+of the (1) the reordering in the issue of memory instructions by the core $P_i$
+if it can issue instructions out-of-order and (2) the properties of the local
+buffers $B_i$ connected to each processor $P_i$. For instance, if $B_i$ was a
+store buffer, into which the stores from an in-order processor $P_i$ that are
+``committed'' are enqueued, and if any ``subsequent'' load from the same
+processor $P_i$ reads the latest value from the store buffer, then one can
+establish that such a multi-processor system implements the TSO
+\cite{x86tsocacm10} memory model.
+
+The memory component is composed of a \emph{hierarchy of caches}, where each
+cache node (labeled like ``L1,'' ``L2,'' etc.) communicates only with its
+neighbors in the graph.  Each processor has a dedicated L1 cache, from which we
+do our best to satisfy all memory requests, to avoid the latency of a
+round-trip with main memory.  With an L1 cache miss, we may still find a hit in
+a parent cache below the level of main memory, realizing a smaller but still
+significant speedup.  Therefore it is the responsibility of the hierarchy of
+caches, which form the memory sub-component, to implement the store atomicity
+property. In fact, as we will prove in Section \ref{sec:cc}, the purpose of the
+cache coherence protocol is to establish this invariant for the memory
+sub-component.  Concretely, we have verified a \emph{directory-based} protocol
+for coordinating an arbitrary tree of caches, where each node stores a
+conservative approximation of its children's states.
+
+As an instance of the above decomposition, we will prove that a multi-processor
+system with no local buffering in between the processor and the memory
+components indeed implements sequential consistency. We implement a highly
+speculative processor which executes instructions and issues loads out of
+order, but commits instructions (once some ``verification'' is done) in order.
+
+The processor itself can be decomposed into several components. In the zoomed
+in version of Figure \ref{zoom}, we show a highly speculative, out-of-order
+issue processor. We have the normal architectural state, such as values of
+registers.  Our proofs are generic over \emph{a family of instruction set
+architectures}, with parameters for opcode sets and functions for executing
+opcodes and decoding them from memory.  Other key components are a \emph{branch
+predictor}, which guesses at the control-flow path that a processor will
+follow, to facilitate speculation; and a \emph{reorder buffer (ROB)}, which
+decides which instructions along that path to try executing ahead of schedule.
+Our proofs apply to an arbitrary branch predictor, and they work for any
+reorder buffer satisfying a simple semantic condition.
+
+Figure~\ref{proofs} gives the overall proof structure that we employ to verify
+the system of Figure~\ref{zoom}. As we will explain later, $P_\text{so}$
+represents the speculative out-of-order processor, $M_c$ the cache memory,
+$M_m$ the simple memory, $P_\text{ref}$ a simple decoupled processor, and
+SC the simple composition of P$^n_\text{ref}+ M_m$ executing each
+instruction atomically thus implementing sequential consistency. Notation
+$A^n$ is for $n$ copies of system $A$ running in parallel.
+$\sqsubseteq$ is a refinement operator, capturing a suitable notion of when
+one system implements another.  We go into more detail on the symbols from
+the figure in the rest of the paper.
+
+Our current framework establishes theorems of the form ``if system $A$ has a run
+with some particular observable behavior, then system $B$ also has a run with
+the same behavior.''  In this sense, we say that $A$ correctly implements $B$.
+Other important properties, such as \emph{deadlock freedom} for $A$ (which
+might get stuck without producing any useful behavior), we leave for future
+work.
+
+\begin{figure}
+\includegraphics[scale=.45]{proofs}
+\caption{Overall proof structure}
+\label{proofs}
+\end{figure}
+
 \section{Specifying Sequential Consistency}\label{sec:sc}
@@
 
-Figure~\ref{Ref} defines an $n$-processor, sequentially consistent system as an
-LTS.  The definition is parametrized over details of an \emph{instruction set
-architecture} (ISA).  In particular, the ISA gives us some domains of
-architectural states $s$ (e.g., register files) and of program counters $\pc$.
-A function $\dec(s, \pc)$ figures out which instruction $\pc$ references in the
-current state, returning the instruction's ``decoded'' form.  A companion
-function $\exec(s, \pc, \mathit{dec})$ actually executes the instruction,
-returning the next program counter $\pc'$ and the new state $s'$.
+\begin{figure}
+\centering
+\begin{boxedminipage}[c]{.48\textwidth}
+\inference
+[Halt]
+{\dec(s,\pc) = H}
+{\io{P$_\text{ref}$}{s,\pc}{H}{}}
+
+\inference
+[NM]
+{\dec(s,\pc) = (\nm, x) & \exec(s, \pc, (\nm, x)) = (\pc', s')}
+{\ioe{P$_\text{ref}$}{s,\pc}{s', \pc'}}
+
+\inference
+[Ld]
+{\dec(s,\pc) = (\ld, x, a) & \exec(s, \pc, (\ld, x, v)) = (\pc', s')}
+{\io{P$_\text{ref}$}{s, \pc}
+{s', \pc'}{(\ld\req, a), (\ld\resp, v)}}
+
+\inference
+[St]
+{\dec(s,\pc) = (\st, a, v) & \exec(s, \pc, (\st)) = (\pc', s')}
+{\ior{P$_\text{ref}$}{s, \pc}
+{s', \pc'}{\st, a, v}}
+\end{boxedminipage}
+
+\caption{LTS for a single reference processor (P$_{\text{ref}}$)}
+\label{Pref$}
+\end{figure}
+
+Figure \ref{Pref$} shows the LTS for a single processor which will be used as a
+reference processor.  The definition is parametrized over details of an
+\emph{instruction set architecture} (ISA).  In particular, the ISA gives us
+some domains of architectural states $s$ (e.g., register files) and of program
+counters $\pc$.  A function $\dec(s, \pc)$ figures out which instruction $\pc$
+references in the current state, returning the instruction's ``decoded'' form.
+A companion function $\exec(s, \pc, \mathit{dec})$ actually executes the
+instruction, returning the next program counter $\pc'$ and the new state $s'$.
 
@@
 
+We can define sequential consistency using the simple memory defined in Figure
+\ref{$M_m$} as follows:
+\begin{defn}
+$SC \triangleq P_\text{ref} + M_m$
+\label{sc}
+\end{defn}
+
 We define the initial state of SC as $(\theta_0, m_0)$, where $m_0$ is
@@
 
-%\medskip
-
 This LTS encodes Lamport's notion of SC, where processors take turns executing
@@
 
-However, we are not interested merely in verifying the particular system $O$.
-Instead, we want to structure our proof \emph{modularly}, so that different
-common hardware components can be verified separately, and, in a black-box
-manner, we may assemble a correctness proof for any composition of these
-components.  For instance, we want to verify CPUs and memory systems
-separately, picking any realization of each to produce a verified system, with
-minimal new proof effort. As a more concrete example, memory systems should be
-verifiable independently of which weak memory model processors assume; we want
-to reuse memory proofs for different kinds of processors.
-
-To support this style of modular decomposition, we need to introduce a few
-intermediate systems.
-
-\begin{figure}
-\centering
-\begin{boxedminipage}[c]{.48\textwidth}
-\inference
-[Halt]
-{\dec(s,\pc) = H}
-{\io{P$_\text{ref}$}{s,\pc,\bot}{H}{}}
-
-\inference
-[NM]
-{\dec(s,\pc) = (\nm, x) & \exec(s, \pc, (\nm, x)) = (\pc', s')}
-{\ioe{P$_\text{ref}$}{s,\pc,\bot}{s', \pc', \bot}}
-
-\inference
-[LdReq]
-{\dec(s,\pc) = (\ld, x, a)}
-{\ior{P$_\text{ref}$}{s, \pc, \bot}
-{s, \pc, \top}{\ld, a}}
-
-\inference
-[StReq]
-{\dec(s,\pc) = (\st, a, v)}
-{\ior{P$_\text{ref}$}{s, \pc, \bot}
-{s, \pc, \top}{\st, a, v}}
-
-\inference
-[LdRep]
-{\dec(s,\pc) = (\ld, x, a) & \exec(s, \pc, (\ld, x, v)) = (\pc', s')}
-{\iol{P$_\text{ref}$}{s, \pc, \top}
-{s', \pc', \bot}{\ld, v}}
-
-\inference
-[StRep]
-{\dec(s,\pc) = (\st, a, v) & \exec(s, \pc, (\st)) = (\pc', s')}
-{\iol{P$_\text{ref}$}{s, \pc, \top}
-{s', \pc', \bot}{\st}}
-\end{boxedminipage}
-
-\caption{LTS for a simple decoupled processor (P$_{\text{ref}}$)}
-\label{Pref$}
-\end{figure}
-
-\begin{figure*}
-\centering
-\begin{boxedminipage}[c]{.48\textwidth}
-\inference
-[Insert]
-{}
-{\iorn{$M_m$}{\inp, \outp, m}{\inp[i \coloneqq \{q\} \uplus \inp(i)], \outp, m}{i}{q}}
-
-\inference
-[Remove]
-{\outp(i) = \mathit{rs} \uplus \{r\}}
-{\ioln{$M_m$}{\inp, \outp, m}{\inp, \outp[i \coloneqq \mathit{rs}], m}{i}{r}}
-
-\inference
-[SpLoad]
-{\inp(i) = \mathit{qs} \uplus \{\spc\ld,t,a\}}
-{\io{$M_m$}{\inp, \outp, m}{i, \inp[i \coloneqq \mathit{qs}], \outp[i \coloneqq
-(\spc\ld, t, m(a)) \uplus \outp(i)], m}{}}
-
-\inference
-[Load]
-{\inp(i) = \mathit{qs} \uplus \{\ld,a\}}
-{\io{$M_m$}{\inp, \outp, m}{i, \inp[i \coloneqq \mathit{qs}], \outp[i \coloneqq
-(\ld, m(a)) \uplus \outp(i)], m}{}}
-
-\inference
-[Store]
-{\inp(i) = \mathit{qs} \uplus \{\st,a,v\}}
-{\io{$M_m$}{\inp, \outp, m}{i, \inp[i \coloneqq \mathit{qs}], \outp[i \coloneqq
-(\st) \uplus \outp(i)], m[a\coloneqq v]}{}}%i, (\st, a,v)}}
-
-\end{boxedminipage}
-\caption{LTS for a simple memory}
-\label{$M_m$}
-\end{figure*}
-
-\section{Respecifying Sequential Consistency with Communication}\label{sec:ref}
-
-Realistic hardware systems do not implement the monolithic SC model of
-Figure~\ref{Ref} directly.  Instead, there is usually a split between
-processors and memory. Here we formalize that split using LTSes that 
-compose to produce a system refining the simple SC model.
-
-Figure~\ref{Pref$} defines an LTS for a simple \emph{decoupled} processor
-(P$_\text{ref}$).  That is, memory does not appear within a processor's state, but
-instead, to load from or store to an address, we send \emph{requests} to the
-memory system and receive its \emph{responses}.  Both kinds of messages are
-encoded as labels, $\text{ToM}$ for requests to memory and $\text{ToP}$ for
-responses from memory back to the processor.
-
-A state of P$_{\text{ref}}$ is a triple $(s, \pc, \wait)$, giving the current architectural
-state $s$ and program counter $\pc$, as well as a Boolean flag $\wait$
-indicating whether the processor is blocked waiting for a response from the
-memory system.
-
-As in the SC model, we change the state of the processor to $H$ whenever
-$\dec$ returns $H$.  Note that there is some subtlety in the design of
-this semantics regarding halting; we plan ahead to be sure we will be able to prove the
-appropriate simulation facts.  Those proofs would be harder if, \eg{}
-a processor could halt while it is still waiting for a memory response.
-
-As initial state for system $P_\text{ref}$, we use $(s_0, \pc_0, \bot)$.
-
-%\medskip
-
-Figure~\ref{$M_m$} defines a simple memory, meant to be composed with P$_{\text{ref}}$
-processors.  The input-output interface of this memory $M_m$ is largely what
-one would expect from the P$_{\text{ref}}$ definition.  However, in anticipation of the
-more complex processor we will define later, we add requests and responses for
-\emph{speculative loads}.  These are load operations that a processor may
-attempt, before it knows that the program will actually need the result.  Many
-speculative loads from a single processor may be in flight at once, improving
-performance via parallelism.  A request to memory like $(\spc\ld, t, a)$ asks
-for the value of memory cell $a$, associating a \emph{tag} $t$ that the
-processor can use to match responses to requests.  Those responses take the
-form $(\spc\ld, t, v)$, giving the value $v$ of the requested memory address.
-
-A memory state is a triple $(\inp, \outp, m)$, giving not just the
-memory $m$ itself, but also queues $\inp$ and $\outp$ for receiving
-processor requests and batching up responses to processors,
-respectively.  We define the initial state of the $M_m$ LTS as
-$(\emptyset, \emptyset, m_0)$, with empty queues.
-
-Note that we have used $\uplus$ for inserting (or removing) from the buffers
-$\inp$ and $\outp$. The intent is to convey that $\inp$ and $\outp$ are treated
-as unordered sets. We will later use $\coloncolon$ for inserting (or removing)
-from FIFO buffers. Intuitively, the unordered requests/responses set suffices
-even for requests/responses for the same address because the processor always
-waits for a response before sending the next request.
-%\medskip
-
-Now we can compose these LTSes to produce an implementation of SC.  Rule SpLoad
-in Figure~\ref{$M_m$} will be irrelevant in this particular composition, since
-P$_{\text{ref}}$ never does speculation.
-
 For a system of $n$ processors, our decoupled SC implementation is
 $\text{P$^n_{\text{ref}}$} + M_m$.
-
-\begin{figure*}[t]
-\begin{displaymath}
-\upd((s, \pc, \_), \mathit{rs}) =
-\left\{
-\begin{array}{ll}
-s', \pc' &: (\ld, v) \in \mathit{rs} \wedge \dec(pc, s) = (\ld, x, a) \wedge
-\exec(pc, s, (\ld, x, v)) = (\pc', s'))\\
-s', \pc' &: (\st) \in \mathit{rs} \wedge \dec(pc, s) = (\st, a, v) \wedge
-\exec(pc, s, (\st)) = (\pc', s')\\
-s, \pc &: \text{otherwise if } \dec(pc,s) \neq H\\
-H &: \text{otherwise}
-\end{array}
-\right.
-\end{displaymath}
-\caption{Mapping of states from $(\text{P$^n_\text{ref}$} + M_m)$ to SC}
-\label{smap}
-\end{figure*}
-
-\begin{theorem}
-$\text{P$^n_{\text{ref}}$} + M_m \sqsubseteq SC$\label{scthm}
-\end{theorem}
-\begin{proof}
-\vspace{-.1cm}
-By induction on traces of the decoupled system.  We need to choose an
-abstraction function $f$ from states of the complex system to states of
-the simple system.  This function must be inductive in the
-appropriate sense: a step from $s$ to $s'$ on the left of the
-simulation relation must be matched by sequences of steps on the right
-from $f(s)$ to $f(s')$.
-
-Figure~\ref{smap} defines a suitable function $\upd$ applying to the
-state of a single processor, and our abstraction function is the
-natural lifting to $n$ processors.
-The single-processor version $\upd$ maps
-the decoupled state of a processor to an SC state, when also given the
-set $rs$ of memory responses that have not been processed yet. If such a
-response exists, then the updated $s', \pc'$ is the result of
-receiving the response and performing the state updates on the current $s,
-\pc$ pair.  We can also prove that multiple responses will not be present in
-the $rs$ set at any point, intuitively because the processor goes into a wait
-state as soon as it sends a request, and the single response, if present, will
-correspond exactly to the instruction currently waiting for the response.
-Therefore, the apparently ambiguous definition of $\upd$ is actually
-well-founded.  The only interesting case in the proof arises when the memory adds a
-response to $rs$; one has to show that the SC system can lead to the same
-(projected) state.
-\end{proof}
 
@@
 \item $\add(\pc, \rob)$, which appends the program instruction at location $\pc$ to the list of instructions that the buffer is allowed to consider executing.
+\item $\get\spc\ld(\rob)$, which returns an optional speculative load to issue
 \item $\compute(\rob)$, which models a
-step of computation inside the buffer, returning both an updated state
-and an optional speculative load to issue. For instance, it invokes the $\dec$
+step of computation inside the buffer. For instance, it invokes the $\dec$
 and $\exec$ functions (as defined for SC) internally to obtain the next program
-counter, state, \etc{} (but the actual states are not updated).
+counter, state, \etc{} (only the speculative states are updated, not the actual states).
 \item $\upd\ld(\rob,t,v)$, which informs the buffer that the memory
@@
 {}
-{\ioe{P$_\text{so}$}{s,\pc,\wait,\rob,\ppc}{s,\pc,\wait,\add(\curPpc(\ppc),\rob),\nextPpc(\ppc)}}
+{\ioe{P$_\text{so}$}{s,\pc,\rob,\ppc}{s,\pc,\add(\curPpc(\ppc),\rob),\nextPpc(\ppc)}}
 
 \inference[Compute]
-{\compute(\rob) = (\rob',\epsilon)}
-{\ioe{\text{P$_\text{so}$}}{s,\pc,\wait,\rob,\ppc}{s,\pc,\wait,\rob',\ppc}}
-
-\inference[SpecLdReq]
-{\compute(\rob) = (\rob',(\spc\ld,t,a))}
-{\ior{\text{P$_\text{so}$}}{s,\pc,\wait,\rob,\ppc}{s,\pc,\wait,\rob',\ppc}
-{\spc\ld, t, a}}
-
-\inference[SpecLdRep]
-{}
-{\iol{\text{P$_\text{so}$}}{s,\pc,\wait,\rob,\ppc}
-{s,\pc,\wait,\upd\ld(\rob, t, v),\ppc
-}{\spc\ld,t,v}}
+{\get\spc\ld(\rob) = \epsilon}
+{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\compute(\rob),\ppc}}
+
+\inference[SpecLd]
+{\get\spc\ld(\rob) = (\spc\ld,t,a)}
+{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\upd\ld(\compute(\rob),t,v),\ppc}
+{(\ld\req, a), (\ld\resp, v)}}
 
@@
 {\commit(\rob) = (\pc',\_,\_) & \pc' \neq \pc}
-{\ioe{\text{P$_\text{so}$}}{s,\pc,\wait,\rob,\ppc}{s,\pc,\wait,\phi,\setBp(\ppc,\pc)}}
+{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\phi,\setBp(\ppc,\pc)}}
 
@@
 {\commit(\rob) = H}
-{\ioe{\text{P$_\text{so}$}}{s,\pc,\bot,\rob,\ppc}{H}}
+{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{H}}
 
@@
 {\commit(\rob) = (\pc,\pc',(\nm,s'))}
-{\ioe{\text{P$_\text{so}$}}{s,\pc,\bot,\rob,\ppc}{s',\pc',\bot,\retire(\rob),\ppc}}
-
-\inference[StReq]
+{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s',\pc',\retire(\rob),\ppc}}
+
+\inference[St]
 {\commit(\rob) = (\pc,\pc',(\st,a,v,s'))}
-{\ior{\text{P$_\text{so}$}}{s,\pc,\bot,\rob,\ppc}{s,\pc,\top,\rob,\ppc}
+{\io{\text{P$_\text{so}$}}{s',\pc',\retire(\rob),\ppc}{s,\pc,\rob,\ppc}
 {\st, a, v}}
 
-\inference[LdReq]
+\inference[LdGood]
 {\commit(\rob) = (\pc,\pc',(\ld,x,a,v,s'))}
-{\ior{\text{P$_\text{so}$}}{s,\pc,\bot,\rob,\ppc}{s,\pc,\top,\rob,\ppc}{\ld, a}}
-\inference[StRep]
-{\commit(\rob) = (\pc,\pc',(\st,a,v,s'))}
-{\iol{\text{P$_\text{so}$}}{s,\pc,\top,\rob,\ppc}{s',\pc',\bot,\retire(\rob),\ppc}{\st}}
-
-\inference[LdRepGood]
-{\commit(\rob) = (\pc,\pc',(\ld,x,a,v,s'))}
-{\iol{\text{P$_\text{so}$}}{s,\pc,\top,\rob,\ppc}{s',\pc',\bot,\retire(\rob),\ppc}{\ld,v}}
-
-\inference[LdRepBad]
+{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s',\pc',\retire(\rob),\ppc}{(\ld\req, a), (\ld\resp, v)}}
+
+\inference[LdBad]
 {\commit(\rob) = (\pc,\pc',(\ld,x,a,v',s')) & v' \neq v & \exec(s,\pc,(\ld,x,v)) = (\pc'', s'')}
-{\iol{\text{P$_\text{so}$}}{s,\pc,\top,\rob,\ppc}{s'',\pc'',\bot,\phi,\setBp(\ppc,\pc)}{\ld, v}}
-
+{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s'',\pc'',\phi,\setBp(\ppc,\pc)}{(\ld\req, v), (\ld\resp,v)}}
 \end{boxedminipage}
@@
 
-A full processor state is $(s, \pc, \wait, \rob, \ppc)$, comprising
-architectural state, the program counter, a Boolean flag indicating
-whether the processor is waiting for a memory response about an
-instruction being committed, and the reorder buffer and branch
-predictor states. Its initial state is given by $(s_0, \pc_0, \bot, \phi, \ppc_0)$.
-The interface of this processor with memory (\ie{}
-communication labels with $\toM, \toP$) is identical
+A full processor state is $(s, \pc, \rob, \ppc)$, comprising
+architectural state, the program counter, and the reorder buffer and branch
+predictor states. Its initial state is given by $(s_0, \pc_0, \phi, \ppc_0)$.
+The interface of this processor with memory is identical
 to that of the reference processor, but now we can exercise a memory
-system's transitions for speculative loads.
-
-%The processor state is represented in the form $(s,\pc,\wait,\rob,\ppc)$, where
-%the extra states compared to the decoupled processor of Figure~\ref{Pref$}. The
-%additional states $\rob$ and $\ppc$ correspond to the reorder buffer holding
-%the speculative operations and the speculative program counter indicating the
-%next instruction to execute speculatively.
+system's transitions even for speculative loads.
 
@@
 \begin{inv}
-If $P_\text{so}$ reaches a state $(s, \pc, \wait, \rob, \ppc)$, \ie{}
-
-$\io{P$_\text{so}^{*}$}{s_0,\pc_0,\bot,\phi,\ppc_0}{s,\pc,\wait,\rob,\ppc}{\eta} \Rightarrow$
+If $P_\text{so}$ reaches a state $(s, \pc, \rob, \ppc)$, \ie{}
+
+$\io{P$_\text{so}^{*}$}{s_0,\pc_0,\phi,\ppc_0}{s,\pc,\rob,\ppc}{\eta} \Rightarrow$
 \begin{displaymath}
@@
 We describe a state of the system using 8 fields: $\dt$, $\ch$, $\s$, $\dst$,
-$\wt$, $\dwt$, $\inp$, $\outp$. The $\inp$ and $\outp$ sets are the interfaces
-to the processors and are exactly the same as in the simple memory LTS
-(Figure~\ref{$M_m$}). The rest are summarized in Figure~\ref{fields}.
+$\wt$, $\dwt$ which are summarized in Figure~\ref{fields}.
 We use $\parent(c,p)$ to denote that $p$ is the parent of $c$ and $\leaf(c)$ to 
@@
 
-\inference
-[Insert]
-{}
-{\iorn{$M_c$}{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}{\dt, \ch,
-    \s, \dst, \wt, \dwt, \inp[i \coloneqq \{q\} \uplus \inp(i)],
-    \outp}{i}{q}}
-\vspace{.1in}
-\inference
-[Remove]
-{\outp(i) = \mathit{rs} \uplus \{r\}}
-{\ioln{$M_c$}{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp[i \coloneqq \mathit{rs}]}{i}{r}}
-\vspace{.1in}
-\inference[SpecLoad]
-{\inp(c) = \{(\spc\ld, t, a)\} \uplus \rs & \s(c,a) \geq S}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
-{
-\dt, \ch, \s, \dst, \wt, \dwt, \inp[c \coloneqq \rs],\\
-\outp[c \coloneqq \outp(c) \uplus \{(\spc\ld, t, \dt(c, a))\}]
-}}
-\vspace{.1in}
 \inference[Load]
-{\inp(c) = \{(\ld, a)\} \uplus \rs & \s(c,a) \geq S}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp[c \coloneqq \rs], \\
-\outp[c \coloneqq \outp(c) \uplus \{(\ld, \dt(c, a))\}]}}
+{\s(c,a) \geq S}
+{\io{$M_c$}
+{\dt, \ch, \s, \dst, \wt, \dwt}
+{\dt, \ch, \s, \dst, \wt, \dwt}{c,((\ld\req, a),(\ld\resp,v))}}
 \vspace{.1in}
 \inference[Store]
-{\inp(c) = \{(\st, a, v)\} \uplus \rs & \s(c,a) \geq M}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
-{\dt[(c, a) \coloneqq v], \ch, \s, \dst, \wt, \dwt, \inp[c \coloneqq \rs],
-  \\
-\outp[c \coloneqq \outp(c) \uplus \{(\st)\}]}}
+{s(c,a) \geq M}
+{\io{$M_c$}
+{\dt, \ch, \s, \dst, \wt, \dwt}
+{\dt[(c, a) \coloneqq v], \ch, \s, \dst, \wt, \dwt}{c,(\st,a,v)}}
 
@@
 {\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt, \inp, \outp}
+{\dt, \ch, \s, \dst, \wt, \\ \dwt}
 {\dt, \ch[ (c, p, \req) \coloneqq (a, \s(c, a), x) \uplus \ch(c,p,\req)], \s, \dst,\\
-\wt[(c,a)\coloneqq x], \dwt, \inp, \outp}}
+\wt[(c,a)\coloneqq x], \dwt}}
 \vspace{.1in}
@@
 {\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt, \inp, \outp}
+{\dt, \ch, \s, \dst,\\ \wt, \dwt}
 {\dt, \ch[(c,p,\req) \coloneqq \rs][(p, c) \coloneqq (\resp, (a, \dst(p,c,a), x,\\ \;\;\;\;
 \ite{\dst(p,c,a) = I}{\dt(p,a)}{\_})) \coloncolon \ch(p,c)], \\ \s, \dst[(p,c, a)\coloneqq x],
-\wt, \dwt, \inp, \outp}}
+\wt, \dwt}}
 \vspace{.1in}
@@
 {\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt, \inp, \outp}
+{\dt, \ch, \s, \dst, \wt, \\ \dwt}
 {\dt[(c,a) \coloneqq \ite{y = I}{v}{\dt(c,a)}], \ch[(p,c) \coloneqq \rs], \s[(c,a) \coloneqq x], \dst,\\
-\wt[(c,a)\coloneqq \ite{\wt(c,a) \le x}{\epsilon}{\wt(c,a)}], \dwt, \inp, \outp}}
+\wt[(c,a)\coloneqq \ite{\wt(c,a) \le x}{\epsilon}{\wt(c,a)}], \dwt}}
 
@@
 {\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt, \inp, \outp}
+{\dt, \ch, \s, \dst, \wt, \\ \dwt}
 {\dt, \ch[(p, c)\coloneqq (\req, (a, \dst(p, c, a), x)) \coloncolon \ch(p,c)], \s, \dst,\\
-\wt, \dwt[(p, c,a)\coloneqq x], \inp, \outp}}
+\wt, \dwt[(p, c,a)\coloneqq x]}}
 \vspace{.1in}
@@
 {\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt, \inp, \outp}
+{\dt, \ch, \s, \dst,\\ \wt, \dwt}
 {\dt, \ch[(p,c) \coloneqq \rs][(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
 \ite{\dst(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
-\wt, \dwt, \inp, \outp}}
+\wt, \dwt}}
 \vspace{.1in}
@@
 {\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt, \inp, \outp}
+{\dt, \ch, \s, \dst, \wt, \\ \dwt}
 {\dt[(p,a) \coloneqq \ite{y = M}{v}{\dt(p,a)}], \ch[(c,p,\resp) \coloneqq \rs], \s, \dst[(p,c,a) \coloneqq x],\\
-\wt, \dwt[(p,c,a)\coloneqq \ite{\dwt(p,c,a) \ge x}{\epsilon}{\dwt(p,c,a)}], \inp, \outp}}
+\wt, \dwt[(p,c,a)\coloneqq \ite{\dwt(p,c,a) \ge x}{\epsilon}{\dwt(p,c,a)}]}}
 
@@
 {\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt, \inp, \outp}
+{\dt, \ch, \s, \dst,\\ \wt, \dwt}
 {\dt, \ch[(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
 \ite{\s(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
-\wt, \dwt, \inp, \outp}}
+\wt, \dwt}}
 
@@
 {\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
+{\dt, \ch, \s, \dst, \wt, \dwt}
 {\dt, \ch[(p,c) \coloneqq \rs], \s, \dst,
-\wt, \dwt, \inp, \outp}}
+\wt, \dwt}}
 \end{boxedminipage}
@@
 We twice apply $\sqsubseteq$ transitivity (Theorem \ref{transitive}) to
-connect Theorems \ref{ocorrect2}, \ref{ocorrect1}, and \ref{scthm}.
+connect Theorems \ref{ocorrect2} and \ref{ocorrect1}, and use Definition \ref{sc} of $SC$
 \end{proof}
-
-%Figure \ref{overall} gives a summary the overall proof structure that we have
-%employed to prove the correctness of the full system.
-%
-%\begin{figure*}
-%\begin{prooftree}
-%\AxiomC{$M_c \sqsubseteq M_m$}
-%\AxiomC{$\sqsubseteq$ is reflexive}
-%\AxiomC{$\begin{array}{l} \forall f, A, A', B, B'. A \sqsubseteq_f A' \wedge B \sqsubseteq_f B' \\\;\;\;\;\Rightarrow A+B \sqsubseteq A'+B'(\text{Theorem 3})\end{array}$}
-%\TrinaryInfC{P$_\text{so}^n+M_c \sqsubseteq $ P$_\text{so}^n+M_m$}
-%\end{prooftree}
-%\begin{prooftree}
-%\AxiomC{P$_\text{so} \sqsubseteq_\noSpec $ P$_\text{ref}$}
-%\AxiomC{$\begin{array}{l}\forall f, A, B, n. A \sqsubseteq_f B \Rightarrow \\ \;\;\;\;\;\;\;A^n \sqsubseteq_{f^n} B^n(\text{Theorem 2})\end{array}$}
-%\BinaryInfC{P$_\text{so}^n \sqsubseteq_\noSpec $ P$_\text{ref}^n$}
-%\AxiomC{$M_m \sqsubseteq_\noSpec M_m$}
-%\AxiomC{$\begin{array}{l} \forall f, A, A', B, B'. A \sqsubseteq_f A' \wedge B \sqsubseteq_f B' \\\;\;\;\;\Rightarrow A+B \sqsubseteq A'+B'(\text{Theorem 3})\end{array}$}
-%\TrinaryInfC{P$_\text{so}^n+M_m \sqsubseteq $ P$_\text{ref}^n+M_m$}
-%\end{prooftree}
-%\begin{prooftree}
-%\AxiomC{P$_\text{so}^n+M_c \sqsubseteq $ P$_\text{so}^n+M_m$}
-%\AxiomC{P$_\text{so}^n+M_m \sqsubseteq $ P$_\text{ref}^n+M_m$}
-%\AxiomC{P$_\text{ref}^n + M_m \sqsubseteq SC$}
-%\AxiomC{$\sqsubseteq $ is transitive}
-%\QuaternaryInfC{P$_\text{so}^n + M_c \sqsubseteq SC$}
-%\end{prooftree}
-%\caption{Overall proof structure}
-%\label{overall}
-%\end{figure*}

### 2014-10-15 RelatedWork.tex
related work

--- 2014-10-15 RelatedWork.tex
@@
 
+\xxx{TODO: Correct This}
+
+There has been a lot of work in providing frameworks for axiomatic
+specifications of memory models based on empirically observed behavior of
+certain ``litmus
+tests''\cite{alglave2012formal,alglave2014herding,Alglave:FMSD,Alglave:2010,Alglave11,Alglave:TACAS}.
+There has been some work in providing operational and axiomatic semantics that
+describes the observed behavior (based on certain ``litmus tests'' of real
+hardware
+\cite{mador2012axiomatic,sarkar2011understanding,sarkar2012synchronising}.
+Finally, there has been work formally relating the axiomatic specification of
+TSO to operational semantics \cite{x86tsocacm10}.
+
 %There has been a lot of work in providing frameworks for axiomatic
@@
 %and in providing axiomatic memory model specifications for real systems
-%through analyis of empirically observed behavior of certain ``litmus tests''
+%through analysis of empirically observed behavior of certain ``litmus tests''
 %\cite{mador2012axiomatic,sarkar2012synchronising}. There has also been some
@@
 %real hardware \cite{sarkar2011understanding}. Finally, there has been work
-%formally relating the axiomatic specification of TSO to operational semantics
-%of real x86 and Sparc machines \cite{x86tsocacm10}.
-%
+%formally relating the axiomatic specification of TSO to operational semantics\cite{x86tsocacm10}.
+
 %The plethora of work in the area of memory models indeed suggests that this is

### 2014-10-15 RelatedWork.tex

@@
 certain ``litmus
-tests''\cite{alglave2012formal,alglave2014herding,Alglave:FMSD,Alglave:2010,Alglave11,Alglave:TACAS}.
+tests''\cite{alglave2012formal,Alglave:TACAS,alglave2014herding,Alglave:FMSD,Alglave:2010,Alglave11}.
 There has been some work in providing operational and axiomatic semantics that

### 2014-10-15 RelatedWork.tex
small edits

@@
 certain ``litmus
-tests''\cite{alglave2012formal,Alglave:TACAS,alglave2014herding,Alglave:FMSD,Alglave:2010,Alglave11}.
+tests''\cite{Alglave11,Alglave:FMSD,alglave2012formal,Alglave:2010,Alglave:TACAS,alglave2014herding}.
 There has been some work in providing operational and axiomatic semantics that

### 2014-10-19 Abstract.tex
Replace abstract

--- 2014-10-15 Abstract.tex
@@
-We provide a modular proof of a high-performance implementation of Lamport's
-sequential consistency in a system with distributed memory. Unlike
-Lamport's reference operational model, a high-performance implementation
-processes many memory accesses concurrently. Processors employ
-\emph{speculative execution} to generate many concurrent loads to pipeline the
-memory system effectively. We define what it means for a speculative processor
-to be correct independently of the memory system. We then define a memory
-system with caches and show that our cache-coherence protocol preserves
-\emph{store atomicity} (independently of the processor), a property that all
-reasonable memories have in common.  We offer the first formal proof that when
-an atomic memory is connected to $n$ speculative processors, the system as a
-whole is sequentially consistent. The proof is composed modularly from proofs
-of the individual components. Additionally, the cache-coherent memory we
-present in this paper is the first fully verified system that supports
-arbitrary cache hierarchies. We describe all our components using labeled
-transition systems and machine-verify all our proofs using the Coq proof
-assistant.
+We present a new framework for modular verification of hardware designs
+in the style of the Bluespec language.  That is, we formalize the idea of
+components in a hardware design, with well-defined input and output channels;
+and we show how to specify and verify components individually, with
+machine-checked proofs in the Coq proof assistant.  In doing so, we import
+to the world of hardware those modular reasoning principles that are
+well-known in software verification.  As a demonstration, we verify a
+fairly realistic implementation of a multicore shared-memory system.  The
+main modularity split is between a memory-system component and a processor
+component.  Both include nontrivial optimizations, with the memory system
+employing an arbitrary hierarchy of cache nodes that communicate with each
+other concurrently, and with the processor doing speculative execution of
+many concurrent read operations.  Nonetheless, we prove that the combined
+system implements sequential consistency.  One interesting aspect of the proof
+is that the memory component is well-specified to be reused for systems
+implementing a variety of weak memory models.  To our knowledge, our proof
+of that memory system is the first machine verification of a cache-coherence
+protocol parameterized over an arbitrary cache hierarchy, and our full-system
+proof is the first machine verification of sequential consistency for a
+multicore hardware design that includes memory and processors.

### 2014-10-19 Introduction.tex
Pass through Introduction

@@
 complex hardware systems. Hardware designers already recognize the challenge of
-implementing such systems, to such an extent that they have already become some
+implementing these systems, to such an extent that they have already become some
 of the most serious real-world adopters of formal methods.  Hardware
@@
 addresses, and one bit of data. Symbolic model checking techniques have fared
-slightly better; McMillan \etal{} have verified a 2-level MSI protocol based on
+slightly better: McMillan \etal{} have verified a 2-level MSI protocol based on
 the Gigamax distributed multiprocessor using SMV~\cite{gigamax}. Optimizations
@@
 verification of such complex systems. For the designer, it allows a separation
-of concerns increasing robustness by allowing the behavior encapsulated by a
-modular boundary to be realized by multiple implementation any of which may be
-safely dropped into the system. It also allows a greater measure of
-parallelization of design improving development time. Similarly verification is
-simplified as modular interface agreements provide a set of natural lemmas to
-verify. This leads us towards a decomposition of the whole system verification
-task into lemmas of sub-systems, which can be composed in a black-box manner to
+of concerns, increasing robustness by allowing the behavior encapsulated by a
+modular boundary to be realized by multiple implementations, any of which may be
+dropped into the system safely. It also allows a greater measure of
+parallelization of human design effort, improving development time. Similarly, verification is
+simplified, as modular interface agreements provide a natural lemma structure.
+This leads us towards a decomposition of the whole system verification
+task into lemmas of subsystems, which can be composed in a black-box manner to
 produce full-system theorems.
 
-This works introduces an interface enabling true modular implementation and
-verification of shared memory systems. We want our modular system to permit
-\emph{black-box refinement}, that is, nothing beyond the interface contract can
-be assumed about other modules.  That is, we should be able to describe a
-modular decomposition where we may replace either the processor nodes, the
-memory with locally verified alternatives without changing the correctness of
-our whole system.  To this end, we identify a key property called \emph{store
-atomicity} shared by the memory component of every memory model. This property
-effectively restricts the behavior of the memory sub-system to that of an
-\emph{atomic instantaneous memory}. We prove that this property is in fact what
-is guaranteed using the cache coherence protocols in a multi-processor system
-containing a hierarchy of caches, enabling replacement of the cache hierarchy
-with an atomic instantaneous memory. Similarly, on the processor side, we
-define what is meant to be a correct single processor. These invariants about
-the processor and the memory sub-system remains constant across memory models.
-We show a concrete instance of this framework by proving that a highly
-speculative multi-processor system connected to a coherent cache hierarchy
-implements the sequential consistency memory model.
+In this paper, we introduce \textbf{a new approach to modular,
+  deductive verification of hardware designs} at the highest level of
+rigor, backed by proofs using the Coq proof assistant.  As a
+challenging case study, we focus on \textbf{verifying that a
+  particular infinite family of multicore, shared-memory systems
+  implements sequential consistency}.  The proof is parametrized over
+an unknown number of processors connected to an arbitrary memory
+hierarchy with an unknown number of caches in an unknown number of
+layers (e.g., L1, L2).  In our design, processors and memory systems
+both independently employ intricate optimizations that exploit
+opportunities for parallelism.  We are able to prove that each of
+those two main components still implements strong enough semantics to
+support SC, and then we compose those theorems into a result for the
+full system.  Either component may be optimized further without
+requiring any changes to the implementation, specification, or proof
+of the other.
 
-Our proof methodology involves describing a reference model of a component and
-an implementation model for that component using labeled transition systems
-(LTS) and formally proving that the implementation is correct (sound) with
-respect to our reference. Using LTS for describing hardware has already been
-studied \cite{HoeArvind:TRSSynthesis1, Hoe:TCAD}, and there are compilers converting an LTS into efficient
-hardware \cite{BSV:LangRef, Bluespec:TFRG}.
-In fact, a cache coherence protocol implementation in Bluespec was discussed in \cite{DNA:CoherenceImplementation}.
-In fact, the cache coherence protocol that we
-study in this work is obtained from a direct transliteration of Bluespec code
-used to implement a multi-processor PowerPC system \cite{Khan:PowerPc}. The
-resulting implementation from that Bluespec code is pretty efficient; an 8-core
-system can run 55 million instructions overall per second on the BEE FPGA
-board, with a 2-level cache hierarchy. All our proofs are mechanized in the Coq
-proof assistant, and are included as the non-anonymous supplement to this
-paper.
+To structure our hardware verification framework, we import
+\emph{labeled transition systems (LTSes)}, a well-studied approach from the
+world of concurrent software modeling.  As our unified notion of
+specification and proof, we adopt \emph{trace refinement}, which
+captures when one concurrent system can only produce observable IO
+behaviors that another system could also produce.  Each of our
+realistic hardware components is associated with a simpler ``reference
+implementation'' that serves as a specification, and we prove that
+each realistic component refines its spec.  Thereafter, it is sound to
+use the simpler spec component in reasoning about the behavior of the
+system.
+
+Each key component of our case-study system has a simple spec: the
+ideal memory responds instantly and atomically to each load or store
+request, and the ideal processor executes instructions in order with
+no speculation.  The optimized implementation of each is much more
+complicated: the cache-hierarchy memory system potentially allows
+every cache to be busy simultaneously computing and sending messages,
+and the optimized speculative processor sends multiple load requests
+to the memory at once, without waiting for responses first.  The meat
+of the verification is in showing that each optimized component
+refines its simpler counterpart.  After each such proof, we may in
+effect substitute the optimized version for the simple version in a
+black-box way within a design, with a full guarantee of soundness.
+
+LTSes as hardware descriptions are an established
+idea~\cite{HoeArvind:TRSSynthesis1, Hoe:TCAD}, and there are
+compilers that convert LTSes into efficient hardware.  Our work is
+based on the Bluespec language~\cite{BSV:LangRef, Bluespec:TFRG},
+which served as the model for the formalism of this paper.  Bluespec
+models hardware components as atomic rules of a transition system over
+state elements, and there is a commercial compiler from such code into
+register transfer-level circuits (i.e., Verilog code) with competitive
+performance.  Our cache-coherent memory system is directly
+transliterated from a Bluespec
+implementation~\cite{DNA:CoherenceImplementation} used to
+implement a multi-processor PowerPC system~\cite{Khan:PowerPc}. The
+hardware synthesized from that implementation is rather efficient:
+an 8-core system can run 55 million instructions per second on
+the BEE FPGA board\cite{}, with a 2-level cache hierarchy.
 
@@
   algorithms, based on \emph{labeled transition systems}.  We are able to verify
-  the several components of Figure~\ref{zoom} against general formal interfaces,
+  the several components of Figure~\ref{zoom} \adam{OK to have a
+    forward reference to this figure?} against general formal interfaces,
   enabling mixing and matching of different realizations of each interface,
@@
 trace refinement. In Section~\ref{sec:store-atomicity}, we show a generic
-decomposition of any multi-processor system independent of the memory model
-that it implements, and discuss the store-atomicity property of the memory
+decomposition of any multi-processor system independently of the memory model
+that it implements, and we discuss the store-atomicity property of the memory
 sub-component. In Section~\ref{sec:sc} we
@@
 Section~\ref{sec:finalresult} we show the whole-system modular proof of our
-complex system and a discussion of related work and other conclusions in
+complex system, finishing with discussion of related work and other conclusions in
 Sections~\ref{relatedWork} and \ref{sec:conclusion}.

### 2014-10-20 TransitionsUseful.tex
more random changes. now figures are correct
small edits
Pass up to Section 4

@@
 
-For LTS $A$, we write $\io{A}{s}{s'}{\ell}$ as shorthand for $(s, \ell, s') \in
-A$, and we write $A_0$ for $A$'s initial state. The intuition is that $A$ is
-one process within a concurrent system. The label $\ell$ from set $\mathcal L$
-of labels is produced when $A$ participates in some IO exchange with another
-process; otherwise it is an empty or ``silent'' label $\epsilon$. As a
-shorthand, we sometimes omit labels for $\epsilon$ steps.
+For LTS $A$, we write $\io{A}{s}{s'}{\ell}$ as shorthand for $(s,
+\ell, s') \in A$, and we write $A_0$ for $A$'s initial state. The
+intuition is that $A$ is one process within a concurrent system. The
+label $\ell$ from set $\mathcal L$ of labels is produced when $A$
+participates in some IO exchange with another process; otherwise it is
+an empty or ``silent'' label $\epsilon$.  As a shorthand, we sometimes
+omit labels for $\epsilon$ steps.  Note that the IO exchange can be
+input, output, or both, and its interpretation is dependent on the
+context.
 
@@
 labels, or erase them altogether. Let LTSes $A$ and $B$ have the same label
-sets $\mathcal L$. We say that
+set $\mathcal L$. We say that
 \textbf{$A$ trace-refines $B$ w.r.t. $f$}, or $A \sqsubseteq_f B$, if:
@@
 \begin{figure}
+\small
 \centering
@@
 
-\section{Decomposing a multi-processor system}
+\section{Decomposing a Multi-Processor System}
 \label{sec:store-atomicity}
@@
 
-Any multi-processor component can be divided into 3 components as shown in
+Most any conventional multi-processor system can be divided into 3 components as shown in
 Figure \ref{zoom}. 
@@
 The processor component $P_i$ can be implemented in a
-variety of ways, from a simple in-order issue, to a complex out-of-order issue
-core, with aggressive speculation. The memory component is normally implemented
+variety of ways, from literal execution of instructions in program order, to complex speculation
+with instructions run simultaneously to exploit parallelism. The memory component is normally implemented
 using a hierarchy of caches, in order to increase the performance of the
@@
 
-Independent of the memory models, the memory component always implements the
-\emph{store atomicity} property. This property can be understood using the
-behavior defined by the operational semantics of Figure \ref{$M_m$}. The
-different behaviors permitted by different memory models arise merely because
-of the (1) the reordering in the issue of memory instructions by the core $P_i$
-if it can issue instructions out-of-order and (2) the properties of the local
-buffers $B_i$ connected to each processor $P_i$. For instance, if $B_i$ was a
-store buffer, into which the stores from an in-order processor $P_i$ that are
-``committed'' are enqueued, and if any ``subsequent'' load from the same
-processor $P_i$ reads the latest value from the store buffer, then one can
-establish that such a multi-processor system implements the TSO
-\cite{x86tsocacm10} memory model.
+Popular instruction set architectures do not guarantee sequential
+consistency.  That is, their behavior makes it unsound to imagine that
+the different processors of a shared-memory system are running
+according to simple interleaving, where each instruction executes
+atomically.  However, we want to emphasize that, in every weak-memory
+system we are aware of, \emph{the main memory still exposes atomic
+  loads and stores}!  The weaker semantics arises only because of (1)
+reordering of memory instructions by the core $P_i$ and/or (2) the
+properties of the local buffers $B_i$ connected to each processor
+$P_i$.  For instance, if $B_i$ were a store buffer, into which the
+committed stores from an in-order processor $P_i$ are enqueued, and if
+any ``subsequent'' load from the same processor $P_i$ reads the latest
+value from the store buffer, then one can establish that such a
+multi-processor system implements the TSO~\cite{x86tsocacm10} memory
+model.
+
+So, we focus on this opportunity to simplify proof decomposition.  We
+will prove that our main memory component satisfies an intuitive
+\emph{store atomicity} property.  Despite the potential red herring of
+the word ``atomicity,'' this specification is appropriate even for
+implementation of weaker memory models than SC.  Store atomicity can
+be understood via the operational semantics of Figure \ref{$M_m$},
+describing an LTS that receives load and store requests ($\ld$ and
+$\st$) from processors and sends back load responses ($\ld\resp$).
 
@@
 significant speedup.  Therefore it is the responsibility of the hierarchy of
-caches, which form the memory sub-component, to implement the store atomicity
+caches, which forms the memory sub-component, to implement the store atomicity
 property. In fact, as we will prove in Section \ref{sec:cc}, the purpose of the
-cache coherence protocol is to establish this invariant for the memory
+cache-coherence protocol is to establish this invariant for the memory
 sub-component.  Concretely, we have verified a \emph{directory-based} protocol
@@
 components indeed implements sequential consistency. We implement a highly
-speculative processor which executes instructions and issues loads out of
+speculative processor that executes instructions and issues loads out of
 order, but commits instructions (once some ``verification'' is done) in order.
 
-The processor itself can be decomposed into several components. In the zoomed
-in version of Figure \ref{zoom}, we show a highly speculative, out-of-order
+The processor itself can be decomposed into several components. In the zoomed-in
+version of Figure \ref{zoom}, we show a highly speculative, out-of-order
 issue processor. We have the normal architectural state, such as values of
@@
 SC the simple composition of P$^n_\text{ref}+ M_m$ executing each
-instruction atomically thus implementing sequential consistency. Notation
-$A^n$ is for $n$ copies of system $A$ running in parallel.
-$\sqsubseteq$ is a refinement operator, capturing a suitable notion of when
-one system implements another.  We go into more detail on the symbols from
-the figure in the rest of the paper.
+instruction atomically, thus implementing sequential consistency.
 
@@
 \begin{figure*}[t]
+\small
 \centering
@@
 \begin{figure*}[t]
+\small
 \begin{inv}
@@
 \begin{figure*}
+\small
 \centering

### 2014-10-20 TransitionsUseful.tex
Pass up to Section 5

@@
 system implements sequential consistency (SC) properly.  We state the theorem
-in terms of the trace refinement relation $\sqsubseteq$ developed in the prior
-section.  Therefore, we need to define an LTS embodying SC.  The simpler this
+in terms of the trace refinement relation $\sqsubseteq$.
+Therefore, we need to define an LTS embodying SC.  The simpler this
 system, the better.  We do not need to worry about its performance properties,
@@
 {\dec(s,\pc) = (\st, a, v) & \exec(s, \pc, (\st)) = (\pc', s')}
-{\ior{P$_\text{ref}$}{s, \pc}
-{s', \pc'}{\st, a, v}}
+{\io{P$_\text{ref}$}{s, \pc}
+{s', \pc'}{(\st, a, v)}}
 \end{boxedminipage}
@@
 
-Figure \ref{Pref$} shows the LTS for a single processor which will be used as a
+Figure \ref{Pref$} shows the LTS for a single processor that will be used as a
 reference processor.  The definition is parametrized over details of an
@@
 
-We can define sequential consistency using the simple memory defined in Figure
-\ref{$M_m$} as follows:
+We can define sequential consistency using the simple memory from Figure
+\ref{$M_m$} as follows, for $n$ processors:
 \begin{defn}
-$SC \triangleq P_\text{ref} + M_m$
+$SC \triangleq P^n_\text{ref} + M_m$
 \label{sc}
@@
 
-For a system of $n$ processors, our decoupled SC implementation is
-$\text{P$^n_{\text{ref}}$} + M_m$.
+Note that, in this setting, an operational specification like the LTS
+for SC is precisely the right characterization of \textbf{full functional
+  correctness} for a hardware design, much as a
+precondition-postcondition pair does that job in a partial-correctness
+Hoare logic.  Our SC LTS fully constrains observable behavior of a
+system to remain consistent with simple interleaving.  Similar
+operational models are possible as top-level specifications for
+systems following weaker memory models.
 

### 2014-10-20 TransitionsUseful.tex
few corrections

@@
 
-\begin{figure}
-\small
-\centering
-\begin{boxedminipage}[c]{.48\textwidth}
+%\begin{figure}
+%\small
 %\centering
-\inference
-[Halt]
-{\theta(i) = (s, \pc) & \dec(s,\pc) = H}
-{\io{SC}{\theta,m}{H}{}}
-
-\inference
-[NonMem]
-{\theta(i) = (s, \pc) & \dec(s,\pc) = (\nm, x) \\ \exec(s, \pc, (\nm, x)) = (\pc', s')}
-{\io{SC}{\theta,m}{\theta[i \coloneqq (s', \pc')], m}{}}
-
-\inference
-[Load]
-{\theta(i) = (s,\pc) & \dec(s,\pc) = (\ld, x, a) \\ \exec(s, \pc, (\ld, x, m(a))) = (\pc', s')}
-{\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m}{}}
-
-\inference
-[Store]
-{\theta(i) = (s,\pc) & \dec(s,\pc) = (\st, a, v) \\ \exec(s, \pc, (\st)) = (\pc', s')}
-{\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m[a\coloneqq v]}{}}
-\end{boxedminipage}
-\caption{LTS for sequential consistency with $n$ simple processors}
-\label{Ref}
-\end{figure}
-
-\section{Decomposing a Multi-Processor System}
-\label{sec:store-atomicity}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.43]{zoom}
-\caption{Components of a multi-processors system}
-\label{zoom}
-\end{figure}
+%\begin{boxedminipage}[c]{.48\textwidth}
+%%\centering
+%\inference
+%[Halt]
+%{\theta(i) = (s, \pc) & \dec(s,\pc) = H}
+%{\io{SC}{\theta,m}{H}{}}
+%
+%\inference
+%[NonMem]
+%{\theta(i) = (s, \pc) & \dec(s,\pc) = (\nm, x) \\ \exec(s, \pc, (\nm, x)) = (\pc', s')}
+%{\io{SC}{\theta,m}{\theta[i \coloneqq (s', \pc')], m}{}}
+%
+%\inference
+%[Load]
+%{\theta(i) = (s,\pc) & \dec(s,\pc) = (\ld, x, a) \\ \exec(s, \pc, (\ld, x, m(a))) = (\pc', s')}
+%{\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m}{}}
+%
+%\inference
+%[Store]
+%{\theta(i) = (s,\pc) & \dec(s,\pc) = (\st, a, v) \\ \exec(s, \pc, (\st)) = (\pc', s')}
+%{\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m[a\coloneqq v]}{}}
+%\end{boxedminipage}
+%\caption{LTS for sequential consistency with $n$ simple processors}
+%\label{Ref}
+%\end{figure}
+%
+%\section{Decomposing a Multi-Processor System}
+%\label{sec:store-atomicity}
+%
+%\begin{figure}
+%\centering
+%\includegraphics[scale=.43]{zoom}
+%\caption{Components of a multi-processors system}
+%\label{zoom}
+%\end{figure}
 
@@
 {}
-{\io{$M_m$}{m}{m}{(\ld\req,a),(\ld\resp,m(a))}}
+{\io{$M_m$}{m}{m}{(\ld\req,a),(\ld\resp,m(a)),\_}}
 
@@
 [NM]
-{\dec(s,\pc) = (\nm, x) & \exec(s, \pc, (\nm, x)) = (\pc', s')}
+{\dec(s,\pc) = (\nm, x) & \exec(s, \pc, (\nm, x)) = (s', \pc')}
 {\ioe{P$_\text{ref}$}{s,\pc}{s', \pc'}}
@@
 [Ld]
-{\dec(s,\pc) = (\ld, x, a) & \exec(s, \pc, (\ld, x, v)) = (\pc', s')}
+{\dec(s,\pc) = (\ld, x, a) & \exec(s, \pc, (\ld, x, v)) = (s', \pc')}
 {\io{P$_\text{ref}$}{s, \pc}
-{s', \pc'}{(\ld\req, a), (\ld\resp, v)}}
+{s', \pc'}{(\ld\req, a), (\ld\resp, v), \Real}}
 
@@
 [St]
-{\dec(s,\pc) = (\st, a, v) & \exec(s, \pc, (\st)) = (\pc', s')}
+{\dec(s,\pc) = (\st, a, v) & \exec(s, \pc, (\st)) = (s', \pc')}
 {\io{P$_\text{ref}$}{s, \pc}
@@
 A companion function $\exec(s, \pc, \mathit{dec})$ actually executes the
-instruction, returning the next program counter $\pc'$ and the new state $s'$.
+instruction, returning the new state $s'$ and the next program counter $\pc'$.
 
@@
 $H$, for a ``halt'' instruction that moves the LTS to state $H$. The parameter
-$x$ above represents the rest of the instruction, including the
-opcode, registers, constants, \etc{}
+$x$ above represents the rest of the instruction, including the opcode,
+registers, constants, \etc{}
+
+The label generated for a load contains the request sent to the
+memory system, and the response obtained, as well as a label suffix $\Real$
+which denotes that the load is non-speculative (We will later discuss the
+alternative, $\Spec$ which arises in the load labels for our optimized
+speculative processor).
 
@@
 address to value $v_0$; and $\theta_0$ maps every processor ID to
-$(s_0, \pc_0)$, using architecture-specific default values $s_0$ and
+$(\pc_0, s_0)$, using architecture-specific default values $s_0$ and
 $\pc_0$.
@@
 operational models are possible as top-level specifications for
-systems following weaker memory models.
+systems following weaker memory models, by giving the LTS for the \emph{Local
+Buffer} component and composing the three components simultaneously.
 
@@
 {\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\upd\ld(\compute(\rob),t,v),\ppc}
-{(\ld\req, a), (\ld\resp, v)}}
+{(\ld\req, a), (\ld\resp, v), \Spec}}
 
@@
 {\commit(\rob) = (\pc,\pc',(\ld,x,a,v,s'))}
-{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s',\pc',\retire(\rob),\ppc}{(\ld\req, a), (\ld\resp, v)}}
+{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s',\pc',\retire(\rob),\ppc}{(\ld\req, a), (\ld\resp, v), \Real}}
 
 \inference[LdBad]
-{\commit(\rob) = (\pc,\pc',(\ld,x,a,v',s')) & v' \neq v & \exec(s,\pc,(\ld,x,v)) = (\pc'', s'')}
-{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s'',\pc'',\phi,\setBp(\ppc,\pc)}{(\ld\req, v), (\ld\resp,v)}}
+{\commit(\rob) = (\pc,\pc',(\ld,x,a,v',s')) & v' \neq v & \exec(s,\pc,(\ld,x,v)) = (s'', \pc'')}
+{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s'',\pc'',\phi,\setBp(\ppc,\pc)}{(\ld\req, v), (\ld\resp,v), \Real}}
 \end{boxedminipage}
@@
 program order. 
-We will see the processor issue two kinds of loads, a speculative
-load ($\spc\ld$) and a commit or real load ($\ld$).  To maintain SC, every speculative
+We will see the processor issue two kinds of loads, a speculative load (with
+the label suffix $\Spec$) and a commit or real load (with the label suffix $\Real$).  To
+maintain SC, every speculative
 load must have a matching verification load later on, and we maintain
@@
 \exists x. \; \dec(s,\pc) = (\nm, x) \wedge \exec(s,\pc, (\nm,x)) =
-(\pc', s')\\
+(s', \pc')\\
 \commit(\rob) = (\pc, \pc', (\ld, x, a, v, s')) & \Rightarrow
-\dec(s,\pc) = (\ld, x,a) \wedge \exec(s,\pc, (\ld,x,v)) = (\pc', s')\\
+\dec(s,\pc) = (\ld, x,a) \wedge \exec(s,\pc, (\ld,x,v)) = (s',\pc')\\
 \commit(\rob) = (\pc, \pc', (\st, a, v, s')) &\Rightarrow
 \dec(s,\pc) = (\st, a,v) \wedge \exec(s,\pc, (\st)) =
-(\pc', s')\\
+(s',\pc')\\
 \commit(\rob) = H & \Rightarrow
@@
 {\dt, \ch, \s, \dst, \wt, \dwt}
-{\dt, \ch, \s, \dst, \wt, \dwt}{c,((\ld\req, a),(\ld\resp,v))}}
+{\dt, \ch, \s, \dst, \wt, \dwt}{c,((\ld\req, a),(\ld\resp,v)), \_}}
 \vspace{.1in}

### 2014-10-22 TransitionsUseful.tex
few corrections
stuff

@@
 %
-%\section{Decomposing a Multi-Processor System}
-%\label{sec:store-atomicity}
-%
-%\begin{figure}
-%\centering
-%\includegraphics[scale=.43]{zoom}
-%\caption{Components of a multi-processors system}
-%\label{zoom}
-%\end{figure}
+\section{Decomposing a Multi-Processor System}
+\label{sec:store-atomicity}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.43]{zoom}
+\caption{Components of a multi-processors system}
+\label{zoom}
+\end{figure}
 
@@
 {}
-{\io{$M_m$}{m}{m}{(\ld\req,a),(\ld\resp,m(a)),\_}}
+{\io{$M_m$}{m}{m}{p, (\ld\req,a),(\ld\resp,m(a), \_)}}
 
@@
 {}
-{\io{$M_m$}{m}{m[a\coloneqq v]}{(\st, a, v)}}
+{\io{$M_m$}{m}{m[a\coloneqq v]}{p, (\st, a, v)}}
 
@@
 describing an LTS that receives load and store requests ($\ld$ and
-$\st$) from processors and sends back load responses ($\ld\resp$).
+$\st$) from processors and sends back load responses ($\ld\resp$);
+$p$ is the processor which sends and receives the request.
 
@@
 \item $\add(\pc, \rob)$, which appends the program instruction at location $\pc$ to the list of instructions that the buffer is allowed to consider executing.
-\item $\get\spc\ld(\rob)$, which returns an optional speculative load to issue
+\item $\get\spc\ld(\rob)$, which returns a set of speculative loads that can be issued to the memory system
 \item $\compute(\rob)$, which models a
@@
 \inference[Compute]
-{\get\spc\ld(\rob) = \epsilon}
+{}
 {\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\compute(\rob),\ppc}}
@@
 \inference[SpecLd]
-{\get\spc\ld(\rob) = (\spc\ld,t,a)}
-{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\upd\ld(\compute(\rob),t,v),\ppc}
+{(\spc\ld,t,a) \in \get\spc\ld(\rob)}
+{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\upd\ld(\rob,t,v),\ppc}
 {(\ld\req, a), (\ld\resp, v), \Spec}}
@@
 {\dt, \ch, \s, \dst, \wt, \dwt}
-{\dt, \ch, \s, \dst, \wt, \dwt}{c,((\ld\req, a),(\ld\resp,v)), \_}}
+{\dt, \ch, \s, \dst, \wt, \dwt}{c,((\ld\req, a),(\ld\resp,v), \_)}}
 \vspace{.1in}

### 2014-10-30 Abstract.tex
put back the old stuff in abstract and intro as comment
stuff

@@
+\xxx{\textbf{Murali's version in red:} We provide a framework for modular verification of multi-processor hardware
+systems with respect to memory model specification. In this framework, we make
+use of a key observation that eventhough different memory models have different
+overall system behaviors, they all share a coherent cache hierarchy. We prove
+that this coherent cache hierarchy implements the \emph{store atomicity}
+property, which makes it behave like an atomic instantaneous memory with
+respect to the rest of the system. We further modularize the proof by defining
+what is meant for a processor to be correct independently of the memory system.
+As a concrete instance of our framework, we walk through a modular proof of
+high-performance multi-processor implementation of Lamport's Sequential
+Consistency model. The processors employ \emph{speculative execution} to
+generate many concurrent loads to the memory sub-system, while obeying our
+correctness criteria. We offer the first formal proof that when an atomic
+memory is connected to $n$ speculative processors, the system as a whole is
+sequentially consistent.  Additionally, the cache-coherent memory we present in
+this paper is the first fully verified system that supports arbitrary cache
+hierarchies. Our framework involves describing all our components using labeled
+transition systems and we machine-verify all our proofs using the Coq proof
+assistant.}
 We present a new framework for modular verification of hardware designs

### 2014-10-30 Introduction.tex
put back the old stuff in abstract and intro as comment
stuff

@@
 modular boundary to be realized by multiple implementations, any of which may be
-dropped into the system safely. It also allows a greater measure of
+dropped into the system safely. It also allows a greater
 parallelization of human design effort, improving development time. Similarly, verification is
@@
 produce full-system theorems.
+
+\xxx{\textbf{Murali's version in red} This works introduces an interface enabling true modular implementation and
+verification of shared memory systems. We want our modular system to permit
+\emph{black-box refinement}, that is, nothing beyond the interface contract can
+be assumed about other modules.  That is, we should be able to describe a
+modular decomposition where we may replace either the processor nodes, the
+memory with locally verified alternatives without changing the correctness of
+our whole system.  To this end, we identify a key property called \emph{store
+atomicity} shared by the memory component of every memory model. This property
+effectively restricts the behavior of the memory sub-system to that of an
+\emph{atomic instantaneous memory}. We prove that this property is in fact what
+is guaranteed using the cache coherence protocols in a multi-processor system
+containing a hierarchy of caches, enabling replacement of the cache hierarchy
+with an atomic instantaneous memory. Similarly, on the processor side, we
+define what is meant to be a correct single processor. These invariants about
+the processor and the memory sub-system remains constant across memory models.
+We show a concrete instance of this framework by proving that a highly
+speculative multi-processor system connected to a coherent cache hierarchy
+implements the sequential consistency memory model.}
 

### 2014-10-30 TransitionsUsefulFirst.tex
stuff

--- 2014-10-22 TransitionsUseful.tex
@@
 \caption{LTS for a simple memory}
-\label{$M_m$}
+\label{M_m}
 \end{figure}
@@
 implementation of weaker memory models than SC.  Store atomicity can
-be understood via the operational semantics of Figure \ref{$M_m$},
+be understood via the operational semantics of Figure \ref{M_m},
 describing an LTS that receives load and store requests ($\ld$ and
@@
 \end{figure}
-
-\section{Specifying Sequential Consistency}\label{sec:sc}
-
-Our final theorem in this paper establishes that a particular complex hardware
-system implements sequential consistency (SC) properly.  We state the theorem
-in terms of the trace refinement relation $\sqsubseteq$.
-Therefore, we need to define an LTS embodying SC.  The simpler this
-system, the better.  We do not need to worry about its performance properties,
-since we will prove that an optimized system remains faithful to it.
-
-\begin{figure}
-\small
-\centering
-\begin{boxedminipage}[c]{.48\textwidth}
-\inference
-[Halt]
-{\dec(s,\pc) = H}
-{\io{P$_\text{ref}$}{s,\pc}{H}{}}
-
-\inference
-[NM]
-{\dec(s,\pc) = (\nm, x) & \exec(s, \pc, (\nm, x)) = (s', \pc')}
-{\ioe{P$_\text{ref}$}{s,\pc}{s', \pc'}}
-
-\inference
-[Ld]
-{\dec(s,\pc) = (\ld, x, a) & \exec(s, \pc, (\ld, x, v)) = (s', \pc')}
-{\io{P$_\text{ref}$}{s, \pc}
-{s', \pc'}{(\ld\req, a), (\ld\resp, v), \Real}}
-
-\inference
-[St]
-{\dec(s,\pc) = (\st, a, v) & \exec(s, \pc, (\st)) = (s', \pc')}
-{\io{P$_\text{ref}$}{s, \pc}
-{s', \pc'}{(\st, a, v)}}
-\end{boxedminipage}
-
-\caption{LTS for a single reference processor (P$_{\text{ref}}$)}
-\label{Pref$}
-\end{figure}
-
-Figure \ref{Pref$} shows the LTS for a single processor that will be used as a
-reference processor.  The definition is parametrized over details of an
-\emph{instruction set architecture} (ISA).  In particular, the ISA gives us
-some domains of architectural states $s$ (e.g., register files) and of program
-counters $\pc$.  A function $\dec(s, \pc)$ figures out which instruction $\pc$
-references in the current state, returning the instruction's ``decoded'' form.
-A companion function $\exec(s, \pc, \mathit{dec})$ actually executes the
-instruction, returning the new state $s'$ and the next program counter $\pc'$.
-
-The legal instruction forms, which are outputs of $\dec$, are $(\nm, x)$, for
-an operation not accessing memory; $(\ld, x, a)$, for a memory load from
-address $a$; $(\st, a, v)$, for a memory store of value $v$ to address $a$; and
-$H$, for a ``halt'' instruction that moves the LTS to state $H$. The parameter
-$x$ above represents the rest of the instruction, including the opcode,
-registers, constants, \etc{}
-
-The label generated for a load contains the request sent to the
-memory system, and the response obtained, as well as a label suffix $\Real$
-which denotes that the load is non-speculative (We will later discuss the
-alternative, $\Spec$ which arises in the load labels for our optimized
-speculative processor).
-
-The legal inputs to $\exec$ encode both a decoded instruction and any relevant
-responses from the memory system.  These inputs are $(\nm, x)$ and $\st$, which
-need no extra input from the memory; and $(\ld, x, v)$, where $v$ gives the
-contents of the requested memory cell.
-
-We can define sequential consistency using the simple memory from Figure
-\ref{$M_m$} as follows, for $n$ processors:
-\begin{defn}
-$SC \triangleq P^n_\text{ref} + M_m$
-\label{sc}
-\end{defn}
-
-We define the initial state of SC as $(\theta_0, m_0)$, where $m_0$ is
-some initial memory fixed throughout our development, mapping every
-address to value $v_0$; and $\theta_0$ maps every processor ID to
-$(\pc_0, s_0)$, using architecture-specific default values $s_0$ and
-$\pc_0$.
-
-This LTS encodes Lamport's notion of SC, where processors take turns executing
-nondeterministically in a simple interleaving.  We need know very few details
-of the ISA to define the SC model.  Our final, optimized system is parametrized
-over an ISA in the same way.  In the course of the rest of this paper, we will
-define an optimized system $O$ and prove $O \sqsubseteq \text{SC}$.
-
-Note that, in this setting, an operational specification like the LTS
-for SC is precisely the right characterization of \textbf{full functional
-  correctness} for a hardware design, much as a
-precondition-postcondition pair does that job in a partial-correctness
-Hoare logic.  Our SC LTS fully constrains observable behavior of a
-system to remain consistent with simple interleaving.  Similar
-operational models are possible as top-level specifications for
-systems following weaker memory models, by giving the LTS for the \emph{Local
-Buffer} component and composing the three components simultaneously.
-
-\section{Speculative Out-Of-Order Processor}\label{sec:ooo}
-
-The previous section divided a multiprocessor system into two rather
-simple components, processors and memories.  We are now ready to begin
-developing more complex, optimized versions of each piece, starting
-with processors.
-
-We implement a \emph{speculative} processor, which may create many
-simultaneous outstanding requests to the memory, as an optimization to
-increase parallelism.  Our processor proof is in some sense generic
-over correct speculation strategies.  We parametrize over two key
-components of a processor design: a \emph{branch predictor}, which
-makes guesses about processor-local control flow in advance of
-%receiving enough memory responses to resolve
-resolving conditional jumps; and a
-\emph{reorder buffer}, which decides what speculative instructions (like memory loads)
-are worth issuing at which moments (in effect \emph{reordering} later
-instructions to happen before earlier instructions have finished).
-
-The branch predictor is the simpler of the two components to model.
-We indicate branch-predictor state with metavariable $\ppc$.
-The operations on such state are $\curPpc(\ppc)$, to extract the
-current program-counter prediction; $\nextPpc(\ppc)$, to ask the
-predictor to step forward by one instruction; and $\setBp(\ppc, \pc)$,
-to reset prediction to begin at a known-accurate position $\pc$.  It
-turns out that we need to impose no explicit correctness criterion on
-branch predictors.  The processor uses predictions only as hints, and
-it always resets the predictor using $\setBp$ after detecting an
-inaccurate hint.
-
-The interface and formal contract of a reorder buffer are more
-involved.  We write $\rob$ as a metavariable for reorder buffer
-state.  The associated operations are:
-\begin{itemize}
-\item $\phi$, the state of an empty buffer.
-\item $\add(\pc, \rob)$, which appends the program instruction at location $\pc$ to the list of instructions that the buffer is allowed to consider executing.
-\item $\get\spc\ld(\rob)$, which returns a set of speculative loads that can be issued to the memory system
-\item $\compute(\rob)$, which models a
-step of computation inside the buffer. For instance, it invokes the $\dec$
-and $\exec$ functions (as defined for SC) internally to obtain the next program
-counter, state, \etc{} (only the speculative states are updated, not the actual states).
-\item $\upd\ld(\rob,t,v)$, which informs the buffer that the memory
-has returned result value $v$ for the speculative load with tag $t$.
-\item $\commit(\rob)$, which returns the next instruction in serial
-program order, if we have accumulated enough memory responses to execute it
-accurately, or returns
-$\epsilon$ otherwise.  When $\commit$ returns an instruction, it also
-returns the associated program counter plus the next program counter
-we would advance to afterward.
- Furthermore, the instruction is
-extended with any relevant response from memory (used only for load
-instructions, obtained through $\upd\ld$) and with the new architectural state (\eg{} register
-file) after execution.
-\item $\retire(\rob)$, which informs the buffer that its $\commit$
-instruction was executed successfully, so it is time to move on to the
-next instruction.
-\end{itemize}
-
-We postpone characterization of correct reorder buffers until after we
-present the LTS for speculating processors.
-
-\begin{figure*}[t]
-\small
-\centering
-\begin{boxedminipage}[c]{.95\textwidth}
-\inference[Fetch]
-{}
-{\ioe{P$_\text{so}$}{s,\pc,\rob,\ppc}{s,\pc,\add(\curPpc(\ppc),\rob),\nextPpc(\ppc)}}
-
-\inference[Compute]
-{}
-{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\compute(\rob),\ppc}}
-
-\inference[SpecLd]
-{(\spc\ld,t,a) \in \get\spc\ld(\rob)}
-{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\upd\ld(\rob,t,v),\ppc}
-{(\ld\req, a), (\ld\resp, v), \Spec}}
-
-\inference[Abort]
-{\commit(\rob) = (\pc',\_,\_) & \pc' \neq \pc}
-{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\phi,\setBp(\ppc,\pc)}}
-
-\inference[Halt]
-{\commit(\rob) = H}
-{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{H}}
-
-\inference[Nm]
-{\commit(\rob) = (\pc,\pc',(\nm,s'))}
-{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s',\pc',\retire(\rob),\ppc}}
-
-\inference[St]
-{\commit(\rob) = (\pc,\pc',(\st,a,v,s'))}
-{\io{\text{P$_\text{so}$}}{s',\pc',\retire(\rob),\ppc}{s,\pc,\rob,\ppc}
-{\st, a, v}}
-
-\inference[LdGood]
-{\commit(\rob) = (\pc,\pc',(\ld,x,a,v,s'))}
-{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s',\pc',\retire(\rob),\ppc}{(\ld\req, a), (\ld\resp, v), \Real}}
-
-\inference[LdBad]
-{\commit(\rob) = (\pc,\pc',(\ld,x,a,v',s')) & v' \neq v & \exec(s,\pc,(\ld,x,v)) = (s'', \pc'')}
-{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s'',\pc'',\phi,\setBp(\ppc,\pc)}{(\ld\req, v), (\ld\resp,v), \Real}}
-\end{boxedminipage}
-
-\caption{Speculating, out-of-order issue processor}
-\label{Ooo}
-\end{figure*}
-
-Figure~\ref{Ooo} defines that LTS,
-P$_{\texttt{so}}$. This processor may issue arbitrary speculative
-loads, but it only ever \emph{commits} the next instruction in serial
-program order. 
-We will see the processor issue two kinds of loads, a speculative load (with
-the label suffix $\Spec$) and a commit or real load (with the label suffix $\Real$).  To
-maintain SC, every speculative
-load must have a matching verification load later on, and we maintain
-the illusion that the program only depends on the results of
-verification loads, which, along with stores, \emph{must be issued in
-serial program order}.
-
-When committing a
-previously issued speculative load instruction, the associated speculative
-memory load response is verified against the new commit load response. If the
-resulting values do not match, the processor terminates all past uncommitted
-speculation, by emptying the reorder buffer and resetting the
-next predicted program counter in the branch predictor to the correct next value. This step mimics
-the mechanism implemented in most modern-day processors, operating on the
-conservative assumption that the processor may have performed arbitrary
-incorrect speculation after receiving the inaccurate response from
-memory.  (Note that here ``inaccurate'' does not mean ``wrong at
-the time the response was received,'' but rather ``wrong at commit
-time'' or ``correct before, but not anymore.'')
-
-A full processor state is $(s, \pc, \rob, \ppc)$, comprising
-architectural state, the program counter, and the reorder buffer and branch
-predictor states. Its initial state is given by $(s_0, \pc_0, \phi, \ppc_0)$.
-The interface of this processor with memory is identical
-to that of the reference processor, but now we can exercise a memory
-system's transitions even for speculative loads.
-
-\begin{figure*}[t]
-\small
-\begin{inv}
-If $P_\text{so}$ reaches a state $(s, \pc, \rob, \ppc)$, \ie{}
-
-$\io{P$_\text{so}^{*}$}{s_0,\pc_0,\phi,\ppc_0}{s,\pc,\rob,\ppc}{\eta} \Rightarrow$
-\begin{displaymath}
-\left\{
-\begin{array}{ll}
-\commit(\rob) = (\pc, \pc', (\nm, s')) &\Rightarrow
-\exists x. \; \dec(s,\pc) = (\nm, x) \wedge \exec(s,\pc, (\nm,x)) =
-(s', \pc')\\
-\commit(\rob) = (\pc, \pc', (\ld, x, a, v, s')) & \Rightarrow
-\dec(s,\pc) = (\ld, x,a) \wedge \exec(s,\pc, (\ld,x,v)) = (s',\pc')\\
-\commit(\rob) = (\pc, \pc', (\st, a, v, s')) &\Rightarrow
-\dec(s,\pc) = (\st, a,v) \wedge \exec(s,\pc, (\st)) =
-(s',\pc')\\
-\commit(\rob) = H & \Rightarrow
-\dec(s, \pc) = H\\
-\end{array}
-\right.\end{displaymath}
-\label{rob}
-\end{inv}
-\vspace{-.5cm}
-\caption{Correctness of reorder buffer}
-\label{robfig}
-\end{figure*}
-
-Finally, we impose a general correctness condition on reorder
-buffers.  They must maintain Invariant~\ref{rob} in Figure~\ref{robfig}.
-Intuitively, whenever the buffer claims in a $\commit$
-output that a particular instruction is next to execute, causing
-certain state changes, that instruction must really be next in line according
-to how the program runs in the SC system, and running it must really cause
-those state changes.
-
-When this condition holds, we may conclude the correctness theorem for
-out-of-order processors.  We use a trace-transformation function
-$\noSpec$ that drops all speculative-load requests and responses.
-See Definition~\ref{refines} for a review of how we use such
-functions in framing trace refinement.  Intuitively, we prove that any
-behavior by the speculating processor can be matched by the simple
-processor, with speculative messages erased.
-
-\begin{theorem}
-\label{ocorrect}
-$\text{P$_\text{so}$} \sqsubseteq_\noSpec$ P$_\text{ref}$
-\end{theorem}
-\begin{proof}
-By induction on P$_\text{so}$ traces, using a proper abstraction
-function (which drops the speculative messages and the $\rob$ and $\ppc$ states) to relate the two systems.  Invariant~\ref{rob} is crucial to
-relate the reorder buffer's behavior with the simple in-order
-execution of P$_\text{ref}$.
-\end{proof}
-
-\begin{corollary}
-\label{ges}
-$\text{P$^n_\text{so}$} \sqsubseteq_{\noSpec^n}$ P$_\text{ref}^n$
-\end{corollary}
-\begin{proof}
-Direct consequence of Theorems~\ref{ocorrect} and~\ref{liftn} (where the
-latter is one of our preliminary results about $n$-repetition).
-\end{proof}
-
-\section{Cache-Based Memory System}\label{sec:cc}
-
-We now turn our attention toward a more efficient implementation of
-memory.  It is physically infeasible to connect many processors
-directly to a single memory with low enough latency to handle most
-memory requests quickly enough.  Instead, each processor will have its
-own \emph{L1 cache} of recently used memory cells.  Whenever possible, we
-want to service memory requests from processors' own caches, only
-contacting memory for addresses not found in the caches.  To further
-avoid contacting main memory, we create additional higher-level
-caches, connecting each L1 cache to an L2 cache, potentially
-connecting L2 caches to L3, etc., bottoming out in main memory at the
-root of our tree. The zoomed-in portion of the memory in Figure~\ref{zoom}
-diagrams the basic layout.
-
-Now we have concurrent interaction of many processors with many caches
-with main memory, and the relationship with the original $M_m$ system
-is far from direct.  However, this intricate concurrent execution is
-crucial to hiding the latency of main-memory access.
-Figure~\ref{cache} formalizes as an LTS $M_c$ the algorithm we implemented,
-for providing the memory abstraction on top of a cache hierarchy.
-We have what is called an \emph{invalidating directory-based hierarchical
-cache-coherence protocol}.  It is a transliteration of a realistic
-algorithm implemented in the Bluespec hardware description
-language~\cite{Bluespec:TFRG}.
-
-We now try to explain aspects of this transition system at an
-intuitive level.  Thanks to the overall modular proof structure in
-this paper, it is not essential to understand the details of the cache
-system, to understand any other part of our development.  Therefore,
-some readers may want to skip this explanation upon first reading,
-skipping to the next section for a key theorem relating $M_c$ and $M_m$.
-
-We describe a state of the system using 8 fields: $\dt$, $\ch$, $\s$, $\dst$,
-$\wt$, $\dwt$ which are summarized in Figure~\ref{fields}.
-We use $\parent(c,p)$ to denote that $p$ is the parent of $c$ and $\leaf(c)$ to 
-denote that $c$ is a leaf or L1 cache.
-
-\begin{figure}
-\centering
-\begin{tabular}{|l|p{5.5cm}|}
-\hline
-$\dt(c,a)$ & Data in cache $c$ for address $a$\\
-$\s(c,a)$ & Coherence state of cache $c$ for address $a$\\
-$\dst(p,c,a)$ & Directory state of cache $p$ for child $c$ and address $a$\\
-$\wt(c,a)$ & Gives cache $c$'s desired coherence state to upgrade to (for address $a$), or $\epsilon$ if none\\
-$\dwt(p,c,a)$ & Gives cache $p$'s desired coherence state for its child $c$ to downgrade to (for address $a$), or $\epsilon$ if none\\
-$\ch$ & Channels between nodes in the system\\
-\hline
-\end{tabular}
-\caption{Fields in a cache-based memory system}
-\label{fields}
-\end{figure}
-
-Fields $\dt, \s$ are maps over pairs of cache names and memory addresses.
-The $\dt$ map assigns each pair a memory-cell value, and the $\s$ map
-assigns each pair a coherence state.  A coherence state is $M, S$ or
-$I$, broadly representing permissions to modify, read, or do nothing
-with an address, respectively, the decreasing permissions denoted by $M > S > I$. This interpretation of permissions is
-accurate at the leaf caches connected to the processors (\ie{} L1 caches),
-but the interpretation is subtly different at the higher level caches.
-
-Mapping $\wt$ goes from cache name and address to either $\epsilon$ or a
-coherence state.  It denotes what permission the cache is waiting for,
-if any.  That is, a cache may have decided to \emph{upgrade} its
-coherence state for some address to a more permissive value, but the
-cache is waiting for acknowledgment from another cache before
-finalizing the upgrade.
-
-Since we have a directory-based protocol, the $\s$ and
-$\wt$ maps have mirror-image counterparts $\dst$ and $\dwt$.
-The former maps triples of parent name, child name, and address to
-coherence state. It represents the parent's notion of the
-coherence state of the child for that address. We later prove that this notion
-is always conservative, \ie{} if the parent assumes that a child does not have
-a particular permission, then it is guaranteed in this system that the child
-will not have that permission.  The $\dwt$ field also represents a mapping from parent
-name, child name, and address to either $\epsilon$ or coherence state. It
-denotes whether the parent is waiting for any downgrade response from the
-corresponding child, and if so, the coherence state that the child must
-downgrade to as well.
-
-All communication channels in the system are represented by field $\ch$.
-The channels are asymmetric as
-follows. Channels going from parent to children (indexed by the names of the
-parent and child) maintain FIFO order of messages between the same parent-child
-pair. They carry both request messages (asking the child to downgrade) and
-response messages (acknowledging that the child can upgrade). Channels going
-from children to their parents are further split into request and response
-channels, thus indexed by tuples (child name, parent name, request or response).
-These channels need not maintain the FIFO order for messages between the same
-child-parent pair (notice the use of $\coloncolon$ to add and remove elements
-from the parent to child channel, as opposed to $\uplus$ for the 
-child to parent channel). The reason is that only one request or one response is
-sent for a particular address between a child-parent pair -- this invariant can
-be proved for the LTS in  Figure \ref{cache}. The order between a request and
-response message sent from the same child to its parent is lost, but our
-protocol ensures that nothing bad happens as a result.  Notice rules
-ParentRecvRsp and ParentRecvReq in Figure \ref{cache}, each of which looks at the
-previous coherence state of the incoming message and compares it with its own
-directory state to decide whether to process the message.
-
-%\medskip
-
-Here is an intuition on how the transitions work in the common case.  A cache
-can decide, spontaneously, to upgrade its coherence state, in which case it
-sends an upgrade request to its parent. The parent then makes a local decision
-on whether to send a response to the requesting child or not, based on its
-directory approximation and its own coherence state $\s$. If $\s$ is lower than
-the requested upgrade, then it cannot handle the request, and instead must
-decide to upgrade $\s$.  Once the parent's $\s$ is not lower than the requested
-upgrade, it makes sure that the rest of its children are ``compatible'' with
-the requested upgrade (given by $\compat$ definition below).  If not, the
-parent must send requests to the incompatible children to downgrade. Finally, when
-the $\s$'s upgrade and children's downgrade responses are all received, the
-original request can be responded to. The transitions for sending requests are
-allowed to happen spontaneously without any trigger. An L1 cache can process a
-load request or a store request only if it is in at least the \Sh{} state or
-\Mo{} state, respectively, for the requested address; otherwise it must
-spontaneously request to upgrade.
-
-\begin{defn}
-\begin{displaymath}
-\compat(p, c, x, a) = \left\{
-\begin{array}{ll}
-\forall c' \neq c. \; \dst(p,c',a) = I &: x = M\\
-\forall c' \neq c. \; \dst(p,c',a) \le S &: x = S\\
-\end{array}
-\right.
-\end{displaymath}
-\end{defn}
-
-The reason for having separate request and response channels for the same
-child-parent pair is to avoid deadlocks. Throughout the protocol, we maintain
-the invariant
-that a downgrade request from the parent always gets priority, \ie{} if a child
-is waiting for an upgrade response, it must still try to satisfy a downgrade
-request from the parent. If both the parent and the child are waiting
-for their respective responses, there will be a deadlock. This tie cannot
-be broken arbitrarily: if the parent concedes and starts processing a child's
-request, then it may have to send downgrade requests to other children to make
-the states compatible, but the other children may have sent upgrade requests
-and remain waiting, leading to a deadlock. The only way to break the
-vicious waiting cycle is for the children to concede and respond when they get
-downgrade requests. Of course, they have to downgrade the states of their own
-children first, and this process goes on recursively until the downgrade requests
-reach the leaf caches, which can respond immediately.
-
-A complication arises because a cache can voluntarily decide to downgrade its
-state.  This transition is used to model invalidation of cache lines to make
-room for a different location (we permit a cache to downgrade from $M$ to $S$
-instead of only to $I$).
-As a result, the parent's $\dst$ and the
-corresponding $\s$ of the child may go out of sync, leading to the parent
-requesting a child to downgrade when it already has. To handle this situation,
-the child has to drop the downgrade request when it has already downgraded to
-the required state (Rule DropReq in Figure \ref{cache}), to avoid deadlocks
-by not dequeuing the request.
-
-%Dropping the request
-%removes the entry from the buffer, allowing the child to process further messages
-%from the parent -- otherwise there would have been a deadlock.
-
-\begin{figure*}
-\small
-\centering
-
-\begin{boxedminipage}[c]{.95\textwidth}
-
-\textbf{Processor/Memory Interface}
-
-\inference[Load]
-{\s(c,a) \geq S}
-{\io{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt}
-{\dt, \ch, \s, \dst, \wt, \dwt}{c,((\ld\req, a),(\ld\resp,v), \_)}}
-\vspace{.1in}
-\inference[Store]
-{s(c,a) \geq M}
-{\io{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt}
-{\dt[(c, a) \coloneqq v], \ch, \s, \dst, \wt, \dwt}{c,(\st,a,v)}}
-
-\textbf{Child Upgrade}
-
-\inference[ChildSendReq]
-{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt}
-{\dt, \ch[ (c, p, \req) \coloneqq (a, \s(c, a), x) \uplus \ch(c,p,\req)], \s, \dst,\\
-\wt[(c,a)\coloneqq x], \dwt}}
-\vspace{.1in}
-\inference[ParentRecvReq]
-{\parent(c,p) & \ch(c, p, \req) = \{(a, y, x)\} \uplus \rs & \s(p, a) \geq x
-\\ \compat(p, c, x, a) & \dwt(p,c,a) = \epsilon & \dst(p,c,a) \le y}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt}
-{\dt, \ch[(c,p,\req) \coloneqq \rs][(p, c) \coloneqq (\resp, (a, \dst(p,c,a), x,\\ \;\;\;\;
-\ite{\dst(p,c,a) = I}{\dt(p,a)}{\_})) \coloncolon \ch(p,c)], \\ \s, \dst[(p,c, a)\coloneqq x],
-\wt, \dwt}}
-\vspace{.1in}
-\inference[ChildRecvRsp]
-{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\resp, (a, y, x, v))}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt}
-{\dt[(c,a) \coloneqq \ite{y = I}{v}{\dt(c,a)}], \ch[(p,c) \coloneqq \rs], \s[(c,a) \coloneqq x], \dst,\\
-\wt[(c,a)\coloneqq \ite{\wt(c,a) \le x}{\epsilon}{\wt(c,a)}], \dwt}}
-
-\textbf{Parent Downgrade}
-
-\inference[ParentSendReq]
-{\parent(c,p) & \dst(p,c,a) > x & \dwt(p,c,a) = \epsilon}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt}
-{\dt, \ch[(p, c)\coloneqq (\req, (a, \dst(p, c, a), x)) \coloncolon \ch(p,c)], \s, \dst,\\
-\wt, \dwt[(p, c,a)\coloneqq x]}}
-\vspace{.1in}
-\inference[ChildRecvReq]
-{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\req, (a, y, x)) & (\forall i. \; \parent(i, c) \Rightarrow \dst(c, i, a) \le x)
-& \s(c,a) > x}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt}
-{\dt, \ch[(p,c) \coloneqq \rs][(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
-\ite{\dst(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
-\wt, \dwt}}
-\vspace{.1in}
-\inference[ParentRecvRsp]
-{\parent(c,p) & \ch(c,p,\resp) = \{(a, y, x, v)\} \uplus \rs & \dst(p,c,a) = y}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt}
-{\dt[(p,a) \coloneqq \ite{y = M}{v}{\dt(p,a)}], \ch[(c,p,\resp) \coloneqq \rs], \s, \dst[(p,c,a) \coloneqq x],\\
-\wt, \dwt[(p,c,a)\coloneqq \ite{\dwt(p,c,a) \ge x}{\epsilon}{\dwt(p,c,a)}]}}
-
-\textbf{Voluntary downgrade for replacement}
-
-\inference[VolResp]
-{\parent(c,p) & (\forall i. \; \parent(i, c) \Rightarrow \dst(c, i, a) \le x)
-& \s(c,a) > x}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt}
-{\dt, \ch[(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
-\ite{\s(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
-\wt, \dwt}}
-
-\textbf{Dropping request because of voluntary downgrade}
-
-\inference[DropReq]
-{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\req, (a, y, x)) & 
- \s(c,a) \le x}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt}
-{\dt, \ch[(p,c) \coloneqq \rs], \s, \dst,
-\wt, \dwt}}
-\end{boxedminipage}
-
-\caption{LTS for cache-coherent shared-memory system}
-\label{cache}
-\end{figure*}
-
-\section{Proof that the MSI Protocol Preserves Store Atomicity}\label{sec:ccproof}
-\label{safety}
-
-We must prove the following theorem, \ie{} the cache-based system is sound with
-respect to the simple memory.
-\begin{theorem}
-\label{ccorrect}
-$M_c \sqsubseteq M_m$
-\end{theorem}
-
-The key property we need for this proof is what we define as
-``store atomicity'' below. Throughout this section, we say
-\emph{time} to denote the number of transitions that occurred before reaching
-the specified state.
-
-\begin{defn}
-Store atomicity is defined as the following property:
-For a load (or speculative load) request $\toM(\ld, a)$ (or $\toM(\spc\ld, t, a)$)
-from any processor, the response sent at time $T$
-$\toP(\ld, v)$ (or $\toP(\spc\ld, t, v)$) should be such that
-\begin{enumerate}
-\item $v = v_0$ (the initial value of any memory address) and no store
-  request $\toM(\st, a, v')$ has been processed at any time $T'$ such
-  that $T' < T$ or
-\item There is a store request $\toM(\st, a, v)$ that was processed at time $T_q$ such that
-$T_q < T$ and no other store
-request $\toM(\st, a, v')$ was processed at any time $T'$ such that $T_q < T' < T$.
-\end{enumerate}
-\label{sa}
-\end{defn}
-
-We can easily show that any system obeying store atomicity
-is sound with respect to $M_m$. Intuitively, the store atomicity
-property precisely describes the property of a memory, \viz{} that if a store
-happens, then that location gets updated such that any subsequent load gets
-back the updated value.
-
-The proof that $M_c$ obeys store atomicity is involved enough to deserve a
-section's worth of explanation, which we now turn to.
-
-We first reduce the proof of store atomicity to several lemmas. We then
-present a key insight that we discovered for proving these lemmas in the form
-of three fundamental invariants. We believe that other invalidation-based
-protocols with inclusive caches (\ie{} those cache hierarchies where a parent
-cache contains all the locations that any of its children contains) like MESI,
-MOSI, and MOESI also satisfy our invariants and lemmas, making their verification
-easier by reusing most of our MSI proof; we omit the discussion of extension of
-this proof to other protocols for space reasons.
-
-%\adamc{...and maybe we don't, for space reasons. \texttt{:-)}}
-
-%We use the terms write to mean handling a store request, \ie{} executing the
-%StoreReq transition and read to mean handling a load request, \ie{} executing
-%the LoadReq transition.
-
-The lemmas and invariants presented below hold after any number of transitions
-starting from the initial state.
-
-While we have proofs in Coq for each lemma, our discussion will be mainly
-informal for brevity. All the proofs can be found in our supplementary material.
-%\adamc{or just remind of the source code URL given earlier}
-
-%\subsection{Invariants}
-%\label{firstLevel}
-Intuitively our protocol is correct because it is guaranteed that any cache that attempts to
-handle a load request has the ``latest'' value.
-
-\begin{lemma}
-\textit{latestValue}:
-At any time $T$, a cache $c$ that is \clean{} (defined below) for an address
-$a$ will have the most up-to-date value for that address, \ie{}
-\begin{enumerate}
-\item $\dt(c,a) = v_0$ and no store request $\toM(\st, a, v)$ has been processed at
-any time $T'$ such that $T' < T$ or
-\item There is a store request $\toM(\st, a, v)$ that was processed at time $T_q$ such that
-$T_q < T \wedge \dt(c,a) = v$ and no other store
-request $\toM(\st, a, v')$ was processed at any time $T'$ such that $T_q < T' < T$.
-\end{enumerate}
-\label{latestValue}
-\end{lemma}
-
-\vspace{-.15in}
-
-\begin{defn}
-\textit{clean}: A cache is said to be \textit{clean} for an address $a$ if and only if
-the state of the cache for that address is at least \Sh{} and the directory 
-state for any of its children is at most \Sh.
-\label{clean}
-\end{defn}
-
-It is pretty intuitive to prove the store atomicity property given
-Lemma~\ref{latestValue}. A leaf cache can respond to a request only when it
-has the required permission, at which point it is guaranteed to have the latest
-value, because it is \textit{clean} (as it has no children).
-
-Lemma~\ref{latestValue} requires further decomposition into more
-lemmas.
-
-\begin{lemma}
-\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
-that neither is an ancestor of the other in the cache-tree hierarchy (where
-the ancestor relation is the reflexive-transitive closure of the parent relation),
-for each address $a$, $\s(c_1,a)$ and $\s(c_2,a)$ are
-\textit{compatible} (defined below).
-\label{nonAncestorCompatible}
-\end{lemma}
-\begin{defn}
-\textit{compatible}: Values of states $x_1$ and $x_2$ are \textit{compatible}
-iff whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In.
-\label{compatible}
-\end{defn}
-
-It is easy to see that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also
-compatible if $x_1$ and $x_2$ are compatible.
-
-\begin{lemma}
-\textit{noTransitWrite}: Whenever data for an address $a$ is in transit (\ie{}
-$\forall T. \; T_s \le T \le T_r$ where $T_s$ is the time of sending the data and
-$T_r$ the time of receiving the data), no cache can process a store request for
-$a$, and the data must be sent from a \textit{clean} cache.
-\label{noTransitWrite}
-\end{lemma}
-
-\begin{lemma}
-\textit{conservative}: The state of an address in a cache is never greater than
-the corresponding entry in its parent's directory.
-\label{conservative}
-\end{lemma}
-
-\begin{lemma}
-\textit{localCompatibility1}:
-The state of a cache is never less than its directory entry for any of
-its children.
-\label{localCompatibility1}
-\end{lemma}
-\begin{lemma}
-\textit{localCompatibility2}: Let $c_1$ and $c_2$ be distinct caches with the
-same parent $p$. Then, $\s(c_1, a)$ and $\s(c_2, a)$ are
-\textit{compatible}.
-\label{localCompatibility2}
-\end{lemma}
-
-Lemmas~\ref{conservative} and~\ref{localCompatibility1} imply a
-$\ge$ relation between the states of a parent and its child for any address.
-Since the ancestor relation is the reflexive-transitive closure of the parent
-relation, the $\ge$ relation is preserved between the states of two caches $p$
-and $c$ where $p$ is the ancestor of $c$.
-
-Any two caches $c_1, c_2$ neither of which is an ancestor of the other have a
-lowest common ancestor $x$, which has
-two distinct children $x_1$ and $x_2$ such that they are the ancestors of $c_1$
-and $c_2$ respectively. For any address $a$, $\s(x_1, a)$ and
-$\s(x_2, a)$ are compatible by Lemma~\ref{localCompatibility2}. From
-the previous paragraph about the relation between the states of a cache and its
-ancestor, it follows that $c_1$ and $c_2$ are also compatible, proving
-Lemma~\ref{nonAncestorCompatible}.
-
-Using Lemmas~\ref{nonAncestorCompatible} and~\ref{noTransitWrite}, one can
-prove Lemma~\ref{latestValue}. The proof is by induction on time $T$; we assume
-that any \clean{} (node, address) pair has the latest value until time
-$T$. Consider a (node, addr) pair at time $T+1$.
-If the pair was \clean{} at $T$, then no cache,
-save itself (because of Lemma~\ref{nonAncestorCompatible}), can process any
-store request for that address, thus continuing to be \clean{} and having the latest value.
-And if the pair was not \clean{} at time $T$, it will receive data from a cache
-that is \clean{} for the (node, address) pair, and no store
-request can be processed when the data is in transit (by
-Lemma~\ref{noTransitWrite}). Therefore, by the induction hypothesis the cache will
-continue to have the latest value whenever it is \clean{} for an address at time $T+1$.
-
-\subsection{Key Insight for Proving Lemmas}
-
-We present what we believe to be the key insight for proving any
-cache-coherence protocol.  In fact, the request/response message semantics for
-the MSI protocol was designed to satisfy the following three invariants.
-
-\begin{inv}
-\textit{stateChange}:
-\begin{enumerate}
-\item A cache downgrades its state for an address iff it sends a response for that address.
-\item A parent upgrades the directory entry for an address for one of its
-children iff it sends a response for that child and address.
-\item The state or directory entry changes only on sending or
-receiving a response $(\resp, (a, y, x, d))$, to $x$.
-\end{enumerate}
-\label{stateChange}
-\end{inv}
-
-\vspace{-.1in}
-
-Invariant~\ref{stateChange} ensures that a response message acts as a sync
-between a parent's directory entry for a child and the state of the child, as
-both the sender and the receiver change their respective directory entry or
-state to the same $x$ in a response $(\resp, (a, y, x, d))$
-
-\begin{inv}
-\textit{noCross}: A response from a parent to its child for an address cannot
-be in transit while another response from the same child to the parent for the
-same address is in transit.
-\label{noCross}
-\end{inv}
-
-\begin{inv}
-\textit{respFIFO}: Responses are received in the order in which they are sent,
-for a source-destination pair, for the same address.
-\label{respFIFO}
-\end{inv}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.45]{resps}
-\caption{An example showing a sequence of responses between a child and its parent}
-\label{resps}
-\end{figure}
-
-Figure~\ref{resps} shows an example of the effect of
-Invariants~\ref{stateChange}, \ref{noCross} and \ref{respFIFO}. The
-directory entry of the parent or the state of the child changes on sending and
-receiving a response, and the new state is shown at the points of sending or
-receiving responses.  Both the cache and its parent follow the same sequence of
-state and directory entry changes, but the receiver always lags slightly behind
-the sender.  The receiver cannot send any response before receiving because of
-Invariant~\ref{noCross}, thereby eliminating any scenarios where the lag
-complicates reasoning.
-
-Using these three invariants (Invariants~\ref{stateChange}, \ref{noCross} and
-\ref{respFIFO}), it is easy to see that Lemma~\ref{conservative} holds.
-Intuitively, going back to Figure~\ref{resps}, we can see that at all times,
-either there is no response in flight, only responses from the parent are in
-flight, or only responses from the child are in flight. A response from child
-downgrades the state, and a response from parent upgrades the directory. Since
-the other side simply lags behind, Lemma~\ref{conservative} holds.
-Since Invariants~\ref{stateChange}, \ref{noCross} and \ref{respFIFO} deal with
-just one cache's state for a single address and the corresponding directory
-entry, these properties can also potentially be verified using model-checking.
-Thus, if some other protocol is verified to obey Invariants~\ref{stateChange},
-\ref{noCross} and \ref{respFIFO} using model-checking, then
-Lemma~\ref{conservative} is already automatically verified for that
-protocol.
-
-Analyzing a parent-child pair will suffice in proving
-Lemmas~\ref{localCompatibility1} and \ref{localCompatibility2}.  Proof of
-Lemma~\ref{noTransitWrite} is similar to the proof of
-Lemma~\ref{nonAncestorCompatible}.
-
-%Figure \ref{something} gives the structure of the proof for Theorem \ref{ccorrect}
-%\begin{figure*}
-%\begin{prooftree}
-%\AxiomC{stateChange}
-%\AxiomC{noCross}
-%\AxiomC{respFifo}
-%\TrinaryInfC{conservative}
-%\AxiomC{localCompatibility1}
-%\AxiomC{localCompatibility2}
-%\TrinaryInfC{nonAncestorCompatible}
-%\AxiomC{noTransitWrite}
-%\BinaryInfC{latestValue}
-%\AxiomC{$\ldots$ (trivial)}
-%\BinaryInfC{StoreAtomicity}
-%\UnaryInfC{$M_c \sqsubseteq M_m$}
-%\end{prooftree}
-%\caption{Proving $M_c \sqsubseteq M_m$}
-%\label{something}
-%\end{figure*}
-
-\section{The Final Result}\label{sec:finalresult}
-
-With our two main results about optimized processors and memories, it
-is now straightforward to complete the correctness proof of the
-composed optimized system.
-
-First, we need to know that, whenever the simple memory can generate
-some trace of messages, it could also generate the same trace with all
-speculative messages removed.  We need this property to justify the
-introduction of speculation, during our final series of refinements
-from the optimized system to SC.
-
-\begin{theorem}
-\label{mcorrect}
-$M_m \sqsubseteq_{\noSpec^n} M_m$
-\end{theorem}
-\begin{proof}
-By induction on traces, with an identity abstraction function.
-\end{proof}
-
-That theorem turns out to be the crucial ingredient to justify placing
-a speculative processor in-context with simple memory.
-
-\begin{theorem}
-\label{ocorrect1}
-$\text{P$^n_\text{so}$} + M_m \sqsubseteq \text{P$^n_\text{ref}$} + M_m$
-\end{theorem}
-\begin{proof}
-Follows from Theorem \ref{liftplus} (one of our preliminary results
-about $+$), Corollary \ref{ges}, and Theorem \ref{mcorrect}.
-\end{proof}
-
-The last theorem kept the memory the same while refining the
-processor.  The next one does the opposite, switching out memory.
-
-\begin{theorem}
-\label{ocorrect2}
-$\text{P$^n_\text{so}$} + M_c \sqsubseteq \text{P$^n_\text{so}$} + M_m$
-\end{theorem}
-\begin{proof}
-Follows from Theorems \ref{ccorrect} and \ref{liftplus} plus reflexivity of $\sqsubseteq$ (Theorem \ref{transitive}).
-\end{proof}
-
-\begin{theorem}
-\label{ofull}
-$\text{P$^n_\text{so}$} + M_c \sqsubseteq SC$
-\end{theorem}
-\begin{proof}
-We twice apply $\sqsubseteq$ transitivity (Theorem \ref{transitive}) to
-connect Theorems \ref{ocorrect2} and \ref{ocorrect1}, and use Definition \ref{sc} of $SC$
-\end{proof}

### 2014-10-30 TransitionsUsefulSecond.tex
stuff

--- 2014-10-22 TransitionsUseful.tex
@@
-\section{Labeled Transition Systems}\label{sec:lts}
-
-We make extensive use of the general theory of labeled transition
-systems, a semantics approach especially relevant to communicating
-concurrent systems.  As we are formalizing processors for
-Turing-complete machine languages, it is challenging to prove that a
-system preserves almost any aspect of processor behavior from a model
-like SC.  To focus our theorems, we pick the time-honored property of
-\emph{termination}.  An optimized system should terminate or diverge
-iff the reference system could also terminate or diverge,
-respectively.  Our basic definitions of transition systems build in
-special treatment of halting, so that we need not mention it
-explicitly in most contexts to come.
-
-\begin{defn}
-A \textbf{labeled transition system (LTS)} is a ternary relation, over
-$\mathcal S^H \times \mathcal L^\epsilon \times \mathcal S^H$, for some sets
-$\mathcal S$ of states and $\mathcal L$ of labels. We usually do not mention
-these sets explicitly, as they tend to be clear from context. We write
-$X^\epsilon$ for lifting of a set $X$ to have an extra ``empty'' element
-$\epsilon$ (like an \texttt{option} type in ML). We write $X^H$ for lifting of
-a set $X$ to have an extra ``halt'' element $H$. We also implicitly consider
-each LTS to be associated with an \emph{initial} state in $\mathcal S$.
-\end{defn}
-
-For LTS $A$, we write $\io{A}{s}{s'}{\ell}$ as shorthand for $(s,
-\ell, s') \in A$, and we write $A_0$ for $A$'s initial state. The
-intuition is that $A$ is one process within a concurrent system. The
-label $\ell$ from set $\mathcal L$ of labels is produced when $A$
-participates in some IO exchange with another process; otherwise it is
-an empty or ``silent'' label $\epsilon$.  As a shorthand, we sometimes
-omit labels for $\epsilon$ steps.  Note that the IO exchange can be
-input, output, or both, and its interpretation is dependent on the
-context.
-
-\subsection{Basic Constructions on LTSes}
-
-From an LTS representing single-step system evolution, we often want to build
-an LTS capturing arbitrary-length evolutions.
-
-\begin{defn}
-The \textbf{transitive-reflexive closure} of $A$, written $A^{*}$, is a derived
-LTS.  Where $A$'s states and labels are $\mathcal S$ and $\mathcal L$, the
-states of $A^{*}$ are $\mathcal S$, and the labels are $\mathcal L^{*}$, or
-\emph{sequences} of labels from the original system.  $A^{*}$ steps from $s$ to
-$s'$ when there exist zero or more transitions in $A$ that move from $s$ to
-$s'$.  The label of this transition is the \emph{concatenation} of all labels
-generated in $A$, where the empty or ``silent'' label $\epsilon$ is treated as
-an identity element for concatenation.
-\end{defn}
-
-We also want to compose $n$ copies of an LTS together, with no explicit
-communication between them.  We apply this construction later to lift a
-single-CPU system to a multi-CPU system.
-
-\begin{defn}
-The \textbf{$n$-repetition} of $A$, written $\smulti{n}{A}$, is a derived LTS.
-Where $A$'s states and labels are $\mathcal S$ and $\mathcal L$, the states of
-$\smulti{n}{A}$ are $\mathcal S^n$, and the labels are $[1, n] \times \mathcal
-L$, or pairs that \emph{tag} labels with which component system generated them.
-These labels are generated only when the component system generates a label.
-The whole system halts whenever one of the components halts.  We define the
-transition relation with the following inference rules. We treat $n$-tuples as
-functions keyed on numeric positions, and we write $\theta[i := v]$ for
-updating $n$-tuple $\theta$ to overwrite the $i$th position with value $v$, and
-we treat any label $(i, \epsilon)$ as $\epsilon$.
-$$\inference[i]
-{\theta(i) = a & \io{$A$}{a}{a'}{\ell} & a' \neq H}
-{\io{$\smulti{n}{A}$}{\theta}{\theta[i:=a']}{i,\ell}}
-\quad \inference[H]
-{\io{$A$}{\theta(i)}{H}{\ell}}
-{\io{$\smulti{n}{A}$}{\theta}{H}{i,\ell}}
-$$
-\end{defn}
-
-Eventually, we need processes to be able to communicate with each
-other, which we formalize via the $+$ composition operator. It
-connects same-label transitions in the two systems, treating the label
-as a cooperative communication event that may now be hidden from the
-outside world, as an $\epsilon$ label.
-
-\begin{defn}\label{plus}
-Where $A$ and $B$ are two LTSes sharing labels set $\mathcal L$,
-and with state sets $\mathcal S_A$ and
-$\mathcal S_B$ respectively, the \textbf{communicating composition} $A + B$ is
-a new LTS with states $\mathcal S_A \times \mathcal S_B$ and an empty label set,
-defined as follows:
-\vspace{-.2in}\begin{center}
-$$\inference[$A$]
-{\io{$A$}{a}{a'}{} & a' \neq H}
-{\io{$A + B$}{a, b}{a', b}{}}
-\quad \inference[$B$]
-{\io{$B$}{b}{b'}{} & b' \neq H}
-{\io{$A + B$}{a, b}{a, b'}{}}$$
-$$\inference[H$_A$]
-{\io{$A$}{a}{H}{}}
-{\io{$A + B$}{a, b}{H}{}}
-\quad \inference[H$_B$]
-{\io{$B$}{b}{H}{}}
-{\io{$A + B$}{a, b}{H}{}}$$
-$$\inference[Join]
-{\io{$A$}{a}{a'}{\ell} & \io{$B$}{b}{b'}{\ell} & a', b' \neq H}
-{\io{$A + B$}{a, b}{a', b'}{}}$$
-\end{center}
-\end{defn}
-
-\subsection{Refinement Between LTSes}
-
-We need a notion of when one LTS ``implements'' another.  Intuitively,
-transition labels and halting are all that the outside world
-can observe. Two systems that produce identical labels and termination behavior
-under all circumstances can be considered as safe substitutes for
-one another. We will only need an asymmetrical notion of compatibility:
-
-\begin{defn}\label{refines}
-For some label domain $\mathcal L$, let $f : \mathcal L \to \mathcal L^\epsilon$ 
-be a function that is able to replace labels with alternative
-labels, or erase them altogether. Let LTSes $A$ and $B$ have the same label
-set $\mathcal L$. We say that
-\textbf{$A$ trace-refines $B$ w.r.t. $f$}, or $A \sqsubseteq_f B$, if:
-\begin{multline*}
-\forall s_A, \eta. \; \io{$A^{*}$}{A_0}{s_A}{\eta} \Rightarrow \exists s_B. \;
-\io{$B^{*}$}{B_0}{s_B}{f(\eta)} \; \wedge \\
-(s_A = H \Leftrightarrow s_B = H)
-\end{multline*}
-Each label in the trace is replaced by the mapping of $f$ on it, and
-labels mapped to $\epsilon$ by $f$ are dropped. $f$ is overloaded
-to denote that multi-label version when applied to $\eta$.
-\end{defn}
-
-As a shorthand, we write $A \sqsubseteq B$ for $A \sqsubseteq_\mathsf{id} B$,
-for $\mathsf{id}$ an identity function, forcing traces in the two systems to
-match exactly. Under this notion of identical traces, we say that $A$
-is sound w.r.t. $B$.
-
-\subsection{A Few Useful Lemmas}
-
-The reflexivity and transitivity properties of refinement are crucial to help us
-structure modular proofs.
-
-\begin{theorem}\label{transitive}
-$\sqsubseteq$ is reflexive and transitive.
-\end{theorem}
-
-Another crucial proof step is to lift a single-process result to the $n$-way
-replication of that process.
-
-\begin{theorem}\label{liftn}
-If $A \sqsubseteq_f B$, then $\smulti{n}{A} \sqsubseteq_{f^n} \smulti{n}{B}$, where
-$f^n$ is $f$ lifted appropriately to deal with indices ($f^n(i, \ell) =
-(i, \ell')$ when $f(\ell) = \ell'$, and $f^n(i, \ell) = \epsilon$ when
-$f(\ell) = \epsilon$).
-\end{theorem}
-
-We will also need a result applying to communicating compositions.
-
-\begin{theorem}\label{liftplus}
-If $A \sqsubseteq_f A'$ and $B \sqsubseteq_f B'$, then $A + B \sqsubseteq A' +
-B'$. In other words, if systems $A$ and $B$ are individually simulated by $A'$ and
-$B'$ on identical alteration of the traces, then the composed
-system $A+B$ will be sound with respect to $A'+B'$.
-\end{theorem}
-
-%\begin{figure}
-%\small
-%\centering
-%\begin{boxedminipage}[c]{.48\textwidth}
-%%\centering
-%\inference
-%[Halt]
-%{\theta(i) = (s, \pc) & \dec(s,\pc) = H}
-%{\io{SC}{\theta,m}{H}{}}
-%
-%\inference
-%[NonMem]
-%{\theta(i) = (s, \pc) & \dec(s,\pc) = (\nm, x) \\ \exec(s, \pc, (\nm, x)) = (\pc', s')}
-%{\io{SC}{\theta,m}{\theta[i \coloneqq (s', \pc')], m}{}}
-%
-%\inference
-%[Load]
-%{\theta(i) = (s,\pc) & \dec(s,\pc) = (\ld, x, a) \\ \exec(s, \pc, (\ld, x, m(a))) = (\pc', s')}
-%{\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m}{}}
-%
-%\inference
-%[Store]
-%{\theta(i) = (s,\pc) & \dec(s,\pc) = (\st, a, v) \\ \exec(s, \pc, (\st)) = (\pc', s')}
-%{\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m[a\coloneqq v]}{}}
-%\end{boxedminipage}
-%\caption{LTS for sequential consistency with $n$ simple processors}
-%\label{Ref}
-%\end{figure}
-%
-\section{Decomposing a Multi-Processor System}
-\label{sec:store-atomicity}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.43]{zoom}
-\caption{Components of a multi-processors system}
-\label{zoom}
-\end{figure}
-
-Most any conventional multi-processor system can be divided into 3 components as shown in
-Figure \ref{zoom}. 
-We show the top-level system design at top right, and we zoom in on boxes standing for the
-memory system and a single processor ($P_i$).
-The processor component $P_i$ can be implemented in a
-variety of ways, from literal execution of instructions in program order, to complex speculation
-with instructions run simultaneously to exploit parallelism. The memory component is normally implemented
-using a hierarchy of caches, in order to increase the performance of the
-overall system, since the latency of accessing a memory directly is very slow
-compared to accessing a much smaller cache. The component between each
-processor and the global memory sub-system contains local buffers, $B_i$, specific to
-the processor $P_i$ they are connected to.
-
-\begin{figure}
-\small
-\centering
-\begin{boxedminipage}[c]{.48\textwidth}
-\inference
-[Load]
-{}
-{\io{$M_m$}{m}{m}{p, (\ld\req,a),(\ld\resp,m(a), \_)}}
-
-\inference
-[Store]
-{}
-{\io{$M_m$}{m}{m[a\coloneqq v]}{p, (\st, a, v)}}
-
-\end{boxedminipage}
-\caption{LTS for a simple memory}
-\label{$M_m$}
-\end{figure}
-
-Popular instruction set architectures do not guarantee sequential
-consistency.  That is, their behavior makes it unsound to imagine that
-the different processors of a shared-memory system are running
-according to simple interleaving, where each instruction executes
-atomically.  However, we want to emphasize that, in every weak-memory
-system we are aware of, \emph{the main memory still exposes atomic
-  loads and stores}!  The weaker semantics arises only because of (1)
-reordering of memory instructions by the core $P_i$ and/or (2) the
-properties of the local buffers $B_i$ connected to each processor
-$P_i$.  For instance, if $B_i$ were a store buffer, into which the
-committed stores from an in-order processor $P_i$ are enqueued, and if
-any ``subsequent'' load from the same processor $P_i$ reads the latest
-value from the store buffer, then one can establish that such a
-multi-processor system implements the TSO~\cite{x86tsocacm10} memory
-model.
-
-So, we focus on this opportunity to simplify proof decomposition.  We
-will prove that our main memory component satisfies an intuitive
-\emph{store atomicity} property.  Despite the potential red herring of
-the word ``atomicity,'' this specification is appropriate even for
-implementation of weaker memory models than SC.  Store atomicity can
-be understood via the operational semantics of Figure \ref{$M_m$},
-describing an LTS that receives load and store requests ($\ld$ and
-$\st$) from processors and sends back load responses ($\ld\resp$);
-$p$ is the processor which sends and receives the request.
-
-The memory component is composed of a \emph{hierarchy of caches}, where each
-cache node (labeled like ``L1,'' ``L2,'' etc.) communicates only with its
-neighbors in the graph.  Each processor has a dedicated L1 cache, from which we
-do our best to satisfy all memory requests, to avoid the latency of a
-round-trip with main memory.  With an L1 cache miss, we may still find a hit in
-a parent cache below the level of main memory, realizing a smaller but still
-significant speedup.  Therefore it is the responsibility of the hierarchy of
-caches, which forms the memory sub-component, to implement the store atomicity
-property. In fact, as we will prove in Section \ref{sec:cc}, the purpose of the
-cache-coherence protocol is to establish this invariant for the memory
-sub-component.  Concretely, we have verified a \emph{directory-based} protocol
-for coordinating an arbitrary tree of caches, where each node stores a
-conservative approximation of its children's states.
-
-As an instance of the above decomposition, we will prove that a multi-processor
-system with no local buffering in between the processor and the memory
-components indeed implements sequential consistency. We implement a highly
-speculative processor that executes instructions and issues loads out of
-order, but commits instructions (once some ``verification'' is done) in order.
-
-The processor itself can be decomposed into several components. In the zoomed-in
-version of Figure \ref{zoom}, we show a highly speculative, out-of-order
-issue processor. We have the normal architectural state, such as values of
-registers.  Our proofs are generic over \emph{a family of instruction set
-architectures}, with parameters for opcode sets and functions for executing
-opcodes and decoding them from memory.  Other key components are a \emph{branch
-predictor}, which guesses at the control-flow path that a processor will
-follow, to facilitate speculation; and a \emph{reorder buffer (ROB)}, which
-decides which instructions along that path to try executing ahead of schedule.
-Our proofs apply to an arbitrary branch predictor, and they work for any
-reorder buffer satisfying a simple semantic condition.
-
-Figure~\ref{proofs} gives the overall proof structure that we employ to verify
-the system of Figure~\ref{zoom}. As we will explain later, $P_\text{so}$
-represents the speculative out-of-order processor, $M_c$ the cache memory,
-$M_m$ the simple memory, $P_\text{ref}$ a simple decoupled processor, and
-SC the simple composition of P$^n_\text{ref}+ M_m$ executing each
-instruction atomically, thus implementing sequential consistency.
-
-Our current framework establishes theorems of the form ``if system $A$ has a run
-with some particular observable behavior, then system $B$ also has a run with
-the same behavior.''  In this sense, we say that $A$ correctly implements $B$.
-Other important properties, such as \emph{deadlock freedom} for $A$ (which
-might get stuck without producing any useful behavior), we leave for future
-work.
-
-\begin{figure}
-\includegraphics[scale=.45]{proofs}
-\caption{Overall proof structure}
-\label{proofs}
-\end{figure}
-
 \section{Specifying Sequential Consistency}\label{sec:sc}
@@
 \end{proof}
-
-\section{Cache-Based Memory System}\label{sec:cc}
-
-We now turn our attention toward a more efficient implementation of
-memory.  It is physically infeasible to connect many processors
-directly to a single memory with low enough latency to handle most
-memory requests quickly enough.  Instead, each processor will have its
-own \emph{L1 cache} of recently used memory cells.  Whenever possible, we
-want to service memory requests from processors' own caches, only
-contacting memory for addresses not found in the caches.  To further
-avoid contacting main memory, we create additional higher-level
-caches, connecting each L1 cache to an L2 cache, potentially
-connecting L2 caches to L3, etc., bottoming out in main memory at the
-root of our tree. The zoomed-in portion of the memory in Figure~\ref{zoom}
-diagrams the basic layout.
-
-Now we have concurrent interaction of many processors with many caches
-with main memory, and the relationship with the original $M_m$ system
-is far from direct.  However, this intricate concurrent execution is
-crucial to hiding the latency of main-memory access.
-Figure~\ref{cache} formalizes as an LTS $M_c$ the algorithm we implemented,
-for providing the memory abstraction on top of a cache hierarchy.
-We have what is called an \emph{invalidating directory-based hierarchical
-cache-coherence protocol}.  It is a transliteration of a realistic
-algorithm implemented in the Bluespec hardware description
-language~\cite{Bluespec:TFRG}.
-
-We now try to explain aspects of this transition system at an
-intuitive level.  Thanks to the overall modular proof structure in
-this paper, it is not essential to understand the details of the cache
-system, to understand any other part of our development.  Therefore,
-some readers may want to skip this explanation upon first reading,
-skipping to the next section for a key theorem relating $M_c$ and $M_m$.
-
-We describe a state of the system using 8 fields: $\dt$, $\ch$, $\s$, $\dst$,
-$\wt$, $\dwt$ which are summarized in Figure~\ref{fields}.
-We use $\parent(c,p)$ to denote that $p$ is the parent of $c$ and $\leaf(c)$ to 
-denote that $c$ is a leaf or L1 cache.
-
-\begin{figure}
-\centering
-\begin{tabular}{|l|p{5.5cm}|}
-\hline
-$\dt(c,a)$ & Data in cache $c$ for address $a$\\
-$\s(c,a)$ & Coherence state of cache $c$ for address $a$\\
-$\dst(p,c,a)$ & Directory state of cache $p$ for child $c$ and address $a$\\
-$\wt(c,a)$ & Gives cache $c$'s desired coherence state to upgrade to (for address $a$), or $\epsilon$ if none\\
-$\dwt(p,c,a)$ & Gives cache $p$'s desired coherence state for its child $c$ to downgrade to (for address $a$), or $\epsilon$ if none\\
-$\ch$ & Channels between nodes in the system\\
-\hline
-\end{tabular}
-\caption{Fields in a cache-based memory system}
-\label{fields}
-\end{figure}
-
-Fields $\dt, \s$ are maps over pairs of cache names and memory addresses.
-The $\dt$ map assigns each pair a memory-cell value, and the $\s$ map
-assigns each pair a coherence state.  A coherence state is $M, S$ or
-$I$, broadly representing permissions to modify, read, or do nothing
-with an address, respectively, the decreasing permissions denoted by $M > S > I$. This interpretation of permissions is
-accurate at the leaf caches connected to the processors (\ie{} L1 caches),
-but the interpretation is subtly different at the higher level caches.
-
-Mapping $\wt$ goes from cache name and address to either $\epsilon$ or a
-coherence state.  It denotes what permission the cache is waiting for,
-if any.  That is, a cache may have decided to \emph{upgrade} its
-coherence state for some address to a more permissive value, but the
-cache is waiting for acknowledgment from another cache before
-finalizing the upgrade.
-
-Since we have a directory-based protocol, the $\s$ and
-$\wt$ maps have mirror-image counterparts $\dst$ and $\dwt$.
-The former maps triples of parent name, child name, and address to
-coherence state. It represents the parent's notion of the
-coherence state of the child for that address. We later prove that this notion
-is always conservative, \ie{} if the parent assumes that a child does not have
-a particular permission, then it is guaranteed in this system that the child
-will not have that permission.  The $\dwt$ field also represents a mapping from parent
-name, child name, and address to either $\epsilon$ or coherence state. It
-denotes whether the parent is waiting for any downgrade response from the
-corresponding child, and if so, the coherence state that the child must
-downgrade to as well.
-
-All communication channels in the system are represented by field $\ch$.
-The channels are asymmetric as
-follows. Channels going from parent to children (indexed by the names of the
-parent and child) maintain FIFO order of messages between the same parent-child
-pair. They carry both request messages (asking the child to downgrade) and
-response messages (acknowledging that the child can upgrade). Channels going
-from children to their parents are further split into request and response
-channels, thus indexed by tuples (child name, parent name, request or response).
-These channels need not maintain the FIFO order for messages between the same
-child-parent pair (notice the use of $\coloncolon$ to add and remove elements
-from the parent to child channel, as opposed to $\uplus$ for the 
-child to parent channel). The reason is that only one request or one response is
-sent for a particular address between a child-parent pair -- this invariant can
-be proved for the LTS in  Figure \ref{cache}. The order between a request and
-response message sent from the same child to its parent is lost, but our
-protocol ensures that nothing bad happens as a result.  Notice rules
-ParentRecvRsp and ParentRecvReq in Figure \ref{cache}, each of which looks at the
-previous coherence state of the incoming message and compares it with its own
-directory state to decide whether to process the message.
-
-%\medskip
-
-Here is an intuition on how the transitions work in the common case.  A cache
-can decide, spontaneously, to upgrade its coherence state, in which case it
-sends an upgrade request to its parent. The parent then makes a local decision
-on whether to send a response to the requesting child or not, based on its
-directory approximation and its own coherence state $\s$. If $\s$ is lower than
-the requested upgrade, then it cannot handle the request, and instead must
-decide to upgrade $\s$.  Once the parent's $\s$ is not lower than the requested
-upgrade, it makes sure that the rest of its children are ``compatible'' with
-the requested upgrade (given by $\compat$ definition below).  If not, the
-parent must send requests to the incompatible children to downgrade. Finally, when
-the $\s$'s upgrade and children's downgrade responses are all received, the
-original request can be responded to. The transitions for sending requests are
-allowed to happen spontaneously without any trigger. An L1 cache can process a
-load request or a store request only if it is in at least the \Sh{} state or
-\Mo{} state, respectively, for the requested address; otherwise it must
-spontaneously request to upgrade.
-
-\begin{defn}
-\begin{displaymath}
-\compat(p, c, x, a) = \left\{
-\begin{array}{ll}
-\forall c' \neq c. \; \dst(p,c',a) = I &: x = M\\
-\forall c' \neq c. \; \dst(p,c',a) \le S &: x = S\\
-\end{array}
-\right.
-\end{displaymath}
-\end{defn}
-
-The reason for having separate request and response channels for the same
-child-parent pair is to avoid deadlocks. Throughout the protocol, we maintain
-the invariant
-that a downgrade request from the parent always gets priority, \ie{} if a child
-is waiting for an upgrade response, it must still try to satisfy a downgrade
-request from the parent. If both the parent and the child are waiting
-for their respective responses, there will be a deadlock. This tie cannot
-be broken arbitrarily: if the parent concedes and starts processing a child's
-request, then it may have to send downgrade requests to other children to make
-the states compatible, but the other children may have sent upgrade requests
-and remain waiting, leading to a deadlock. The only way to break the
-vicious waiting cycle is for the children to concede and respond when they get
-downgrade requests. Of course, they have to downgrade the states of their own
-children first, and this process goes on recursively until the downgrade requests
-reach the leaf caches, which can respond immediately.
-
-A complication arises because a cache can voluntarily decide to downgrade its
-state.  This transition is used to model invalidation of cache lines to make
-room for a different location (we permit a cache to downgrade from $M$ to $S$
-instead of only to $I$).
-As a result, the parent's $\dst$ and the
-corresponding $\s$ of the child may go out of sync, leading to the parent
-requesting a child to downgrade when it already has. To handle this situation,
-the child has to drop the downgrade request when it has already downgraded to
-the required state (Rule DropReq in Figure \ref{cache}), to avoid deadlocks
-by not dequeuing the request.
-
-%Dropping the request
-%removes the entry from the buffer, allowing the child to process further messages
-%from the parent -- otherwise there would have been a deadlock.
-
-\begin{figure*}
-\small
-\centering
-
-\begin{boxedminipage}[c]{.95\textwidth}
-
-\textbf{Processor/Memory Interface}
-
-\inference[Load]
-{\s(c,a) \geq S}
-{\io{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt}
-{\dt, \ch, \s, \dst, \wt, \dwt}{c,((\ld\req, a),(\ld\resp,v), \_)}}
-\vspace{.1in}
-\inference[Store]
-{s(c,a) \geq M}
-{\io{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt}
-{\dt[(c, a) \coloneqq v], \ch, \s, \dst, \wt, \dwt}{c,(\st,a,v)}}
-
-\textbf{Child Upgrade}
-
-\inference[ChildSendReq]
-{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt}
-{\dt, \ch[ (c, p, \req) \coloneqq (a, \s(c, a), x) \uplus \ch(c,p,\req)], \s, \dst,\\
-\wt[(c,a)\coloneqq x], \dwt}}
-\vspace{.1in}
-\inference[ParentRecvReq]
-{\parent(c,p) & \ch(c, p, \req) = \{(a, y, x)\} \uplus \rs & \s(p, a) \geq x
-\\ \compat(p, c, x, a) & \dwt(p,c,a) = \epsilon & \dst(p,c,a) \le y}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt}
-{\dt, \ch[(c,p,\req) \coloneqq \rs][(p, c) \coloneqq (\resp, (a, \dst(p,c,a), x,\\ \;\;\;\;
-\ite{\dst(p,c,a) = I}{\dt(p,a)}{\_})) \coloncolon \ch(p,c)], \\ \s, \dst[(p,c, a)\coloneqq x],
-\wt, \dwt}}
-\vspace{.1in}
-\inference[ChildRecvRsp]
-{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\resp, (a, y, x, v))}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt}
-{\dt[(c,a) \coloneqq \ite{y = I}{v}{\dt(c,a)}], \ch[(p,c) \coloneqq \rs], \s[(c,a) \coloneqq x], \dst,\\
-\wt[(c,a)\coloneqq \ite{\wt(c,a) \le x}{\epsilon}{\wt(c,a)}], \dwt}}
-
-\textbf{Parent Downgrade}
-
-\inference[ParentSendReq]
-{\parent(c,p) & \dst(p,c,a) > x & \dwt(p,c,a) = \epsilon}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt}
-{\dt, \ch[(p, c)\coloneqq (\req, (a, \dst(p, c, a), x)) \coloncolon \ch(p,c)], \s, \dst,\\
-\wt, \dwt[(p, c,a)\coloneqq x]}}
-\vspace{.1in}
-\inference[ChildRecvReq]
-{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\req, (a, y, x)) & (\forall i. \; \parent(i, c) \Rightarrow \dst(c, i, a) \le x)
-& \s(c,a) > x}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt}
-{\dt, \ch[(p,c) \coloneqq \rs][(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
-\ite{\dst(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
-\wt, \dwt}}
-\vspace{.1in}
-\inference[ParentRecvRsp]
-{\parent(c,p) & \ch(c,p,\resp) = \{(a, y, x, v)\} \uplus \rs & \dst(p,c,a) = y}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt}
-{\dt[(p,a) \coloneqq \ite{y = M}{v}{\dt(p,a)}], \ch[(c,p,\resp) \coloneqq \rs], \s, \dst[(p,c,a) \coloneqq x],\\
-\wt, \dwt[(p,c,a)\coloneqq \ite{\dwt(p,c,a) \ge x}{\epsilon}{\dwt(p,c,a)}]}}
-
-\textbf{Voluntary downgrade for replacement}
-
-\inference[VolResp]
-{\parent(c,p) & (\forall i. \; \parent(i, c) \Rightarrow \dst(c, i, a) \le x)
-& \s(c,a) > x}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt}
-{\dt, \ch[(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
-\ite{\s(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
-\wt, \dwt}}
-
-\textbf{Dropping request because of voluntary downgrade}
-
-\inference[DropReq]
-{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\req, (a, y, x)) & 
- \s(c,a) \le x}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt}
-{\dt, \ch[(p,c) \coloneqq \rs], \s, \dst,
-\wt, \dwt}}
-\end{boxedminipage}
-
-\caption{LTS for cache-coherent shared-memory system}
-\label{cache}
-\end{figure*}
-
-\section{Proof that the MSI Protocol Preserves Store Atomicity}\label{sec:ccproof}
-\label{safety}
-
-We must prove the following theorem, \ie{} the cache-based system is sound with
-respect to the simple memory.
-\begin{theorem}
-\label{ccorrect}
-$M_c \sqsubseteq M_m$
-\end{theorem}
-
-The key property we need for this proof is what we define as
-``store atomicity'' below. Throughout this section, we say
-\emph{time} to denote the number of transitions that occurred before reaching
-the specified state.
-
-\begin{defn}
-Store atomicity is defined as the following property:
-For a load (or speculative load) request $\toM(\ld, a)$ (or $\toM(\spc\ld, t, a)$)
-from any processor, the response sent at time $T$
-$\toP(\ld, v)$ (or $\toP(\spc\ld, t, v)$) should be such that
-\begin{enumerate}
-\item $v = v_0$ (the initial value of any memory address) and no store
-  request $\toM(\st, a, v')$ has been processed at any time $T'$ such
-  that $T' < T$ or
-\item There is a store request $\toM(\st, a, v)$ that was processed at time $T_q$ such that
-$T_q < T$ and no other store
-request $\toM(\st, a, v')$ was processed at any time $T'$ such that $T_q < T' < T$.
-\end{enumerate}
-\label{sa}
-\end{defn}
-
-We can easily show that any system obeying store atomicity
-is sound with respect to $M_m$. Intuitively, the store atomicity
-property precisely describes the property of a memory, \viz{} that if a store
-happens, then that location gets updated such that any subsequent load gets
-back the updated value.
-
-The proof that $M_c$ obeys store atomicity is involved enough to deserve a
-section's worth of explanation, which we now turn to.
-
-We first reduce the proof of store atomicity to several lemmas. We then
-present a key insight that we discovered for proving these lemmas in the form
-of three fundamental invariants. We believe that other invalidation-based
-protocols with inclusive caches (\ie{} those cache hierarchies where a parent
-cache contains all the locations that any of its children contains) like MESI,
-MOSI, and MOESI also satisfy our invariants and lemmas, making their verification
-easier by reusing most of our MSI proof; we omit the discussion of extension of
-this proof to other protocols for space reasons.
-
-%\adamc{...and maybe we don't, for space reasons. \texttt{:-)}}
-
-%We use the terms write to mean handling a store request, \ie{} executing the
-%StoreReq transition and read to mean handling a load request, \ie{} executing
-%the LoadReq transition.
-
-The lemmas and invariants presented below hold after any number of transitions
-starting from the initial state.
-
-While we have proofs in Coq for each lemma, our discussion will be mainly
-informal for brevity. All the proofs can be found in our supplementary material.
-%\adamc{or just remind of the source code URL given earlier}
-
-%\subsection{Invariants}
-%\label{firstLevel}
-Intuitively our protocol is correct because it is guaranteed that any cache that attempts to
-handle a load request has the ``latest'' value.
-
-\begin{lemma}
-\textit{latestValue}:
-At any time $T$, a cache $c$ that is \clean{} (defined below) for an address
-$a$ will have the most up-to-date value for that address, \ie{}
-\begin{enumerate}
-\item $\dt(c,a) = v_0$ and no store request $\toM(\st, a, v)$ has been processed at
-any time $T'$ such that $T' < T$ or
-\item There is a store request $\toM(\st, a, v)$ that was processed at time $T_q$ such that
-$T_q < T \wedge \dt(c,a) = v$ and no other store
-request $\toM(\st, a, v')$ was processed at any time $T'$ such that $T_q < T' < T$.
-\end{enumerate}
-\label{latestValue}
-\end{lemma}
-
-\vspace{-.15in}
-
-\begin{defn}
-\textit{clean}: A cache is said to be \textit{clean} for an address $a$ if and only if
-the state of the cache for that address is at least \Sh{} and the directory 
-state for any of its children is at most \Sh.
-\label{clean}
-\end{defn}
-
-It is pretty intuitive to prove the store atomicity property given
-Lemma~\ref{latestValue}. A leaf cache can respond to a request only when it
-has the required permission, at which point it is guaranteed to have the latest
-value, because it is \textit{clean} (as it has no children).
-
-Lemma~\ref{latestValue} requires further decomposition into more
-lemmas.
-
-\begin{lemma}
-\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
-that neither is an ancestor of the other in the cache-tree hierarchy (where
-the ancestor relation is the reflexive-transitive closure of the parent relation),
-for each address $a$, $\s(c_1,a)$ and $\s(c_2,a)$ are
-\textit{compatible} (defined below).
-\label{nonAncestorCompatible}
-\end{lemma}
-\begin{defn}
-\textit{compatible}: Values of states $x_1$ and $x_2$ are \textit{compatible}
-iff whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In.
-\label{compatible}
-\end{defn}
-
-It is easy to see that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also
-compatible if $x_1$ and $x_2$ are compatible.
-
-\begin{lemma}
-\textit{noTransitWrite}: Whenever data for an address $a$ is in transit (\ie{}
-$\forall T. \; T_s \le T \le T_r$ where $T_s$ is the time of sending the data and
-$T_r$ the time of receiving the data), no cache can process a store request for
-$a$, and the data must be sent from a \textit{clean} cache.
-\label{noTransitWrite}
-\end{lemma}
-
-\begin{lemma}
-\textit{conservative}: The state of an address in a cache is never greater than
-the corresponding entry in its parent's directory.
-\label{conservative}
-\end{lemma}
-
-\begin{lemma}
-\textit{localCompatibility1}:
-The state of a cache is never less than its directory entry for any of
-its children.
-\label{localCompatibility1}
-\end{lemma}
-\begin{lemma}
-\textit{localCompatibility2}: Let $c_1$ and $c_2$ be distinct caches with the
-same parent $p$. Then, $\s(c_1, a)$ and $\s(c_2, a)$ are
-\textit{compatible}.
-\label{localCompatibility2}
-\end{lemma}
-
-Lemmas~\ref{conservative} and~\ref{localCompatibility1} imply a
-$\ge$ relation between the states of a parent and its child for any address.
-Since the ancestor relation is the reflexive-transitive closure of the parent
-relation, the $\ge$ relation is preserved between the states of two caches $p$
-and $c$ where $p$ is the ancestor of $c$.
-
-Any two caches $c_1, c_2$ neither of which is an ancestor of the other have a
-lowest common ancestor $x$, which has
-two distinct children $x_1$ and $x_2$ such that they are the ancestors of $c_1$
-and $c_2$ respectively. For any address $a$, $\s(x_1, a)$ and
-$\s(x_2, a)$ are compatible by Lemma~\ref{localCompatibility2}. From
-the previous paragraph about the relation between the states of a cache and its
-ancestor, it follows that $c_1$ and $c_2$ are also compatible, proving
-Lemma~\ref{nonAncestorCompatible}.
-
-Using Lemmas~\ref{nonAncestorCompatible} and~\ref{noTransitWrite}, one can
-prove Lemma~\ref{latestValue}. The proof is by induction on time $T$; we assume
-that any \clean{} (node, address) pair has the latest value until time
-$T$. Consider a (node, addr) pair at time $T+1$.
-If the pair was \clean{} at $T$, then no cache,
-save itself (because of Lemma~\ref{nonAncestorCompatible}), can process any
-store request for that address, thus continuing to be \clean{} and having the latest value.
-And if the pair was not \clean{} at time $T$, it will receive data from a cache
-that is \clean{} for the (node, address) pair, and no store
-request can be processed when the data is in transit (by
-Lemma~\ref{noTransitWrite}). Therefore, by the induction hypothesis the cache will
-continue to have the latest value whenever it is \clean{} for an address at time $T+1$.
-
-\subsection{Key Insight for Proving Lemmas}
-
-We present what we believe to be the key insight for proving any
-cache-coherence protocol.  In fact, the request/response message semantics for
-the MSI protocol was designed to satisfy the following three invariants.
-
-\begin{inv}
-\textit{stateChange}:
-\begin{enumerate}
-\item A cache downgrades its state for an address iff it sends a response for that address.
-\item A parent upgrades the directory entry for an address for one of its
-children iff it sends a response for that child and address.
-\item The state or directory entry changes only on sending or
-receiving a response $(\resp, (a, y, x, d))$, to $x$.
-\end{enumerate}
-\label{stateChange}
-\end{inv}
-
-\vspace{-.1in}
-
-Invariant~\ref{stateChange} ensures that a response message acts as a sync
-between a parent's directory entry for a child and the state of the child, as
-both the sender and the receiver change their respective directory entry or
-state to the same $x$ in a response $(\resp, (a, y, x, d))$
-
-\begin{inv}
-\textit{noCross}: A response from a parent to its child for an address cannot
-be in transit while another response from the same child to the parent for the
-same address is in transit.
-\label{noCross}
-\end{inv}
-
-\begin{inv}
-\textit{respFIFO}: Responses are received in the order in which they are sent,
-for a source-destination pair, for the same address.
-\label{respFIFO}
-\end{inv}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.45]{resps}
-\caption{An example showing a sequence of responses between a child and its parent}
-\label{resps}
-\end{figure}
-
-Figure~\ref{resps} shows an example of the effect of
-Invariants~\ref{stateChange}, \ref{noCross} and \ref{respFIFO}. The
-directory entry of the parent or the state of the child changes on sending and
-receiving a response, and the new state is shown at the points of sending or
-receiving responses.  Both the cache and its parent follow the same sequence of
-state and directory entry changes, but the receiver always lags slightly behind
-the sender.  The receiver cannot send any response before receiving because of
-Invariant~\ref{noCross}, thereby eliminating any scenarios where the lag
-complicates reasoning.
-
-Using these three invariants (Invariants~\ref{stateChange}, \ref{noCross} and
-\ref{respFIFO}), it is easy to see that Lemma~\ref{conservative} holds.
-Intuitively, going back to Figure~\ref{resps}, we can see that at all times,
-either there is no response in flight, only responses from the parent are in
-flight, or only responses from the child are in flight. A response from child
-downgrades the state, and a response from parent upgrades the directory. Since
-the other side simply lags behind, Lemma~\ref{conservative} holds.
-Since Invariants~\ref{stateChange}, \ref{noCross} and \ref{respFIFO} deal with
-just one cache's state for a single address and the corresponding directory
-entry, these properties can also potentially be verified using model-checking.
-Thus, if some other protocol is verified to obey Invariants~\ref{stateChange},
-\ref{noCross} and \ref{respFIFO} using model-checking, then
-Lemma~\ref{conservative} is already automatically verified for that
-protocol.
-
-Analyzing a parent-child pair will suffice in proving
-Lemmas~\ref{localCompatibility1} and \ref{localCompatibility2}.  Proof of
-Lemma~\ref{noTransitWrite} is similar to the proof of
-Lemma~\ref{nonAncestorCompatible}.
-
-%Figure \ref{something} gives the structure of the proof for Theorem \ref{ccorrect}
-%\begin{figure*}
-%\begin{prooftree}
-%\AxiomC{stateChange}
-%\AxiomC{noCross}
-%\AxiomC{respFifo}
-%\TrinaryInfC{conservative}
-%\AxiomC{localCompatibility1}
-%\AxiomC{localCompatibility2}
-%\TrinaryInfC{nonAncestorCompatible}
-%\AxiomC{noTransitWrite}
-%\BinaryInfC{latestValue}
-%\AxiomC{$\ldots$ (trivial)}
-%\BinaryInfC{StoreAtomicity}
-%\UnaryInfC{$M_c \sqsubseteq M_m$}
-%\end{prooftree}
-%\caption{Proving $M_c \sqsubseteq M_m$}
-%\label{something}
-%\end{figure*}
-
-\section{The Final Result}\label{sec:finalresult}
-
-With our two main results about optimized processors and memories, it
-is now straightforward to complete the correctness proof of the
-composed optimized system.
-
-First, we need to know that, whenever the simple memory can generate
-some trace of messages, it could also generate the same trace with all
-speculative messages removed.  We need this property to justify the
-introduction of speculation, during our final series of refinements
-from the optimized system to SC.
-
-\begin{theorem}
-\label{mcorrect}
-$M_m \sqsubseteq_{\noSpec^n} M_m$
-\end{theorem}
-\begin{proof}
-By induction on traces, with an identity abstraction function.
-\end{proof}
-
-That theorem turns out to be the crucial ingredient to justify placing
-a speculative processor in-context with simple memory.
-
-\begin{theorem}
-\label{ocorrect1}
-$\text{P$^n_\text{so}$} + M_m \sqsubseteq \text{P$^n_\text{ref}$} + M_m$
-\end{theorem}
-\begin{proof}
-Follows from Theorem \ref{liftplus} (one of our preliminary results
-about $+$), Corollary \ref{ges}, and Theorem \ref{mcorrect}.
-\end{proof}
-
-The last theorem kept the memory the same while refining the
-processor.  The next one does the opposite, switching out memory.
-
-\begin{theorem}
-\label{ocorrect2}
-$\text{P$^n_\text{so}$} + M_c \sqsubseteq \text{P$^n_\text{so}$} + M_m$
-\end{theorem}
-\begin{proof}
-Follows from Theorems \ref{ccorrect} and \ref{liftplus} plus reflexivity of $\sqsubseteq$ (Theorem \ref{transitive}).
-\end{proof}
-
-\begin{theorem}
-\label{ofull}
-$\text{P$^n_\text{so}$} + M_c \sqsubseteq SC$
-\end{theorem}
-\begin{proof}
-We twice apply $\sqsubseteq$ transitivity (Theorem \ref{transitive}) to
-connect Theorems \ref{ocorrect2} and \ref{ocorrect1}, and use Definition \ref{sc} of $SC$
-\end{proof}

### 2014-10-30 TransitionsUsefulSecond1.tex
stuff

--- 2014-10-15 TransitionsUseful.tex
@@
-\section{Labeled Transition Systems}\label{sec:lts}
-
-We make extensive use of the general theory of labeled transition
-systems, a semantics approach especially relevant to communicating
-concurrent systems.  As we are formalizing processors for
-Turing-complete machine languages, it is challenging to prove that a
-system preserves almost any aspect of processor behavior from a model
-like SC.  To focus our theorems, we pick the time-honored property of
-\emph{termination}.  An optimized system should terminate or diverge
-iff the reference system could also terminate or diverge,
-respectively.  Our basic definitions of transition systems build in
-special treatment of halting, so that we need not mention it
-explicitly in most contexts to come.
-
-\begin{defn}
-A \textbf{labeled transition system (LTS)} is a ternary relation, over
-$\mathcal S^H \times \mathcal L^\epsilon \times \mathcal S^H$, for some sets
-$\mathcal S$ of states and $\mathcal L$ of labels. We usually do not mention
-these sets explicitly, as they tend to be clear from context. We write
-$X^\epsilon$ for lifting of a set $X$ to have an extra ``empty'' element
-$\epsilon$ (like an \texttt{option} type in ML). We write $X^H$ for lifting of
-a set $X$ to have an extra ``halt'' element $H$. We also implicitly consider
-each LTS to be associated with an \emph{initial} state in $\mathcal S$.
-\end{defn}
-
-For LTS $A$, we write $\io{A}{s}{s'}{\ell}$ as shorthand for $(s, \ell, s') \in
-A$, and we write $A_0$ for $A$'s initial state. The intuition is that $A$ is
-one process within a concurrent system. The label $\ell$ from set $\mathcal L$
-of labels is produced when $A$ participates in some IO exchange with another
-process; otherwise it is an empty or ``silent'' label $\epsilon$. As a
-shorthand, we sometimes omit labels for $\epsilon$ steps.
-
-\subsection{Basic Constructions on LTSes}
-
-From an LTS representing single-step system evolution, we often want to build
-an LTS capturing arbitrary-length evolutions.
-
-\begin{defn}
-The \textbf{transitive-reflexive closure} of $A$, written $A^{*}$, is a derived
-LTS.  Where $A$'s states and labels are $\mathcal S$ and $\mathcal L$, the
-states of $A^{*}$ are $\mathcal S$, and the labels are $\mathcal L^{*}$, or
-\emph{sequences} of labels from the original system.  $A^{*}$ steps from $s$ to
-$s'$ when there exist zero or more transitions in $A$ that move from $s$ to
-$s'$.  The label of this transition is the \emph{concatenation} of all labels
-generated in $A$, where the empty or ``silent'' label $\epsilon$ is treated as
-an identity element for concatenation.
-\end{defn}
-
-We also want to compose $n$ copies of an LTS together, with no explicit
-communication between them.  We apply this construction later to lift a
-single-CPU system to a multi-CPU system.
-
-\begin{defn}
-The \textbf{$n$-repetition} of $A$, written $\smulti{n}{A}$, is a derived LTS.
-Where $A$'s states and labels are $\mathcal S$ and $\mathcal L$, the states of
-$\smulti{n}{A}$ are $\mathcal S^n$, and the labels are $[1, n] \times \mathcal
-L$, or pairs that \emph{tag} labels with which component system generated them.
-These labels are generated only when the component system generates a label.
-The whole system halts whenever one of the components halts.  We define the
-transition relation with the following inference rules. We treat $n$-tuples as
-functions keyed on numeric positions, and we write $\theta[i := v]$ for
-updating $n$-tuple $\theta$ to overwrite the $i$th position with value $v$, and we treat any label $(i, \epsilon)$ as $\epsilon$.
-$$\inference[i]
-{\theta(i) = a & \io{$A$}{a}{a'}{\ell} & a' \neq H}
-{\io{$\smulti{n}{A}$}{\theta}{\theta[i:=a']}{i,\ell}}
-\quad \inference[H]
-{\io{$A$}{\theta(i)}{H}{\ell}}
-{\io{$\smulti{n}{A}$}{\theta}{H}{i,\ell}}
-$$
-\end{defn}
-
-Eventually, we need processes to be able to communicate with each
-other, which we formalize via the $+$ composition operator. It
-connects same-label transitions in the two systems, treating the label
-as a cooperative communication event that may now be hidden from the
-outside world, as an $\epsilon$ label.
-
-\begin{defn}\label{plus}
-Where $A$ and $B$ are two LTSes sharing labels set $\mathcal L$,
-and with state sets $\mathcal S_A$ and
-$\mathcal S_B$ respectively, the \textbf{communicating composition} $A + B$ is
-a new LTS with states $\mathcal S_A \times \mathcal S_B$ and an empty label set,
-defined as follows:
-\vspace{-.2in}\begin{center}
-$$\inference[$A$]
-{\io{$A$}{a}{a'}{} & a' \neq H}
-{\io{$A + B$}{a, b}{a', b}{}}
-\quad \inference[$B$]
-{\io{$B$}{b}{b'}{} & b' \neq H}
-{\io{$A + B$}{a, b}{a, b'}{}}$$
-$$\inference[H$_A$]
-{\io{$A$}{a}{H}{}}
-{\io{$A + B$}{a, b}{H}{}}
-\quad \inference[H$_B$]
-{\io{$B$}{b}{H}{}}
-{\io{$A + B$}{a, b}{H}{}}$$
-$$\inference[Join]
-{\io{$A$}{a}{a'}{\ell} & \io{$B$}{b}{b'}{\ell} & a', b' \neq H}
-{\io{$A + B$}{a, b}{a', b'}{}}$$
-\end{center}
-\end{defn}
-
-\subsection{Refinement Between LTSes}
-
-We need a notion of when one LTS ``implements'' another.  Intuitively,
-transition labels and halting are all that the outside world
-can observe. Two systems that produce identical labels and termination behavior
-under all circumstances can be considered as safe substitutes for
-one another. We will only need an asymmetrical notion of compatibility:
-
-\begin{defn}\label{refines}
-For some label domain $\mathcal L$, let $f : \mathcal L \to \mathcal L^\epsilon$ 
-be a function that is able to replace labels with alternative
-labels, or erase them altogether. Let LTSes $A$ and $B$ have the same label
-sets $\mathcal L$. We say that
-\textbf{$A$ trace-refines $B$ w.r.t. $f$}, or $A \sqsubseteq_f B$, if:
-\begin{multline*}
-\forall s_A, \eta. \; \io{$A^{*}$}{A_0}{s_A}{\eta} \Rightarrow \exists s_B. \;
-\io{$B^{*}$}{B_0}{s_B}{f(\eta)} \; \wedge \\
-(s_A = H \Leftrightarrow s_B = H)
-\end{multline*}
-Each label in the trace is replaced by the mapping of $f$ on it, and
-labels mapped to $\epsilon$ by $f$ are dropped. $f$ is overloaded
-to denote that multi-label version when applied to $\eta$.
-\end{defn}
-
-As a shorthand, we write $A \sqsubseteq B$ for $A \sqsubseteq_\mathsf{id} B$,
-for $\mathsf{id}$ an identity function, forcing traces in the two systems to
-match exactly. Under this notion of identical traces, we say that $A$
-is sound w.r.t. $B$.
-
-\subsection{A Few Useful Lemmas}
-
-The reflexivity and transitivity properties of refinement are crucial to help us
-structure modular proofs.
-
-\begin{theorem}\label{transitive}
-$\sqsubseteq$ is reflexive and transitive.
-\end{theorem}
-
-Another crucial proof step is to lift a single-process result to the $n$-way
-replication of that process.
-
-\begin{theorem}\label{liftn}
-If $A \sqsubseteq_f B$, then $\smulti{n}{A} \sqsubseteq_{f^n} \smulti{n}{B}$, where
-$f^n$ is $f$ lifted appropriately to deal with indices ($f^n(i, \ell) =
-(i, \ell')$ when $f(\ell) = \ell'$, and $f^n(i, \ell) = \epsilon$ when
-$f(\ell) = \epsilon$).
-\end{theorem}
-
-We will also need a result applying to communicating compositions.
-
-\begin{theorem}\label{liftplus}
-If $A \sqsubseteq_f A'$ and $B \sqsubseteq_f B'$, then $A + B \sqsubseteq A' +
-B'$. In other words, if systems $A$ and $B$ are individually simulated by $A'$ and
-$B'$ on identical alteration of the traces, then the composed
-system $A+B$ will be sound with respect to $A'+B'$.
-\end{theorem}
+\section{Specifying Sequential Consistency}\label{sec:sc}
 
@@
 \end{figure}
-
-\section{Specifying Sequential Consistency}\label{sec:sc}
 
@@
 define an optimized system $O$ and prove $O \sqsubseteq \text{SC}$.
+
+Note that, in this setting, an operational specification like the LTS
+for SC is precisely the right characterization of \textbf{full functional
+  correctness} for a hardware design, much as a
+precondition-postcondition pair does that job in a partial-correctness
+Hoare logic.  Our SC LTS fully constrains observable behavior of a
+system to remain consistent with simple interleaving.  Similar
+operational models are possible as top-level specifications for
+systems following weaker memory models, by giving the LTS for the \emph{Local
+Buffer} component and composing the three components simultaneously.
 
@@
 \end{proof}
-
-\section{Cache-Based Memory System}\label{sec:cc}
-
-We now turn our attention toward a more efficient implementation of
-memory.  It is physically infeasible to connect many processors
-directly to a single memory with low enough latency to handle most
-memory requests quickly enough.  Instead, each processor will have its
-own \emph{L1 cache} of recently used memory cells.  Whenever possible, we
-want to service memory requests from processors' own caches, only
-contacting memory for addresses not found in the caches.  To further
-avoid contacting main memory, we create additional higher-level
-caches, connecting each L1 cache to an L2 cache, potentially
-connecting L2 caches to L3, etc., bottoming out in main memory at the
-root of our tree. The zoomed-in portion of the memory in Figure~\ref{zoom}
-diagrams the basic layout.
-
-Now we have concurrent interaction of many processors with many caches
-with main memory, and the relationship with the original $M_m$ system
-is far from direct.  However, this intricate concurrent execution is
-crucial to hiding the latency of main-memory access.
-Figure~\ref{cache} formalizes as an LTS $M_c$ the algorithm we implemented,
-for providing the memory abstraction on top of a cache hierarchy.
-We have what is called an \emph{invalidating directory-based hierarchical
-cache-coherence protocol}.  It is a transliteration of a realistic
-algorithm implemented in the Bluespec hardware description
-language~\cite{Bluespec:TFRG}.
-
-We now try to explain aspects of this transition system at an
-intuitive level.  Thanks to the overall modular proof structure in
-this paper, it is not essential to understand the details of the cache
-system, to understand any other part of our development.  Therefore,
-some readers may want to skip this explanation upon first reading,
-skipping to the next section for a key theorem relating $M_c$ and $M_m$.
-
-We describe a state of the system using 8 fields: $\dt$, $\ch$, $\s$, $\dst$,
-$\wt$, $\dwt$, $\inp$, $\outp$. The $\inp$ and $\outp$ sets are the interfaces
-to the processors and are exactly the same as in the simple memory LTS
-(Figure~\ref{$M_m$}). The rest are summarized in Figure~\ref{fields}.
-We use $\parent(c,p)$ to denote that $p$ is the parent of $c$ and $\leaf(c)$ to 
-denote that $c$ is a leaf or L1 cache.
-
-\begin{figure}
-\centering
-\begin{tabular}{|l|p{5.5cm}|}
-\hline
-$\dt(c,a)$ & Data in cache $c$ for address $a$\\
-$\s(c,a)$ & Coherence state of cache $c$ for address $a$\\
-$\dst(p,c,a)$ & Directory state of cache $p$ for child $c$ and address $a$\\
-$\wt(c,a)$ & Gives cache $c$'s desired coherence state to upgrade to (for address $a$), or $\epsilon$ if none\\
-$\dwt(p,c,a)$ & Gives cache $p$'s desired coherence state for its child $c$ to downgrade to (for address $a$), or $\epsilon$ if none\\
-$\ch$ & Channels between nodes in the system\\
-\hline
-\end{tabular}
-\caption{Fields in a cache-based memory system}
-\label{fields}
-\end{figure}
-
-Fields $\dt, \s$ are maps over pairs of cache names and memory addresses.
-The $\dt$ map assigns each pair a memory-cell value, and the $\s$ map
-assigns each pair a coherence state.  A coherence state is $M, S$ or
-$I$, broadly representing permissions to modify, read, or do nothing
-with an address, respectively, the decreasing permissions denoted by $M > S > I$. This interpretation of permissions is
-accurate at the leaf caches connected to the processors (\ie{} L1 caches),
-but the interpretation is subtly different at the higher level caches.
-
-Mapping $\wt$ goes from cache name and address to either $\epsilon$ or a
-coherence state.  It denotes what permission the cache is waiting for,
-if any.  That is, a cache may have decided to \emph{upgrade} its
-coherence state for some address to a more permissive value, but the
-cache is waiting for acknowledgment from another cache before
-finalizing the upgrade.
-
-Since we have a directory-based protocol, the $\s$ and
-$\wt$ maps have mirror-image counterparts $\dst$ and $\dwt$.
-The former maps triples of parent name, child name, and address to
-coherence state. It represents the parent's notion of the
-coherence state of the child for that address. We later prove that this notion
-is always conservative, \ie{} if the parent assumes that a child does not have
-a particular permission, then it is guaranteed in this system that the child
-will not have that permission.  The $\dwt$ field also represents a mapping from parent
-name, child name, and address to either $\epsilon$ or coherence state. It
-denotes whether the parent is waiting for any downgrade response from the
-corresponding child, and if so, the coherence state that the child must
-downgrade to as well.
-
-All communication channels in the system are represented by field $\ch$.
-The channels are asymmetric as
-follows. Channels going from parent to children (indexed by the names of the
-parent and child) maintain FIFO order of messages between the same parent-child
-pair. They carry both request messages (asking the child to downgrade) and
-response messages (acknowledging that the child can upgrade). Channels going
-from children to their parents are further split into request and response
-channels, thus indexed by tuples (child name, parent name, request or response).
-These channels need not maintain the FIFO order for messages between the same
-child-parent pair (notice the use of $\coloncolon$ to add and remove elements
-from the parent to child channel, as opposed to $\uplus$ for the 
-child to parent channel). The reason is that only one request or one response is
-sent for a particular address between a child-parent pair -- this invariant can
-be proved for the LTS in  Figure \ref{cache}. The order between a request and
-response message sent from the same child to its parent is lost, but our
-protocol ensures that nothing bad happens as a result.  Notice rules
-ParentRecvRsp and ParentRecvReq in Figure \ref{cache}, each of which looks at the
-previous coherence state of the incoming message and compares it with its own
-directory state to decide whether to process the message.
-
-%\medskip
-
-Here is an intuition on how the transitions work in the common case.  A cache
-can decide, spontaneously, to upgrade its coherence state, in which case it
-sends an upgrade request to its parent. The parent then makes a local decision
-on whether to send a response to the requesting child or not, based on its
-directory approximation and its own coherence state $\s$. If $\s$ is lower than
-the requested upgrade, then it cannot handle the request, and instead must
-decide to upgrade $\s$.  Once the parent's $\s$ is not lower than the requested
-upgrade, it makes sure that the rest of its children are ``compatible'' with
-the requested upgrade (given by $\compat$ definition below).  If not, the
-parent must send requests to the incompatible children to downgrade. Finally, when
-the $\s$'s upgrade and children's downgrade responses are all received, the
-original request can be responded to. The transitions for sending requests are
-allowed to happen spontaneously without any trigger. An L1 cache can process a
-load request or a store request only if it is in at least the \Sh{} state or
-\Mo{} state, respectively, for the requested address; otherwise it must
-spontaneously request to upgrade.
-
-\begin{defn}
-\begin{displaymath}
-\compat(p, c, x, a) = \left\{
-\begin{array}{ll}
-\forall c' \neq c. \; \dst(p,c',a) = I &: x = M\\
-\forall c' \neq c. \; \dst(p,c',a) \le S &: x = S\\
-\end{array}
-\right.
-\end{displaymath}
-\end{defn}
-
-The reason for having separate request and response channels for the same
-child-parent pair is to avoid deadlocks. Throughout the protocol, we maintain
-the invariant
-that a downgrade request from the parent always gets priority, \ie{} if a child
-is waiting for an upgrade response, it must still try to satisfy a downgrade
-request from the parent. If both the parent and the child are waiting
-for their respective responses, there will be a deadlock. This tie cannot
-be broken arbitrarily: if the parent concedes and starts processing a child's
-request, then it may have to send downgrade requests to other children to make
-the states compatible, but the other children may have sent upgrade requests
-and remain waiting, leading to a deadlock. The only way to break the
-vicious waiting cycle is for the children to concede and respond when they get
-downgrade requests. Of course, they have to downgrade the states of their own
-children first, and this process goes on recursively until the downgrade requests
-reach the leaf caches, which can respond immediately.
-
-A complication arises because a cache can voluntarily decide to downgrade its
-state.  This transition is used to model invalidation of cache lines to make
-room for a different location (we permit a cache to downgrade from $M$ to $S$
-instead of only to $I$).
-As a result, the parent's $\dst$ and the
-corresponding $\s$ of the child may go out of sync, leading to the parent
-requesting a child to downgrade when it already has. To handle this situation,
-the child has to drop the downgrade request when it has already downgraded to
-the required state (Rule DropReq in Figure \ref{cache}), to avoid deadlocks
-by not dequeuing the request.
-
-%Dropping the request
-%removes the entry from the buffer, allowing the child to process further messages
-%from the parent -- otherwise there would have been a deadlock.
-
-\begin{figure*}
-\centering
-
-\begin{boxedminipage}[c]{.95\textwidth}
-
-\textbf{Processor/Memory Interface}
-
-\inference
-[Insert]
-{}
-{\iorn{$M_c$}{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}{\dt, \ch,
-    \s, \dst, \wt, \dwt, \inp[i \coloneqq \{q\} \uplus \inp(i)],
-    \outp}{i}{q}}
-\vspace{.1in}
-\inference
-[Remove]
-{\outp(i) = \mathit{rs} \uplus \{r\}}
-{\ioln{$M_c$}{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp[i \coloneqq \mathit{rs}]}{i}{r}}
-\vspace{.1in}
-\inference[SpecLoad]
-{\inp(c) = \{(\spc\ld, t, a)\} \uplus \rs & \s(c,a) \geq S}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
-{
-\dt, \ch, \s, \dst, \wt, \dwt, \inp[c \coloneqq \rs],\\
-\outp[c \coloneqq \outp(c) \uplus \{(\spc\ld, t, \dt(c, a))\}]
-}}
-\vspace{.1in}
-\inference[Load]
-{\inp(c) = \{(\ld, a)\} \uplus \rs & \s(c,a) \geq S}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp[c \coloneqq \rs], \\
-\outp[c \coloneqq \outp(c) \uplus \{(\ld, \dt(c, a))\}]}}
-\vspace{.1in}
-\inference[Store]
-{\inp(c) = \{(\st, a, v)\} \uplus \rs & \s(c,a) \geq M}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
-{\dt[(c, a) \coloneqq v], \ch, \s, \dst, \wt, \dwt, \inp[c \coloneqq \rs],
-  \\
-\outp[c \coloneqq \outp(c) \uplus \{(\st)\}]}}
-
-\textbf{Child Upgrade}
-
-\inference[ChildSendReq]
-{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt, \inp, \outp}
-{\dt, \ch[ (c, p, \req) \coloneqq (a, \s(c, a), x) \uplus \ch(c,p,\req)], \s, \dst,\\
-\wt[(c,a)\coloneqq x], \dwt, \inp, \outp}}
-\vspace{.1in}
-\inference[ParentRecvReq]
-{\parent(c,p) & \ch(c, p, \req) = \{(a, y, x)\} \uplus \rs & \s(p, a) \geq x
-\\ \compat(p, c, x, a) & \dwt(p,c,a) = \epsilon & \dst(p,c,a) \le y}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt, \inp, \outp}
-{\dt, \ch[(c,p,\req) \coloneqq \rs][(p, c) \coloneqq (\resp, (a, \dst(p,c,a), x,\\ \;\;\;\;
-\ite{\dst(p,c,a) = I}{\dt(p,a)}{\_})) \coloncolon \ch(p,c)], \\ \s, \dst[(p,c, a)\coloneqq x],
-\wt, \dwt, \inp, \outp}}
-\vspace{.1in}
-\inference[ChildRecvRsp]
-{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\resp, (a, y, x, v))}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt, \inp, \outp}
-{\dt[(c,a) \coloneqq \ite{y = I}{v}{\dt(c,a)}], \ch[(p,c) \coloneqq \rs], \s[(c,a) \coloneqq x], \dst,\\
-\wt[(c,a)\coloneqq \ite{\wt(c,a) \le x}{\epsilon}{\wt(c,a)}], \dwt, \inp, \outp}}
-
-\textbf{Parent Downgrade}
-
-\inference[ParentSendReq]
-{\parent(c,p) & \dst(p,c,a) > x & \dwt(p,c,a) = \epsilon}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt, \inp, \outp}
-{\dt, \ch[(p, c)\coloneqq (\req, (a, \dst(p, c, a), x)) \coloncolon \ch(p,c)], \s, \dst,\\
-\wt, \dwt[(p, c,a)\coloneqq x], \inp, \outp}}
-\vspace{.1in}
-\inference[ChildRecvReq]
-{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\req, (a, y, x)) & (\forall i. \; \parent(i, c) \Rightarrow \dst(c, i, a) \le x)
-& \s(c,a) > x}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt, \inp, \outp}
-{\dt, \ch[(p,c) \coloneqq \rs][(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
-\ite{\dst(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
-\wt, \dwt, \inp, \outp}}
-\vspace{.1in}
-\inference[ParentRecvRsp]
-{\parent(c,p) & \ch(c,p,\resp) = \{(a, y, x, v)\} \uplus \rs & \dst(p,c,a) = y}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt, \inp, \outp}
-{\dt[(p,a) \coloneqq \ite{y = M}{v}{\dt(p,a)}], \ch[(c,p,\resp) \coloneqq \rs], \s, \dst[(p,c,a) \coloneqq x],\\
-\wt, \dwt[(p,c,a)\coloneqq \ite{\dwt(p,c,a) \ge x}{\epsilon}{\dwt(p,c,a)}], \inp, \outp}}
-
-\textbf{Voluntary downgrade for replacement}
-
-\inference[VolResp]
-{\parent(c,p) & (\forall i. \; \parent(i, c) \Rightarrow \dst(c, i, a) \le x)
-& \s(c,a) > x}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt, \inp, \outp}
-{\dt, \ch[(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
-\ite{\s(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
-\wt, \dwt, \inp, \outp}}
-
-\textbf{Dropping request because of voluntary downgrade}
-
-\inference[DropReq]
-{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\req, (a, y, x)) & 
- \s(c,a) \le x}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
-{\dt, \ch[(p,c) \coloneqq \rs], \s, \dst,
-\wt, \dwt, \inp, \outp}}
-\end{boxedminipage}
-
-\caption{LTS for cache-coherent shared-memory system}
-\label{cache}
-\end{figure*}
-
-\section{Proof that the MSI Protocol Preserves Store Atomicity}\label{sec:ccproof}
-\label{safety}
-
-We must prove the following theorem, \ie{} the cache-based system is sound with
-respect to the simple memory.
-\begin{theorem}
-\label{ccorrect}
-$M_c \sqsubseteq M_m$
-\end{theorem}
-
-The key property we need for this proof is what we define as
-``store atomicity'' below. Throughout this section, we say
-\emph{time} to denote the number of transitions that occurred before reaching
-the specified state.
-
-\begin{defn}
-Store atomicity is defined as the following property:
-For a load (or speculative load) request $\toM(\ld, a)$ (or $\toM(\spc\ld, t, a)$)
-from any processor, the response sent at time $T$
-$\toP(\ld, v)$ (or $\toP(\spc\ld, t, v)$) should be such that
-\begin{enumerate}
-\item $v = v_0$ (the initial value of any memory address) and no store
-  request $\toM(\st, a, v')$ has been processed at any time $T'$ such
-  that $T' < T$ or
-\item There is a store request $\toM(\st, a, v)$ that was processed at time $T_q$ such that
-$T_q < T$ and no other store
-request $\toM(\st, a, v')$ was processed at any time $T'$ such that $T_q < T' < T$.
-\end{enumerate}
-\label{sa}
-\end{defn}
-
-We can easily show that any system obeying store atomicity
-is sound with respect to $M_m$. Intuitively, the store atomicity
-property precisely describes the property of a memory, \viz{} that if a store
-happens, then that location gets updated such that any subsequent load gets
-back the updated value.
-
-The proof that $M_c$ obeys store atomicity is involved enough to deserve a
-section's worth of explanation, which we now turn to.
-
-We first reduce the proof of store atomicity to several lemmas. We then
-present a key insight that we discovered for proving these lemmas in the form
-of three fundamental invariants. We believe that other invalidation-based
-protocols with inclusive caches (\ie{} those cache hierarchies where a parent
-cache contains all the locations that any of its children contains) like MESI,
-MOSI, and MOESI also satisfy our invariants and lemmas, making their verification
-easier by reusing most of our MSI proof; we omit the discussion of extension of
-this proof to other protocols for space reasons.
-
-%\adamc{...and maybe we don't, for space reasons. \texttt{:-)}}
-
-%We use the terms write to mean handling a store request, \ie{} executing the
-%StoreReq transition and read to mean handling a load request, \ie{} executing
-%the LoadReq transition.
-
-The lemmas and invariants presented below hold after any number of transitions
-starting from the initial state.
-
-While we have proofs in Coq for each lemma, our discussion will be mainly
-informal for brevity. All the proofs can be found in our supplementary material.
-%\adamc{or just remind of the source code URL given earlier}
-
-%\subsection{Invariants}
-%\label{firstLevel}
-Intuitively our protocol is correct because it is guaranteed that any cache that attempts to
-handle a load request has the ``latest'' value.
-
-\begin{lemma}
-\textit{latestValue}:
-At any time $T$, a cache $c$ that is \clean{} (defined below) for an address
-$a$ will have the most up-to-date value for that address, \ie{}
-\begin{enumerate}
-\item $\dt(c,a) = v_0$ and no store request $\toM(\st, a, v)$ has been processed at
-any time $T'$ such that $T' < T$ or
-\item There is a store request $\toM(\st, a, v)$ that was processed at time $T_q$ such that
-$T_q < T \wedge \dt(c,a) = v$ and no other store
-request $\toM(\st, a, v')$ was processed at any time $T'$ such that $T_q < T' < T$.
-\end{enumerate}
-\label{latestValue}
-\end{lemma}
-
-\vspace{-.15in}
-
-\begin{defn}
-\textit{clean}: A cache is said to be \textit{clean} for an address $a$ if and only if
-the state of the cache for that address is at least \Sh{} and the directory 
-state for any of its children is at most \Sh.
-\label{clean}
-\end{defn}
-
-It is pretty intuitive to prove the store atomicity property given
-Lemma~\ref{latestValue}. A leaf cache can respond to a request only when it
-has the required permission, at which point it is guaranteed to have the latest
-value, because it is \textit{clean} (as it has no children).
-
-Lemma~\ref{latestValue} requires further decomposition into more
-lemmas.
-
-\begin{lemma}
-\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
-that neither is an ancestor of the other in the cache-tree hierarchy (where
-the ancestor relation is the reflexive-transitive closure of the parent relation),
-for each address $a$, $\s(c_1,a)$ and $\s(c_2,a)$ are
-\textit{compatible} (defined below).
-\label{nonAncestorCompatible}
-\end{lemma}
-\begin{defn}
-\textit{compatible}: Values of states $x_1$ and $x_2$ are \textit{compatible}
-iff whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In.
-\label{compatible}
-\end{defn}
-
-It is easy to see that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also
-compatible if $x_1$ and $x_2$ are compatible.
-
-\begin{lemma}
-\textit{noTransitWrite}: Whenever data for an address $a$ is in transit (\ie{}
-$\forall T. \; T_s \le T \le T_r$ where $T_s$ is the time of sending the data and
-$T_r$ the time of receiving the data), no cache can process a store request for
-$a$, and the data must be sent from a \textit{clean} cache.
-\label{noTransitWrite}
-\end{lemma}
-
-\begin{lemma}
-\textit{conservative}: The state of an address in a cache is never greater than
-the corresponding entry in its parent's directory.
-\label{conservative}
-\end{lemma}
-
-\begin{lemma}
-\textit{localCompatibility1}:
-The state of a cache is never less than its directory entry for any of
-its children.
-\label{localCompatibility1}
-\end{lemma}
-\begin{lemma}
-\textit{localCompatibility2}: Let $c_1$ and $c_2$ be distinct caches with the
-same parent $p$. Then, $\s(c_1, a)$ and $\s(c_2, a)$ are
-\textit{compatible}.
-\label{localCompatibility2}
-\end{lemma}
-
-Lemmas~\ref{conservative} and~\ref{localCompatibility1} imply a
-$\ge$ relation between the states of a parent and its child for any address.
-Since the ancestor relation is the reflexive-transitive closure of the parent
-relation, the $\ge$ relation is preserved between the states of two caches $p$
-and $c$ where $p$ is the ancestor of $c$.
-
-Any two caches $c_1, c_2$ neither of which is an ancestor of the other have a
-lowest common ancestor $x$, which has
-two distinct children $x_1$ and $x_2$ such that they are the ancestors of $c_1$
-and $c_2$ respectively. For any address $a$, $\s(x_1, a)$ and
-$\s(x_2, a)$ are compatible by Lemma~\ref{localCompatibility2}. From
-the previous paragraph about the relation between the states of a cache and its
-ancestor, it follows that $c_1$ and $c_2$ are also compatible, proving
-Lemma~\ref{nonAncestorCompatible}.
-
-Using Lemmas~\ref{nonAncestorCompatible} and~\ref{noTransitWrite}, one can
-prove Lemma~\ref{latestValue}. The proof is by induction on time $T$; we assume
-that any \clean{} (node, address) pair has the latest value until time
-$T$. Consider a (node, addr) pair at time $T+1$.
-If the pair was \clean{} at $T$, then no cache,
-save itself (because of Lemma~\ref{nonAncestorCompatible}), can process any
-store request for that address, thus continuing to be \clean{} and having the latest value.
-And if the pair was not \clean{} at time $T$, it will receive data from a cache
-that is \clean{} for the (node, address) pair, and no store
-request can be processed when the data is in transit (by
-Lemma~\ref{noTransitWrite}). Therefore, by the induction hypothesis the cache will
-continue to have the latest value whenever it is \clean{} for an address at time $T+1$.
-
-\subsection{Key Insight for Proving Lemmas}
-
-We present what we believe to be the key insight for proving any
-cache-coherence protocol.  In fact, the request/response message semantics for
-the MSI protocol was designed to satisfy the following three invariants.
-
-\begin{inv}
-\textit{stateChange}:
-\begin{enumerate}
-\item A cache downgrades its state for an address iff it sends a response for that address.
-\item A parent upgrades the directory entry for an address for one of its
-children iff it sends a response for that child and address.
-\item The state or directory entry changes only on sending or
-receiving a response $(\resp, (a, y, x, d))$, to $x$.
-\end{enumerate}
-\label{stateChange}
-\end{inv}
-
-\vspace{-.1in}
-
-Invariant~\ref{stateChange} ensures that a response message acts as a sync
-between a parent's directory entry for a child and the state of the child, as
-both the sender and the receiver change their respective directory entry or
-state to the same $x$ in a response $(\resp, (a, y, x, d))$
-
-\begin{inv}
-\textit{noCross}: A response from a parent to its child for an address cannot
-be in transit while another response from the same child to the parent for the
-same address is in transit.
-\label{noCross}
-\end{inv}
-
-\begin{inv}
-\textit{respFIFO}: Responses are received in the order in which they are sent,
-for a source-destination pair, for the same address.
-\label{respFIFO}
-\end{inv}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.45]{resps}
-\caption{An example showing a sequence of responses between a child and its parent}
-\label{resps}
-\end{figure}
-
-Figure~\ref{resps} shows an example of the effect of
-Invariants~\ref{stateChange}, \ref{noCross} and \ref{respFIFO}. The
-directory entry of the parent or the state of the child changes on sending and
-receiving a response, and the new state is shown at the points of sending or
-receiving responses.  Both the cache and its parent follow the same sequence of
-state and directory entry changes, but the receiver always lags slightly behind
-the sender.  The receiver cannot send any response before receiving because of
-Invariant~\ref{noCross}, thereby eliminating any scenarios where the lag
-complicates reasoning.
-
-Using these three invariants (Invariants~\ref{stateChange}, \ref{noCross} and
-\ref{respFIFO}), it is easy to see that Lemma~\ref{conservative} holds.
-Intuitively, going back to Figure~\ref{resps}, we can see that at all times,
-either there is no response in flight, only responses from the parent are in
-flight, or only responses from the child are in flight. A response from child
-downgrades the state, and a response from parent upgrades the directory. Since
-the other side simply lags behind, Lemma~\ref{conservative} holds.
-Since Invariants~\ref{stateChange}, \ref{noCross} and \ref{respFIFO} deal with
-just one cache's state for a single address and the corresponding directory
-entry, these properties can also potentially be verified using model-checking.
-Thus, if some other protocol is verified to obey Invariants~\ref{stateChange},
-\ref{noCross} and \ref{respFIFO} using model-checking, then
-Lemma~\ref{conservative} is already automatically verified for that
-protocol.
-
-Analyzing a parent-child pair will suffice in proving
-Lemmas~\ref{localCompatibility1} and \ref{localCompatibility2}.  Proof of
-Lemma~\ref{noTransitWrite} is similar to the proof of
-Lemma~\ref{nonAncestorCompatible}.
-
-%Figure \ref{something} gives the structure of the proof for Theorem \ref{ccorrect}
-%\begin{figure*}
-%\begin{prooftree}
-%\AxiomC{stateChange}
-%\AxiomC{noCross}
-%\AxiomC{respFifo}
-%\TrinaryInfC{conservative}
-%\AxiomC{localCompatibility1}
-%\AxiomC{localCompatibility2}
-%\TrinaryInfC{nonAncestorCompatible}
-%\AxiomC{noTransitWrite}
-%\BinaryInfC{latestValue}
-%\AxiomC{$\ldots$ (trivial)}
-%\BinaryInfC{StoreAtomicity}
-%\UnaryInfC{$M_c \sqsubseteq M_m$}
-%\end{prooftree}
-%\caption{Proving $M_c \sqsubseteq M_m$}
-%\label{something}
-%\end{figure*}
-
-\section{The Final Result}\label{sec:finalresult}
-
-With our two main results about optimized processors and memories, it
-is now straightforward to complete the correctness proof of the
-composed optimized system.
-
-First, we need to know that, whenever the simple memory can generate
-some trace of messages, it could also generate the same trace with all
-speculative messages removed.  We need this property to justify the
-introduction of speculation, during our final series of refinements
-from the optimized system to SC.
-
-\begin{theorem}
-\label{mcorrect}
-$M_m \sqsubseteq_{\noSpec^n} M_m$
-\end{theorem}
-\begin{proof}
-By induction on traces, with an identity abstraction function.
-\end{proof}
-
-That theorem turns out to be the crucial ingredient to justify placing
-a speculative processor in-context with simple memory.
-
-\begin{theorem}
-\label{ocorrect1}
-$\text{P$^n_\text{so}$} + M_m \sqsubseteq \text{P$^n_\text{ref}$} + M_m$
-\end{theorem}
-\begin{proof}
-Follows from Theorem \ref{liftplus} (one of our preliminary results
-about $+$), Corollary \ref{ges}, and Theorem \ref{mcorrect}.
-\end{proof}
-
-The last theorem kept the memory the same while refining the
-processor.  The next one does the opposite, switching out memory.
-
-\begin{theorem}
-\label{ocorrect2}
-$\text{P$^n_\text{so}$} + M_c \sqsubseteq \text{P$^n_\text{so}$} + M_m$
-\end{theorem}
-\begin{proof}
-Follows from Theorems \ref{ccorrect} and \ref{liftplus} plus reflexivity of $\sqsubseteq$ (Theorem \ref{transitive}).
-\end{proof}
-
-\begin{theorem}
-\label{ofull}
-$\text{P$^n_\text{so}$} + M_c \sqsubseteq SC$
-\end{theorem}
-\begin{proof}
-We twice apply $\sqsubseteq$ transitivity (Theorem \ref{transitive}) to
-connect Theorems \ref{ocorrect2}, \ref{ocorrect1}, and \ref{scthm}.
-\end{proof}
-
-%Figure \ref{overall} gives a summary the overall proof structure that we have
-%employed to prove the correctness of the full system.
-%
-%\begin{figure*}
-%\begin{prooftree}
-%\AxiomC{$M_c \sqsubseteq M_m$}
-%\AxiomC{$\sqsubseteq$ is reflexive}
-%\AxiomC{$\begin{array}{l} \forall f, A, A', B, B'. A \sqsubseteq_f A' \wedge B \sqsubseteq_f B' \\\;\;\;\;\Rightarrow A+B \sqsubseteq A'+B'(\text{Theorem 3})\end{array}$}
-%\TrinaryInfC{P$_\text{so}^n+M_c \sqsubseteq $ P$_\text{so}^n+M_m$}
-%\end{prooftree}
-%\begin{prooftree}
-%\AxiomC{P$_\text{so} \sqsubseteq_\noSpec $ P$_\text{ref}$}
-%\AxiomC{$\begin{array}{l}\forall f, A, B, n. A \sqsubseteq_f B \Rightarrow \\ \;\;\;\;\;\;\;A^n \sqsubseteq_{f^n} B^n(\text{Theorem 2})\end{array}$}
-%\BinaryInfC{P$_\text{so}^n \sqsubseteq_\noSpec $ P$_\text{ref}^n$}
-%\AxiomC{$M_m \sqsubseteq_\noSpec M_m$}
-%\AxiomC{$\begin{array}{l} \forall f, A, A', B, B'. A \sqsubseteq_f A' \wedge B \sqsubseteq_f B' \\\;\;\;\;\Rightarrow A+B \sqsubseteq A'+B'(\text{Theorem 3})\end{array}$}
-%\TrinaryInfC{P$_\text{so}^n+M_m \sqsubseteq $ P$_\text{ref}^n+M_m$}
-%\end{prooftree}
-%\begin{prooftree}
-%\AxiomC{P$_\text{so}^n+M_c \sqsubseteq $ P$_\text{so}^n+M_m$}
-%\AxiomC{P$_\text{so}^n+M_m \sqsubseteq $ P$_\text{ref}^n+M_m$}
-%\AxiomC{P$_\text{ref}^n + M_m \sqsubseteq SC$}
-%\AxiomC{$\sqsubseteq $ is transitive}
-%\QuaternaryInfC{P$_\text{so}^n + M_c \sqsubseteq SC$}
-%\end{prooftree}
-%\caption{Overall proof structure}
-%\label{overall}
-%\end{figure*}

### 2014-10-30 TransitionsUsefulThird.tex
stuff

--- 2014-10-20 TransitionsUseful.tex
@@
-\section{Labeled Transition Systems}\label{sec:lts}
-
-We make extensive use of the general theory of labeled transition
-systems, a semantics approach especially relevant to communicating
-concurrent systems.  As we are formalizing processors for
-Turing-complete machine languages, it is challenging to prove that a
-system preserves almost any aspect of processor behavior from a model
-like SC.  To focus our theorems, we pick the time-honored property of
-\emph{termination}.  An optimized system should terminate or diverge
-iff the reference system could also terminate or diverge,
-respectively.  Our basic definitions of transition systems build in
-special treatment of halting, so that we need not mention it
-explicitly in most contexts to come.
-
-\begin{defn}
-A \textbf{labeled transition system (LTS)} is a ternary relation, over
-$\mathcal S^H \times \mathcal L^\epsilon \times \mathcal S^H$, for some sets
-$\mathcal S$ of states and $\mathcal L$ of labels. We usually do not mention
-these sets explicitly, as they tend to be clear from context. We write
-$X^\epsilon$ for lifting of a set $X$ to have an extra ``empty'' element
-$\epsilon$ (like an \texttt{option} type in ML). We write $X^H$ for lifting of
-a set $X$ to have an extra ``halt'' element $H$. We also implicitly consider
-each LTS to be associated with an \emph{initial} state in $\mathcal S$.
-\end{defn}
-
-For LTS $A$, we write $\io{A}{s}{s'}{\ell}$ as shorthand for $(s,
-\ell, s') \in A$, and we write $A_0$ for $A$'s initial state. The
-intuition is that $A$ is one process within a concurrent system. The
-label $\ell$ from set $\mathcal L$ of labels is produced when $A$
-participates in some IO exchange with another process; otherwise it is
-an empty or ``silent'' label $\epsilon$.  As a shorthand, we sometimes
-omit labels for $\epsilon$ steps.  Note that the IO exchange can be
-input, output, or both, and its interpretation is dependent on the
-context.
-
-\subsection{Basic Constructions on LTSes}
-
-From an LTS representing single-step system evolution, we often want to build
-an LTS capturing arbitrary-length evolutions.
-
-\begin{defn}
-The \textbf{transitive-reflexive closure} of $A$, written $A^{*}$, is a derived
-LTS.  Where $A$'s states and labels are $\mathcal S$ and $\mathcal L$, the
-states of $A^{*}$ are $\mathcal S$, and the labels are $\mathcal L^{*}$, or
-\emph{sequences} of labels from the original system.  $A^{*}$ steps from $s$ to
-$s'$ when there exist zero or more transitions in $A$ that move from $s$ to
-$s'$.  The label of this transition is the \emph{concatenation} of all labels
-generated in $A$, where the empty or ``silent'' label $\epsilon$ is treated as
-an identity element for concatenation.
-\end{defn}
-
-We also want to compose $n$ copies of an LTS together, with no explicit
-communication between them.  We apply this construction later to lift a
-single-CPU system to a multi-CPU system.
-
-\begin{defn}
-The \textbf{$n$-repetition} of $A$, written $\smulti{n}{A}$, is a derived LTS.
-Where $A$'s states and labels are $\mathcal S$ and $\mathcal L$, the states of
-$\smulti{n}{A}$ are $\mathcal S^n$, and the labels are $[1, n] \times \mathcal
-L$, or pairs that \emph{tag} labels with which component system generated them.
-These labels are generated only when the component system generates a label.
-The whole system halts whenever one of the components halts.  We define the
-transition relation with the following inference rules. We treat $n$-tuples as
-functions keyed on numeric positions, and we write $\theta[i := v]$ for
-updating $n$-tuple $\theta$ to overwrite the $i$th position with value $v$, and
-we treat any label $(i, \epsilon)$ as $\epsilon$.
-$$\inference[i]
-{\theta(i) = a & \io{$A$}{a}{a'}{\ell} & a' \neq H}
-{\io{$\smulti{n}{A}$}{\theta}{\theta[i:=a']}{i,\ell}}
-\quad \inference[H]
-{\io{$A$}{\theta(i)}{H}{\ell}}
-{\io{$\smulti{n}{A}$}{\theta}{H}{i,\ell}}
-$$
-\end{defn}
-
-Eventually, we need processes to be able to communicate with each
-other, which we formalize via the $+$ composition operator. It
-connects same-label transitions in the two systems, treating the label
-as a cooperative communication event that may now be hidden from the
-outside world, as an $\epsilon$ label.
-
-\begin{defn}\label{plus}
-Where $A$ and $B$ are two LTSes sharing labels set $\mathcal L$,
-and with state sets $\mathcal S_A$ and
-$\mathcal S_B$ respectively, the \textbf{communicating composition} $A + B$ is
-a new LTS with states $\mathcal S_A \times \mathcal S_B$ and an empty label set,
-defined as follows:
-\vspace{-.2in}\begin{center}
-$$\inference[$A$]
-{\io{$A$}{a}{a'}{} & a' \neq H}
-{\io{$A + B$}{a, b}{a', b}{}}
-\quad \inference[$B$]
-{\io{$B$}{b}{b'}{} & b' \neq H}
-{\io{$A + B$}{a, b}{a, b'}{}}$$
-$$\inference[H$_A$]
-{\io{$A$}{a}{H}{}}
-{\io{$A + B$}{a, b}{H}{}}
-\quad \inference[H$_B$]
-{\io{$B$}{b}{H}{}}
-{\io{$A + B$}{a, b}{H}{}}$$
-$$\inference[Join]
-{\io{$A$}{a}{a'}{\ell} & \io{$B$}{b}{b'}{\ell} & a', b' \neq H}
-{\io{$A + B$}{a, b}{a', b'}{}}$$
-\end{center}
-\end{defn}
-
-\subsection{Refinement Between LTSes}
-
-We need a notion of when one LTS ``implements'' another.  Intuitively,
-transition labels and halting are all that the outside world
-can observe. Two systems that produce identical labels and termination behavior
-under all circumstances can be considered as safe substitutes for
-one another. We will only need an asymmetrical notion of compatibility:
-
-\begin{defn}\label{refines}
-For some label domain $\mathcal L$, let $f : \mathcal L \to \mathcal L^\epsilon$ 
-be a function that is able to replace labels with alternative
-labels, or erase them altogether. Let LTSes $A$ and $B$ have the same label
-set $\mathcal L$. We say that
-\textbf{$A$ trace-refines $B$ w.r.t. $f$}, or $A \sqsubseteq_f B$, if:
-\begin{multline*}
-\forall s_A, \eta. \; \io{$A^{*}$}{A_0}{s_A}{\eta} \Rightarrow \exists s_B. \;
-\io{$B^{*}$}{B_0}{s_B}{f(\eta)} \; \wedge \\
-(s_A = H \Leftrightarrow s_B = H)
-\end{multline*}
-Each label in the trace is replaced by the mapping of $f$ on it, and
-labels mapped to $\epsilon$ by $f$ are dropped. $f$ is overloaded
-to denote that multi-label version when applied to $\eta$.
-\end{defn}
-
-As a shorthand, we write $A \sqsubseteq B$ for $A \sqsubseteq_\mathsf{id} B$,
-for $\mathsf{id}$ an identity function, forcing traces in the two systems to
-match exactly. Under this notion of identical traces, we say that $A$
-is sound w.r.t. $B$.
-
-\subsection{A Few Useful Lemmas}
-
-The reflexivity and transitivity properties of refinement are crucial to help us
-structure modular proofs.
-
-\begin{theorem}\label{transitive}
-$\sqsubseteq$ is reflexive and transitive.
-\end{theorem}
-
-Another crucial proof step is to lift a single-process result to the $n$-way
-replication of that process.
-
-\begin{theorem}\label{liftn}
-If $A \sqsubseteq_f B$, then $\smulti{n}{A} \sqsubseteq_{f^n} \smulti{n}{B}$, where
-$f^n$ is $f$ lifted appropriately to deal with indices ($f^n(i, \ell) =
-(i, \ell')$ when $f(\ell) = \ell'$, and $f^n(i, \ell) = \epsilon$ when
-$f(\ell) = \epsilon$).
-\end{theorem}
-
-We will also need a result applying to communicating compositions.
-
-\begin{theorem}\label{liftplus}
-If $A \sqsubseteq_f A'$ and $B \sqsubseteq_f B'$, then $A + B \sqsubseteq A' +
-B'$. In other words, if systems $A$ and $B$ are individually simulated by $A'$ and
-$B'$ on identical alteration of the traces, then the composed
-system $A+B$ will be sound with respect to $A'+B'$.
-\end{theorem}
-
-\begin{figure}
-\small
-\centering
-\begin{boxedminipage}[c]{.48\textwidth}
-%\centering
-\inference
-[Halt]
-{\theta(i) = (s, \pc) & \dec(s,\pc) = H}
-{\io{SC}{\theta,m}{H}{}}
-
-\inference
-[NonMem]
-{\theta(i) = (s, \pc) & \dec(s,\pc) = (\nm, x) \\ \exec(s, \pc, (\nm, x)) = (\pc', s')}
-{\io{SC}{\theta,m}{\theta[i \coloneqq (s', \pc')], m}{}}
-
-\inference
-[Load]
-{\theta(i) = (s,\pc) & \dec(s,\pc) = (\ld, x, a) \\ \exec(s, \pc, (\ld, x, m(a))) = (\pc', s')}
-{\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m}{}}
-
-\inference
-[Store]
-{\theta(i) = (s,\pc) & \dec(s,\pc) = (\st, a, v) \\ \exec(s, \pc, (\st)) = (\pc', s')}
-{\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m[a\coloneqq v]}{}}
-\end{boxedminipage}
-\caption{LTS for sequential consistency with $n$ simple processors}
-\label{Ref}
-\end{figure}
-
-\section{Decomposing a Multi-Processor System}
-\label{sec:store-atomicity}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.43]{zoom}
-\caption{Components of a multi-processors system}
-\label{zoom}
-\end{figure}
-
-Most any conventional multi-processor system can be divided into 3 components as shown in
-Figure \ref{zoom}. 
-We show the top-level system design at top right, and we zoom in on boxes standing for the
-memory system and a single processor ($P_i$).
-The processor component $P_i$ can be implemented in a
-variety of ways, from literal execution of instructions in program order, to complex speculation
-with instructions run simultaneously to exploit parallelism. The memory component is normally implemented
-using a hierarchy of caches, in order to increase the performance of the
-overall system, since the latency of accessing a memory directly is very slow
-compared to accessing a much smaller cache. The component between each
-processor and the global memory sub-system contains local buffers, $B_i$, specific to
-the processor $P_i$ they are connected to.
-
-\begin{figure}
-\small
-\centering
-\begin{boxedminipage}[c]{.48\textwidth}
-\inference
-[Load]
-{}
-{\io{$M_m$}{m}{m}{(\ld\req,a),(\ld\resp,m(a))}}
-
-\inference
-[Store]
-{}
-{\io{$M_m$}{m}{m[a\coloneqq v]}{(\st, a, v)}}
-
-\end{boxedminipage}
-\caption{LTS for a simple memory}
-\label{$M_m$}
-\end{figure}
-
-Popular instruction set architectures do not guarantee sequential
-consistency.  That is, their behavior makes it unsound to imagine that
-the different processors of a shared-memory system are running
-according to simple interleaving, where each instruction executes
-atomically.  However, we want to emphasize that, in every weak-memory
-system we are aware of, \emph{the main memory still exposes atomic
-  loads and stores}!  The weaker semantics arises only because of (1)
-reordering of memory instructions by the core $P_i$ and/or (2) the
-properties of the local buffers $B_i$ connected to each processor
-$P_i$.  For instance, if $B_i$ were a store buffer, into which the
-committed stores from an in-order processor $P_i$ are enqueued, and if
-any ``subsequent'' load from the same processor $P_i$ reads the latest
-value from the store buffer, then one can establish that such a
-multi-processor system implements the TSO~\cite{x86tsocacm10} memory
-model.
-
-So, we focus on this opportunity to simplify proof decomposition.  We
-will prove that our main memory component satisfies an intuitive
-\emph{store atomicity} property.  Despite the potential red herring of
-the word ``atomicity,'' this specification is appropriate even for
-implementation of weaker memory models than SC.  Store atomicity can
-be understood via the operational semantics of Figure \ref{$M_m$},
-describing an LTS that receives load and store requests ($\ld$ and
-$\st$) from processors and sends back load responses ($\ld\resp$).
-
-The memory component is composed of a \emph{hierarchy of caches}, where each
-cache node (labeled like ``L1,'' ``L2,'' etc.) communicates only with its
-neighbors in the graph.  Each processor has a dedicated L1 cache, from which we
-do our best to satisfy all memory requests, to avoid the latency of a
-round-trip with main memory.  With an L1 cache miss, we may still find a hit in
-a parent cache below the level of main memory, realizing a smaller but still
-significant speedup.  Therefore it is the responsibility of the hierarchy of
-caches, which forms the memory sub-component, to implement the store atomicity
-property. In fact, as we will prove in Section \ref{sec:cc}, the purpose of the
-cache-coherence protocol is to establish this invariant for the memory
-sub-component.  Concretely, we have verified a \emph{directory-based} protocol
-for coordinating an arbitrary tree of caches, where each node stores a
-conservative approximation of its children's states.
-
-As an instance of the above decomposition, we will prove that a multi-processor
-system with no local buffering in between the processor and the memory
-components indeed implements sequential consistency. We implement a highly
-speculative processor that executes instructions and issues loads out of
-order, but commits instructions (once some ``verification'' is done) in order.
-
-The processor itself can be decomposed into several components. In the zoomed-in
-version of Figure \ref{zoom}, we show a highly speculative, out-of-order
-issue processor. We have the normal architectural state, such as values of
-registers.  Our proofs are generic over \emph{a family of instruction set
-architectures}, with parameters for opcode sets and functions for executing
-opcodes and decoding them from memory.  Other key components are a \emph{branch
-predictor}, which guesses at the control-flow path that a processor will
-follow, to facilitate speculation; and a \emph{reorder buffer (ROB)}, which
-decides which instructions along that path to try executing ahead of schedule.
-Our proofs apply to an arbitrary branch predictor, and they work for any
-reorder buffer satisfying a simple semantic condition.
-
-Figure~\ref{proofs} gives the overall proof structure that we employ to verify
-the system of Figure~\ref{zoom}. As we will explain later, $P_\text{so}$
-represents the speculative out-of-order processor, $M_c$ the cache memory,
-$M_m$ the simple memory, $P_\text{ref}$ a simple decoupled processor, and
-SC the simple composition of P$^n_\text{ref}+ M_m$ executing each
-instruction atomically, thus implementing sequential consistency.
-
-Our current framework establishes theorems of the form ``if system $A$ has a run
-with some particular observable behavior, then system $B$ also has a run with
-the same behavior.''  In this sense, we say that $A$ correctly implements $B$.
-Other important properties, such as \emph{deadlock freedom} for $A$ (which
-might get stuck without producing any useful behavior), we leave for future
-work.
-
-\begin{figure}
-\includegraphics[scale=.45]{proofs}
-\caption{Overall proof structure}
-\label{proofs}
-\end{figure}
-
-\section{Specifying Sequential Consistency}\label{sec:sc}
-
-Our final theorem in this paper establishes that a particular complex hardware
-system implements sequential consistency (SC) properly.  We state the theorem
-in terms of the trace refinement relation $\sqsubseteq$.
-Therefore, we need to define an LTS embodying SC.  The simpler this
-system, the better.  We do not need to worry about its performance properties,
-since we will prove that an optimized system remains faithful to it.
-
-\begin{figure}
-\small
-\centering
-\begin{boxedminipage}[c]{.48\textwidth}
-\inference
-[Halt]
-{\dec(s,\pc) = H}
-{\io{P$_\text{ref}$}{s,\pc}{H}{}}
-
-\inference
-[NM]
-{\dec(s,\pc) = (\nm, x) & \exec(s, \pc, (\nm, x)) = (\pc', s')}
-{\ioe{P$_\text{ref}$}{s,\pc}{s', \pc'}}
-
-\inference
-[Ld]
-{\dec(s,\pc) = (\ld, x, a) & \exec(s, \pc, (\ld, x, v)) = (\pc', s')}
-{\io{P$_\text{ref}$}{s, \pc}
-{s', \pc'}{(\ld\req, a), (\ld\resp, v)}}
-
-\inference
-[St]
-{\dec(s,\pc) = (\st, a, v) & \exec(s, \pc, (\st)) = (\pc', s')}
-{\io{P$_\text{ref}$}{s, \pc}
-{s', \pc'}{(\st, a, v)}}
-\end{boxedminipage}
-
-\caption{LTS for a single reference processor (P$_{\text{ref}}$)}
-\label{Pref$}
-\end{figure}
-
-Figure \ref{Pref$} shows the LTS for a single processor that will be used as a
-reference processor.  The definition is parametrized over details of an
-\emph{instruction set architecture} (ISA).  In particular, the ISA gives us
-some domains of architectural states $s$ (e.g., register files) and of program
-counters $\pc$.  A function $\dec(s, \pc)$ figures out which instruction $\pc$
-references in the current state, returning the instruction's ``decoded'' form.
-A companion function $\exec(s, \pc, \mathit{dec})$ actually executes the
-instruction, returning the next program counter $\pc'$ and the new state $s'$.
-
-The legal instruction forms, which are outputs of $\dec$, are $(\nm, x)$, for
-an operation not accessing memory; $(\ld, x, a)$, for a memory load from
-address $a$; $(\st, a, v)$, for a memory store of value $v$ to address $a$; and
-$H$, for a ``halt'' instruction that moves the LTS to state $H$. The parameter
-$x$ above represents the rest of the instruction, including the
-opcode, registers, constants, \etc{}
-
-The legal inputs to $\exec$ encode both a decoded instruction and any relevant
-responses from the memory system.  These inputs are $(\nm, x)$ and $\st$, which
-need no extra input from the memory; and $(\ld, x, v)$, where $v$ gives the
-contents of the requested memory cell.
-
-We can define sequential consistency using the simple memory from Figure
-\ref{$M_m$} as follows, for $n$ processors:
-\begin{defn}
-$SC \triangleq P^n_\text{ref} + M_m$
-\label{sc}
-\end{defn}
-
-We define the initial state of SC as $(\theta_0, m_0)$, where $m_0$ is
-some initial memory fixed throughout our development, mapping every
-address to value $v_0$; and $\theta_0$ maps every processor ID to
-$(s_0, \pc_0)$, using architecture-specific default values $s_0$ and
-$\pc_0$.
-
-This LTS encodes Lamport's notion of SC, where processors take turns executing
-nondeterministically in a simple interleaving.  We need know very few details
-of the ISA to define the SC model.  Our final, optimized system is parametrized
-over an ISA in the same way.  In the course of the rest of this paper, we will
-define an optimized system $O$ and prove $O \sqsubseteq \text{SC}$.
-
-Note that, in this setting, an operational specification like the LTS
-for SC is precisely the right characterization of \textbf{full functional
-  correctness} for a hardware design, much as a
-precondition-postcondition pair does that job in a partial-correctness
-Hoare logic.  Our SC LTS fully constrains observable behavior of a
-system to remain consistent with simple interleaving.  Similar
-operational models are possible as top-level specifications for
-systems following weaker memory models.
-
-\section{Speculative Out-Of-Order Processor}\label{sec:ooo}
-
-The previous section divided a multiprocessor system into two rather
-simple components, processors and memories.  We are now ready to begin
-developing more complex, optimized versions of each piece, starting
-with processors.
-
-We implement a \emph{speculative} processor, which may create many
-simultaneous outstanding requests to the memory, as an optimization to
-increase parallelism.  Our processor proof is in some sense generic
-over correct speculation strategies.  We parametrize over two key
-components of a processor design: a \emph{branch predictor}, which
-makes guesses about processor-local control flow in advance of
-%receiving enough memory responses to resolve
-resolving conditional jumps; and a
-\emph{reorder buffer}, which decides what speculative instructions (like memory loads)
-are worth issuing at which moments (in effect \emph{reordering} later
-instructions to happen before earlier instructions have finished).
-
-The branch predictor is the simpler of the two components to model.
-We indicate branch-predictor state with metavariable $\ppc$.
-The operations on such state are $\curPpc(\ppc)$, to extract the
-current program-counter prediction; $\nextPpc(\ppc)$, to ask the
-predictor to step forward by one instruction; and $\setBp(\ppc, \pc)$,
-to reset prediction to begin at a known-accurate position $\pc$.  It
-turns out that we need to impose no explicit correctness criterion on
-branch predictors.  The processor uses predictions only as hints, and
-it always resets the predictor using $\setBp$ after detecting an
-inaccurate hint.
-
-The interface and formal contract of a reorder buffer are more
-involved.  We write $\rob$ as a metavariable for reorder buffer
-state.  The associated operations are:
-\begin{itemize}
-\item $\phi$, the state of an empty buffer.
-\item $\add(\pc, \rob)$, which appends the program instruction at location $\pc$ to the list of instructions that the buffer is allowed to consider executing.
-\item $\get\spc\ld(\rob)$, which returns an optional speculative load to issue
-\item $\compute(\rob)$, which models a
-step of computation inside the buffer. For instance, it invokes the $\dec$
-and $\exec$ functions (as defined for SC) internally to obtain the next program
-counter, state, \etc{} (only the speculative states are updated, not the actual states).
-\item $\upd\ld(\rob,t,v)$, which informs the buffer that the memory
-has returned result value $v$ for the speculative load with tag $t$.
-\item $\commit(\rob)$, which returns the next instruction in serial
-program order, if we have accumulated enough memory responses to execute it
-accurately, or returns
-$\epsilon$ otherwise.  When $\commit$ returns an instruction, it also
-returns the associated program counter plus the next program counter
-we would advance to afterward.
- Furthermore, the instruction is
-extended with any relevant response from memory (used only for load
-instructions, obtained through $\upd\ld$) and with the new architectural state (\eg{} register
-file) after execution.
-\item $\retire(\rob)$, which informs the buffer that its $\commit$
-instruction was executed successfully, so it is time to move on to the
-next instruction.
-\end{itemize}
-
-We postpone characterization of correct reorder buffers until after we
-present the LTS for speculating processors.
-
-\begin{figure*}[t]
-\small
-\centering
-\begin{boxedminipage}[c]{.95\textwidth}
-\inference[Fetch]
-{}
-{\ioe{P$_\text{so}$}{s,\pc,\rob,\ppc}{s,\pc,\add(\curPpc(\ppc),\rob),\nextPpc(\ppc)}}
-
-\inference[Compute]
-{\get\spc\ld(\rob) = \epsilon}
-{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\compute(\rob),\ppc}}
-
-\inference[SpecLd]
-{\get\spc\ld(\rob) = (\spc\ld,t,a)}
-{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\upd\ld(\compute(\rob),t,v),\ppc}
-{(\ld\req, a), (\ld\resp, v)}}
-
-\inference[Abort]
-{\commit(\rob) = (\pc',\_,\_) & \pc' \neq \pc}
-{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s,\pc,\phi,\setBp(\ppc,\pc)}}
-
-\inference[Halt]
-{\commit(\rob) = H}
-{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{H}}
-
-\inference[Nm]
-{\commit(\rob) = (\pc,\pc',(\nm,s'))}
-{\ioe{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s',\pc',\retire(\rob),\ppc}}
-
-\inference[St]
-{\commit(\rob) = (\pc,\pc',(\st,a,v,s'))}
-{\io{\text{P$_\text{so}$}}{s',\pc',\retire(\rob),\ppc}{s,\pc,\rob,\ppc}
-{\st, a, v}}
-
-\inference[LdGood]
-{\commit(\rob) = (\pc,\pc',(\ld,x,a,v,s'))}
-{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s',\pc',\retire(\rob),\ppc}{(\ld\req, a), (\ld\resp, v)}}
-
-\inference[LdBad]
-{\commit(\rob) = (\pc,\pc',(\ld,x,a,v',s')) & v' \neq v & \exec(s,\pc,(\ld,x,v)) = (\pc'', s'')}
-{\io{\text{P$_\text{so}$}}{s,\pc,\rob,\ppc}{s'',\pc'',\phi,\setBp(\ppc,\pc)}{(\ld\req, v), (\ld\resp,v)}}
-\end{boxedminipage}
-
-\caption{Speculating, out-of-order issue processor}
-\label{Ooo}
-\end{figure*}
-
-Figure~\ref{Ooo} defines that LTS,
-P$_{\texttt{so}}$. This processor may issue arbitrary speculative
-loads, but it only ever \emph{commits} the next instruction in serial
-program order. 
-We will see the processor issue two kinds of loads, a speculative
-load ($\spc\ld$) and a commit or real load ($\ld$).  To maintain SC, every speculative
-load must have a matching verification load later on, and we maintain
-the illusion that the program only depends on the results of
-verification loads, which, along with stores, \emph{must be issued in
-serial program order}.
-
-When committing a
-previously issued speculative load instruction, the associated speculative
-memory load response is verified against the new commit load response. If the
-resulting values do not match, the processor terminates all past uncommitted
-speculation, by emptying the reorder buffer and resetting the
-next predicted program counter in the branch predictor to the correct next value. This step mimics
-the mechanism implemented in most modern-day processors, operating on the
-conservative assumption that the processor may have performed arbitrary
-incorrect speculation after receiving the inaccurate response from
-memory.  (Note that here ``inaccurate'' does not mean ``wrong at
-the time the response was received,'' but rather ``wrong at commit
-time'' or ``correct before, but not anymore.'')
-
-A full processor state is $(s, \pc, \rob, \ppc)$, comprising
-architectural state, the program counter, and the reorder buffer and branch
-predictor states. Its initial state is given by $(s_0, \pc_0, \phi, \ppc_0)$.
-The interface of this processor with memory is identical
-to that of the reference processor, but now we can exercise a memory
-system's transitions even for speculative loads.
-
-\begin{figure*}[t]
-\small
-\begin{inv}
-If $P_\text{so}$ reaches a state $(s, \pc, \rob, \ppc)$, \ie{}
-
-$\io{P$_\text{so}^{*}$}{s_0,\pc_0,\phi,\ppc_0}{s,\pc,\rob,\ppc}{\eta} \Rightarrow$
-\begin{displaymath}
-\left\{
-\begin{array}{ll}
-\commit(\rob) = (\pc, \pc', (\nm, s')) &\Rightarrow
-\exists x. \; \dec(s,\pc) = (\nm, x) \wedge \exec(s,\pc, (\nm,x)) =
-(\pc', s')\\
-\commit(\rob) = (\pc, \pc', (\ld, x, a, v, s')) & \Rightarrow
-\dec(s,\pc) = (\ld, x,a) \wedge \exec(s,\pc, (\ld,x,v)) = (\pc', s')\\
-\commit(\rob) = (\pc, \pc', (\st, a, v, s')) &\Rightarrow
-\dec(s,\pc) = (\st, a,v) \wedge \exec(s,\pc, (\st)) =
-(\pc', s')\\
-\commit(\rob) = H & \Rightarrow
-\dec(s, \pc) = H\\
-\end{array}
-\right.\end{displaymath}
-\label{rob}
-\end{inv}
-\vspace{-.5cm}
-\caption{Correctness of reorder buffer}
-\label{robfig}
-\end{figure*}
-
-Finally, we impose a general correctness condition on reorder
-buffers.  They must maintain Invariant~\ref{rob} in Figure~\ref{robfig}.
-Intuitively, whenever the buffer claims in a $\commit$
-output that a particular instruction is next to execute, causing
-certain state changes, that instruction must really be next in line according
-to how the program runs in the SC system, and running it must really cause
-those state changes.
-
-When this condition holds, we may conclude the correctness theorem for
-out-of-order processors.  We use a trace-transformation function
-$\noSpec$ that drops all speculative-load requests and responses.
-See Definition~\ref{refines} for a review of how we use such
-functions in framing trace refinement.  Intuitively, we prove that any
-behavior by the speculating processor can be matched by the simple
-processor, with speculative messages erased.
-
-\begin{theorem}
-\label{ocorrect}
-$\text{P$_\text{so}$} \sqsubseteq_\noSpec$ P$_\text{ref}$
-\end{theorem}
-\begin{proof}
-By induction on P$_\text{so}$ traces, using a proper abstraction
-function (which drops the speculative messages and the $\rob$ and $\ppc$ states) to relate the two systems.  Invariant~\ref{rob} is crucial to
-relate the reorder buffer's behavior with the simple in-order
-execution of P$_\text{ref}$.
-\end{proof}
-
-\begin{corollary}
-\label{ges}
-$\text{P$^n_\text{so}$} \sqsubseteq_{\noSpec^n}$ P$_\text{ref}^n$
-\end{corollary}
-\begin{proof}
-Direct consequence of Theorems~\ref{ocorrect} and~\ref{liftn} (where the
-latter is one of our preliminary results about $n$-repetition).
-\end{proof}
-
 \section{Cache-Based Memory System}\label{sec:cc}
@@
 We describe a state of the system using 8 fields: $\dt$, $\ch$, $\s$, $\dst$,
-$\wt$, $\dwt$ which are summarized in Figure~\ref{fields}.
+$\wt$, $\dwt$, $\inp$, $\outp$. The $\inp$ and $\outp$ sets are the interfaces
+to the processors and are exactly the same as in the simple memory LTS
+(Figure~\ref{$M_m$}). The rest are summarized in Figure~\ref{fields}.
+
+%We describe a state of the system using 6 fields: $\dt$, $\ch$, $\s$, $\dst$,
+%$\wt$, $\dwt$ which are summarized in Figure~\ref{fields}.
 We use $\parent(c,p)$ to denote that $p$ is the parent of $c$ and $\leaf(c)$ to 
@@
 
-%\medskip
-
 Here is an intuition on how the transitions work in the common case.  A cache
@@
 
-%Dropping the request
-%removes the entry from the buffer, allowing the child to process further messages
-%from the parent -- otherwise there would have been a deadlock.
-
-\begin{figure*}
-\small
-\centering
-
-\begin{boxedminipage}[c]{.95\textwidth}
-
-\textbf{Processor/Memory Interface}
-
-\inference[Load]
-{\s(c,a) \geq S}
-{\io{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt}
-{\dt, \ch, \s, \dst, \wt, \dwt}{c,((\ld\req, a),(\ld\resp,v))}}
-\vspace{.1in}
-\inference[Store]
-{s(c,a) \geq M}
-{\io{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt}
-{\dt[(c, a) \coloneqq v], \ch, \s, \dst, \wt, \dwt}{c,(\st,a,v)}}
-
-\textbf{Child Upgrade}
-
-\inference[ChildSendReq]
-{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt}
-{\dt, \ch[ (c, p, \req) \coloneqq (a, \s(c, a), x) \uplus \ch(c,p,\req)], \s, \dst,\\
-\wt[(c,a)\coloneqq x], \dwt}}
-\vspace{.1in}
-\inference[ParentRecvReq]
-{\parent(c,p) & \ch(c, p, \req) = \{(a, y, x)\} \uplus \rs & \s(p, a) \geq x
-\\ \compat(p, c, x, a) & \dwt(p,c,a) = \epsilon & \dst(p,c,a) \le y}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt}
-{\dt, \ch[(c,p,\req) \coloneqq \rs][(p, c) \coloneqq (\resp, (a, \dst(p,c,a), x,\\ \;\;\;\;
-\ite{\dst(p,c,a) = I}{\dt(p,a)}{\_})) \coloncolon \ch(p,c)], \\ \s, \dst[(p,c, a)\coloneqq x],
-\wt, \dwt}}
-\vspace{.1in}
-\inference[ChildRecvRsp]
-{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\resp, (a, y, x, v))}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt}
-{\dt[(c,a) \coloneqq \ite{y = I}{v}{\dt(c,a)}], \ch[(p,c) \coloneqq \rs], \s[(c,a) \coloneqq x], \dst,\\
-\wt[(c,a)\coloneqq \ite{\wt(c,a) \le x}{\epsilon}{\wt(c,a)}], \dwt}}
-
-\textbf{Parent Downgrade}
-
-\inference[ParentSendReq]
-{\parent(c,p) & \dst(p,c,a) > x & \dwt(p,c,a) = \epsilon}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt}
-{\dt, \ch[(p, c)\coloneqq (\req, (a, \dst(p, c, a), x)) \coloncolon \ch(p,c)], \s, \dst,\\
-\wt, \dwt[(p, c,a)\coloneqq x]}}
-\vspace{.1in}
-\inference[ChildRecvReq]
-{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\req, (a, y, x)) & (\forall i. \; \parent(i, c) \Rightarrow \dst(c, i, a) \le x)
-& \s(c,a) > x}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt}
-{\dt, \ch[(p,c) \coloneqq \rs][(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
-\ite{\dst(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
-\wt, \dwt}}
-\vspace{.1in}
-\inference[ParentRecvRsp]
-{\parent(c,p) & \ch(c,p,\resp) = \{(a, y, x, v)\} \uplus \rs & \dst(p,c,a) = y}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt}
-{\dt[(p,a) \coloneqq \ite{y = M}{v}{\dt(p,a)}], \ch[(c,p,\resp) \coloneqq \rs], \s, \dst[(p,c,a) \coloneqq x],\\
-\wt, \dwt[(p,c,a)\coloneqq \ite{\dwt(p,c,a) \ge x}{\epsilon}{\dwt(p,c,a)}]}}
-
-\textbf{Voluntary downgrade for replacement}
-
-\inference[VolResp]
-{\parent(c,p) & (\forall i. \; \parent(i, c) \Rightarrow \dst(c, i, a) \le x)
-& \s(c,a) > x}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt}
-{\dt, \ch[(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
-\ite{\s(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
-\wt, \dwt}}
-
-\textbf{Dropping request because of voluntary downgrade}
-
-\inference[DropReq]
-{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\req, (a, y, x)) & 
- \s(c,a) \le x}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt}
-{\dt, \ch[(p,c) \coloneqq \rs], \s, \dst,
-\wt, \dwt}}
-\end{boxedminipage}
-
-\caption{LTS for cache-coherent shared-memory system}
-\label{cache}
-\end{figure*}
+\input{TransitionsUsefulCache1}
 
@@
 
-We first reduce the proof of store atomicity to several lemmas. We then
-present a key insight that we discovered for proving these lemmas in the form
-of three fundamental invariants. We believe that other invalidation-based
+We first reduce the proof of store atomicity to several lemmas.
+We believe that other invalidation-based
 protocols with inclusive caches (\ie{} those cache hierarchies where a parent
 cache contains all the locations that any of its children contains) like MESI,
-MOSI, and MOESI also satisfy our invariants and lemmas, making their verification
+MOSI, and MOESI also satisfy our lemmas, making their verification
 easier by reusing most of our MSI proof; we omit the discussion of extension of
@@
 
-%\adamc{...and maybe we don't, for space reasons. \texttt{:-)}}
-
-%We use the terms write to mean handling a store request, \ie{} executing the
-%StoreReq transition and read to mean handling a load request, \ie{} executing
-%the LoadReq transition.
-
-The lemmas and invariants presented below hold after any number of transitions
+The lemmas
+presented below hold after any number of transitions
 starting from the initial state.
@@
 informal for brevity. All the proofs can be found in our supplementary material.
-%\adamc{or just remind of the source code URL given earlier}
 
-%\subsection{Invariants}
-%\label{firstLevel}
 Intuitively our protocol is correct because it is guaranteed that any cache that attempts to
@@
 
-\subsection{Key Insight for Proving Lemmas}
-
-We present what we believe to be the key insight for proving any
-cache-coherence protocol.  In fact, the request/response message semantics for
-the MSI protocol was designed to satisfy the following three invariants.
-
-\begin{inv}
-\textit{stateChange}:
-\begin{enumerate}
-\item A cache downgrades its state for an address iff it sends a response for that address.
-\item A parent upgrades the directory entry for an address for one of its
-children iff it sends a response for that child and address.
-\item The state or directory entry changes only on sending or
-receiving a response $(\resp, (a, y, x, d))$, to $x$.
-\end{enumerate}
-\label{stateChange}
-\end{inv}
-
-\vspace{-.1in}
-
-Invariant~\ref{stateChange} ensures that a response message acts as a sync
-between a parent's directory entry for a child and the state of the child, as
-both the sender and the receiver change their respective directory entry or
-state to the same $x$ in a response $(\resp, (a, y, x, d))$
-
-\begin{inv}
-\textit{noCross}: A response from a parent to its child for an address cannot
-be in transit while another response from the same child to the parent for the
-same address is in transit.
-\label{noCross}
-\end{inv}
-
-\begin{inv}
-\textit{respFIFO}: Responses are received in the order in which they are sent,
-for a source-destination pair, for the same address.
-\label{respFIFO}
-\end{inv}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.45]{resps}
-\caption{An example showing a sequence of responses between a child and its parent}
-\label{resps}
-\end{figure}
-
-Figure~\ref{resps} shows an example of the effect of
-Invariants~\ref{stateChange}, \ref{noCross} and \ref{respFIFO}. The
-directory entry of the parent or the state of the child changes on sending and
-receiving a response, and the new state is shown at the points of sending or
-receiving responses.  Both the cache and its parent follow the same sequence of
-state and directory entry changes, but the receiver always lags slightly behind
-the sender.  The receiver cannot send any response before receiving because of
-Invariant~\ref{noCross}, thereby eliminating any scenarios where the lag
-complicates reasoning.
-
-Using these three invariants (Invariants~\ref{stateChange}, \ref{noCross} and
-\ref{respFIFO}), it is easy to see that Lemma~\ref{conservative} holds.
-Intuitively, going back to Figure~\ref{resps}, we can see that at all times,
-either there is no response in flight, only responses from the parent are in
-flight, or only responses from the child are in flight. A response from child
-downgrades the state, and a response from parent upgrades the directory. Since
-the other side simply lags behind, Lemma~\ref{conservative} holds.
-Since Invariants~\ref{stateChange}, \ref{noCross} and \ref{respFIFO} deal with
-just one cache's state for a single address and the corresponding directory
-entry, these properties can also potentially be verified using model-checking.
-Thus, if some other protocol is verified to obey Invariants~\ref{stateChange},
-\ref{noCross} and \ref{respFIFO} using model-checking, then
-Lemma~\ref{conservative} is already automatically verified for that
-protocol.
-
 Analyzing a parent-child pair will suffice in proving
@@
 Lemma~\ref{nonAncestorCompatible}.
-
-%Figure \ref{something} gives the structure of the proof for Theorem \ref{ccorrect}
-%\begin{figure*}
-%\begin{prooftree}
-%\AxiomC{stateChange}
-%\AxiomC{noCross}
-%\AxiomC{respFifo}
-%\TrinaryInfC{conservative}
-%\AxiomC{localCompatibility1}
-%\AxiomC{localCompatibility2}
-%\TrinaryInfC{nonAncestorCompatible}
-%\AxiomC{noTransitWrite}
-%\BinaryInfC{latestValue}
-%\AxiomC{$\ldots$ (trivial)}
-%\BinaryInfC{StoreAtomicity}
-%\UnaryInfC{$M_c \sqsubseteq M_m$}
-%\end{prooftree}
-%\caption{Proving $M_c \sqsubseteq M_m$}
-%\label{something}
-%\end{figure*}
 
@@
 We twice apply $\sqsubseteq$ transitivity (Theorem \ref{transitive}) to
-connect Theorems \ref{ocorrect2} and \ref{ocorrect1}, and use Definition \ref{sc} of $SC$
+%connect Theorems \ref{ocorrect2} and \ref{ocorrect1}, and use Definition \ref{sc} of $SC$
+connect Theorems \ref{ocorrect2}, \ref{ocorrect1}, and \ref{scthm}.
 \end{proof}

### 2014-10-31 Introduction.tex
stuff

@@
 \item A general \emph{modular verification methodology} applicable to hardware
-  algorithms, based on \emph{labeled transition systems}.  We are able to verify
-  the several components of Figure~\ref{zoom} \adam{OK to have a
-    forward reference to this figure?} against general formal interfaces,
+  algorithms, based on \emph{labeled transition systems}.
+We are able to verify
+  the several components of a modern processor 
+\adam{(Figure~\ref{zoom} OK to have a forward reference to this figure?}
+against general formal interfaces,
   enabling mixing and matching of different realizations of each interface,

### 2014-10-31 TransitionsUsefulFirst.tex
stuff

@@
 %
+It is worth noting that the LTS notation described above, along with the
+composition operators are very similar to the semantics of hardware designs
+written in Bluespec. As an example consider an inference rule in LTS and the
+corresponding code in Bluespec in Figure \ref{both}. This code was actually
+written for the PowerPC system \cite{Khan:PowerPc} (we will be using this rule
+later in the paper when we describe the coherent caches).
+
+\begin{figure*}
+\centering
+\begin{boxedminipage}{.53\textwidth}
+\noindent \textbf{rule} $(c.\parent == p \; \mathit{\&\&} \; c.\s[a] < x \; \mathit{\&\&} \; c.\wt[a]==None);$
+
+\hspace{1cm} $\ch[c][p][\req].\enq(a,c.\s[a],x); c.\wt[a] \Leftarrow x;$
+
+\textbf{endrule}
+\end{boxedminipage}
+\begin{boxedminipage}{\textwidth}
+\inference[]
+{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
+{\ioxe{$M_c$}
+{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
+{\dt, \ch[ (c, p, \req) \coloneqq
+(a, \s(c, a), x) \uplus \ch(c,p,\req)], \s, \dst,\\
+\wt[(c,a)\coloneqq x], \dwt, \inp, \outp}}
+\end{boxedminipage}
+\caption{Rule in Bluespec and its corresponding LTS transliteration}
+\label{both}
+\end{figure*}
+
 \section{Decomposing a Multi-Processor System}

### 2014-10-31 Introduction.tex
stuff

--- 2014-10-19 Introduction.tex
@@
 produce full-system theorems.
+
+%\xxx{\textbf{Murali's version in red} This works introduces an interface enabling true modular implementation and
+%verification of shared memory systems. We want our modular system to permit
+%\emph{black-box refinement}, that is, nothing beyond the interface contract can
+%be assumed about other modules.  That is, we should be able to describe a
+%modular decomposition where we may replace either the processor nodes, the
+%memory with locally verified alternatives without changing the correctness of
+%our whole system.  To this end, we identify a key property called \emph{store
+%atomicity} shared by the memory component of every memory model. This property
+%effectively restricts the behavior of the memory sub-system to that of an
+%\emph{atomic instantaneous memory}. We prove that this property is in fact what
+%is guaranteed using the cache coherence protocols in a multi-processor system
+%containing a hierarchy of caches, enabling replacement of the cache hierarchy
+%with an atomic instantaneous memory. Similarly, on the processor side, we
+%define what is meant to be a correct single processor. These invariants about
+%the processor and the memory sub-system remains constant across memory models.
+%We show a concrete instance of this framework by proving that a highly
+%speculative multi-processor system connected to a coherent cache hierarchy
+%implements the sequential consistency memory model.}
 

### 2014-11-05 Introduction.tex
lots of modifications

@@
 on these techniques (\eg{} partial order reduction~\cite{part}, symmetry
-reduction~\cite{sym1, sym2}, compositional reasoning~\cite{comp}, \etc{}) scale
+reduction~\cite{sym1, sym2}, compositional reasoning~\cite{comp, Mccomp, mcc}, \etc{}) scale
 the approach further, but they are still unable to handle realistic
@@
   the several components of a modern processor 
-\adam{(Figure~\ref{zoom} OK to have a forward reference to this figure?}
+%\adam{(Figure~\ref{zoom} OK to have a forward reference to this figure?}
 against general formal interfaces,

### 2014-11-05 TransitionsUsefulSecond.tex
lots of modifications

@@
 We can define sequential consistency using the simple memory from Figure
-\ref{$M_m$} as follows, for $n$ processors:
+\ref{M_m} as follows, for $n$ processors:
 \begin{defn}
@@
 
-Figure~\ref{Ooo} defines that LTS,
-P$_{\texttt{so}}$. This processor may issue arbitrary speculative
-loads, but it only ever \emph{commits} the next instruction in serial
-program order. 
-We will see the processor issue two kinds of loads, a speculative load (with
-the label suffix $\Spec$) and a commit or real load (with the label suffix $\Real$).  To
-maintain SC, every speculative
-load must have a matching verification load later on, and we maintain
-the illusion that the program only depends on the results of
-verification loads, which, along with stores, \emph{must be issued in
-serial program order}.
+Figure~\ref{Ooo} defines that LTS, P$_{\texttt{so}}$. This processor may issue
+arbitrary speculative loads (with label suffix $\Spec$), but it only ever
+\emph{commits} the next instruction in serial program order. These loads are
+issued non-deterministically from any load instruction after its address is
+computed.  This non-deterministic issue is going to be necessary when the
+memory LTS $M_m$ is replaced by caches.  To maintain SC, every speculative load
+must have a matching verification load later on (with label suffix $\Real$),
+and we maintain the illusion that the program only depends on the results of
+verification loads, which, along with stores, \emph{must be issued in serial
+program order}.
 

### 2014-11-05 TransitionsUsefulThird1.tex
lots of modifications

--- 2014-10-30 TransitionsUsefulThird.tex
@@
 \centering
-\begin{tabular}{|l|p{5.5cm}|}
+\begin{tabular}{|l|p{6cm}|}
 \hline
@@
 
-Fields $\dt, \s$ are maps over pairs of cache names and memory addresses.
-The $\dt$ map assigns each pair a memory-cell value, and the $\s$ map
-assigns each pair a coherence state.  A coherence state is $M, S$ or
-$I$, broadly representing permissions to modify, read, or do nothing
-with an address, respectively, the decreasing permissions denoted by $M > S > I$. This interpretation of permissions is
-accurate at the leaf caches connected to the processors (\ie{} L1 caches),
-but the interpretation is subtly different at the higher level caches.
+A coherence state is $M, S$ or $I$, broadly representing permissions to modify,
+read, or do nothing with an address, respectively, the decreasing permissions
+denoted by $M > S > I$. This interpretation of permissions is accurate at the
+leaf caches connected to the processors, but the
+interpretation is subtly different at the higher level caches.
 
-Mapping $\wt$ goes from cache name and address to either $\epsilon$ or a
-coherence state.  It denotes what permission the cache is waiting for,
-if any.  That is, a cache may have decided to \emph{upgrade} its
-coherence state for some address to a more permissive value, but the
-cache is waiting for acknowledgment from another cache before
+$\wt$ denotes what permission the cache is waiting for,
+if any.  That is, a cache has decided to \emph{upgrade} its
+coherence state for some address to a more permissive value, but
+it is waiting for acknowledgment from its parent before
 finalizing the upgrade.
 
-Since we have a directory-based protocol, the $\s$ and
-$\wt$ maps have mirror-image counterparts $\dst$ and $\dwt$.
-The former maps triples of parent name, child name, and address to
-coherence state. It represents the parent's notion of the
+$\dst$ represents the parent's notion of the
 coherence state of the child for that address. We later prove that this notion
@@
 a particular permission, then it is guaranteed in this system that the child
-will not have that permission.  The $\dwt$ field also represents a mapping from parent
-name, child name, and address to either $\epsilon$ or coherence state. It
+will not have that permission.  $\dwt$
 denotes whether the parent is waiting for any downgrade response from the
@@
 
-All communication channels in the system are represented by field $\ch$.
-The channels are asymmetric as
-follows. Channels going from parent to children (indexed by the names of the
-parent and child) maintain FIFO order of messages between the same parent-child
-pair. They carry both request messages (asking the child to downgrade) and
-response messages (acknowledging that the child can upgrade). Channels going
-from children to their parents are further split into request and response
-channels, thus indexed by tuples (child name, parent name, request or response).
-These channels need not maintain the FIFO order for messages between the same
-child-parent pair (notice the use of $\coloncolon$ to add and remove elements
-from the parent to child channel, as opposed to $\uplus$ for the 
-child to parent channel). The reason is that only one request or one response is
-sent for a particular address between a child-parent pair -- this invariant can
-be proved for the LTS in  Figure \ref{cache}. The order between a request and
-response message sent from the same child to its parent is lost, but our
-protocol ensures that nothing bad happens as a result.  Notice rules
-ParentRecvRsp and ParentRecvReq in Figure \ref{cache}, each of which looks at the
-previous coherence state of the incoming message and compares it with its own
-directory state to decide whether to process the message.
+All communication channels in the system are represented by field $\ch$.  The
+channels are asymmetric as follows. Channels going from parent to children
+(indexed by the names of the parent and child) maintain FIFO order of messages
+between the same parent-child pair. They carry both request messages (asking
+the child to downgrade) and response messages (acknowledging that the child can
+upgrade). Channels going from children to their parents are further split into
+request and response channels, thus indexed by tuples (child name, parent name,
+request or response).  These channels need not maintain the FIFO order for
+messages between the same child-parent pair (notice the use of $\coloncolon$ to
+add and remove elements from the parent to child channel, as opposed to
+$\uplus$ for the child to parent channel). The reason is that only one request
+or one response is sent for a particular address between a child-parent pair --
+this invariant can be proved for the LTS in  Figure \ref{cache}. The order
+between a request and response message sent from the same child to its parent
+is lost, but our protocol ensures that they are processed in order by the parent, by
+comparing the previous coherence state of the incoming messages with the
+parent's directory state (in rules ParentRecvRsp
+and ParentRecvReq).
 
@@
 original request can be responded to. The transitions for sending requests are
-allowed to happen spontaneously without any trigger. An L1 cache can process a
-load request or a store request only if it is in at least the \Sh{} state or
-\Mo{} state, respectively, for the requested address; otherwise it must
-spontaneously request to upgrade.
+allowed to happen spontaneously without any trigger. $P_\text{ref}$ issues speculative
+loads only for addresses which are in \Sh{} or \Mo{} state in L1; otherwise L1
+spontaneously requests an upgrade for that address.
 
@@
 child-parent pair is to avoid deadlocks. Throughout the protocol, we maintain
-the invariant
-that a downgrade request from the parent always gets priority, \ie{} if a child
-is waiting for an upgrade response, it must still try to satisfy a downgrade
-request from the parent. If both the parent and the child are waiting
-for their respective responses, there will be a deadlock. This tie cannot
-be broken arbitrarily: if the parent concedes and starts processing a child's
-request, then it may have to send downgrade requests to other children to make
-the states compatible, but the other children may have sent upgrade requests
-and remain waiting, leading to a deadlock. The only way to break the
+the invariant that a downgrade request from the parent always gets priority,
+\ie{} if a child is waiting for an upgrade response, it must still try to
+satisfy a downgrade request from the parent. If both the parent and the child
+are waiting for their respective responses, there will be a deadlock. This tie
+cannot be broken arbitrarily: if the parent concedes and starts processing a
+child's request, then it may have to send downgrade requests to other children
+to make the states compatible, but the other children may have sent upgrade
+requests and remain waiting, leading to a deadlock. The only way to break the
 vicious waiting cycle is for the children to concede and respond when they get
 downgrade requests. Of course, they have to downgrade the states of their own
-children first, and this process goes on recursively until the downgrade requests
-reach the leaf caches, which can respond immediately.
+children first, and this process goes on recursively until the downgrade
+requests reach the leaf caches, which can respond immediately.
 
@@
 
-\section{Proof that the MSI Protocol Preserves Store Atomicity}\label{sec:ccproof}
+\subsection{Proof that the MSI Protocol Preserves Store Atomicity}\label{sec:ccproof}
 \label{safety}
@@
 
-We can easily show that any system obeying store atomicity
-is sound with respect to $M_m$. Intuitively, the store atomicity
-property precisely describes the property of a memory, \viz{} that if a store
-happens, then that location gets updated such that any subsequent load gets
-back the updated value.
+We can easily show that any system obeying store atomicity is sound with
+respect to $M_m$. Intuitively, the store atomicity property precisely describes
+the property of a memory, \viz{} that if a store happens, then that location
+gets updated such that any subsequent load gets back the updated value.
 
-The proof that $M_c$ obeys store atomicity is involved enough to deserve a
-section's worth of explanation, which we now turn to.
+The proof that $M_c$ obeys store atomicity is involved enough that we will only
+state the lemmas used to prove this. The proofs in Coq for each theorem and
+lemma in this section can be found in our supplementary material.
 
-We first reduce the proof of store atomicity to several lemmas.
-We believe that other invalidation-based
-protocols with inclusive caches (\ie{} those cache hierarchies where a parent
-cache contains all the locations that any of its children contains) like MESI,
-MOSI, and MOESI also satisfy our lemmas, making their verification
-easier by reusing most of our MSI proof; we omit the discussion of extension of
-this proof to other protocols for space reasons.
+We believe that other invalidation-based protocols with inclusive caches (\ie{}
+those cache hierarchies where a parent cache contains all the locations that
+any of its children contains) like MESI, MOSI, and MOESI also satisfy our
+lemmas, making their verification easier by reusing most of our MSI proof; we
+omit the discussion of extension of this proof to other protocols for space
+reasons.
 
-The lemmas
-presented below hold after any number of transitions
+The following lemmas hold in the state reached after any number of transitions
 starting from the initial state.
-
-While we have proofs in Coq for each lemma, our discussion will be mainly
-informal for brevity. All the proofs can be found in our supplementary material.
-
-Intuitively our protocol is correct because it is guaranteed that any cache that attempts to
-handle a load request has the ``latest'' value.
 
@@
 
-It is easy to see that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also
-compatible if $x_1$ and $x_2$ are compatible.
-
 \begin{lemma}
@@
 \end{lemma}
-
-Lemmas~\ref{conservative} and~\ref{localCompatibility1} imply a
-$\ge$ relation between the states of a parent and its child for any address.
-Since the ancestor relation is the reflexive-transitive closure of the parent
-relation, the $\ge$ relation is preserved between the states of two caches $p$
-and $c$ where $p$ is the ancestor of $c$.
-
-Any two caches $c_1, c_2$ neither of which is an ancestor of the other have a
-lowest common ancestor $x$, which has
-two distinct children $x_1$ and $x_2$ such that they are the ancestors of $c_1$
-and $c_2$ respectively. For any address $a$, $\s(x_1, a)$ and
-$\s(x_2, a)$ are compatible by Lemma~\ref{localCompatibility2}. From
-the previous paragraph about the relation between the states of a cache and its
-ancestor, it follows that $c_1$ and $c_2$ are also compatible, proving
-Lemma~\ref{nonAncestorCompatible}.
-
-Using Lemmas~\ref{nonAncestorCompatible} and~\ref{noTransitWrite}, one can
-prove Lemma~\ref{latestValue}. The proof is by induction on time $T$; we assume
-that any \clean{} (node, address) pair has the latest value until time
-$T$. Consider a (node, addr) pair at time $T+1$.
-If the pair was \clean{} at $T$, then no cache,
-save itself (because of Lemma~\ref{nonAncestorCompatible}), can process any
-store request for that address, thus continuing to be \clean{} and having the latest value.
-And if the pair was not \clean{} at time $T$, it will receive data from a cache
-that is \clean{} for the (node, address) pair, and no store
-request can be processed when the data is in transit (by
-Lemma~\ref{noTransitWrite}). Therefore, by the induction hypothesis the cache will
-continue to have the latest value whenever it is \clean{} for an address at time $T+1$.
-
-Analyzing a parent-child pair will suffice in proving
-Lemmas~\ref{localCompatibility1} and \ref{localCompatibility2}.  Proof of
-Lemma~\ref{noTransitWrite} is similar to the proof of
-Lemma~\ref{nonAncestorCompatible}.
 

### 2014-11-05 Introduction.tex
We still need to cut two pages. If you want to go back to the earlier version, use TransitionUsefulSecond and Third instead of Second1 and Third1

@@
 Section~\ref{sec:ooo} discusses definition and verification of a speculative
-processor model, and Sections~\ref{sec:cc} and~\ref{sec:ccproof} respectively
-define and prove our hierarchical cache-coherence protocol.  Finally, in
+processor model, and Section~\ref{sec:cc}
+defines and proves our hierarchical cache-coherence protocol.  Finally, in
 Section~\ref{sec:finalresult} we show the whole-system modular proof of our

### 2014-11-09 Abstract.tex
stuff

--- 2014-10-19 Abstract.tex
@@
+%\xxx{\textbf{Murali's version in red:} We provide a framework for modular verification of multi-processor hardware
+%systems with respect to memory model specification. In this framework, we make
+%use of a key observation that eventhough different memory models have different
+%overall system behaviors, they all share a coherent cache hierarchy. We prove
+%that this coherent cache hierarchy implements the \emph{store atomicity}
+%property, which makes it behave like an atomic instantaneous memory with
+%respect to the rest of the system. We further modularize the proof by defining
+%what is meant for a processor to be correct independently of the memory system.
+%As a concrete instance of our framework, we walk through a modular proof of
+%high-performance multi-processor implementation of Lamport's Sequential
+%Consistency model. The processors employ \emph{speculative execution} to
+%generate many concurrent loads to the memory sub-system, while obeying our
+%correctness criteria. We offer the first formal proof that when an atomic
+%memory is connected to $n$ speculative processors, the system as a whole is
+%sequentially consistent.  Additionally, the cache-coherent memory we present in
+%this paper is the first fully verified system that supports arbitrary cache
+%hierarchies. Our framework involves describing all our components using labeled
+%transition systems and we machine-verify all our proofs using the Coq proof
+%assistant.}
 We present a new framework for modular verification of hardware designs
@@
 system implements sequential consistency.  One interesting aspect of the proof
-is that the memory component is well-specified to be reused for systems
+is that the memory component is well-specified to be reusable for systems
 implementing a variety of weak memory models.  To our knowledge, our proof

### 2014-11-09 TransitionsUsefulFirst.tex
stuff
lots of modifications

@@
 \section{Labeled Transition Systems}\label{sec:lts}
+
+\begin{figure*}
+\centering
+\begin{boxedminipage}{.53\textwidth}
+\noindent \textbf{rule} $(c.\parent == p \; \mathit{\&\&} \; c.\s[a] < x \; \mathit{\&\&} \; c.\wt[a]==$None$);$
+
+\hspace{1cm} $\ch[c][p][\req].\enq(a,c.\s[a],x); c.\wt[a] \Leftarrow x;$
+
+\textbf{endrule}
+\end{boxedminipage}
+\begin{boxedminipage}{\textwidth}
+\small
+\inference[]
+{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
+{\io{$M_c$}
+{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
+{\dt, \ch[ (c, p, \req) \coloneqq
+(a, \s(c, a), x) \uplus \ch(c,p,\req)], \s, \dst,
+\wt[(c,a)\coloneqq x], \dwt, \inp, \outp}{}}
+\end{boxedminipage}
+\caption{Rule in Bluespec and its corresponding LTS transliteration}
+\label{both}
+\end{figure*}
 
@@
 an empty or ``silent'' label $\epsilon$.  As a shorthand, we sometimes
-omit labels for $\epsilon$ steps.  Note that the IO exchange can be
-input, output, or both, and its interpretation is dependent on the
-context.
+omit labels for $\epsilon$ steps.
 
@@
 
-\begin{figure*}
-\centering
-\begin{boxedminipage}{.53\textwidth}
-\noindent \textbf{rule} $(c.\parent == p \; \mathit{\&\&} \; c.\s[a] < x \; \mathit{\&\&} \; c.\wt[a]==None);$
-
-\hspace{1cm} $\ch[c][p][\req].\enq(a,c.\s[a],x); c.\wt[a] \Leftarrow x;$
-
-\textbf{endrule}
-\end{boxedminipage}
-\begin{boxedminipage}{\textwidth}
-\inference[]
-{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
-{\dt, \ch[ (c, p, \req) \coloneqq
-(a, \s(c, a), x) \uplus \ch(c,p,\req)], \s, \dst,\\
-\wt[(c,a)\coloneqq x], \dwt, \inp, \outp}}
-\end{boxedminipage}
-\caption{Rule in Bluespec and its corresponding LTS transliteration}
-\label{both}
-\end{figure*}
-
-\section{Decomposing a Multi-Processor System}
+\section{Decomposing a Share Memory Multi-Processor System}
 \label{sec:store-atomicity}
@@
 
-Most any conventional multi-processor system can be divided into 3 components as shown in
+Any conventional multi-processor system can be logically divided into 3 components as shown in
 Figure \ref{zoom}. 
@@
 using a hierarchy of caches, in order to increase the performance of the
-overall system, since the latency of accessing a memory directly is very slow
-compared to accessing a much smaller cache. The component between each
+overall system, since the latency of accessing a memory directly is large
+compared to that of accessing a much smaller cache. The component between each
 processor and the global memory sub-system contains local buffers, $B_i$, specific to
@@
 {}
-{\io{$M_m$}{m}{m}{p, (\ld\req,a),(\ld\resp,m(a), \_)}}
+{\io{$M_m$}{m}{m}{p, ((\ld\req,a),(\ld\resp,m(a)), \_)}}
 

### 2014-11-09 TransitionsUsefulSecond1.tex
stuff

@@
 
-Note that, in this setting, an operational specification like the LTS
-for SC is precisely the right characterization of \textbf{full functional
-  correctness} for a hardware design, much as a
-precondition-postcondition pair does that job in a partial-correctness
-Hoare logic.  Our SC LTS fully constrains observable behavior of a
-system to remain consistent with simple interleaving.  Similar
-operational models are possible as top-level specifications for
-systems following weaker memory models, by giving the LTS for the \emph{Local
-Buffer} component and composing the three components simultaneously.
-
 However, we are not interested merely in verifying the particular system $O$.
@@
 
+Note that, in this setting, an operational specification like the LTS
+for SC is precisely the right characterization of \textbf{full functional
+  correctness} for a hardware design, much as a
+precondition-postcondition pair does that job in a partial-correctness
+Hoare logic.  Our SC LTS fully constrains observable behavior of a
+system to remain consistent with simple interleaving.  Similar
+operational models are possible as top-level specifications for
+systems following weaker memory models, by giving the LTS for the \emph{Local
+Buffer} component and composing the three components simultaneously.
+
 \begin{figure}
+\small
 \centering
@@
 \inference
-[LdRep]
+[LdRp]
 {\dec(s,\pc) = (\ld, x, a) & \exec(s, \pc, (\ld, x, v)) = (\pc', s')}
@@
 \inference
-[StRep]
+[StRp]
 {\dec(s,\pc) = (\st, a, v) & \exec(s, \pc, (\st)) = (\pc', s')}
@@
 indicating whether the processor is blocked waiting for a response from the
-memory system.
-
-As in the SC model, we change the state of the processor to $H$ whenever
-$\dec$ returns $H$.  Note that there is some subtlety in the design of
-this semantics regarding halting; we plan ahead to be sure we will be able to prove the
-appropriate simulation facts.  Those proofs would be harder if, \eg{}
-a processor could halt while it is still waiting for a memory response.
+memory system. As in the SC model, we change the state of the processor to $H$
+whenever $\dec$ returns $H$. 
+%Note that there is some subtlety in the design of
+%this semantics regarding halting; we plan ahead to be sure we will be able to prove the
+%appropriate simulation facts.  Those proofs would be harder if, \eg{}
+%a processor could halt while it is still waiting for a memory response.
 
@@
 as unordered sets. We will later use $\coloncolon$ for inserting (or removing)
-from FIFO buffers. Intuitively, the unordered requests/responses set suffices
-even for requests/responses for the same address because the processor always
-waits for a response before sending the next request.
+from FIFO buffers. %Intuitively, the unordered requests/responses set suffices
+%even for requests/responses for the same address because the processor always
+%waits for a response before sending the next request.
 %\medskip

### 2014-11-09 TransitionsUsefulFirst.tex
stuff

@@
 \section{Labeled Transition Systems}\label{sec:lts}
-
-\begin{figure*}
-\centering
-\begin{boxedminipage}{.53\textwidth}
-\noindent \textbf{rule} $(c.\parent == p \; \mathit{\&\&} \; c.\s[a] < x \; \mathit{\&\&} \; c.\wt[a]==$None$);$
-
-\hspace{1cm} $\ch[c][p][\req].\enq(a,c.\s[a],x); c.\wt[a] \Leftarrow x;$
-
-\textbf{endrule}
-\end{boxedminipage}
-\begin{boxedminipage}{\textwidth}
-\small
-\inference[]
-{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
-{\io{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
-{\dt, \ch[ (c, p, \req) \coloneqq
-(a, \s(c, a), x) \uplus \ch(c,p,\req)], \s, \dst,
-\wt[(c,a)\coloneqq x], \dwt, \inp, \outp}{}}
-\end{boxedminipage}
-\caption{Rule in Bluespec and its corresponding LTS transliteration}
-\label{both}
-\end{figure*}
 
@@
 %
-It is worth noting that the LTS notation described above, along with the
-composition operators are very similar to the semantics of hardware designs
-written in Bluespec. As an example consider an inference rule in LTS and the
-corresponding code in Bluespec in Figure \ref{both}. This code was actually
-written for the PowerPC system \cite{Khan:PowerPc} (we will be using this rule
-later in the paper when we describe the coherent caches).
+%It is worth noting that the LTS notation described above, along with the
+%composition operators are very similar to the semantics of hardware designs
+%written in Bluespec. As an example consider an inference rule in LTS and the
+%corresponding code in Bluespec in Figure \ref{both}. This code was actually
+%written for the PowerPC system \cite{Khan:PowerPc} (we will be using this rule
+%later in the paper when we describe the coherent caches).
 

### 2014-11-09 Introduction.tex
stuff

@@
 of the most serious real-world adopters of formal methods.  Hardware
-verification is dominated by model-checking, where abstraction techniques are
+verification is dominated by model-checking; for instance, processor
+verification ~\cite{burch1994automatic, mcmillan1998verification} and 
+more recently, Intel's execution cluster verification~\cite{kaivola2009replacing}.
+Many abstraction techniques are
 used to reduce designs to finite state spaces, which can be explored
-exhaustively.  There are limits to the construction of sound abstractions, so
+exhaustively.
+ There are limits to the construction of sound abstractions, so
 verifications of, e.g., cache-coherence protocols have almost always treated
@@
 an 8-core system can run 55 million instructions per second on
-the BEE FPGA board\cite{}, with a 2-level cache hierarchy.
+the BEE FPGA board\cite{}, with a 2-level cache hierarchy. Figure \ref{both} gives an example
+of an LTS inference rule transliterated directly from an actual Bluespec rule used
+in the PowerPC system.
+
+\begin{figure}[t]
+\centering
+\begin{boxedminipage}{\columnwidth}
+\small
+\noindent \textbf{rule} $(c.\parent == p \; \mathit{\&\&} \; c.\s[a] < x \; \mathit{\&\&} \; c.\wt[a]==$None$);$
+
+\hspace{1cm} $\ch[c][p][\req].\enq(a,c.\s[a],x); c.\wt[a] \Leftarrow x;$
+
+\textbf{endrule}
+\end{boxedminipage}
+\begin{boxedminipage}{\columnwidth}
+\small
+\inference[]
+{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
+{\ioxe{$M_c$}
+{\dt, \ch, \s,\\ \dst, \wt, \dwt,\\ \inp, \outp}
+{\dt, \ch[ (c, p, \req) \coloneqq(a, \s(c, a),x)\\
+\;\;\uplus \ch(c,p,\req)], \s, \dst,\\
+\wt[(c,a)\coloneqq x], \dwt,\inp, \outp}{}}
+\end{boxedminipage}
+\caption{Rule in Bluespec and its corresponding LTS transliteration}
+\label{both}
+\end{figure}
 

### 2014-11-09 RelatedWork.tex
lots of modifications
stuff

@@
 
-Formal verification has been used to verify multiprocessor
-systems, especially with model checking
-\cite{burch1994automatic,mcmillan1998verification}.
+%Formal verification has been used to verify multiprocessor
+%systems, especially with model checking
+%\cite{burch1994automatic,mcmillan1998verification}.
 
-More recently, Intel's execution cluster (\ie{} the ALU and its
-interface) has been formally verified using model checking
-\cite{kaivola2009replacing}.
+%More recently, Intel's execution cluster (\ie{} the ALU and its
+%interface) has been formally verified using model checking
+%\cite{kaivola2009replacing}.
 
@@
 the goal of making them more amenable to verification \cite{shen1999using}.
-
-%\murali{Next paragraph is all I know about this stuff, should probably be
-%removed}
-%\adam{What are you referring to with ``it'' in your comment?  I don't understand what this paragraph refers to.}
-%It has been verified using model checking \cite{qadeer2003verifying},
-%with a fixed number of processors and memory locations.  Model-checking has also
-%been used to verify communication fabrics \cite{chatterjee2012automatic}.
 
@@
 invariants, making it more adaptable to different protocols.
+Cachet~\cite{StoyShenArvind:Proofs}, another single-level cache coherence
+protocol, was verified to be obey the CRF memory
+model~\cite{Shen:CRF}, and parts of the proof were implemented using PVS.
 
-Cachet~\cite{StoyShenArvind:Proofs}, another single-level cache coherence
-protocol, was proven to be implementable in the CRF memory
-model~\cite{Shen:CRF}. Shen \etal{} provided a paper-and-pencil proof showing
-Cachet does not violate the CRF memory model, while Stoy \etal{} implemented
-portions of the paper-and-pencil proof using PVS. However, neither proof
-establishes that the protocol is store atomic. Rather, they show that Cachet
-obeys the CRF memory model, which is not restricted to store-atomic behavior.
-
-\xxx{TODO: Correct This}
+% Shen \etal{} provided a paper-and-pencil proof showing
+%Cachet does not violate the CRF memory model, while Stoy \etal{} implemented
+%portions of the paper-and-pencil proof using PVS. However, neither proof
+%establishes that the protocol is store atomic. Rather, they show that Cachet
+%obeys the CRF memory model, which is not restricted to store-atomic behavior.
 

### 2014-11-09 TransitionsUsefulSecond1.tex
stuff

@@
 {\ior{P$_\text{ref}$}{s, \pc, \bot}
-{s, \pc, \top}{\ld, a}}
+{s, \pc, \top}{\epsilon, \ld, a}}
 
@@
 {\iol{P$_\text{ref}$}{s, \pc, \top}
-{s', \pc', \bot}{\ld, v}}
+{s', \pc', \bot}{\epsilon, \ld, v}}
 
@@
 
-\inference
-[SpLoad]
-{\inp(i) = \mathit{qs} \uplus \{\spc\ld,t,a\}}
-{\io{$M_m$}{\inp, \outp, m}{i, \inp[i \coloneqq \mathit{qs}], \outp[i \coloneqq
-(\spc\ld, t, m(a)) \uplus \outp(i)], m}{}}
+%\inference
+%[SpLoad]
+%{\inp(i) = \mathit{qs} \uplus \{\spc\ld,t,a\}}
+%{\io{$M_m$}{\inp, \outp, m}{i, \inp[i \coloneqq \mathit{qs}], \outp[i \coloneqq
+%(\spc\ld, t, m(a)) \uplus \outp(i)], m}{}}
 
@@
 [Load]
-{\inp(i) = \mathit{qs} \uplus \{\ld,a\}}
+{\inp(i) = \mathit{qs} \uplus \{t, \ld,a\}}
 {\io{$M_m$}{\inp, \outp, m}{i, \inp[i \coloneqq \mathit{qs}], \outp[i \coloneqq
-(\ld, m(a)) \uplus \outp(i)], m}{}}
+(t, \ld, m(a)) \uplus \outp(i)], m}{}}
 
@@
 processors.  The input-output interface of this memory $M_m$ is largely what
-one would expect from the P$_{\text{ref}}$ definition.  However, in anticipation of the
-more complex processor we will define later, we add requests and responses for
-\emph{speculative loads}.  These are load operations that a processor may
-attempt, before it knows that the program will actually need the result.  Many
-speculative loads from a single processor may be in flight at once, improving
-performance via parallelism.  A request to memory like $(\spc\ld, t, a)$ asks
+one would expect from the P$_{\text{ref}}$ definition. 
+%However, in anticipation of the
+%more complex processor we will define later, we add requests and responses for
+%\emph{speculative loads}.  These are load operations that a processor may
+a%ttempt, before it knows that the program will actually need the result.  Many
+s%peculative loads from a single processor may be in flight at once, improving
+p%erformance via parallelism. 
+A request to memory like $(t, \ld, a)$ asks
 for the value of memory cell $a$, associating a \emph{tag} $t$ that the
 processor can use to match responses to requests.  Those responses take the
-form $(\spc\ld, t, v)$, giving the value $v$ of the requested memory address.
+form $(t, \ld, v)$, giving the value $v$ of the requested memory address.
 
@@
 
-Note that we have used $\uplus$ for inserting (or removing) from the buffers
-$\inp$ and $\outp$. The intent is to convey that $\inp$ and $\outp$ are treated
-as unordered sets. We will later use $\coloncolon$ for inserting (or removing)
-from FIFO buffers. %Intuitively, the unordered requests/responses set suffices
+%Note that we have used $\uplus$ for inserting (or removing) from the buffers
+%$\inp$ and $\outp$. The intent is to convey that $\inp$ and $\outp$ are treated
+%as unordered sets. We will later use $\coloncolon$ for inserting (or removing)
+%from FIFO buffers. Intuitively, the unordered requests/responses set suffices
 %even for requests/responses for the same address because the processor always
@@
 
-Now we can compose these LTSes to produce an implementation of SC.  Rule SpLoad
-in Figure~\ref{$M_m$} will be irrelevant in this particular composition, since
-P$_{\text{ref}}$ never does speculation.
+Now we can compose these LTSes to produce an implementation of SC. 
+% Rule SpLoad
+%in Figure~\ref{$M_m$} will be irrelevant in this particular composition, since
+%P$_{\text{ref}}$ never does speculation.
 
@@
 \item $\upd\ld(\rob,t,v)$, which informs the buffer that the memory
-has returned result value $v$ for the speculative load with tag $t$.
+has returned result value $v$ for the speculative load with tag $t \neq \epsilon$.
 \item $\commit(\rob)$, which returns the next instruction in serial
@@
 {\ior{\text{P$_\text{so}$}}{s,\pc,\wait,\rob,\ppc}{s,\pc,\wait,\rob',\ppc}
-{\spc\ld, t, a}}
+{t,\ld, a}}
 
 \inference[SpecLdRep]
-{}
+{t \neq \epsilon}
 {\iol{\text{P$_\text{so}$}}{s,\pc,\wait,\rob,\ppc}
 {s,\pc,\wait,\upd\ld(\rob, t, v),\ppc
-}{\spc\ld,t,v}}
+}{t,\ld,v}}
 
@@
 {\commit(\rob) = (\pc,\pc',(\ld,x,a,v,s'))}
-{\ior{\text{P$_\text{so}$}}{s,\pc,\bot,\rob,\ppc}{s,\pc,\top,\rob,\ppc}{\ld, a}}
+{\ior{\text{P$_\text{so}$}}{s,\pc,\bot,\rob,\ppc}{s,\pc,\top,\rob,\ppc}{\epsilon, \ld, a}}
 \inference[StRep]
@@
 {\commit(\rob) = (\pc,\pc',(\ld,x,a,v,s'))}
-{\iol{\text{P$_\text{so}$}}{s,\pc,\top,\rob,\ppc}{s',\pc',\bot,\retire(\rob),\ppc}{\ld,v}}
+{\iol{\text{P$_\text{so}$}}{s,\pc,\top,\rob,\ppc}{s',\pc',\bot,\retire(\rob),\ppc}{\epsilon, \ld,v}}
 
@@
 {\commit(\rob) = (\pc,\pc',(\ld,x,a,v',s')) & v' \neq v & \exec(s,\pc,(\ld,x,v)) = (\pc'', s'')}
-{\iol{\text{P$_\text{so}$}}{s,\pc,\top,\rob,\ppc}{s'',\pc'',\bot,\phi,\setBp(\ppc,\pc)}{\ld, v}}
+{\iol{\text{P$_\text{so}$}}{s,\pc,\top,\rob,\ppc}{s'',\pc'',\bot,\phi,\setBp(\ppc,\pc)}{\epsilon, \ld, v}}
 
@@
 We will see the processor issue two kinds of loads, a speculative
-load ($\spc\ld$) and a commit or real load ($\ld$).  To maintain SC, every speculative
+load (whose tag is not $\epsilon$) and a commit or real load (whose tag is $\epsilon$).  To maintain SC, every speculative
 load must have a matching verification load later on, and we maintain
@@
 communication labels with $\toM, \toP$) is identical
-to that of the reference processor, but now we can exercise a memory
-system's transitions for speculative loads.
+to that of the reference processor.
+% but now we can exercise a memory
+%system's transitions for speculative loads.
 
@@
 $\io{P$_\text{so}^{*}$}{s_0,\pc_0,\bot,\phi,\ppc_0}{s,\pc,\wait,\rob,\ppc}{\eta} \Rightarrow$
-\begin{displaymath}
-\left\{
+
+\begin{math}
+\;\;\;\;\;\;\;\;\left\{
 \begin{array}{ll}
@@
 \end{array}
-\right.\end{displaymath}
+\right.\end{math}
 \label{rob}
@@
 out-of-order processors.  We use a trace-transformation function
-$\noSpec$ that drops all speculative-load requests and responses.
+$\noSpec$ that drops all speculative-load requests and responses (\ie{} those load requests and responses whose tags are $\epsilon$).
 See Definition~\ref{refines} for a review of how we use such
@@
 \begin{proof}
-Direct consequence of Theorems~\ref{ocorrect} and~\ref{liftn} (where the
+Direct consequence of Theorems~\ref{ocorrect},~\ref{liftn} (where the
 latter is one of our preliminary results about $n$-repetition).

### 2014-11-09 TransitionsUsefulSecond1.tex
stuff

@@
 \end{theorem}
-\begin{proof}
+\begin{prf}
 \vspace{-.1cm}
@@
 (projected) state.
-\end{proof}
+\end{prf}
 
@@
 \end{theorem}
-\begin{proof}
+\begin{prf}
 By induction on P$_\text{so}$ traces, using a proper abstraction
@@
 execution of P$_\text{ref}$.
-\end{proof}
+\end{prf}
 
@@
 \end{corollary}
-\begin{proof}
-Direct consequence of Theorems~\ref{ocorrect},~\ref{liftn} (where the
+\begin{prf}
+Direct consequence of Theorems~\ref{ocorrect},~\ref{liftn} (the
 latter is one of our preliminary results about $n$-repetition).
-\end{proof}
+\end{prf}

### 2014-11-09 TransitionsUsefulThird1.tex
stuff

@@
 $\dst(p,c,a)$ & Directory state of cache $p$ for child $c$ and address $a$\\
-$\wt(c,a)$ & Gives cache $c$'s desired coherence state to upgrade to (for address $a$), or $\epsilon$ if none\\
-$\dwt(p,c,a)$ & Gives cache $p$'s desired coherence state for its child $c$ to downgrade to (for address $a$), or $\epsilon$ if none\\
+$\wt(c,a)$ & State cache $c$ is waiting to upgrade to for address $a$, or $\epsilon$\\ %Gives cache $c$'s desired coherence state to upgrade to (for address $a$), or $\epsilon$ if none\\
+$\dwt(p,c,a)$ & State cache $p$ is waiting its child $c$ to downgrade to for address $a$, or $\epsilon$\\
+%$\dwt(p,c,a)$ & Gives cache $p$'s desired coherence state for its child $c$ to downgrade to (for address $a$), or $\epsilon$ if none\\
 $\ch$ & Channels between nodes in the system\\
@@
 \begin{defn}
-\begin{displaymath}
-\compat(p, c, x, a) = \left\{
+$\compat(p, c, x, a) =$
+%\vspace{.3cm}
+
+\begin{math}
+\left\{
 \begin{array}{ll}
@@
 \right.
-\end{displaymath}
+\end{math}
 \end{defn}
@@
 \end{theorem}
-\begin{proof}
+\begin{prf}
 By induction on traces, with an identity abstraction function.
-\end{proof}
+\end{prf}
 
@@
 \end{theorem}
-\begin{proof}
+\begin{prf}
 Follows from Theorem \ref{liftplus} (one of our preliminary results
 about $+$), Corollary \ref{ges}, and Theorem \ref{mcorrect}.
-\end{proof}
+\end{prf}
 
@@
 \end{theorem}
-\begin{proof}
+\begin{prf}
 Follows from Theorems \ref{ccorrect} and \ref{liftplus} plus reflexivity of $\sqsubseteq$ (Theorem \ref{transitive}).
-\end{proof}
+\end{prf}
 
@@
 \end{theorem}
-\begin{proof}
+\begin{prf}
 We twice apply $\sqsubseteq$ transitivity (Theorem \ref{transitive}) to
@@
 connect Theorems \ref{ocorrect2}, \ref{ocorrect1}, and \ref{scthm}.
-\end{proof}
+\end{prf}

### 2014-11-09 TransitionsUsefulSecond1.tex
stuff

@@
 \end{boxedminipage}
-\caption{LTS for sequential consistency with $n$ simple processors}
+\caption{LTS for SC with $n$ simple processors}
 \label{Ref}
@@
 
-%\medskip
+\medskip
 
@@
 \begin{figure*}[t]
-\begin{displaymath}
+\begin{math}
 \upd((s, \pc, \_), \mathit{rs}) =
@@
 \right.
-\end{displaymath}
+\end{math}
 \caption{Mapping of states from $(\text{P$^n_\text{ref}$} + M_m)$ to SC}
@@
 \begin{prf}
-\vspace{-.1cm}
+%\vspace{-.1cm}
 By induction on traces of the decoupled system.  We need to choose an
@@
 \begin{inv}
-If $P_\text{so}$ reaches a state $(s, \pc, \wait, \rob, \ppc)$, \ie{}
-
+If $P_\text{so}$ reaches a state $(s, \pc, \wait, \rob, \ppc)$, \ie{}\hspace{.2cm}
 $\io{P$_\text{so}^{*}$}{s_0,\pc_0,\bot,\phi,\ppc_0}{s,\pc,\wait,\rob,\ppc}{\eta} \Rightarrow$

### 2014-11-09 TransitionsUsefulThird1.tex
stuff

@@
 
-The following lemmas hold in the state reached after any number of transitions
-starting from the initial state.
-
 \begin{lemma}
@@
 has the required permission, at which point it is guaranteed to have the latest
-value, because it is \textit{clean} (as it has no children).
-
-Lemma~\ref{latestValue} requires further decomposition into more
-lemmas.
+value, because it is \textit{clean} (as it has no children). To prove Lemma
+\ref{latestValue}, it has to be decomposed futher into the following, each of
+which holds at any time.
 
@@
 \begin{defn}
-\textit{compatible}: Values of states $x_1$ and $x_2$ are \textit{compatible}
+\textit{compatible}: States $x_1$ and $x_2$ are \textit{compatible}
 iff whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In.
@@
 \begin{lemma}
-\textit{conservative}: The state of an address in a cache is never greater than
-the corresponding entry in its parent's directory.
+\textit{conservative}: A cache's state for an address is never greater than
+the corresponding state in its parent's directory.
 \label{conservative}

### 2014-11-09 TransitionsUsefulThird1.tex
stuff

@@
 $\dst(p,c,a)$ & Directory state of cache $p$ for child $c$ and address $a$\\
-$\wt(c,a)$ & State cache $c$ is waiting to upgrade to for address $a$, or $\epsilon$\\ %Gives cache $c$'s desired coherence state to upgrade to (for address $a$), or $\epsilon$ if none\\
-$\dwt(p,c,a)$ & State cache $p$ is waiting its child $c$ to downgrade to for address $a$, or $\epsilon$\\
+$\wt(c,a)$ & State that cache $c$ is waiting to upgrade to, for address $a$, or $\epsilon$\\ %Gives cache $c$'s desired coherence state to upgrade to (for address $a$), or $\epsilon$ if none\\
+$\dwt(p,c,a)$ & State that cache $p$ is waiting its child $c$ to downgrade to, for address $a$, or $\epsilon$\\
 %$\dwt(p,c,a)$ & Gives cache $p$'s desired coherence state for its child $c$ to downgrade to (for address $a$), or $\epsilon$ if none\\
@@
 \begin{lemma}
-\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
-that neither is an ancestor of the other in the cache-tree hierarchy (where
-the ancestor relation is the reflexive-transitive closure of the parent relation),
-for each address $a$, $\s(c_1,a)$ and $\s(c_2,a)$ are
-\textit{compatible} (defined below).
+\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$, neither of which is the
+ancestor of the other in the cache-tree,
+%such
+%that neither is an ancestor of the other in the cache-tree hierarchy (where
+%the ancestor relation is the reflexive-transitive closure of the parent relation),
+coherence states $\s(c_1,a)$ and $\s(c_2,a)$ are
+\textit{compatible} (defined below) for every address $a$.
 \label{nonAncestorCompatible}
@@
 \begin{defn}
-\textit{compatible}: States $x_1$ and $x_2$ are \textit{compatible}
+\textit{compatible}: Coherence states $x_1$ and $x_2$ are \textit{compatible}
 iff whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In.

### 2014-11-09 TransitionsUsefulFirst.tex
stuff

@@
 
-\begin{figure}
-\small
-\centering
-\begin{boxedminipage}[c]{.48\textwidth}
-\inference
-[Load]
-{}
-{\io{$M_m$}{m}{m}{p, ((\ld\req,a),(\ld\resp,m(a)), \_)}}
-
-\inference
-[Store]
-{}
-{\io{$M_m$}{m}{m[a\coloneqq v]}{p, (\st, a, v)}}
-
-\end{boxedminipage}
-\caption{LTS for a simple memory}
-\label{M_m}
-\end{figure}
+%\begin{figure}
+%\small
+%\centering
+%\begin{boxedminipage}[c]{.48\textwidth}
+%\inference
+%[Load]
+%{}
+%{\io{$M_m$}{m}{m}{p, ((\ld\req,a),(\ld\resp,m(a)), \_)}}
+%
+%\inference
+%[Store]
+%{}
+%{\io{$M_m$}{m}{m[a\coloneqq v]}{p, (\st, a, v)}}
+%
+%\end{boxedminipage}
+%\caption{LTS for a simple memory}
+%\label{M_m}
+%\end{figure}
 
@@
 
-So, we focus on this opportunity to simplify proof decomposition.  We
-will prove that our main memory component satisfies an intuitive
-\emph{store atomicity} property.  Despite the potential red herring of
-the word ``atomicity,'' this specification is appropriate even for
-implementation of weaker memory models than SC.  Store atomicity can
-be understood via the operational semantics of Figure \ref{M_m},
-describing an LTS that receives load and store requests ($\ld$ and
-$\st$) from processors and sends back load responses ($\ld\resp$);
-$p$ is the processor which sends and receives the request.
+\begin{figure}
+\small
+\centering
+\begin{boxedminipage}[c]{.48\textwidth}
+\inference
+[Ins]
+{}
+{\iorn{$M_m$}{\inp, \outp, m}{\inp[p \coloneqq \{q\} \uplus \inp(p)], \outp, m}{p}{q}}
+
+\inference
+[Rm]
+{\outp(p) = \mathit{rs} \uplus \{r\}}
+{\ioln{$M_m$}{\inp, \outp, m}{\inp, \outp[p \coloneqq \mathit{rs}], m}{p}{r}}
+
+\inference
+[Ld]
+{\inp(p) = \mathit{qs} \uplus \{t, \ld,a\}}
+{\ioxe{$M_m$}{\inp, \\\outp, m}{\inp[p \coloneqq \mathit{qs}], \outp[p \coloneqq\\
+\;\;\;(t, \ld, m(a)) \uplus \outp(p)], m}{}}
+
+\inference
+[St]
+{\inp(p) = \mathit{qs} \uplus \{\st,a,v\}}
+{\ioxe{$M_m$}{\inp, \\\outp, m}{\inp[p \coloneqq \mathit{qs}], \outp[p \coloneqq
+(\st) \\\;\;\;\uplus \outp(p)], m[a\coloneqq v]}{}}%p, (\st, a,v)}}
+
+\end{boxedminipage}
+\caption{LTS for a simple memory}
+\label{$M_m$}
+\end{figure}
+
+So, we focus on this opportunity to simplify proof decomposition.  We will
+prove that our main memory component satisfies an intuitive \emph{store
+atomicity} property.  Despite the potential red herring of the word
+``atomicity,'' this specification is appropriate even for implementation of
+weaker memory models than SC.  Store atomicity can be understood via the
+operational semantics of Figure \ref{$M_m$}, describing an LTS that receives
+load and store requests ($\ld$ and $\st$) from processors and sends back load
+responses ($\ld\resp$). The transfer happens via input buffers $\inp(p)$ from
+processor $p$ and output buffers $\outp(p)$ to processor $p$.
 

### 2014-11-09 TransitionsUsefulSecond1.tex
stuff

@@
 \inference
-[LdReq]
+[LdRq]
 {\dec(s,\pc) = (\ld, x, a)}
@@
 \inference
-[StReq]
+[StRq]
 {\dec(s,\pc) = (\st, a, v)}
@@
 \end{figure}
-
-\begin{figure*}
-\centering
-\begin{boxedminipage}[c]{.48\textwidth}
-\inference
-[Insert]
-{}
-{\iorn{$M_m$}{\inp, \outp, m}{\inp[i \coloneqq \{q\} \uplus \inp(i)], \outp, m}{i}{q}}
-
-\inference
-[Remove]
-{\outp(i) = \mathit{rs} \uplus \{r\}}
-{\ioln{$M_m$}{\inp, \outp, m}{\inp, \outp[i \coloneqq \mathit{rs}], m}{i}{r}}
-
-%\inference
-%[SpLoad]
-%{\inp(i) = \mathit{qs} \uplus \{\spc\ld,t,a\}}
-%{\io{$M_m$}{\inp, \outp, m}{i, \inp[i \coloneqq \mathit{qs}], \outp[i \coloneqq
-%(\spc\ld, t, m(a)) \uplus \outp(i)], m}{}}
-
-\inference
-[Load]
-{\inp(i) = \mathit{qs} \uplus \{t, \ld,a\}}
-{\io{$M_m$}{\inp, \outp, m}{i, \inp[i \coloneqq \mathit{qs}], \outp[i \coloneqq
-(t, \ld, m(a)) \uplus \outp(i)], m}{}}
-
-\inference
-[Store]
-{\inp(i) = \mathit{qs} \uplus \{\st,a,v\}}
-{\io{$M_m$}{\inp, \outp, m}{i, \inp[i \coloneqq \mathit{qs}], \outp[i \coloneqq
-(\st) \uplus \outp(i)], m[a\coloneqq v]}{}}%i, (\st, a,v)}}
-
-\end{boxedminipage}
-\caption{LTS for a simple memory}
-\label{$M_m$}
-\end{figure*}
 
@@
 
-Figure~\ref{$M_m$} defines a simple memory, meant to be composed with P$_{\text{ref}}$
-processors.  The input-output interface of this memory $M_m$ is largely what
-one would expect from the P$_{\text{ref}}$ definition. 
+The simple memory defined earlier in Figure~\ref{$M_m$} is meant to be composed with P$_{\text{ref}}$
+processors.  %The input-output interface of this memory $M_m$ is largely what
+%one would expect from the P$_{\text{ref}}$ definition. 
 %However, in anticipation of the
@@
 A memory state is a triple $(\inp, \outp, m)$, giving not just the
-memory $m$ itself, but also queues $\inp$ and $\outp$ for receiving
+memory $m$ itself, but also buffers $\inp$ and $\outp$ for receiving
 processor requests and batching up responses to processors,
@@
 
-\begin{figure*}[t]
+\begin{figure}[t]
+\small
 \centering
@@
 {}
-{\ioe{P$_\text{so}$}{s,\pc,\wait,\rob,\ppc}{s,\pc,\wait,\add(\curPpc(\ppc),\rob),\nextPpc(\ppc)}}
+{\ioxe{P$_\text{so}$}{\hspace{-.2cm}s,\pc,\wait,\\\rob,\ppc}{\hspace{-.2cm}s,\pc,\wait,\add(\curPpc(\ppc),\\\;\;\;\rob),\nextPpc(\ppc)}}
 
@@
 
-\inference[SpecLdReq]
+\inference[SpLdRq]
 {\compute(\rob) = (\rob',(\spc\ld,t,a))}
@@
 
-\inference[SpecLdRep]
+\inference[SpLdRp]
 {t \neq \epsilon}
-{\iol{\text{P$_\text{so}$}}{s,\pc,\wait,\rob,\ppc}
-{s,\pc,\wait,\upd\ld(\rob, t, v),\ppc
-}{t,\ld,v}}
+{\iox{\text{P$_\text{so}$}}{s,\pc,\wait,\\\rob,\ppc}
+{s,\pc,\wait,\\\upd\ld(\rob, t, v),\ppc
+}{\text{ToP}(t,\ld,v)}}
 
@@
 {\commit(\rob) = (\pc',\_,\_) & \pc' \neq \pc}
-{\ioe{\text{P$_\text{so}$}}{s,\pc,\wait,\rob,\ppc}{s,\pc,\wait,\phi,\setBp(\ppc,\pc)}}
+{\ioxe{\text{P$_\text{so}$}}{s,\pc,\wait,\\\rob,\ppc}{s,\pc,\wait,\phi,\\\setBp(\ppc,\pc)}}
 
@@
 
-\inference[StReq]
+\inference[StRq]
 {\commit(\rob) = (\pc,\pc',(\st,a,v,s'))}
@@
 
-\inference[LdReq]
+\inference[LdRq]
 {\commit(\rob) = (\pc,\pc',(\ld,x,a,v,s'))}
 {\ior{\text{P$_\text{so}$}}{s,\pc,\bot,\rob,\ppc}{s,\pc,\top,\rob,\ppc}{\epsilon, \ld, a}}
-\inference[StRep]
+\inference[StRp]
 {\commit(\rob) = (\pc,\pc',(\st,a,v,s'))}
@@
 
-\inference[LdRepGood]
+\inference[LdRpGood]
 {\commit(\rob) = (\pc,\pc',(\ld,x,a,v,s'))}
-{\iol{\text{P$_\text{so}$}}{s,\pc,\top,\rob,\ppc}{s',\pc',\bot,\retire(\rob),\ppc}{\epsilon, \ld,v}}
+{\iox{\text{P$_\text{so}$}}{s,\pc,\top,\\\rob,\ppc}{s',\pc',\bot,\\\retire(\rob),\ppc}{\text{ToP}(\epsilon, \ld,v)}}
 
-\inference[LdRepBad]
-{\commit(\rob) = (\pc,\pc',(\ld,x,a,v',s')) & v' \neq v & \exec(s,\pc,(\ld,x,v)) = (\pc'', s'')}
-{\iol{\text{P$_\text{so}$}}{s,\pc,\top,\rob,\ppc}{s'',\pc'',\bot,\phi,\setBp(\ppc,\pc)}{\epsilon, \ld, v}}
+\inference[LdRpBad]
+{\commit(\rob) = (\pc,\pc',(\ld,x,a,v',s')) \\ v' \neq v & \exec(s,\pc,(\ld,x,v)) = (\pc'', s'')}
+{\iox{\text{P$_\text{so}$}}{s,\pc,\top,\\\rob,\ppc}{s'',\pc'',\bot,\phi,\\\setBp(\ppc,\pc)}{\text{ToP}(\epsilon, \ld, v)}}
 
@@
 \label{Ooo}
-\end{figure*}
+\end{figure}
 

### 2014-11-09 TransitionsUsefulSecond1.tex
stuff

@@
 \begin{figure*}[t]
+\small
 \begin{math}
@@
 
-The branch predictor is the simpler of the two components to model.
+The branch predictor is the simpler of the two components.
 We indicate branch-predictor state with metavariable $\ppc$.
@@
 current program-counter prediction; $\nextPpc(\ppc)$, to ask the
-predictor to step forward by one instruction; and $\setBp(\ppc, \pc)$,
+predictor to get the program-counter for the next instruction; and $\setBp(\ppc, \pc)$,
 to reset prediction to begin at a known-accurate position $\pc$.  It
@@
 \begin{math}
+\small
 \;\;\;\;\;\;\;\;\left\{

### 2014-11-09 TransitionsUsefulSecond1.tex
stuff

--- 2014-11-09 TransitionsUsefulSecond1.tex
@@
 
-\begin{figure*}[t]
+\begin{figure}[h]
+\begin{boxedminipage}{\columnwidth}
+\small
+$\upd((s, \pc, \_), \mathit{rs}) =$
+
 \begin{math}
-\upd((s, \pc, \_), \mathit{rs}) =
-\left\{
+\;\;\;\;\;\left\{
 \begin{array}{ll}
-s', \pc' &: (\ld, v) \in \mathit{rs} \wedge \dec(pc, s) = (\ld, x, a) \wedge
-\exec(pc, s, (\ld, x, v)) = (\pc', s'))\\
-s', \pc' &: (\st) \in \mathit{rs} \wedge \dec(pc, s) = (\st, a, v) \wedge
-\exec(pc, s, (\st)) = (\pc', s')\\
+s', \pc' &: (\ld, v) \in \mathit{rs} \wedge \dec(pc, s) = (\ld, x, a) \wedge\\
+\multicolumn{2}{r}{\exec(pc, s, (\ld, x, v)) = (\pc', s'))}\\
+s', \pc' &: (\st) \in \mathit{rs} \wedge \dec(pc, s) = (\st, a, v) \wedge\\
+\multicolumn{2}{r}{\exec(pc, s, (\st)) = (\pc', s')}\\
 s, \pc &: \text{otherwise if } \dec(pc,s) \neq H\\
@@
 \end{math}
+\end{boxedminipage}
 \caption{Mapping of states from $(\text{P$^n_\text{ref}$} + M_m)$ to SC}
 \label{smap}
-\end{figure*}
+\end{figure}
 
@@
 \begin{prf}
-%\vspace{-.1cm}
 By induction on traces of the decoupled system.  We need to choose an
@@
 
-The branch predictor is the simpler of the two components to model.
-We indicate branch-predictor state with metavariable $\ppc$.
+The branch predictor is the simpler of the two components, whose state
+is indicated with metavariable $\ppc$.
 The operations on such state are $\curPpc(\ppc)$, to extract the
-current program-counter prediction; $\nextPpc(\ppc)$, to ask the
-predictor to step forward by one instruction; and $\setBp(\ppc, \pc)$,
-to reset prediction to begin at a known-accurate position $\pc$.  It
-turns out that we need to impose no explicit correctness criterion on
-branch predictors.  The processor uses predictions only as hints, and
+current program-counter prediction; $\nextPpc(\ppc)$, to get the predicted
+program-counter for the next instruction; and $\setBp(\ppc, \pc)$,
+to reset prediction to begin at a known-accurate position $\pc$. We need to impose no explicit correctness criterion on
+branch predictors; the processor uses predictions only as hints, and
 it always resets the predictor using $\setBp$ after detecting an
@@
 involved.  We write $\rob$ as a metavariable for reorder buffer
-state.  The associated operations are:
+state. $\phi$ denotes the state of an empty buffer. The operations associated with $\rob$ are:
 \begin{itemize}
-\item $\phi$, the state of an empty buffer.
+%\item $\phi$, the state of an empty buffer.
 \item $\add(\pc, \rob)$, which appends the program instruction at location $\pc$ to the list of instructions that the buffer is allowed to consider executing.
@@
 processor, with speculative messages erased.
-
+%\vspace{-.6cm}
 \begin{theorem}

### 2014-11-09 TransitionsUsefulThird1.tex
stuff

@@
 
-\begin{figure}
+\begin{figure}[h]
 \centering
@@
 
-\subsection{Proof that the MSI Protocol Preserves Store Atomicity}\label{sec:ccproof}
+\subsection{Proving $M_c$ is Store Atomicity}\label{sec:ccproof}
 \label{safety}
@@
 Store atomicity is defined as the following property:
-For a load (or speculative load) request $\toM(\ld, a)$ (or $\toM(\spc\ld, t, a)$)
+For a load request $\toM(t, \ld, a)$
 from any processor, the response sent at time $T$
-$\toP(\ld, v)$ (or $\toP(\spc\ld, t, v)$) should be such that
+$\toP(t, \ld, v)$ should be such that
 \begin{enumerate}

### 2014-11-09 TransitionsUsefulThird1.tex
stuff

@@
 For a load request $\toM(t, \ld, a)$
-from any processor, the response sent at time $T$
+from any processor, the response sent at time $T$,
 $\toP(t, \ld, v)$ should be such that

### 2014-11-09 TransitionsUsefulThird1.tex
stuff

--- 2014-11-09 TransitionsUsefulThird1.tex
@@
 \begin{defn}
-Store atomicity is defined as the following property:
+\textit{Store atomicity}:
 For a load request $\toM(t, \ld, a)$
-from any processor, the response sent at time $T$
+from any processor, the response sent at time $T$,
 $\toP(t, \ld, v)$ should be such that

### 2014-11-09 TransitionsUsefulSecond1.tex
stuff

@@
 [NonMem]
-{\theta(i) = (s, \pc) & \dec(s,\pc) = (\nm, x) \\ \exec(s, \pc, (\nm, x)) = (\pc', s')}
+{\theta(i) = (s, \pc) & \dec(s,\pc) = (\nm, x) \\ \exec(s, \pc, (\nm, x)) = (s',\pc')}
 {\io{SC}{\theta,m}{\theta[i \coloneqq (s', \pc')], m}{}}
@@
 [Load]
-{\theta(i) = (s,\pc) & \dec(s,\pc) = (\ld, x, a) \\ \exec(s, \pc, (\ld, x, m(a))) = (\pc', s')}
+{\theta(i) = (s,\pc) & \dec(s,\pc) = (\ld, x, a) \\ \exec(s, \pc, (\ld, x, m(a))) = (s',\pc')}
 {\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m}{}}
@@
 [Store]
-{\theta(i) = (s,\pc) & \dec(s,\pc) = (\st, a, v) \\ \exec(s, \pc, (\st)) = (\pc', s')}
+{\theta(i) = (s,\pc) & \dec(s,\pc) = (\st, a, v) \\ \exec(s, \pc, (\st)) = (s',\pc')}
 {\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m[a\coloneqq v]}{}}
@@
 function $\exec(s, \pc, \mathit{dec})$ actually executes the instruction,
-returning the next program counter $\pc'$ and the new state $s'$.
+returning the new state $s'$ and the next program counter $\pc'$.
 
@@
 [NM]
-{\dec(s,\pc) = (\nm, x) & \exec(s, \pc, (\nm, x)) = (\pc', s')}
+{\dec(s,\pc) = (\nm, x) & \exec(s, \pc, (\nm, x)) = (s', \pc')}
 {\ioe{P$_\text{ref}$}{s,\pc,\bot}{s', \pc', \bot}}
@@
 [LdRp]
-{\dec(s,\pc) = (\ld, x, a) & \exec(s, \pc, (\ld, x, v)) = (\pc', s')}
+{\dec(s,\pc) = (\ld, x, a) & \exec(s, \pc, (\ld, x, v)) = (s',\pc')}
 {\iol{P$_\text{ref}$}{s, \pc, \top}
@@
 [StRp]
-{\dec(s,\pc) = (\st, a, v) & \exec(s, \pc, (\st)) = (\pc', s')}
+{\dec(s,\pc) = (\st, a, v) & \exec(s, \pc, (\st)) = (s',\pc')}
 {\iol{P$_\text{ref}$}{s, \pc, \top}
@@
 s', \pc' &: (\ld, v) \in \mathit{rs} \wedge \dec(pc, s) = (\ld, x, a) \wedge\\
-\multicolumn{2}{r}{\exec(pc, s, (\ld, x, v)) = (\pc', s'))}\\
+\multicolumn{2}{r}{\exec(pc, s, (\ld, x, v)) = (s',\pc'))}\\
 s', \pc' &: (\st) \in \mathit{rs} \wedge \dec(pc, s) = (\st, a, v) \wedge\\
-\multicolumn{2}{r}{\exec(pc, s, (\st)) = (\pc', s')}\\
+\multicolumn{2}{r}{\exec(pc, s, (\st)) = (s',\pc')}\\
 s, \pc &: \text{otherwise if } \dec(pc,s) \neq H\\
@@
 \inference[LdRpBad]
-{\commit(\rob) = (\pc,\pc',(\ld,x,a,v',s')) \\ v' \neq v & \exec(s,\pc,(\ld,x,v)) = (\pc'', s'')}
+{\commit(\rob) = (\pc,\pc',(\ld,x,a,v',s')) \\ v' \neq v & \exec(s,\pc,(\ld,x,v)) = (s'',\pc'')}
 {\iox{\text{P$_\text{so}$}}{s,\pc,\top,\\\rob,\ppc}{s'',\pc'',\bot,\phi,\\\setBp(\ppc,\pc)}{\text{ToP}(\epsilon, \ld, v)}}
@@
 \exists x. \; \dec(s,\pc) = (\nm, x) \wedge \exec(s,\pc, (\nm,x)) =
-(\pc', s')\\
+(s',\pc')\\
 \commit(\rob) = (\pc, \pc', (\ld, x, a, v, s')) & \Rightarrow
-\dec(s,\pc) = (\ld, x,a) \wedge \exec(s,\pc, (\ld,x,v)) = (\pc', s')\\
+\dec(s,\pc) = (\ld, x,a) \wedge \exec(s,\pc, (\ld,x,v)) = (s',\pc')\\
 \commit(\rob) = (\pc, \pc', (\st, a, v, s')) &\Rightarrow
 \dec(s,\pc) = (\st, a,v) \wedge \exec(s,\pc, (\st)) =
-(\pc', s')\\
+(s',\pc')\\
 \commit(\rob) = H & \Rightarrow

### 2014-11-09 TransitionsUsefulThird1.tex
stuff

@@
 
-We believe that other invalidation-based protocols with inclusive caches (\ie{}
-those cache hierarchies where a parent cache contains all the locations that
-any of its children contains) like MESI, MOSI, and MOESI also satisfy our
-lemmas, making their verification easier by reusing most of our MSI proof; we
-omit the discussion of extension of this proof to other protocols for space
-reasons.
-
 \begin{lemma}
@@
 
-\begin{lemma}
-\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$, neither of which is the
-ancestor of the other in the cache-tree,
-%such
-%that neither is an ancestor of the other in the cache-tree hierarchy (where
-%the ancestor relation is the reflexive-transitive closure of the parent relation),
-coherence states $\s(c_1,a)$ and $\s(c_2,a)$ are
-\textit{compatible} (defined below) for every address $a$.
-\label{nonAncestorCompatible}
-\end{lemma}
-\begin{defn}
-\textit{compatible}: Coherence states $x_1$ and $x_2$ are \textit{compatible}
-iff whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In.
-\label{compatible}
-\end{defn}
+%\begin{lemma}
+%\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$, neither of which is the
+%ancestor of the other in the cache-tree,
+%%such
+%%that neither is an ancestor of the other in the cache-tree hierarchy (where
+%%the ancestor relation is the reflexive-transitive closure of the parent relation),
+%coherence states $\s(c_1,a)$ and $\s(c_2,a)$ are
+%\textit{compatible} (defined below) for every address $a$.
+%\label{nonAncestorCompatible}
+%\end{lemma}
+%\begin{defn}
+%\textit{compatible}: Coherence states $x_1$ and $x_2$ are \textit{compatible}
+%iff whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In.
+%\label{compatible}
+%\end{defn}
 
-\begin{lemma}
-\textit{noTransitWrite}: Whenever data for an address $a$ is in transit (\ie{}
+\begin{lemma}[noTransitWrite] Whenever data for an address $a$ is in transit (\ie{}
 $\forall T. \; T_s \le T \le T_r$ where $T_s$ is the time of sending the data and
@@
 
-\begin{lemma}
-\textit{conservative}: A cache's state for an address is never greater than
-the corresponding state in its parent's directory.
+\begin{lemma}%[Conservative directory] 
+$\forall p, \forall c, \forall a\;\s(c, a) \le \dst(p, c, a)$
 \label{conservative}
@@
 
-\begin{lemma}
-\textit{localCompatibility1}:
-The state of a cache is never less than its directory entry for any of
-its children.
-\label{localCompatibility1}
+\begin{lemma}%[Compatible directory]
+$\forall p, \forall c, \forall a.\; \compat(p, c, \dst(p,c,a), a)$
 \end{lemma}
-\begin{lemma}
-\textit{localCompatibility2}: Let $c_1$ and $c_2$ be distinct caches with the
-same parent $p$. Then, $\s(c_1, a)$ and $\s(c_2, a)$ are
-\textit{compatible}.
-\label{localCompatibility2}
+
+\begin{lemma}%[Coherence state vs Directory state]
+$\forall p, \forall c, \forall a.\; \dst(p,c,a) \le \s(p, a)$
 \end{lemma}
+%\begin{lemma}
+%\textit{localCompatibility1}:
+%The state of a cache is never less than its directory entry for any of
+%its children.
+%\label{localCompatibility1}
+%\end{lemma}
+%\begin{lemma}
+%\textit{localCompatibility2}: Let $c_1$ and $c_2$ be distinct caches with the
+%same parent $p$. Then, $\s(c_1, a)$ and $\s(c_2, a)$ are
+%\textit{compatible}.
+%\label{localCompatibility2}
+%\end{lemma}
+
+The same proof structure can be used to prove other invalidation-based
+protocols with inclusive caches (\ie{} those cache hierarchies in which a
+parent cache contains all the locations that any of its children contains) like
+MESI, MOSI, and MOESI. We omit the discussion of extension of this proof to
+other protocols for space reasons.
 

### 2014-11-09 Abstract.tex
stuff

--- 2014-10-19 Abstract.tex
@@
-We present a new framework for modular verification of hardware designs
-in the style of the Bluespec language.  That is, we formalize the idea of
-components in a hardware design, with well-defined input and output channels;
-and we show how to specify and verify components individually, with
-machine-checked proofs in the Coq proof assistant.  In doing so, we import
-to the world of hardware those modular reasoning principles that are
-well-known in software verification.  As a demonstration, we verify a
-fairly realistic implementation of a multicore shared-memory system.  The
-main modularity split is between a memory-system component and a processor
-component.  Both include nontrivial optimizations, with the memory system
-employing an arbitrary hierarchy of cache nodes that communicate with each
-other concurrently, and with the processor doing speculative execution of
-many concurrent read operations.  Nonetheless, we prove that the combined
-system implements sequential consistency.  One interesting aspect of the proof
-is that the memory component is well-specified to be reused for systems
-implementing a variety of weak memory models.  To our knowledge, our proof
-of that memory system is the first machine verification of a cache-coherence
-protocol parameterized over an arbitrary cache hierarchy, and our full-system
-proof is the first machine verification of sequential consistency for a
-multicore hardware design that includes memory and processors.
+We present a new framework for modular verification of hardware designs in the
+style of the Bluespec language.  That is, we formalize the idea of components
+in a hardware design, with well-defined input and output channels; and we show
+how to specify and verify components individually, with machine-checked proofs
+in the Coq proof assistant.  As a demonstration, we verify a fairly realistic
+implementation of a multicore shared-memory system consisting of two types of
+components:  memory-system and processor.  Both components include nontrivial
+optimizations, with the memory system employing an arbitrary hierarchy of cache
+nodes that communicate with each other concurrently, and with the processor
+doing speculative execution of many concurrent read operations.  Nonetheless,
+we prove that the combined system implements sequential consistency.  One
+interesting aspect of the proof is that the memory component is well-specified
+to be reusable for systems implementing a variety of weak memory models.  To
+our knowledge, our memory-system system proof is the first machine verification
+of a cache-coherence protocol parameterized over an arbitrary cache hierarchy,
+and our full-system proof is the first machine verification of sequential
+consistency for a multicore hardware design that includes caches and
+speculative processors.

### 2014-11-09 Introduction.tex
stuff

@@
 
-A modern high-performance, cache-coherent, distributed-memory hardware system is
-inherently complex. Such a system implements memory models like sequential
-consistency (SC)~\cite{lamport1979make}, Total Store Order (TSO), and
-Relaxed Memory Order (RMO)~\cite{weaver1994sparc}, each
-of which characterizes how loads and stores may be issued by a processor and
-how the shared-memory component responds to these requests.  These systems by
-their nature are highly concurrent and nondeterministic.
-
-The goal of this work is to provide a framework for full verification of such
-complex hardware systems. Hardware designers already recognize the challenge of
-implementing these systems, to such an extent that they have already become some
-of the most serious real-world adopters of formal methods.  Hardware
-verification is dominated by model-checking; for instance, processor
-verification ~\cite{burch1994automatic, mcmillan1998verification} and 
-more recently, Intel's execution cluster verification~\cite{kaivola2009replacing}.
-Many abstraction techniques are
-used to reduce designs to finite state spaces, which can be explored
-exhaustively.
- There are limits to the construction of sound abstractions, so
-verifications of, e.g., cache-coherence protocols have almost always treated
-systems with concrete topologies, involving particular finite numbers of caches
-and processors. For instance, explicit-state model checking tools like
-Murphi~\cite{murphi} or TLC~\cite{tlc} are only able to handle small and
-simplified systems, \eg{} a single-level cache hierarchy with three CPUs, two
-addresses, and one bit of data. Symbolic model checking techniques have fared
-slightly better: McMillan \etal{} have verified a 2-level MSI protocol based on
-the Gigamax distributed multiprocessor using SMV~\cite{gigamax}. Optimizations
-on these techniques (\eg{} partial order reduction~\cite{part}, symmetry
-reduction~\cite{sym1, sym2}, compositional reasoning~\cite{comp, Mccomp, mcc}, \etc{}) scale
-the approach further, but they are still unable to handle realistic
-hierarchical protocols. We might hope to achieve higher assurance and
-understanding of design ideas by verifying \emph{infinite families} of hardware
-designs, which resist reduction to finite-state models.  A potential related
-benefit is reducing the computational resources required to formally verify a
-new design in the family, applying a more general theorem without much new
-analysis.
+A modern high-performance, cache-coherent, distributed-memory hardware system
+is inherently complex. Such a system implements memory models like sequential
+consistency (SC)~\cite{lamport1979make}, Total Store Order (TSO), and Relaxed
+Memory Order (RMO)~\cite{weaver1994sparc}, each of which characterizes how
+loads and stores may be issued by a processor and how the shared-memory
+component responds to these requests.  These systems by their nature are highly
+concurrent and nondeterministic.  The goal of this work is to provide a
+framework for full verification of such complex hardware systems.
 
@@
 of concerns, increasing robustness by allowing the behavior encapsulated by a
-modular boundary to be realized by multiple implementations, any of which may be
-dropped into the system safely. It also allows a greater
-parallelization of human design effort, improving development time. Similarly, verification is
+modular boundary to be realized by multiple implementations, any of which may
+be dropped into the system safely. It also allows a greater parallelization of
+human design effort, improving development time. Similarly, verification is
 simplified, as modular interface agreements provide a natural lemma structure.
-This leads us towards a decomposition of the whole system verification
-task into lemmas of subsystems, which can be composed in a black-box manner to
+This leads us towards a decomposition of the whole system verification task
+into lemmas of subsystems, which can be composed in a black-box manner to
 produce full-system theorems.
 
-%\xxx{\textbf{Murali's version in red} This works introduces an interface enabling true modular implementation and
-%verification of shared memory systems. We want our modular system to permit
-%\emph{black-box refinement}, that is, nothing beyond the interface contract can
-%be assumed about other modules.  That is, we should be able to describe a
-%modular decomposition where we may replace either the processor nodes, the
-%memory with locally verified alternatives without changing the correctness of
-%our whole system.  To this end, we identify a key property called \emph{store
-%atomicity} shared by the memory component of every memory model. This property
-%effectively restricts the behavior of the memory sub-system to that of an
-%\emph{atomic instantaneous memory}. We prove that this property is in fact what
-%is guaranteed using the cache coherence protocols in a multi-processor system
-%containing a hierarchy of caches, enabling replacement of the cache hierarchy
-%with an atomic instantaneous memory. Similarly, on the processor side, we
-%define what is meant to be a correct single processor. These invariants about
-%the processor and the memory sub-system remains constant across memory models.
-%We show a concrete instance of this framework by proving that a highly
-%speculative multi-processor system connected to a coherent cache hierarchy
-%implements the sequential consistency memory model.}
+In this paper, we introduce \emph{a new approach to modular, deductive
+verification of hardware designs} at the highest level of rigor, backed by
+proofs using the Coq proof assistant.  As a challenging case study, we focus on
+\textbf{verifying that a particular infinite family of multicore, shared-memory
+systems implements sequential consistency}.  The proof is parametrized over an
+unknown number of processors connected to an arbitrary memory hierarchy with an
+unknown number of caches in an unknown number of layers (e.g., L1, L2).  In our
+design, processors and memory systems both independently employ intricate
+optimizations that exploit opportunities for parallelism.  We are able to prove
+that each of those two main components still provides strong enough guarantees
+to support SC, and then we compose those theorems into a result for the full
+system.  Either component may be optimized further without requiring any
+changes to the implementation, specification, or proof of the other.
 
-In this paper, we introduce \textbf{a new approach to modular,
-  deductive verification of hardware designs} at the highest level of
-rigor, backed by proofs using the Coq proof assistant.  As a
-challenging case study, we focus on \textbf{verifying that a
-  particular infinite family of multicore, shared-memory systems
-  implements sequential consistency}.  The proof is parametrized over
-an unknown number of processors connected to an arbitrary memory
-hierarchy with an unknown number of caches in an unknown number of
-layers (e.g., L1, L2).  In our design, processors and memory systems
-both independently employ intricate optimizations that exploit
-opportunities for parallelism.  We are able to prove that each of
-those two main components still implements strong enough semantics to
-support SC, and then we compose those theorems into a result for the
-full system.  Either component may be optimized further without
-requiring any changes to the implementation, specification, or proof
-of the other.
-
-To structure our hardware verification framework, we import
-\emph{labeled transition systems (LTSes)}, a well-studied approach from the
-world of concurrent software modeling.  As our unified notion of
-specification and proof, we adopt \emph{trace refinement}, which
-captures when one concurrent system can only produce observable IO
-behaviors that another system could also produce.  Each of our
-realistic hardware components is associated with a simpler ``reference
-implementation'' that serves as a specification, and we prove that
-each realistic component refines its spec.  Thereafter, it is sound to
-use the simpler spec component in reasoning about the behavior of the
+To structure our hardware verification framework, we use \emph{labeled
+transition systems (LTSes)}, a well-studied approach in the world of
+concurrent software modeling.  As our unified notion of specification and
+proof, we adopt \emph{trace refinement}, which captures when one concurrent
+system can only produce observable IO behaviors that another system could also
+produce.  Each of our realistic hardware components is associated with a
+simpler ``reference implementation'' that serves as a specification, and we
+prove that each realistic component refines its spec.  Thereafter, it is sound
+to use the simpler spec component in reasoning about the behavior of the
 system.
 
-Each key component of our case-study system has a simple spec: the
-ideal memory responds instantly and atomically to each load or store
-request, and the ideal processor executes instructions in order with
-no speculation.  The optimized implementation of each is much more
-complicated: the cache-hierarchy memory system potentially allows
-every cache to be busy simultaneously computing and sending messages,
-and the optimized speculative processor sends multiple load requests
-to the memory at once, without waiting for responses first.  The meat
-of the verification is in showing that each optimized component
-refines its simpler counterpart.  After each such proof, we may in
-effect substitute the optimized version for the simple version in a
-black-box way within a design, with a full guarantee of soundness.
+Each key component of our case-study system has a simple spec: the ideal memory
+responds instantly and atomically to each load or store request, and the ideal
+processor executes instructions in order with no speculation.  The optimized
+implementation of each is much more complicated: the cache-hierarchy memory
+system potentially allows every cache to be busy simultaneously computing and
+sending messages, and the optimized speculative processor sends multiple load
+requests to the memory concurrently, without waiting for responses first.  The meat
+of the verification is in showing that each optimized component refines its
+simpler counterpart.  After each such proof, we may in effect substitute the
+optimized version for the simple version in a black-box way within a design,
+with a full guarantee of soundness.
 
 LTSes as hardware descriptions are an established
-idea~\cite{HoeArvind:TRSSynthesis1, Hoe:TCAD}, and there are
-compilers that convert LTSes into efficient hardware.  Our work is
-based on the Bluespec language~\cite{BSV:LangRef, Bluespec:TFRG},
-which served as the model for the formalism of this paper.  Bluespec
-models hardware components as atomic rules of a transition system over
-state elements, and there is a commercial compiler from such code into
-register transfer-level circuits (i.e., Verilog code) with competitive
-performance.  Our cache-coherent memory system is directly
+idea~\cite{HoeArvind:TRSSynthesis1, Hoe:TCAD}, and there are compilers that
+convert LTSes into efficient hardware.  Our work is based on the Bluespec
+language~\cite{BSV:LangRef, Bluespec:TFRG}, which served as the model for the
+formalism of this paper.  Bluespec models hardware components as atomic rules
+of a transition system over state elements, and there is a commercial compiler
+to synthesize such codes into register transfer-level circuits (i.e., Verilog code) with
+competitive performance.  Our cache-coherent memory system is directly
 transliterated from a Bluespec
-implementation~\cite{DNA:CoherenceImplementation} used to
-implement a multi-processor PowerPC system~\cite{Khan:PowerPc}. The
-hardware synthesized from that implementation is rather efficient:
-an 8-core system can run 55 million instructions per second on
-the BEE FPGA board\cite{}, with a 2-level cache hierarchy. Figure \ref{both} gives an example
-of an LTS inference rule transliterated directly from an actual Bluespec rule used
-in the PowerPC system.
+implementation~\cite{DNA:CoherenceImplementation} used to implement a
+cycle-accurate simulator for a multi-processor PowerPC system~\cite{Khan:PowerPc}. The hardware synthesized
+from that implementation is rather efficient: an 8-core system with a 2-level
+cache hierarchy can run 55 million instructions per second on the BEE FPGA
+board\cite{bee2}. Figure \ref{both} gives an example of an LTS inference rule
+transliterated directly from an actual Bluespec rule used in the PowerPC
+system. In Bluespec notation, only those state elements whose value changes are
+mentioned on the RHS of a rule; in
+contrast LTSes employ a more functional notation. (The meaning of this rule
+will become clear in Section \ref{sec:caches})
 
@@
 \end{boxedminipage}
-\caption{Rule in Bluespec and its corresponding LTS transliteration}
+\caption{Bluespec rule and its corresponding LTS rule}
 \label{both}
@@
 \item A general \emph{modular verification methodology} applicable to hardware
-  algorithms, based on \emph{labeled transition systems}.
-We are able to verify
-  the several components of a modern processor 
-%\adam{(Figure~\ref{zoom} OK to have a forward reference to this figure?}
-against general formal interfaces,
+algorithms, based on \emph{labeled transition systems}.  We are able to verify
+the several components of a modern processor against general formal interfaces,
 enabling mixing and matching of different realizations of each interface,
@@
 \item The \emph{first machine-checked proof} of an invalidation-based
-  cache-coherence protocol for distributed memories comprised of arbitrary
-  cache hierarchies. We believe such a memory is also the correct abstraction for
+cache-coherence protocol for distributed memories comprised of arbitrary cache
+hierarchies. We believe such a memory is also the correct abstraction for
 implementing relaxed memory models.
@@
 
-\paragraph{Paper Organization:} In Section~\ref{sec:lts} we introduce our
-flavor of the labeled transition systems formalism, including a definition of
-trace refinement. In Section~\ref{sec:store-atomicity}, we show a generic
+\paragraph{Paper Organization:} We begin with a discussion of related work in
+Section~\ref{relatedWork}. In Section~\ref{sec:lts} we introduce our flavor of
+the labeled transition systems formalism, including a definition of trace
+refinement. In Section~\ref{sec:store-atomicity}, we show a generic
 decomposition of any multi-processor system independently of the memory model
 that it implements, and we discuss the store-atomicity property of the memory
-sub-component. In Section~\ref{sec:sc} we
-give a simple formal model of sequential consistency.  The
-following sections refine the two main subcomponents from Figure~\ref{zoom}.
-Section~\ref{sec:ooo} discusses definition and verification of a speculative
-processor model, and Section~\ref{sec:cc}
-defines and proves our hierarchical cache-coherence protocol.  Finally, in
-Section~\ref{sec:finalresult} we show the whole-system modular proof of our
-complex system, finishing with discussion of related work and other conclusions in
-Sections~\ref{relatedWork} and \ref{sec:conclusion}.
+sub-component. In Section~\ref{sec:sc} we give a simple formal model of
+sequential consistency.  The following sections refine the two main
+subcomponents of our multi-processor system.  Section~\ref{sec:ooo} discusses
+definition and verification of a speculative processor model, and
+Section~\ref{sec:cc} defines and proves our hierarchical cache-coherence
+protocol.  Finally, in Section~\ref{sec:finalresult} we show the whole-system
+modular proof of our complex system, and end with some conclusions in
+Sections~\ref{sec:conclusion}.
 

### 2014-11-09 RelatedWork.tex
stuff

@@
 
-%Formal verification has been used to verify multiprocessor
-%systems, especially with model checking
-%\cite{burch1994automatic,mcmillan1998verification}.
+Hardware verification is dominated by model-checking; for instance, processor
+verification ~\cite{burch1994automatic, mcmillan1998verification} and more
+recently, Intel's execution cluster verification~\cite{kaivola2009replacing}.
+Many abstraction techniques are used to reduce designs to finite state spaces,
+which can be explored exhaustively. There are limits to the construction of
+sound abstractions, so verifications of, e.g., cache-coherence protocols have
+almost always treated systems with concrete topologies, involving particular
+finite numbers of caches and processors. For instance, explicit-state model
+checking tools like Murphi~\cite{murphi} or TLC~\cite{tlc} are only able to
+handle small and simplified systems, \eg{} a single-level cache hierarchy with
+three CPUs, two addresses, and one bit of data. Symbolic model checking
+techniques have fared better: McMillan \etal{} have verified a 2-level MSI
+protocol based on the Gigamax distributed multiprocessor using
+SMV~\cite{gigamax}. Optimizations on these techniques (\eg{} partial order
+reduction~\cite{part}, symmetry reduction~\cite{sym1, sym2}, compositional
+reasoning~\cite{comp, Mccomp, mcc}, \etc{}) scale the approach further, but
+they are still unable to handle realistic hierarchical protocols. We might hope
+to achieve higher assurance and understanding of design ideas by verifying
+\emph{infinite families} of hardware designs, which resist reduction to
+finite-state models.  A potential related benefit is reducing the computational
+resources required to formally verify a new design in the family, applying a
+more general theorem without much new analysis.
 
-%More recently, Intel's execution cluster (\ie{} the ALU and its
-%interface) has been formally verified using model checking
-%\cite{kaivola2009replacing}.
-
-Theorem provers have also been used to verify microprocessors, \eg{} HOL has been
-used to verify an academic microprocessor AVI-1 \cite{windley1995formal}.
+Theorem provers have also been used to verify microprocessors, \eg{} HOL has
+been used to verify an academic microprocessor AVI-1 \cite{windley1995formal}.
 Term-rewriting based techniques have been used to specify microprocessors, with
@@
 
-Theorem provers have also been used in verifying cache coherence protocols.
-The correctness of the FLASH cache coherence protocol~\cite{flash} for a
+Theorem provers have also been used in verifying cache coherence protocols. The
+correctness of the FLASH cache coherence protocol~\cite{flash} for a
 single-level hierarchy was proven with PVS~\cite{park} using a mapping, called
 ``aggregation of transactions,'' from multiple localized transactions to a
-single global transaction, and finally using model checking to prove the correctness of
-the global transactions model. However, they do not provide a methodology to
-prove the aggregation of transactions for other protocols, which is where, we
-believe, the difficulty lies; in contrast, our modular proof relies on generic
-invariants, making it more adaptable to different protocols.
+single global transaction, and finally using model checking to prove the
+correctness of the global transactions model. However, they do not provide a
+methodology to prove the aggregation of transactions for other protocols, which
+is where, we believe, the difficulty lies; in contrast, our modular proof
+relies on generic invariants, making it more adaptable to different protocols.
 Cachet~\cite{StoyShenArvind:Proofs}, another single-level cache coherence
-protocol, was verified to be obey the CRF memory
-model~\cite{Shen:CRF}, and parts of the proof were implemented using PVS.
+protocol, was verified to be obey the CRF memory model~\cite{Shen:CRF}, and
+parts of the proof were implemented using PVS.
 

### 2014-11-09 Abstract.tex
stuff

@@
 in the Coq proof assistant.  As a demonstration, we verify a fairly realistic
-implementation of a multicore shared-memory system consisting of two types of
+implementation of a multicore shared-memory system with two types of
 components:  memory-system and processor.  Both components include nontrivial

### 2014-11-09 Introduction.tex
stuff

@@
 LTSes as hardware descriptions are an established
-idea~\cite{HoeArvind:TRSSynthesis1, Hoe:TCAD}, and there are compilers that
+idea \cite{HoeArvind:TRSSynthesis1, Hoe:TCAD}, and there are compilers that
 convert LTSes into efficient hardware.  Our work is based on the Bluespec
-language~\cite{BSV:LangRef, Bluespec:TFRG}, which served as the model for the
+language \cite{BSV:LangRef, Bluespec:TFRG}, which served as the model for the
 formalism of this paper.  Bluespec models hardware components as atomic rules
@@
 transliterated from a Bluespec
-implementation~\cite{DNA:CoherenceImplementation} used to implement a
-cycle-accurate simulator for a multi-processor PowerPC system~\cite{Khan:PowerPc}. The hardware synthesized
+implementation \cite{DNA:CoherenceImplementation} used to implement a
+cycle-accurate simulator for a multi-processor PowerPC system \cite{Khan:PowerPc}. The hardware synthesized
 from that implementation is rather efficient: an 8-core system with a 2-level
 cache hierarchy can run 55 million instructions per second on the BEE FPGA
-board\cite{bee2}. Figure \ref{both} gives an example of an LTS inference rule
+board \cite{bee2}. Figure \ref{both} gives an example of an LTS inference rule
 transliterated directly from an actual Bluespec rule used in the PowerPC
-system. In Bluespec notation, only those state elements whose value changes are
-mentioned on the RHS of a rule; in
-contrast LTSes employ a more functional notation. (The meaning of this rule
-will become clear in Section \ref{sec:caches})
-
-\begin{figure}[t]
-\centering
-\begin{boxedminipage}{\columnwidth}
-\small
-\noindent \textbf{rule} $(c.\parent == p \; \mathit{\&\&} \; c.\s[a] < x \; \mathit{\&\&} \; c.\wt[a]==$None$);$
-
-\hspace{1cm} $\ch[c][p][\req].\enq(a,c.\s[a],x); c.\wt[a] \Leftarrow x;$
-
-\textbf{endrule}
-\end{boxedminipage}
-\begin{boxedminipage}{\columnwidth}
-\small
-\inference[]
-{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
-{\ioxe{$M_c$}
-{\dt, \ch, \s,\\ \dst, \wt, \dwt,\\ \inp, \outp}
-{\dt, \ch[ (c, p, \req) \coloneqq(a, \s(c, a),x)\\
-\;\;\uplus \ch(c,p,\req)], \s, \dst,\\
-\wt[(c,a)\coloneqq x], \dwt,\inp, \outp}{}}
-\end{boxedminipage}
-\caption{Bluespec rule and its corresponding LTS rule}
-\label{both}
-\end{figure}
+system; its meaning will become clear in Section \ref{sec:cc}. In Bluespec
+notation, only those state elements whose values change are mentioned in the
+rule; in contrast LTSes employ a more functional notation.
 
 The main contributions of this paper are:
-
+\vspace{-.3cm}
 \begin{itemize} 
@@
 Sections~\ref{sec:conclusion}.
+
+\begin{figure}[t]
+\centering
+\begin{boxedminipage}{\columnwidth}
+\small
+\noindent \textbf{rule} $(c.\parent == p \; \mathit{\&\&} \; c.\s[a] < x \; \mathit{\&\&} \; c.\wt[a]==$None$);$
+
+\hspace{1cm} $\ch[c][p][\req].\enq(a,c.\s[a],x); c.\wt[a] \Leftarrow x;$
+
+\textbf{endrule}
+\end{boxedminipage}
+\begin{boxedminipage}{\columnwidth}
+\small
+\inference[]
+{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
+{\ioxe{$M_c$}
+{\dt, \ch, \s,\\ \dst, \wt, \dwt,\\ \inp, \outp}
+{\dt, \ch[ (c, p, \req) \coloneqq(a, \s(c, a),x)\\
+\;\;\uplus \ch(c,p,\req)], \s, \dst,\\
+\wt[(c,a)\coloneqq x], \dwt,\inp, \outp}{}}
+\end{boxedminipage}
+\caption{Bluespec rule and its corresponding LTS rule}
+\label{both}
+\end{figure}
 

### 2014-11-10 Arvind.tex
stuff

--- 2014-10-15 TransitionsUseful.tex
@@
 
-For LTS $A$, we write $\io{A}{s}{s'}{\ell}$ as shorthand for $(s, \ell, s') \in
-A$, and we write $A_0$ for $A$'s initial state. The intuition is that $A$ is
-one process within a concurrent system. The label $\ell$ from set $\mathcal L$
-of labels is produced when $A$ participates in some IO exchange with another
-process; otherwise it is an empty or ``silent'' label $\epsilon$. As a
-shorthand, we sometimes omit labels for $\epsilon$ steps.
+For LTS $A$, we write $\io{A}{s}{s'}{\ell}$ as shorthand for $(s,
+\ell, s') \in A$, and we write $A_0$ for $A$'s initial state. The
+intuition is that $A$ is one process within a concurrent system. The
+label $\ell$ from set $\mathcal L$ of labels is produced when $A$
+participates in some IO exchange with another process; otherwise it is
+an empty or ``silent'' label $\epsilon$.  As a shorthand, we sometimes
+omit labels for $\epsilon$ steps.
 
@@
 
-From an LTS representing single-step system evolution, we often want to build
+From an LTS representing single-step system evolution, we can build %often want to build
 an LTS capturing arbitrary-length evolutions.
@@
 functions keyed on numeric positions, and we write $\theta[i := v]$ for
-updating $n$-tuple $\theta$ to overwrite the $i$th position with value $v$, and we treat any label $(i, \epsilon)$ as $\epsilon$.
+updating $n$-tuple $\theta$ to overwrite the $i$th position with value $v$, and
+we treat any label $(i, \epsilon)$ as $\epsilon$.
+\small
 $$\inference[i]
@@
 \vspace{-.2in}\begin{center}
+\small
 $$\inference[$A$]
@@
 
+%\begin{figure}
+%\small
+%\centering
+%\begin{boxedminipage}[c]{.48\textwidth}
+%%\centering
+%\inference
+%[Halt]
+%{\theta(i) = (s, \pc) & \dec(s,\pc) = H}
+%{\io{SC}{\theta,m}{H}{}}
+%
+%\inference
+%[NonMem]
+%{\theta(i) = (s, \pc) & \dec(s,\pc) = (\nm, x) \\ \exec(s, \pc, (\nm, x)) = (\pc', s')}
+%{\io{SC}{\theta,m}{\theta[i \coloneqq (s', \pc')], m}{}}
+%
+%\inference
+%[Load]
+%{\theta(i) = (s,\pc) & \dec(s,\pc) = (\ld, x, a) \\ \exec(s, \pc, (\ld, x, m(a))) = (\pc', s')}
+%{\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m}{}}
+%
+%\inference
+%[Store]
+%{\theta(i) = (s,\pc) & \dec(s,\pc) = (\st, a, v) \\ \exec(s, \pc, (\st)) = (\pc', s')}
+%{\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m[a\coloneqq v]}{}}
+%\end{boxedminipage}
+%\caption{LTS for sequential consistency with $n$ simple processors}
+%\label{Ref}
+%\end{figure}
+%
+%It is worth noting that the LTS notation described above, along with the
+%composition operators are very similar to the semantics of hardware designs
+%written in Bluespec. As an example consider an inference rule in LTS and the
+%corresponding code in Bluespec in Figure \ref{both}. This code was actually
+%written for the PowerPC system \cite{Khan:PowerPc} (we will be using this rule
+%later in the paper when we describe the coherent caches).
+\section{Decomposing a Share Memory Multi-Processor System}
+\label{sec:store-atomicity}
+
 \begin{figure}
+\centering
+\includegraphics[scale=.43]{zoom}
+\caption{Components of a multi-processors system}
+\label{zoom}
+\end{figure}
+
+Any conventional multi-processor system can be logically divided into 3 components as shown in Figure \ref{zoom}. 
+The top-level system design is shown at top right, while the details of its components, the memory system and the processor ($P_i$), are shown in the magnified boxes.
+The processor component $P_i$ can be implemented in a variety of ways, from one which executes instructions one-by-one in program order, to a complex one which speculatively executes many instructions concurrently to exploit parallelism.
+The memory component is normally implemented using a hierarchy of caches, in order to increase the performance of the overall system, since the latency of accessing memory directly is large compared to that of accessing a much smaller cache.
+The component between each processor and the global memory sub-system contains local buffers, $B_i$, specific to
+the processor $P_i$.
+
+%\begin{figure}
+%\small
+%\centering
+%\begin{boxedminipage}[c]{.48\textwidth}
+%\inference
+%[Load]
+%{}
+%{\io{$M_m$}{m}{m}{p, ((\ld\req,a),(\ld\resp,m(a)), \_)}}
+%
+%\inference
+%[Store]
+%{}
+%{\io{$M_m$}{m}{m[a\coloneqq v]}{p, (\st, a, v)}}
+%
+%\end{boxedminipage}
+%\caption{LTS for a simple memory}
+%\label{M_m}
+%\end{figure}
+
+Popular instruction set architectures, such as Intel x86, ARM, PowerPC, do not guarantee sequential consistency.  That is, their behavior makes it unsound to imagine that
+the different processors of a shared-memory system are running
+according to simple interleaving of instructions, where each instruction executes atomically.
+However, we want to emphasize that, in every weak-memory system we are aware of, \emph{the main memory still exposes atomic loads and stores}!
+The weaker semantics arises only because of (1)
+reordering of memory instructions by the core $P_i$ and/or (2) the
+properties of the local buffers $B_i$ connected to each processor
+$P_i$.  For instance, if $B_i$ were a store buffer, into which the
+committed stores from an in-order processor $P_i$ are enqueued, and if
+any ``subsequent'' load from the same processor $P_i$ is allowed to read and commit the latest value from the store buffer, then such a multi-processor system is said to implement the TSO~\cite{x86tsocacm10} memory
+model.
+
+\begin{figure}
+\small
+\centering
+\begin{boxedminipage}[c]{.48\textwidth}
+\inference
+[Ins]
+{}
+{\iorn{$M_m$}{\inp, \outp, m}{\inp[p \coloneqq \{q\} \uplus \inp(p)], \outp, m}{p}{q}}
+
+\inference
+[Rm]
+{\outp(p) = \mathit{rs} \uplus \{r\}}
+{\ioln{$M_m$}{\inp, \outp, m}{\inp, \outp[p \coloneqq \mathit{rs}], m}{p}{r}}
+
+\inference
+[Ld]
+{\inp(p) = \mathit{qs} \uplus \{t, \ld,a\}}
+{\ioxe{$M_m$}{\inp, \\\outp, m}{\inp[p \coloneqq \mathit{qs}], \outp[p \coloneqq\\
+\;\;\;(t, \ld, m(a)) \uplus \outp(p)], m}{}}
+
+\inference
+[St]
+{\inp(p) = \mathit{qs} \uplus \{\st,a,v\}}
+{\ioxe{$M_m$}{\inp, \\\outp, m}{\inp[p \coloneqq \mathit{qs}], \outp[p \coloneqq
+(\st) \\\;\;\;\uplus \outp(p)], m[a\coloneqq v]}{}}%p, (\st, a,v)}}
+
+\end{boxedminipage}
+\caption{LTS for a simple memory}
+\label{$M_m$}
+\end{figure}
+
+So, we focus on this opportunity to simplify proof decomposition.  We will
+prove that our main memory component satisfies an intuitive \emph{store
+atomicity} property.  Despite the potential red herring of the word
+``atomicity,'' this specification is appropriate even for implementation of
+weaker memory models than SC.  Store atomicity can be understood via the
+operational semantics of Figure \ref{$M_m$}, describing an LTS that receives
+load and store requests ($\ld$ and $\st$) from processors and sends back load
+responses ($\ld\resp$). The transfer happens via input buffers $\inp(p)$ from
+processor $p$ and output buffers $\outp(p)$ to processor $p$.
+
+The memory component is composed of a \emph{hierarchy of caches}, where each
+cache node (labeled like ``L1,'' ``L2,'' etc.) communicates only with its
+neighbors in the graph.  Each processor has a dedicated L1 cache, from which we
+do our best to satisfy all memory requests, to avoid the latency of a
+round-trip with main memory.  With an L1 cache miss, we may still find a hit in
+a parent cache below the level of main memory, realizing a smaller but still
+significant speedup.  Therefore it is the responsibility of the hierarchy of
+caches, which forms the memory sub-component, to implement the store atomicity
+property. In fact, as we will prove in Section \ref{sec:cc}, the purpose of the
+cache-coherence protocol is to establish this invariant for the memory
+sub-component.  Concretely, we have verified a \emph{directory-based} protocol
+for coordinating an arbitrary tree of caches, where each node stores a
+conservative approximation of its children's states.
+
+As an instance of the above decomposition, we will prove that a multi-processor system with no local buffering in between the processor and the memory components indeed implements sequential consistency.
+We implement a highly speculative processor that executes instructions and issues loads out of order, but commits instructions (once some ``verification'' is done) in order.
+
+The processor itself can be decomposed into several components.
+In the zoomed-in version of Figure \ref{zoom}, we show a highly speculative, out-of-order issue processor.
+We have the normal architectural state, such as values of registers.
+Our proofs are generic over \emph{a family of instruction set architectures}, with parameters for opcode sets and functions for executing opcodes and decoding them from memory.
+Other key components are a \emph{branch predictor}, which guesses at the control-flow path that a processor will follow, to facilitate speculation; and a \emph{reorder buffer (ROB)}, which decides which instructions along that path to try executing ahead of schedule.
+Our proofs apply to an arbitrary branch predictor, and they work for any reorder buffer satisfying a simple semantic condition.
+
+Our framework establishes theorems of the form ``if system $A$ has a run with some particular observable behavior, then system $B$ also has a run with the same behavior.''  In this sense, we say that $A$ correctly implements $B$.
+Other important properties, such as \emph{deadlock freedom} for $A$ (which might get stuck without producing any useful behavior), we leave for future work.
+
+\section{Specifying Sequential Consistency}\label{sec:sc}
+
+\begin{figure}
+\small
 \centering
@@
 [Store]
-{\theta(i) = (s,\pc) & \dec(s,\pc) = (\st, a, v) \\ \exec(s, \pc, (\st)) = (\pc', s')}
+{\theta(i) = (s,\pc) & \dec(s,\pc) = (\st, a, v) \\ \exec(s, \pc, (\st)) = (s',\pc')}
 {\io{SC}{\theta,m}{\theta[i\coloneqq(s',\pc')],m[a\coloneqq v]}{}}
 \end{boxedminipage}
-\caption{LTS for sequential consistency with $n$ simple processors}
+\caption{LTS for SC with $n$ simple processors}
 \label{Ref}
 \end{figure}
-
-\section{Specifying Sequential Consistency}\label{sec:sc}
 
@@
 
+Note that, in this setting, an operational specification like the LTS
+for SC is precisely the right characterization of \emph{full functional
+  correctness} for a hardware design, much as a
+precondition-postcondition pair does that job in a partial-correctness
+Hoare logic.  Our SC LTS fully constrains observable behavior of a
+system to remain consistent with simple interleaving.  Similar
+operational models are possible as top-level specifications for
+systems following weaker memory models, by giving the LTS for the \emph{Local
+Buffer} component and composing the three components simultaneously.
+
 \begin{figure}
+\small
 \centering
@@
 \inference
-[LdReq]
+[LdRq]
 {\dec(s,\pc) = (\ld, x, a)}
 {\ior{P$_\text{ref}$}{s, \pc, \bot}
-{s, \pc, \top}{\ld, a}}
-
-\inference
-[StReq]
+{s, \pc, \top}{\epsilon, \ld, a}}
+
+\inference
+[StRq]
 {\dec(s,\pc) = (\st, a, v)}
@@
 \inference
-[LdRep]
-{\dec(s,\pc) = (\ld, x, a) & \exec(s, \pc, (\ld, x, v)) = (\pc', s')}
+[LdRp]
+{\dec(s,\pc) = (\ld, x, a) & \exec(s, \pc, (\ld, x, v)) = (s',\pc')}
 {\iol{P$_\text{ref}$}{s, \pc, \top}
-{s', \pc', \bot}{\ld, v}}
-
-\inference
-[StRep]
-{\dec(s,\pc) = (\st, a, v) & \exec(s, \pc, (\st)) = (\pc', s')}
+{s', \pc', \bot}{\epsilon, \ld, v}}
+
+\inference
+[StRp]
+{\dec(s,\pc) = (\st, a, v) & \exec(s, \pc, (\st)) = (s',\pc')}
 {\iol{P$_\text{ref}$}{s, \pc, \top}
@@
 \end{figure}
-
-\begin{figure*}
-\centering
-\begin{boxedminipage}[c]{.48\textwidth}
-\inference
-[Insert]
-{}
-{\iorn{$M_m$}{\inp, \outp, m}{\inp[i \coloneqq \{q\} \uplus \inp(i)], \outp, m}{i}{q}}
-
-\inference
-[Remove]
-{\outp(i) = \mathit{rs} \uplus \{r\}}
-{\ioln{$M_m$}{\inp, \outp, m}{\inp, \outp[i \coloneqq \mathit{rs}], m}{i}{r}}
-
-\inference
-[SpLoad]
-{\inp(i) = \mathit{qs} \uplus \{\spc\ld,t,a\}}
-{\io{$M_m$}{\inp, \outp, m}{i, \inp[i \coloneqq \mathit{qs}], \outp[i \coloneqq
-(\spc\ld, t, m(a)) \uplus \outp(i)], m}{}}
-
-\inference
-[Load]
-{\inp(i) = \mathit{qs} \uplus \{\ld,a\}}
-{\io{$M_m$}{\inp, \outp, m}{i, \inp[i \coloneqq \mathit{qs}], \outp[i \coloneqq
-(\ld, m(a)) \uplus \outp(i)], m}{}}
-
-\inference
-[Store]
-{\inp(i) = \mathit{qs} \uplus \{\st,a,v\}}
-{\io{$M_m$}{\inp, \outp, m}{i, \inp[i \coloneqq \mathit{qs}], \outp[i \coloneqq
-(\st) \uplus \outp(i)], m[a\coloneqq v]}{}}%i, (\st, a,v)}}
-
-\end{boxedminipage}
-\caption{LTS for a simple memory}
-\label{$M_m$}
-\end{figure*}
 
@@
 
-Figure~\ref{$M_m$} defines a simple memory, meant to be composed with P$_{\text{ref}}$
-processors.  The input-output interface of this memory $M_m$ is largely what
-one would expect from the P$_{\text{ref}}$ definition.  However, in anticipation of the
-more complex processor we will define later, we add requests and responses for
-\emph{speculative loads}.  These are load operations that a processor may
-attempt, before it knows that the program will actually need the result.  Many
-speculative loads from a single processor may be in flight at once, improving
-performance via parallelism.  A request to memory like $(\spc\ld, t, a)$ asks
+The simple memory defined earlier in Figure~\ref{$M_m$} is meant to be composed with P$_{\text{ref}}$
+processors.  %The input-output interface of this memory $M_m$ is largely what
+%one would expect from the P$_{\text{ref}}$ definition. 
+%However, in anticipation of the
+%more complex processor we will define later, we add requests and responses for
+%\emph{speculative loads}.  These are load operations that a processor may
+a%ttempt, before it knows that the program will actually need the result.  Many
+s%peculative loads from a single processor may be in flight at once, improving
+p%erformance via parallelism. 
+A request to memory like $(t, \ld, a)$ asks
 for the value of memory cell $a$, associating a \emph{tag} $t$ that the
 processor can use to match responses to requests.  Those responses take the
-form $(\spc\ld, t, v)$, giving the value $v$ of the requested memory address.
+form $(t, \ld, v)$, giving the value $v$ of the requested memory address.
 
 A memory state is a triple $(\inp, \outp, m)$, giving not just the
-memory $m$ itself, but also queues $\inp$ and $\outp$ for receiving
+memory $m$ itself, but also buffers $\inp$ and $\outp$ for receiving
 processor requests and batching up responses to processors,
@@
 
-Note that we have used $\uplus$ for inserting (or removing) from the buffers
-$\inp$ and $\outp$. The intent is to convey that $\inp$ and $\outp$ are treated
-as unordered sets. We will later use $\coloncolon$ for inserting (or removing)
-from FIFO buffers. Intuitively, the unordered requests/responses set suffices
-even for requests/responses for the same address because the processor always
-waits for a response before sending the next request.
+%Note that we have used $\uplus$ for inserting (or removing) from the buffers
+%$\inp$ and $\outp$. The intent is to convey that $\inp$ and $\outp$ are treated
+%as unordered sets. We will later use $\coloncolon$ for inserting (or removing)
+%from FIFO buffers. Intuitively, the unordered requests/responses set suffices
+%even for requests/responses for the same address because the processor always
+%waits for a response before sending the next request.
 %\medskip
 
-Now we can compose these LTSes to produce an implementation of SC.  Rule SpLoad
-in Figure~\ref{$M_m$} will be irrelevant in this particular composition, since
-P$_{\text{ref}}$ never does speculation.
+Now we can compose these LTSes to produce an implementation of SC. 
+% Rule SpLoad
+%in Figure~\ref{$M_m$} will be irrelevant in this particular composition, since
+%P$_{\text{ref}}$ never does speculation.
 
@@
 
-\begin{figure*}[t]
-\begin{displaymath}
-\upd((s, \pc, \_), \mathit{rs}) =
-\left\{
+\begin{figure}[h]
+\begin{boxedminipage}{\columnwidth}
+\small
+$\upd((s, \pc, \_), \mathit{rs}) =$
+
+\begin{math}
+\;\;\;\;\;\left\{
 \begin{array}{ll}
-s', \pc' &: (\ld, v) \in \mathit{rs} \wedge \dec(pc, s) = (\ld, x, a) \wedge
-\exec(pc, s, (\ld, x, v)) = (\pc', s'))\\
-s', \pc' &: (\st) \in \mathit{rs} \wedge \dec(pc, s) = (\st, a, v) \wedge
-\exec(pc, s, (\st)) = (\pc', s')\\
+s', \pc' &: (\ld, v) \in \mathit{rs} \wedge \dec(s, \pc) = (\ld, x, a) \wedge\\
+\multicolumn{2}{r}{\exec(s, \pc, (\ld, x, v)) = (s',\pc'))}\\
+s', \pc' &: (\st) \in \mathit{rs} \wedge \dec(s, \pc) = (\st, a, v) \wedge\\
+\multicolumn{2}{r}{\exec(s, \pc, (\st)) = (s',\pc')}\\
 s, \pc &: \text{otherwise if } \dec(pc,s) \neq H\\
@@
 \right.
-\end{displaymath}
+\end{math}
+\end{boxedminipage}
 \caption{Mapping of states from $(\text{P$^n_\text{ref}$} + M_m)$ to SC}
 \label{smap}
-\end{figure*}
+\end{figure}
 
@@
 \end{theorem}
-\begin{proof}
-\vspace{-.1cm}
+\begin{prf}
 By induction on traces of the decoupled system.  We need to choose an
@@
 (projected) state.
-\end{proof}
+\end{prf}
+
+\paragraph{The final theorem and the proof plan:} 
+In this section we divided a multiprocessor system into two rather simple
+components, processor represented by $P_\text{ref}$ and memory represented by
+$M_m$. In the following two sections we will develop more complex, optimized
+versions of each piece and prove their correctness, that is, $P_\text{so}$, the
+speculative out-of-order processor, is sound with respect to $P_\text{ref}$ and
+$M_c$, the cache memory, is sound with respect to $M_m$. In proving the
+soundness of $P_\text{so}$ with respect to $P_\text{ref}$, we need to focus
+only on non-speculative communications. The main theorem we are after is to
+show that P$^n_\text{so}+ M_c$ is SC, by showing that it is sound with respect
+to P$^n_\text{ref}+ M_m$. Figure~\ref{proofs} gives the overall proof structure
+that we will employ.
+
+\begin{figure}
+\includegraphics[scale=.45]{proofs}
+\caption{Overall proof structure}
+\label{proofs}
+\end{figure}
 
 \section{Speculative Out-Of-Order Processor}\label{sec:ooo}
-
-The previous section divided a multiprocessor system into two rather
-simple components, processors and memories.  We are now ready to begin
-developing more complex, optimized versions of each piece, starting
-with processors.
 
@@
 
-\inference[SpecLdReq]
+\inference[SpLdRq]
 {\compute(\rob) = (\rob',(\spc\ld,t,a))}
 {\ior{\text{P$_\text{so}$}}{s,\pc,\wait,\rob,\ppc}{s,\pc,\wait,\rob',\ppc}
-{\spc\ld, t, a}}
-
-\inference[SpecLdRep]
-{}
-{\iol{\text{P$_\text{so}$}}{s,\pc,\wait,\rob,\ppc}
-{s,\pc,\wait,\upd\ld(\rob, t, v),\ppc
-}{\spc\ld,t,v}}
+{t,\ld, a}}
+
+\inference[SpLdRp]
+{t \neq \epsilon}
+{\iox{\text{P$_\text{so}$}}{s,\pc,\wait,\\\rob,\ppc}
+{s,\pc,\wait,\\\upd\ld(\rob, t, v),\ppc
+}{\text{ToP}(t,\ld,v)}}
 
@@
 
-\inference[LdReq]
+\inference[LdRq]
 {\commit(\rob) = (\pc,\pc',(\ld,x,a,v,s'))}
-{\ior{\text{P$_\text{so}$}}{s,\pc,\bot,\rob,\ppc}{s,\pc,\top,\rob,\ppc}{\ld, a}}
-\inference[StRep]
+{\ior{\text{P$_\text{so}$}}{s,\pc,\bot,\rob,\ppc}{s,\pc,\top,\rob,\ppc}{\epsilon, \ld, a}}
+\inference[StRp]
 {\commit(\rob) = (\pc,\pc',(\st,a,v,s'))}
@@
 
-\inference[LdRepGood]
+\inference[LdRpGood]
 {\commit(\rob) = (\pc,\pc',(\ld,x,a,v,s'))}
-{\iol{\text{P$_\text{so}$}}{s,\pc,\top,\rob,\ppc}{s',\pc',\bot,\retire(\rob),\ppc}{\ld,v}}
-
-\inference[LdRepBad]
-{\commit(\rob) = (\pc,\pc',(\ld,x,a,v',s')) & v' \neq v & \exec(s,\pc,(\ld,x,v)) = (\pc'', s'')}
-{\iol{\text{P$_\text{so}$}}{s,\pc,\top,\rob,\ppc}{s'',\pc'',\bot,\phi,\setBp(\ppc,\pc)}{\ld, v}}
+{\iox{\text{P$_\text{so}$}}{s,\pc,\top,\\\rob,\ppc}{s',\pc',\bot,\\\retire(\rob),\ppc}{\text{ToP}(\epsilon, \ld,v)}}
+
+\inference[LdRpBad]
+{\commit(\rob) = (\pc,\pc',(\ld,x,a,v',s')) \\ v' \neq v & \exec(s,\pc,(\ld,x,v)) = (s'',\pc'')}
+{\iox{\text{P$_\text{so}$}}{s,\pc,\top,\\\rob,\ppc}{s'',\pc'',\bot,\phi,\\\setBp(\ppc,\pc)}{\text{ToP}(\epsilon, \ld, v)}}
 
@@
 \begin{inv}
-If $P_\text{so}$ reaches a state $(s, \pc, \wait, \rob, \ppc)$, \ie{}
-
+If $P_\text{so}$ reaches a state $(s, \pc, \wait, \rob, \ppc)$, \ie{}\hspace{.2cm}
 $\io{P$_\text{so}^{*}$}{s_0,\pc_0,\bot,\phi,\ppc_0}{s,\pc,\wait,\rob,\ppc}{\eta} \Rightarrow$
-\begin{displaymath}
-\left\{
+
+\begin{math}
+\small
+\;\;\;\;\;\;\;\;\left\{
 \begin{array}{ll}
@@
 \end{corollary}
-\begin{proof}
-Direct consequence of Theorems~\ref{ocorrect} and~\ref{liftn} (where the
+\begin{prf}
+Direct consequence of Theorems~\ref{ocorrect},~\ref{liftn} (the
 latter is one of our preliminary results about $n$-repetition).
-\end{proof}
+\end{prf}
 
@@
 (Figure~\ref{$M_m$}). The rest are summarized in Figure~\ref{fields}.
+
+%We describe a state of the system using 6 fields: $\dt$, $\ch$, $\s$, $\dst$,
+%$\wt$, $\dwt$ which are summarized in Figure~\ref{fields}.
 We use $\parent(c,p)$ to denote that $p$ is the parent of $c$ and $\leaf(c)$ to 
@@
 
-\begin{figure}
+\begin{figure}[h]
 \centering
-\begin{tabular}{|l|p{5.5cm}|}
+\begin{tabular}{|l|p{6cm}|}
 \hline
@@
 $\dst(p,c,a)$ & Directory state of cache $p$ for child $c$ and address $a$\\
-$\wt(c,a)$ & Gives cache $c$'s desired coherence state to upgrade to (for address $a$), or $\epsilon$ if none\\
-$\dwt(p,c,a)$ & Gives cache $p$'s desired coherence state for its child $c$ to downgrade to (for address $a$), or $\epsilon$ if none\\
+$\wt(c,a)$ & State that cache $c$ is waiting to upgrade to, for address $a$, or $\epsilon$\\ %Gives cache $c$'s desired coherence state to upgrade to (for address $a$), or $\epsilon$ if none\\
+$\dwt(p,c,a)$ & State that cache $p$ is waiting its child $c$ to downgrade to, for address $a$, or $\epsilon$\\
+%$\dwt(p,c,a)$ & Gives cache $p$'s desired coherence state for its child $c$ to downgrade to (for address $a$), or $\epsilon$ if none\\
 $\ch$ & Channels between nodes in the system\\
@@
 
-Fields $\dt, \s$ are maps over pairs of cache names and memory addresses.
-The $\dt$ map assigns each pair a memory-cell value, and the $\s$ map
-assigns each pair a coherence state.  A coherence state is $M, S$ or
-$I$, broadly representing permissions to modify, read, or do nothing
-with an address, respectively, the decreasing permissions denoted by $M > S > I$. This interpretation of permissions is
-accurate at the leaf caches connected to the processors (\ie{} L1 caches),
-but the interpretation is subtly different at the higher level caches.
-
-Mapping $\wt$ goes from cache name and address to either $\epsilon$ or a
-coherence state.  It denotes what permission the cache is waiting for,
-if any.  That is, a cache may have decided to \emph{upgrade} its
-coherence state for some address to a more permissive value, but the
-cache is waiting for acknowledgment from another cache before
+A coherence state is $M, S$ or $I$, broadly representing permissions to modify,
+read, or do nothing with an address, respectively, the decreasing permissions
+denoted by $M > S > I$. This interpretation of permissions is accurate at the
+leaf caches connected to the processors, but the
+interpretation is subtly different at the higher level caches.
+
+$\wt$ denotes what permission the cache is waiting for,
+if any.  That is, a cache has decided to \emph{upgrade} its
+coherence state for some address to a more permissive value, but
+it is waiting for acknowledgment from its parent before
 finalizing the upgrade.
 
-Since we have a directory-based protocol, the $\s$ and
-$\wt$ maps have mirror-image counterparts $\dst$ and $\dwt$.
-The former maps triples of parent name, child name, and address to
-coherence state. It represents the parent's notion of the
+$\dst$ represents the parent's notion of the
 coherence state of the child for that address. We later prove that this notion
@@
 
-All communication channels in the system are represented by field $\ch$.
-The channels are asymmetric as
-follows. Channels going from parent to children (indexed by the names of the
-parent and child) maintain FIFO order of messages between the same parent-child
-pair. They carry both request messages (asking the child to downgrade) and
-response messages (acknowledging that the child can upgrade). Channels going
-from children to their parents are further split into request and response
-channels, thus indexed by tuples (child name, parent name, request or response).
-These channels need not maintain the FIFO order for messages between the same
-child-parent pair (notice the use of $\coloncolon$ to add and remove elements
-from the parent to child channel, as opposed to $\uplus$ for the 
-child to parent channel). The reason is that only one request or one response is
-sent for a particular address between a child-parent pair -- this invariant can
-be proved for the LTS in  Figure \ref{cache}. The order between a request and
-response message sent from the same child to its parent is lost, but our
-protocol ensures that nothing bad happens as a result.  Notice rules
-ParentRecvRsp and ParentRecvReq in Figure \ref{cache}, each of which looks at the
-previous coherence state of the incoming message and compares it with its own
-directory state to decide whether to process the message.
-
-%\medskip
+All communication channels in the system are represented by field $\ch$.  The
+channels are asymmetric as follows. Channels going from parent to children
+(indexed by the names of the parent and child) maintain FIFO order of messages
+between the same parent-child pair. They carry both request messages (asking
+the child to downgrade) and response messages (acknowledging that the child can
+upgrade). Channels going from children to their parents are further split into
+request and response channels, thus indexed by tuples (child name, parent name,
+request or response).  These channels need not maintain the FIFO order for
+messages between the same child-parent pair (notice the use of $\coloncolon$ to
+add and remove elements from the parent to child channel, as opposed to
+$\uplus$ for the child to parent channel). The reason is that only one request
+or one response is sent for a particular address between a child-parent pair --
+this invariant can be proved for the LTS in  Figure \ref{cache}. The order
+between a request and response message sent from the same child to its parent
+is lost, but our protocol ensures that they are processed in order by the parent, by
+comparing the previous coherence state of the incoming messages with the
+parent's directory state (in rules ParentRecvRsp
+and ParentRecvReq).
 
@@
 original request can be responded to. The transitions for sending requests are
-allowed to happen spontaneously without any trigger. An L1 cache can process a
-load request or a store request only if it is in at least the \Sh{} state or
-\Mo{} state, respectively, for the requested address; otherwise it must
-spontaneously request to upgrade.
+allowed to happen spontaneously without any trigger. $P_\text{ref}$ issues speculative
+loads only for addresses which are in \Sh{} or \Mo{} state in L1; otherwise L1
+spontaneously requests an upgrade for that address.
 
 \begin{defn}
-\begin{displaymath}
-\compat(p, c, x, a) = \left\{
+$\compat(p, c, x, a) =$
+%\vspace{.3cm}
+
+\begin{math}
+\left\{
 \begin{array}{ll}
@@
 
-%Dropping the request
-%removes the entry from the buffer, allowing the child to process further messages
-%from the parent -- otherwise there would have been a deadlock.
-
-\begin{figure*}
-\centering
-
-\begin{boxedminipage}[c]{.95\textwidth}
-
-\textbf{Processor/Memory Interface}
-
-\inference
-[Insert]
-{}
-{\iorn{$M_c$}{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}{\dt, \ch,
-    \s, \dst, \wt, \dwt, \inp[i \coloneqq \{q\} \uplus \inp(i)],
-    \outp}{i}{q}}
-\vspace{.1in}
-\inference
-[Remove]
-{\outp(i) = \mathit{rs} \uplus \{r\}}
-{\ioln{$M_c$}{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp[i \coloneqq \mathit{rs}]}{i}{r}}
-\vspace{.1in}
-\inference[SpecLoad]
-{\inp(c) = \{(\spc\ld, t, a)\} \uplus \rs & \s(c,a) \geq S}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
-{
-\dt, \ch, \s, \dst, \wt, \dwt, \inp[c \coloneqq \rs],\\
-\outp[c \coloneqq \outp(c) \uplus \{(\spc\ld, t, \dt(c, a))\}]
-}}
-\vspace{.1in}
-\inference[Load]
-{\inp(c) = \{(\ld, a)\} \uplus \rs & \s(c,a) \geq S}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp[c \coloneqq \rs], \\
-\outp[c \coloneqq \outp(c) \uplus \{(\ld, \dt(c, a))\}]}}
-\vspace{.1in}
-\inference[Store]
-{\inp(c) = \{(\st, a, v)\} \uplus \rs & \s(c,a) \geq M}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
-{\dt[(c, a) \coloneqq v], \ch, \s, \dst, \wt, \dwt, \inp[c \coloneqq \rs],
-  \\
-\outp[c \coloneqq \outp(c) \uplus \{(\st)\}]}}
-
-\textbf{Child Upgrade}
-
-\inference[ChildSendReq]
-{\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt, \inp, \outp}
-{\dt, \ch[ (c, p, \req) \coloneqq (a, \s(c, a), x) \uplus \ch(c,p,\req)], \s, \dst,\\
-\wt[(c,a)\coloneqq x], \dwt, \inp, \outp}}
-\vspace{.1in}
-\inference[ParentRecvReq]
-{\parent(c,p) & \ch(c, p, \req) = \{(a, y, x)\} \uplus \rs & \s(p, a) \geq x
-\\ \compat(p, c, x, a) & \dwt(p,c,a) = \epsilon & \dst(p,c,a) \le y}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt, \inp, \outp}
-{\dt, \ch[(c,p,\req) \coloneqq \rs][(p, c) \coloneqq (\resp, (a, \dst(p,c,a), x,\\ \;\;\;\;
-\ite{\dst(p,c,a) = I}{\dt(p,a)}{\_})) \coloncolon \ch(p,c)], \\ \s, \dst[(p,c, a)\coloneqq x],
-\wt, \dwt, \inp, \outp}}
-\vspace{.1in}
-\inference[ChildRecvRsp]
-{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\resp, (a, y, x, v))}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt, \inp, \outp}
-{\dt[(c,a) \coloneqq \ite{y = I}{v}{\dt(c,a)}], \ch[(p,c) \coloneqq \rs], \s[(c,a) \coloneqq x], \dst,\\
-\wt[(c,a)\coloneqq \ite{\wt(c,a) \le x}{\epsilon}{\wt(c,a)}], \dwt, \inp, \outp}}
-
-\textbf{Parent Downgrade}
-
-\inference[ParentSendReq]
-{\parent(c,p) & \dst(p,c,a) > x & \dwt(p,c,a) = \epsilon}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt, \inp, \outp}
-{\dt, \ch[(p, c)\coloneqq (\req, (a, \dst(p, c, a), x)) \coloncolon \ch(p,c)], \s, \dst,\\
-\wt, \dwt[(p, c,a)\coloneqq x], \inp, \outp}}
-\vspace{.1in}
-\inference[ChildRecvReq]
-{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\req, (a, y, x)) & (\forall i. \; \parent(i, c) \Rightarrow \dst(c, i, a) \le x)
-& \s(c,a) > x}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt, \inp, \outp}
-{\dt, \ch[(p,c) \coloneqq \rs][(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
-\ite{\dst(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
-\wt, \dwt, \inp, \outp}}
-\vspace{.1in}
-\inference[ParentRecvRsp]
-{\parent(c,p) & \ch(c,p,\resp) = \{(a, y, x, v)\} \uplus \rs & \dst(p,c,a) = y}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \\ \dwt, \inp, \outp}
-{\dt[(p,a) \coloneqq \ite{y = M}{v}{\dt(p,a)}], \ch[(c,p,\resp) \coloneqq \rs], \s, \dst[(p,c,a) \coloneqq x],\\
-\wt, \dwt[(p,c,a)\coloneqq \ite{\dwt(p,c,a) \ge x}{\epsilon}{\dwt(p,c,a)}], \inp, \outp}}
-
-\textbf{Voluntary downgrade for replacement}
-
-\inference[VolResp]
-{\parent(c,p) & (\forall i. \; \parent(i, c) \Rightarrow \dst(c, i, a) \le x)
-& \s(c,a) > x}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst,\\ \wt, \dwt, \inp, \outp}
-{\dt, \ch[(c, p,\resp) \coloneqq (a, \s(c,a), x,\\ \;\;\;\;
-\ite{\s(c,a) = M}{\dt(c,a)}{\_}) \uplus \ch(c,p,\resp)], \\ \s[(c,a)\coloneqq x], \dst,
-\wt, \dwt, \inp, \outp}}
-
-\textbf{Dropping request because of voluntary downgrade}
-
-\inference[DropReq]
-{\parent(c,p) & \ch(p, c) = \rs \coloncolon (\req, (a, y, x)) & 
- \s(c,a) \le x}
-{\ioxe{$M_c$}
-{\dt, \ch, \s, \dst, \wt, \dwt, \inp, \outp}
-{\dt, \ch[(p,c) \coloneqq \rs], \s, \dst,
-\wt, \dwt, \inp, \outp}}
-\end{boxedminipage}
-
-\caption{LTS for cache-coherent shared-memory system}
-\label{cache}
-\end{figure*}
-
-\section{Proof that the MSI Protocol Preserves Store Atomicity}\label{sec:ccproof}
+\input{TransitionsUsefulCache1}
+
+\subsection{Proving $M_c$ is Store Atomicity}\label{sec:ccproof}
 \label{safety}
@@
 \begin{defn}
-Store atomicity is defined as the following property:
-For a load (or speculative load) request $\toM(\ld, a)$ (or $\toM(\spc\ld, t, a)$)
-from any processor, the response sent at time $T$
-$\toP(\ld, v)$ (or $\toP(\spc\ld, t, v)$) should be such that
+\textit{Store atomicity}:
+For a load request $\toM(t, \ld, a)$
+from any processor, the response sent at time $T$,
+$\toP(t, \ld, v)$ should be such that
 \begin{enumerate}
@@
 
-We can easily show that any system obeying store atomicity
-is sound with respect to $M_m$. Intuitively, the store atomicity
-property precisely describes the property of a memory, \viz{} that if a store
-happens, then that location gets updated such that any subsequent load gets
-back the updated value.
-
-The proof that $M_c$ obeys store atomicity is involved enough to deserve a
-section's worth of explanation, which we now turn to.
-
-We first reduce the proof of store atomicity to several lemmas. We then
-present a key insight that we discovered for proving these lemmas in the form
-of three fundamental invariants. We believe that other invalidation-based
-protocols with inclusive caches (\ie{} those cache hierarchies where a parent
-cache contains all the locations that any of its children contains) like MESI,
-MOSI, and MOESI also satisfy our invariants and lemmas, making their verification
-easier by reusing most of our MSI proof; we omit the discussion of extension of
-this proof to other protocols for space reasons.
-
-%\adamc{...and maybe we don't, for space reasons. \texttt{:-)}}
-
-%We use the terms write to mean handling a store request, \ie{} executing the
-%StoreReq transition and read to mean handling a load request, \ie{} executing
-%the LoadReq transition.
-
-The lemmas and invariants presented below hold after any number of transitions
-starting from the initial state.
-
-While we have proofs in Coq for each lemma, our discussion will be mainly
-informal for brevity. All the proofs can be found in our supplementary material.
-%\adamc{or just remind of the source code URL given earlier}
-
-%\subsection{Invariants}
-%\label{firstLevel}
-Intuitively our protocol is correct because it is guaranteed that any cache that attempts to
-handle a load request has the ``latest'' value.
+We can easily show that any system obeying store atomicity is sound with
+respect to $M_m$. Intuitively, the store atomicity property precisely describes
+the property of a memory, \viz{} that if a store happens, then that location
+gets updated such that any subsequent load gets back the updated value.
+
+The proof that $M_c$ obeys store atomicity is involved enough that we will only
+state the lemmas used to prove this. The proofs in Coq for each theorem and
+lemma in this section can be found in our supplementary material.
 
@@
 has the required permission, at which point it is guaranteed to have the latest
-value, because it is \textit{clean} (as it has no children).
-
-Lemma~\ref{latestValue} requires further decomposition into more
-lemmas.
-
-\begin{lemma}
-\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$ such
-that neither is an ancestor of the other in the cache-tree hierarchy (where
-the ancestor relation is the reflexive-transitive closure of the parent relation),
-for each address $a$, $\s(c_1,a)$ and $\s(c_2,a)$ are
-\textit{compatible} (defined below).
-\label{nonAncestorCompatible}
-\end{lemma}
-\begin{defn}
-\textit{compatible}: Values of states $x_1$ and $x_2$ are \textit{compatible}
-iff whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In.
-\label{compatible}
-\end{defn}
-
-It is easy to see that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also
-compatible if $x_1$ and $x_2$ are compatible.
-
-\begin{lemma}
-\textit{noTransitWrite}: Whenever data for an address $a$ is in transit (\ie{}
+value, because it is \textit{clean} (as it has no children). To prove Lemma
+\ref{latestValue}, it has to be decomposed futher into the following, each of
+which holds at any time.
+
+%\begin{lemma}
+%\textit{nonAncestorCompatible}: For two distinct caches $c_1$ and $c_2$, neither of which is the
+%ancestor of the other in the cache-tree,
+%%such
+%%that neither is an ancestor of the other in the cache-tree hierarchy (where
+%%the ancestor relation is the reflexive-transitive closure of the parent relation),
+%coherence states $\s(c_1,a)$ and $\s(c_2,a)$ are
+%\textit{compatible} (defined below) for every address $a$.
+%\label{nonAncestorCompatible}
+%\end{lemma}
+%\begin{defn}
+%\textit{compatible}: Coherence states $x_1$ and $x_2$ are \textit{compatible}
+%iff whenever either $x_1$ or $x_2$ is \Mo{}, the other is \In.
+%\label{compatible}
+%\end{defn}
+
+\begin{lemma}[noTransitWrite] Whenever data for an address $a$ is in transit (\ie{}
 $\forall T. \; T_s \le T \le T_r$ where $T_s$ is the time of sending the data and
@@
 
-\begin{lemma}
-\textit{conservative}: The state of an address in a cache is never greater than
-the corresponding entry in its parent's directory.
+\begin{lemma}%[Conservative directory] 
+$\forall p, \forall c, \forall a\;\s(c, a) \le \dst(p, c, a)$
 \label{conservative}
@@
 
-\begin{lemma}
-\textit{localCompatibility1}:
-The state of a cache is never less than its directory entry for any of
-its children.
-\label{localCompatibility1}
+\begin{lemma}%[Compatible directory]
+$\forall p, \forall c, \forall a.\; \compat(p, c, \dst(p,c,a), a)$
 \end{lemma}
-\begin{lemma}
-\textit{localCompatibility2}: Let $c_1$ and $c_2$ be distinct caches with the
-same parent $p$. Then, $\s(c_1, a)$ and $\s(c_2, a)$ are
-\textit{compatible}.
-\label{localCompatibility2}
+
+\begin{lemma}%[Coherence state vs Directory state]
+$\forall p, \forall c, \forall a.\; \dst(p,c,a) \le \s(p, a)$
 \end{lemma}
-
-Lemmas~\ref{conservative} and~\ref{localCompatibility1} imply a
-$\ge$ relation between the states of a parent and its child for any address.
-Since the ancestor relation is the reflexive-transitive closure of the parent
-relation, the $\ge$ relation is preserved between the states of two caches $p$
-and $c$ where $p$ is the ancestor of $c$.
-
-Any two caches $c_1, c_2$ neither of which is an ancestor of the other have a
-lowest common ancestor $x$, which has
-two distinct children $x_1$ and $x_2$ such that they are the ancestors of $c_1$
-and $c_2$ respectively. For any address $a$, $\s(x_1, a)$ and
-$\s(x_2, a)$ are compatible by Lemma~\ref{localCompatibility2}. From
-the previous paragraph about the relation between the states of a cache and its
-ancestor, it follows that $c_1$ and $c_2$ are also compatible, proving
-Lemma~\ref{nonAncestorCompatible}.
-
-Using Lemmas~\ref{nonAncestorCompatible} and~\ref{noTransitWrite}, one can
-prove Lemma~\ref{latestValue}. The proof is by induction on time $T$; we assume
-that any \clean{} (node, address) pair has the latest value until time
-$T$. Consider a (node, addr) pair at time $T+1$.
-If the pair was \clean{} at $T$, then no cache,
-save itself (because of Lemma~\ref{nonAncestorCompatible}), can process any
-store request for that address, thus continuing to be \clean{} and having the latest value.
-And if the pair was not \clean{} at time $T$, it will receive data from a cache
-that is \clean{} for the (node, address) pair, and no store
-request can be processed when the data is in transit (by
-Lemma~\ref{noTransitWrite}). Therefore, by the induction hypothesis the cache will
-continue to have the latest value whenever it is \clean{} for an address at time $T+1$.
-
-\subsection{Key Insight for Proving Lemmas}
-
-We present what we believe to be the key insight for proving any
-cache-coherence protocol.  In fact, the request/response message semantics for
-the MSI protocol was designed to satisfy the following three invariants.
-
-\begin{inv}
-\textit{stateChange}:
-\begin{enumerate}
-\item A cache downgrades its state for an address iff it sends a response for that address.
-\item A parent upgrades the directory entry for an address for one of its
-children iff it sends a response for that child and address.
-\item The state or directory entry changes only on sending or
-receiving a response $(\resp, (a, y, x, d))$, to $x$.
-\end{enumerate}
-\label{stateChange}
-\end{inv}
-
-\vspace{-.1in}
-
-Invariant~\ref{stateChange} ensures that a response message acts as a sync
-between a parent's directory entry for a child and the state of the child, as
-both the sender and the receiver change their respective directory entry or
-state to the same $x$ in a response $(\resp, (a, y, x, d))$
-
-\begin{inv}
-\textit{noCross}: A response from a parent to its child for an address cannot
-be in transit while another response from the same child to the parent for the
-same address is in transit.
-\label{noCross}
-\end{inv}
-
-\begin{inv}
-\textit{respFIFO}: Responses are received in the order in which they are sent,
-for a source-destination pair, for the same address.
-\label{respFIFO}
-\end{inv}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.45]{resps}
-\caption{An example showing a sequence of responses between a child and its parent}
-\label{resps}
-\end{figure}
-
-Figure~\ref{resps} shows an example of the effect of
-Invariants~\ref{stateChange}, \ref{noCross} and \ref{respFIFO}. The
-directory entry of the parent or the state of the child changes on sending and
-receiving a response, and the new state is shown at the points of sending or
-receiving responses.  Both the cache and its parent follow the same sequence of
-state and directory entry changes, but the receiver always lags slightly behind
-the sender.  The receiver cannot send any response before receiving because of
-Invariant~\ref{noCross}, thereby eliminating any scenarios where the lag
-complicates reasoning.
-
-Using these three invariants (Invariants~\ref{stateChange}, \ref{noCross} and
-\ref{respFIFO}), it is easy to see that Lemma~\ref{conservative} holds.
-Intuitively, going back to Figure~\ref{resps}, we can see that at all times,
-either there is no response in flight, only responses from the parent are in
-flight, or only responses from the child are in flight. A response from child
-downgrades the state, and a response from parent upgrades the directory. Since
-the other side simply lags behind, Lemma~\ref{conservative} holds.
-Since Invariants~\ref{stateChange}, \ref{noCross} and \ref{respFIFO} deal with
-just one cache's state for a single address and the corresponding directory
-entry, these properties can also potentially be verified using model-checking.
-Thus, if some other protocol is verified to obey Invariants~\ref{stateChange},
-\ref{noCross} and \ref{respFIFO} using model-checking, then
-Lemma~\ref{conservative} is already automatically verified for that
-protocol.
-
-Analyzing a parent-child pair will suffice in proving
-Lemmas~\ref{localCompatibility1} and \ref{localCompatibility2}.  Proof of
-Lemma~\ref{noTransitWrite} is similar to the proof of
-Lemma~\ref{nonAncestorCompatible}.
-
-%Figure \ref{something} gives the structure of the proof for Theorem \ref{ccorrect}
-%\begin{figure*}
-%\begin{prooftree}
-%\AxiomC{stateChange}
-%\AxiomC{noCross}
-%\AxiomC{respFifo}
-%\TrinaryInfC{conservative}
-%\AxiomC{localCompatibility1}
-%\AxiomC{localCompatibility2}
-%\TrinaryInfC{nonAncestorCompatible}
-%\AxiomC{noTransitWrite}
-%\BinaryInfC{latestValue}
-%\AxiomC{$\ldots$ (trivial)}
-%\BinaryInfC{StoreAtomicity}
-%\UnaryInfC{$M_c \sqsubseteq M_m$}
-%\end{prooftree}
-%\caption{Proving $M_c \sqsubseteq M_m$}
-%\label{something}
-%\end{figure*}
+%\begin{lemma}
+%\textit{localCompatibility1}:
+%The state of a cache is never less than its directory entry for any of
+%its children.
+%\label{localCompatibility1}
+%\end{lemma}
+%\begin{lemma}
+%\textit{localCompatibility2}: Let $c_1$ and $c_2$ be distinct caches with the
+%same parent $p$. Then, $\s(c_1, a)$ and $\s(c_2, a)$ are
+%\textit{compatible}.
+%\label{localCompatibility2}
+%\end{lemma}
+
+The same proof structure can be used to prove other invalidation-based
+protocols with inclusive caches (\ie{} those cache hierarchies in which a
+parent cache contains all the locations that any of its children contains) like
+MESI, MOSI, and MOESI. We omit the discussion of extension of this proof to
+other protocols for space reasons.
 
@@
 \end{theorem}
-\begin{proof}
+\begin{prf}
 We twice apply $\sqsubseteq$ transitivity (Theorem \ref{transitive}) to
+%connect Theorems \ref{ocorrect2} and \ref{ocorrect1}, and use Definition \ref{sc} of $SC$
 connect Theorems \ref{ocorrect2}, \ref{ocorrect1}, and \ref{scthm}.
-\end{proof}
-
-%Figure \ref{overall} gives a summary the overall proof structure that we have
-%employed to prove the correctness of the full system.
-%
-%\begin{figure*}
-%\begin{prooftree}
-%\AxiomC{$M_c \sqsubseteq M_m$}
-%\AxiomC{$\sqsubseteq$ is reflexive}
-%\AxiomC{$\begin{array}{l} \forall f, A, A', B, B'. A \sqsubseteq_f A' \wedge B \sqsubseteq_f B' \\\;\;\;\;\Rightarrow A+B \sqsubseteq A'+B'(\text{Theorem 3})\end{array}$}
-%\TrinaryInfC{P$_\text{so}^n+M_c \sqsubseteq $ P$_\text{so}^n+M_m$}
-%\end{prooftree}
-%\begin{prooftree}
-%\AxiomC{P$_\text{so} \sqsubseteq_\noSpec $ P$_\text{ref}$}
-%\AxiomC{$\begin{array}{l}\forall f, A, B, n. A \sqsubseteq_f B \Rightarrow \\ \;\;\;\;\;\;\;A^n \sqsubseteq_{f^n} B^n(\text{Theorem 2})\end{array}$}
-%\BinaryInfC{P$_\text{so}^n \sqsubseteq_\noSpec $ P$_\text{ref}^n$}
-%\AxiomC{$M_m \sqsubseteq_\noSpec M_m$}
-%\AxiomC{$\begin{array}{l} \forall f, A, A', B, B'. A \sqsubseteq_f A' \wedge B \sqsubseteq_f B' \\\;\;\;\;\Rightarrow A+B \sqsubseteq A'+B'(\text{Theorem 3})\end{array}$}
-%\TrinaryInfC{P$_\text{so}^n+M_m \sqsubseteq $ P$_\text{ref}^n+M_m$}
-%\end{prooftree}
-%\begin{prooftree}
-%\AxiomC{P$_\text{so}^n+M_c \sqsubseteq $ P$_\text{so}^n+M_m$}
-%\AxiomC{P$_\text{so}^n+M_m \sqsubseteq $ P$_\text{ref}^n+M_m$}
-%\AxiomC{P$_\text{ref}^n + M_m \sqsubseteq SC$}
-%\AxiomC{$\sqsubseteq $ is transitive}
-%\QuaternaryInfC{P$_\text{so}^n + M_c \sqsubseteq SC$}
-%\end{prooftree}
-%\caption{Overall proof structure}
-%\label{overall}
-%\end{figure*}
+\end{prf}

### 2014-11-10 Abstract.tex
Pass up to end of Introduction

@@
 implementation of a multicore shared-memory system with two types of
-components:  memory-system and processor.  Both components include nontrivial
+components: memory system and processor.  Both components include nontrivial
 optimizations, with the memory system employing an arbitrary hierarchy of cache
@@
 to be reusable for systems implementing a variety of weak memory models.  To
-our knowledge, our memory-system system proof is the first machine verification
+our knowledge, our memory-system proof is the first machine verification
 of a cache-coherence protocol parameterized over an arbitrary cache hierarchy,

### 2014-11-10 Introduction.tex
Pass up to end of Introduction

@@
 simplified, as modular interface agreements provide a natural lemma structure.
-This leads us towards a decomposition of the whole system verification task
-into lemmas of subsystems, which can be composed in a black-box manner to
+They lead us towards a decomposition of the whole-system verification task
+into lemmas about subsystems, which can be composed in a black-box manner to
 produce full-system theorems.
@@
 of a transition system over state elements, and there is a commercial compiler
-to synthesize such codes into register transfer-level circuits (i.e., Verilog code) with
+to synthesize such codes into register-transfer-level circuits (i.e., Verilog code) with
 competitive performance.  Our cache-coherent memory system is directly
@@
 implementation \cite{DNA:CoherenceImplementation} used to implement a
-cycle-accurate simulator for a multi-processor PowerPC system \cite{Khan:PowerPc}. The hardware synthesized
+cycle-accurate simulator for a multiprocessor PowerPC system \cite{Khan:PowerPc}. The hardware synthesized
 from that implementation is rather efficient: an 8-core system with a 2-level
@@
 transliterated directly from an actual Bluespec rule used in the PowerPC
-system; its meaning will become clear in Section \ref{sec:cc}. In Bluespec
-notation, only those state elements whose values change are mentioned in the
-rule; in contrast LTSes employ a more functional notation.
+system.  Its meaning will become clear in Section \ref{sec:cc};
+for now we just note that, while Bluespec rules employ Boolean \emph{guard}
+conditions paired with sequences of imperative \emph{actions} to update state,
+our (equivalent) formalization follows a natural-deduction inference-rule style,
+where we explain a state transition as a relation between two tuples of state
+elements, using rule premises where Bluespec uses guards.  The high-level point
+is that the natural-deduction style is a semantically trivial rephrasing of the
+Bluespec notation, which has been demonstrated as practical for synthesizing
+realistic, high-performance hardware.  Inference rules like the one in
+Figure \ref{both} fit naturally with the inductive-definition mechanisms of
+proof systems like Coq, providing an effective starting point for formal
+verification of hardware designs.
 
@@
 \item Identification of an interface between the processor and the memory
-components of a multi-processor system that enables modular implementation and
+components of a multiprocessor system that enables modular implementation and
 verification.
@@
 refinement. In Section~\ref{sec:store-atomicity}, we show a generic
-decomposition of any multi-processor system independently of the memory model
+decomposition of any multiprocessor system independently of the memory model
 that it implements, and we discuss the store-atomicity property of the memory
-sub-component. In Section~\ref{sec:sc} we give a simple formal model of
+subcomponent. In Section~\ref{sec:sc} we give a simple formal model of
 sequential consistency.  The following sections refine the two main
-subcomponents of our multi-processor system.  Section~\ref{sec:ooo} discusses
+subcomponents of our multiprocessor system.  Section~\ref{sec:ooo} discusses
 definition and verification of a speculative processor model, and
@@
 protocol.  Finally, in Section~\ref{sec:finalresult} we show the whole-system
-modular proof of our complex system, and end with some conclusions in
-Sections~\ref{sec:conclusion}.
+modular proof of our complex system, and we end with some conclusions in
+Section~\ref{sec:conclusion}.
 
@@
 \small
-\noindent \textbf{rule} $(c.\parent == p \; \mathit{\&\&} \; c.\s[a] < x \; \mathit{\&\&} \; c.\wt[a]==$None$);$
+\noindent \textbf{rule} $(c.\parent == p \; \mathit{\&\&} \; c.\s[a] < x \; \mathit{\&\&} \; c.\wt[a]== \textrm{None});$
 

### 2014-11-10 RelatedWork.tex
stuff
Pass through RelatedWork

@@
+\vspace{-.4cm}
 \section{Related Work}
@@
 
-Hardware verification is dominated by model-checking; for instance, processor
+Hardware verification is dominated by model checking; for instance, processor
 verification ~\cite{burch1994automatic, mcmillan1998verification} and more
@@
 handle small and simplified systems, \eg{} a single-level cache hierarchy with
-three CPUs, two addresses, and one bit of data. Symbolic model checking
+three CPUs, two addresses, and one bit of data. Symbolic model-checking
 techniques have fared better: McMillan \etal{} have verified a 2-level MSI
@@
 been used to verify an academic microprocessor AVI-1 \cite{windley1995formal}.
-Term-rewriting based techniques have been used to specify microprocessors, with
+Term-rewriting techniques have been used to specify microprocessors, with
 the goal of making them more amenable to verification \cite{shen1999using}.
 
-Theorem provers have also been used in verifying cache coherence protocols. The
-correctness of the FLASH cache coherence protocol~\cite{flash} for a
+Cache-coherence proofs have also been done with mechanized theorem provers. The
+correctness of the FLASH cache-coherence protocol~\cite{flash} for a
 single-level hierarchy was proven with PVS~\cite{park} using a mapping, called
@@
 
-There has been a lot of work in providing frameworks for axiomatic
-specifications of memory models based on empirically observed behavior of
+Before verifying that a hardware design implements an ISA-level semantics, it
+is necessary to decide what the correct semantics is.  Much recent work has
+studied widely deployed hardware platforms empirically to derive formal
+semantics.  While our case study in this paper hews to the uncontroversial
+sequential-consistency model, the empirical challenges arise in characterizing
+weaker memory models.  Approaches have included building axiomatic
+memory-model specifications based on empirically observed behavior on
 certain ``litmus
-tests''\cite{Alglave11,Alglave:FMSD,alglave2012formal,Alglave:2010,Alglave:TACAS,alglave2014herding}.
-There has been some work in providing operational and axiomatic semantics that
-describes the observed behavior (based on certain ``litmus tests'' of real
-hardware
-\cite{mador2012axiomatic,sarkar2011understanding,sarkar2012synchronising}.
-Finally, there has been work formally relating the axiomatic specification of
-TSO to operational semantics \cite{x86tsocacm10}.
+tests''~\cite{Alglave11,Alglave:FMSD,alglave2012formal,Alglave:2010,Alglave:TACAS,alglave2014herding},
+building models that include operational semantics~\cite{mador2012axiomatic,sarkar2011understanding,sarkar2012synchronising}, and relating the axiomatic specification of
+TSO to operational semantics~\cite{x86tsocacm10}.
 

### 2014-11-10 Arvind.tex
stuff
A pass through LTS preliminaries

@@
 iff the reference system could also terminate or diverge,
-respectively.  Our basic definitions of transition systems build in
+respectively.  All sorts of other interesting program properties are
+reducible to this one, in the style of computability theory.
+Our basic definitions of transition systems build in
 special treatment of halting, so that we need not mention it
@@
 labels mapped to $\epsilon$ by $f$ are dropped. $f$ is overloaded
-to denote that multi-label version when applied to $\eta$.
+to denote that multilabel version when applied to $\eta$.
 \end{defn}
@@
 %later in the paper when we describe the coherent caches).
-\section{Decomposing a Share Memory Multi-Processor System}
+\section{Decomposing a Shared-Memory Multiprocessor System}
 \label{sec:store-atomicity}
@@
 
-As an instance of the above decomposition, we will prove that a multi-processor system with no local buffering in between the processor and the memory components indeed implements sequential consistency.
-We implement a highly speculative processor that executes instructions and issues loads out of order, but commits instructions (once some ``verification'' is done) in order.
-
-The processor itself can be decomposed into several components.
-In the zoomed-in version of Figure \ref{zoom}, we show a highly speculative, out-of-order issue processor.
-We have the normal architectural state, such as values of registers.
-Our proofs are generic over \emph{a family of instruction set architectures}, with parameters for opcode sets and functions for executing opcodes and decoding them from memory.
-Other key components are a \emph{branch predictor}, which guesses at the control-flow path that a processor will follow, to facilitate speculation; and a \emph{reorder buffer (ROB)}, which decides which instructions along that path to try executing ahead of schedule.
-Our proofs apply to an arbitrary branch predictor, and they work for any reorder buffer satisfying a simple semantic condition.
-
-Our framework establishes theorems of the form ``if system $A$ has a run with some particular observable behavior, then system $B$ also has a run with the same behavior.''  In this sense, we say that $A$ correctly implements $B$.
-Other important properties, such as \emph{deadlock freedom} for $A$ (which might get stuck without producing any useful behavior), we leave for future work.
+As an instance of the above decomposition, we will prove that a multi-processor
+system with no local buffering in between the processor and the memory
+components indeed implements sequential consistency. We implement a highly
+speculative processor that executes instructions and issues loads out of order,
+but commits instructions (once some ``verification'' is done) in order. This idea
+of verification loads for a speculative processor is not new \cite{cain}.
+
+The processor itself can be decomposed into several components. In the
+zoomed-in version of Figure \ref{zoom}, we show a highly speculative,
+out-of-order issue processor. We have the normal architectural state, such as
+values of registers.  Our proofs are generic over \emph{a family of instruction
+set architectures}, with parameters for opcode sets and functions for executing
+opcodes and decoding them from memory.  Other key components are a \emph{branch
+predictor}, which guesses at the control-flow path that a processor will
+follow, to facilitate speculation; and a \emph{reorder buffer (ROB)}, which
+decides which instructions along that path to try executing ahead of schedule.
+Our proofs apply to an arbitrary branch predictor, and they work for any
+reorder buffer satisfying a simple semantic condition.
+
+Our framework establishes theorems of the form ``if system $A$ has a run with
+some particular observable behavior, then system $B$ also has a run with the
+same behavior.''  In this sense, we say that $A$ correctly implements $B$.
+Other important properties, such as \emph{deadlock freedom} for $A$ (which
+might get stuck without producing any useful behavior), we leave for future
+work.
 

### 2014-11-10 Arvind.tex
Pass through Decomposing

@@
 \includegraphics[scale=.43]{zoom}
-\caption{Components of a multi-processors system}
+\caption{Components of a multiprocessor system}
 \label{zoom}
@@
 
-Any conventional multi-processor system can be logically divided into 3 components as shown in Figure \ref{zoom}. 
+Any conventional multiprocessor system can be divided logically into 3 components as shown in Figure \ref{zoom}. 
 The top-level system design is shown at top right, while the details of its components, the memory system and the processor ($P_i$), are shown in the magnified boxes.
-The processor component $P_i$ can be implemented in a variety of ways, from one which executes instructions one-by-one in program order, to a complex one which speculatively executes many instructions concurrently to exploit parallelism.
+The processor component $P_i$ can be implemented in a variety of ways, from one executing instructions one-by-one in program order, to a complex one speculatively executing many instructions concurrently to exploit parallelism.
 The memory component is normally implemented using a hierarchy of caches, in order to increase the performance of the overall system, since the latency of accessing memory directly is large compared to that of accessing a much smaller cache.
-The component between each processor and the global memory sub-system contains local buffers, $B_i$, specific to
-the processor $P_i$.
+The component between each processor and the global memory subsystem contains local load/store buffers, $B_i$, each specific to a processor $P_i$.
 
@@
 
-Popular instruction set architectures, such as Intel x86, ARM, PowerPC, do not guarantee sequential consistency.  That is, their behavior makes it unsound to imagine that
+Popular instruction set architectures, such as Intel x86, ARM, and PowerPC, do not guarantee sequential consistency.  That is, their behavior makes it unsound to imagine that
 the different processors of a shared-memory system are running
@@
 committed stores from an in-order processor $P_i$ are enqueued, and if
-any ``subsequent'' load from the same processor $P_i$ is allowed to read and commit the latest value from the store buffer, then such a multi-processor system is said to implement the TSO~\cite{x86tsocacm10} memory
+any ``subsequent'' load from the same processor $P_i$ is allowed to read and commit the latest value from the store buffer, then such a multiprocessor system is said to implement the TSO~\cite{x86tsocacm10} memory
 model.
@@
 responses ($\ld\resp$). The transfer happens via input buffers $\inp(p)$ from
-processor $p$ and output buffers $\outp(p)$ to processor $p$.
+processor $p$ and output buffers $\outp(p)$ to processor $p$.  Note that this
+model allows the memory system to answer pending memory requests in any order,
+even potentially reordering requests from a single processor,
+so long as, whenever it does process a request, that action appears to take
+place \emph{atomically}.
 
@@
 significant speedup.  Therefore it is the responsibility of the hierarchy of
-caches, which forms the memory sub-component, to implement the store atomicity
+caches, which forms the memory subcomponent, to implement the store atomicity
 property. In fact, as we will prove in Section \ref{sec:cc}, the purpose of the
 cache-coherence protocol is to establish this invariant for the memory
-sub-component.  Concretely, we have verified a \emph{directory-based} protocol
+subcomponent.  Concretely, we have verified a \emph{directory-based} protocol
 for coordinating an arbitrary tree of caches, where each node stores a
@@
 
-As an instance of the above decomposition, we will prove that a multi-processor
+As an instance of the above decomposition, we will prove that a multiprocessor
 system with no local buffering in between the processor and the memory

### 2014-11-10 Arvind.tex
Pass through Respecifying

@@
 system implements sequential consistency (SC) properly.  We state the theorem
-in terms of the trace refinement relation $\sqsubseteq$ developed in the prior
-section.  Therefore, we need to define an LTS embodying SC.  The simpler this
+in terms of the trace refinement relation $\sqsubseteq$ developed in
+Section~\ref{sec:lts}.
+Therefore, we need to define an LTS embodying SC.  The simpler this
 system, the better.  We do not need to worry about its performance properties,
@@
 operational models are possible as top-level specifications for
-systems following weaker memory models, by giving the LTS for the \emph{Local
-Buffer} component and composing the three components simultaneously.
+systems following weaker memory models, by giving the LTS for the \emph{local
+buffer} component and composing the three components simultaneously.
 
@@
 %\emph{speculative loads}.  These are load operations that a processor may
-a%ttempt, before it knows that the program will actually need the result.  Many
-s%peculative loads from a single processor may be in flight at once, improving
-p%erformance via parallelism. 
+%attempt, before it knows that the program will actually need the result.  Many
+%speculative loads from a single processor may be in flight at once, improving
+%performance via parallelism. 
 A request to memory like $(t, \ld, a)$ asks
@@
 $M_m$. In the following two sections we will develop more complex, optimized
-versions of each piece and prove their correctness, that is, $P_\text{so}$, the
-speculative out-of-order processor, is sound with respect to $P_\text{ref}$ and
-$M_c$, the cache memory, is sound with respect to $M_m$. In proving the
+versions of each piece and prove their correctness.  That is, we define $P_\text{so}$, the
+speculative out-of-order processor, and prove it sound with respect to $P_\text{ref}$;
+and we define $M_c$, the cache memory, and prove it sound with respect to $M_m$. In proving the
 soundness of $P_\text{so}$ with respect to $P_\text{ref}$, we need to focus
-only on non-speculative communications. The main theorem we are after is to
+only on nonspeculative communications. The main theorem we are after is to
 show that P$^n_\text{so}+ M_c$ is SC, by showing that it is sound with respect

### 2014-11-10 Arvind.tex
Pass through speculation

@@
 The operations on such state are $\curPpc(\ppc)$, to extract the
-current program-counter prediction; $\nextPpc(\ppc)$, to get the predicted
-program-counter for the next instruction; and $\setBp(\ppc, \pc)$,
+current program-counter prediction; $\nextPpc(\ppc)$, to advance to predicting
+the next instruction; and $\setBp(\ppc, \pc)$,
 to reset prediction to begin at a known-accurate position $\pc$. We need to impose no explicit correctness criterion on
@@
 The interface and formal contract of a reorder buffer are more
-involved.  We write $\rob$ as a metavariable for reorder buffer
+involved.  We write $\rob$ as a metavariable for reorder-buffer
 state. $\phi$ denotes the state of an empty buffer. The operations associated with $\rob$ are:
@@
 {\commit(\rob) = (\pc,\pc',(\ld,x,a,v',s')) \\ v' \neq v & \exec(s,\pc,(\ld,x,v)) = (s'',\pc'')}
-{\iox{\text{P$_\text{so}$}}{s,\pc,\top,\\\rob,\ppc}{s'',\pc'',\bot,\phi,\\\setBp(\ppc,\pc)}{\text{ToP}(\epsilon, \ld, v)}}
+{\iox{\text{P$_\text{so}$}}{s,\pc,\top,\\\rob,\ppc}{s'',\pc'',\bot,\phi,\\\setBp(\ppc,\pc'')}{\text{ToP}(\epsilon, \ld, v)}}
 
@@
 whether the processor is waiting for a memory response about an
-instruction being committed, and the reorder buffer and branch
-predictor states. Its initial state is given by $(s_0, \pc_0, \bot, \phi, \ppc_0)$.
+instruction being committed, and the reorder-buffer and branch-predictor
+states. Its initial state is given by $(s_0, \pc_0, \bot, \phi, \ppc_0)$.
 The interface of this processor with memory (\ie{}
@@
 \begin{prf}
-Direct consequence of Theorems~\ref{ocorrect},~\ref{liftn} (the
+Direct consequence of Theorems~\ref{ocorrect} and \ref{liftn} (the
 latter is one of our preliminary results about $n$-repetition).

### 2014-11-10 Arvind.tex
Pass through cache system

@@
 cache-coherence protocol}.  It is a transliteration of a realistic
-algorithm implemented in the Bluespec hardware description
-language~\cite{Bluespec:TFRG}.
+algorithm~\cite{DNA:CoherenceImplementation} implemented in the
+Bluespec hardware description language~\cite{Bluespec:TFRG}.
 
@@
 some readers may want to skip this explanation upon first reading,
-skipping to the next section for a key theorem relating $M_c$ and $M_m$.
+advancing to the next subsection for a key theorem relating $M_c$ and $M_m$.
 
@@
 $\wt(c,a)$ & State that cache $c$ is waiting to upgrade to, for address $a$, or $\epsilon$\\ %Gives cache $c$'s desired coherence state to upgrade to (for address $a$), or $\epsilon$ if none\\
-$\dwt(p,c,a)$ & State that cache $p$ is waiting its child $c$ to downgrade to, for address $a$, or $\epsilon$\\
+$\dwt(p,c,a)$ & State that cache $p$ is waiting for its child $c$ to downgrade to, for address $a$, or $\epsilon$\\
 %$\dwt(p,c,a)$ & Gives cache $p$'s desired coherence state for its child $c$ to downgrade to (for address $a$), or $\epsilon$ if none\\
@@
 leaf caches connected to the processors, but the
-interpretation is subtly different at the higher level caches.
-
-$\wt$ denotes what permission the cache is waiting for,
+interpretation is subtly different at the higher-level caches.
+
+$\wt$ stores what permission a cache is waiting for,
 if any.  That is, a cache has decided to \emph{upgrade} its
@@
 
-The reason for having separate request and response channels for the same
-child-parent pair is to avoid deadlocks. Throughout the protocol, we maintain
-the invariant that a downgrade request from the parent always gets priority,
-\ie{} if a child is waiting for an upgrade response, it must still try to
-satisfy a downgrade request from the parent. If both the parent and the child
-are waiting for their respective responses, there will be a deadlock. This tie
-cannot be broken arbitrarily: if the parent concedes and starts processing a
-child's request, then it may have to send downgrade requests to other children
-to make the states compatible, but the other children may have sent upgrade
-requests and remain waiting, leading to a deadlock. The only way to break the
-vicious waiting cycle is for the children to concede and respond when they get
-downgrade requests. Of course, they have to downgrade the states of their own
-children first, and this process goes on recursively until the downgrade
-requests reach the leaf caches, which can respond immediately.
+%% The reason for having separate request and response channels for the same
+%% child-parent pair is to avoid deadlocks. Throughout the protocol, we maintain
+%% the invariant that a downgrade request from the parent always gets priority,
+%% \ie{} if a child is waiting for an upgrade response, it must still try to
+%% satisfy a downgrade request from the parent. If both the parent and the child
+%% are waiting for their respective responses, there will be a deadlock. This tie
+%% cannot be broken arbitrarily: if the parent concedes and starts processing a
+%% child's request, then it may have to send downgrade requests to other children
+%% to make the states compatible, but the other children may have sent upgrade
+%% requests and remain waiting, leading to a deadlock. The only way to break the
+%% vicious waiting cycle is for the children to concede and respond when they get
+%% downgrade requests. Of course, they have to downgrade the states of their own
+%% children first, and this process goes on recursively until the downgrade
+%% requests reach the leaf caches, which can respond immediately.
 
@@
 The proof that $M_c$ obeys store atomicity is involved enough that we will only
-state the lemmas used to prove this. The proofs in Coq for each theorem and
-lemma in this section can be found in our supplementary material.
+state key lemmas that we used. The proofs in Coq for each theorem and
+lemma in this section are available in our implementation.
 
@@
 \begin{lemma}%[Conservative directory] 
-$\forall p, \forall c, \forall a\;\s(c, a) \le \dst(p, c, a)$
+$\forall p, \forall c, \forall a.\;\s(c, a) \le \dst(p, c, a)$
 \label{conservative}

### 2014-11-10 Conclusion.tex
Finished a pass through whole paper; now need to think about shrinking

@@
 shared-memory hardware systems, corresponding naturally to the modularization seen in
-hardware implementation. This makes it feasible for verification of the
+hardware implementations. Our approach allows verification of the
 system to be carried out throughout the implementation process without

### 2014-11-10 Arvind.tex
Add a 'why PLDI' paragraph to intro; get us back within the page limit

@@
 \centering
-\includegraphics[scale=.43]{zoom}
+\includegraphics[scale=.4]{zoom}
 \caption{Components of a multiprocessor system}
@@
 but commits instructions (once some ``verification'' is done) in order. This idea
-of verification loads for a speculative processor is not new \cite{cain}.
+of verification loads for a speculative processor is not new \cite{cain}.  In
+common cases, performance of executing loads twice (first to speculate and then
+to check the answers in program order) is good, because it is likely that the
+second load finds the address already in a local cache, thanks to the recent
+processing of the first load.
 
@@
 This LTS encodes Lamport's notion of SC, where processors take turns executing
-nondeterministically in a simple interleaving.  We need know very few details
-of the ISA to define the SC model.  Our final, optimized system is parametrized
-over an ISA in the same way.  In the course of the rest of this paper, we will
-define an optimized system $O$ and prove $O \sqsubseteq \text{SC}$.
-
-However, we are not interested merely in verifying the particular system $O$.
-Instead, we want to structure our proof \emph{modularly}, so that different
-common hardware components can be verified separately, and, in a black-box
-manner, we may assemble a correctness proof for any composition of these
-components.  For instance, we want to verify CPUs and memory systems
-separately, picking any realization of each to produce a verified system, with
-minimal new proof effort. As a more concrete example, memory systems should be
-verifiable independently of which weak memory model processors assume; we want
-to reuse memory proofs for different kinds of processors.
-
-To support this style of modular decomposition, we need to introduce a few
-intermediate systems.
-
+nondeterministically in a simple interleaving.
 Note that, in this setting, an operational specification like the LTS
@@
 buffer} component and composing the three components simultaneously.
+
+Our final, optimized system is parametrized over an ISA in the same way as SC is.
+In the course of the rest of this paper, we will
+define an optimized system $O$ and prove $O \sqsubseteq \text{SC}$.
+To support a modular proof decomposition, however, we need to introduce a few
+intermediate systems first.
 
@@
 speculation, by emptying the reorder buffer and resetting the
-next predicted program counter in the branch predictor to the correct next value. This step mimics
-the mechanism implemented in most modern-day processors, operating on the
-conservative assumption that the processor may have performed arbitrary
-incorrect speculation after receiving the inaccurate response from
-memory.  (Note that here ``inaccurate'' does not mean ``wrong at
-the time the response was received,'' but rather ``wrong at commit
-time'' or ``correct before, but not anymore.'')
+next predicted program counter in the branch predictor to the correct next value.
+
+%% This step mimics
+%% the mechanism implemented in most modern-day processors, operating on the
+%% conservative assumption that the processor may have performed arbitrary
+%% incorrect speculation after receiving the inaccurate response from
+%% memory.  (Note that here ``inaccurate'' does not mean ``wrong at
+%% the time the response was received,'' but rather ``wrong at commit
+%% time'' or ``correct before, but not anymore.'')
 
@@
 out-of-order processors.  We use a trace-transformation function
-$\noSpec$ that drops all speculative-load requests and responses (\ie{} those load requests and responses whose tags are $\epsilon$).
+$\noSpec$ that drops all speculative-load requests and responses (\ie{} those load requests and responses whose tags are not $\epsilon$).
 See Definition~\ref{refines} for a review of how we use such
@@
 
+\input{TransitionsUsefulCache1}
+
 We now turn our attention toward a more efficient implementation of
-memory.  It is physically infeasible to connect many processors
-directly to a single memory with low enough latency to handle most
-memory requests quickly enough.  Instead, each processor will have its
-own \emph{L1 cache} of recently used memory cells.  Whenever possible, we
-want to service memory requests from processors' own caches, only
-contacting memory for addresses not found in the caches.  To further
-avoid contacting main memory, we create additional higher-level
-caches, connecting each L1 cache to an L2 cache, potentially
-connecting L2 caches to L3, etc., bottoming out in main memory at the
-root of our tree. The zoomed-in portion of the memory in Figure~\ref{zoom}
-diagrams the basic layout.
-
-Now we have concurrent interaction of many processors with many caches
+memory.  With the cache hierarchy of Figure~\ref{zoom}, we have concurrent
+interaction of many processors with many caches
 with main memory, and the relationship with the original $M_m$ system
@@
 We now try to explain aspects of this transition system at an
-intuitive level.  Thanks to the overall modular proof structure in
-this paper, it is not essential to understand the details of the cache
-system, to understand any other part of our development.  Therefore,
-some readers may want to skip this explanation upon first reading,
-advancing to the next subsection for a key theorem relating $M_c$ and $M_m$.
+intuitive level, though the rest of the paper may be understood starting just
+from the final theorem of the next subsection.
 
@@
 add and remove elements from the parent to child channel, as opposed to
-$\uplus$ for the child to parent channel). The reason is that only one request
-or one response is sent for a particular address between a child-parent pair --
-this invariant can be proved for the LTS in  Figure \ref{cache}. The order
-between a request and response message sent from the same child to its parent
-is lost, but our protocol ensures that they are processed in order by the parent, by
-comparing the previous coherence state of the incoming messages with the
-parent's directory state (in rules ParentRecvRsp
-and ParentRecvReq).
+$\uplus$ for the child to parent channel). This asymmetry is motivated by subtle
+concerns of avoiding deadlock and taking advantage of communication regularities.
 
@@
 the $\s$'s upgrade and children's downgrade responses are all received, the
-original request can be responded to. The transitions for sending requests are
-allowed to happen spontaneously without any trigger. $P_\text{ref}$ issues speculative
-loads only for addresses which are in \Sh{} or \Mo{} state in L1; otherwise L1
+original request can be responded to. $P_\text{ref}$ issues speculative
+loads only for addresses in \Sh{} or \Mo{} state in L1; otherwise L1
 spontaneously requests an upgrade for that address.
@@
 state.  This transition is used to model invalidation of cache lines to make
-room for a different location (we permit a cache to downgrade from $M$ to $S$
-instead of only to $I$).
+room for a different location.
+%(we permit a cache to downgrade from $M$ to $S$ instead of only to $I$)
 As a result, the parent's $\dst$ and the
@@
 by not dequeuing the request.
-
-\input{TransitionsUsefulCache1}
 
@@
 value, because it is \textit{clean} (as it has no children). To prove Lemma
-\ref{latestValue}, it has to be decomposed futher into the following, each of
+\ref{latestValue}, it has to be decomposed further into the following, each of
 which holds at any time.

### 2014-11-10 Introduction.tex
Add a 'why PLDI' paragraph to intro; get us back within the page limit

@@
 \end{itemize}
+
+Overall, our agenda is to import, to the architecture domain, good ideas from
+the worlds of programming-language semantics and formal software verification.
+Wearing our architecture hats, we have found some of the proof insights to yield
+surprising wisdom about hardware design, and we are optimistic that the
+programming-languages perspective on modular verification will pay even more
+dividends for reliable hardware design and verification.
 

### 2014-11-10 Arvind.tex
stuff

@@
 memory.  With the cache hierarchy of Figure~\ref{zoom}, we have concurrent
-interaction of many processors with many caches
-with main memory, and the relationship with the original $M_m$ system
+interaction of many processors with many caches, and the relationship with the original $M_m$ system
 is far from direct.  However, this intricate concurrent execution is

### 2014-11-10 Arvind.tex
stuff

@@
 committed stores from an in-order processor $P_i$ are enqueued, and if
-any ``subsequent'' load from the same processor $P_i$ is allowed to read and commit the latest value from the store buffer, then such a multiprocessor system is said to implement the TSO~\cite{x86tsocacm10} memory
-model.
+any ``subsequent'' load from the same processor $P_i$ is allowed to read and commit the latest value from the store buffer, then such a multiprocessor system is said to implement the TSO memory model \cite{x86tsocacm10}.
 

### 2014-11-10 Introduction.tex
stuff

@@
 proofs using the Coq proof assistant.  As a challenging case study, we focus on
-\textbf{verifying that a particular infinite family of multicore, shared-memory
+\emph{verifying that a particular infinite family of multicore, shared-memory
 systems implements sequential consistency}.  The proof is parametrized over an
@@
 implementation \cite{DNA:CoherenceImplementation} used to implement a
-cycle-accurate simulator for a multiprocessor PowerPC system \cite{Khan:PowerPc}. The hardware synthesized
+cycle-accurate simulator for a cache-coherent multiprocessor PowerPC system \cite{Khan:PowerPc}. The hardware synthesized
 from that implementation is rather efficient: an 8-core system with a 2-level
 cache hierarchy can run 55 million instructions per second on the BEE FPGA
-board \cite{bee2}. Figure \ref{both} gives an example of an LTS inference rule
+board \cite{bee2}.
+
+Figure \ref{both} gives an example of an LTS inference rule
 transliterated directly from an actual Bluespec rule used in the PowerPC
@@
 {\parent(c,p) & \s(c,a) < x & \wt(c,a) = \epsilon}
-{\ioxe{$M_c$}
-{\dt, \ch, \s,\\ \dst, \wt, \dwt,\\ \inp, \outp}
-{\dt, \ch[ (c, p, \req) \coloneqq(a, \s(c, a),x)\\
-\;\;\uplus \ch(c,p,\req)], \s, \dst,\\
-\wt[(c,a)\coloneqq x], \dwt,\inp, \outp}{}}
+{\hspace{-.4cm}\ioxe{$M_c$}
+{\hspace{-.2cm}\dt, \ch, \s,\hspace{-.2cm}\\
+\hspace{-.2cm}\dst, \wt, \dwt,\hspace{-.2cm}\\
+\hspace{-.2cm}\inp, \outp\hspace{-.2cm}}
+{\hspace{-.2cm}\dt, \ch[ (c, p, \req) \coloneqq(a, \s(c, a),x)\hspace{-.2cm}\\
+\;\;\;\;\;\;\uplus \ch(c,p,\req)], \s, \dst,\hspace{-.2cm}\\
+\hspace{-.2cm}\wt[(c,a)\coloneqq x], \dwt,\inp, \outp\hspace{-.2cm}}{}\hspace{-.4cm}}
 \end{boxedminipage}

### 2014-12-14 Rebuttal.txt
added reviews

Some of the questions raised by the first three reviewers are addressed in the
comments of the Reviewer D.

We have submitted the source code of our Coq development as a suplement to our
paper in the submission website. It can also be obtained from the public repository
http://github.com/vmurali/SeqConsistency.

In terms of code-reuse, we have structured the proofs modularly as described in
the paper. Thus, the proofs of the different components can be reused while
changing the other components. For example, the proof of the cache-coherence
protocol can be reused as is for verifying other (non-sequentially consistent)
systems. Moreover, the proof of the cache-coherence is itself modular and can
be adopted to other protocols (like MESI, MOSI); The proofs for Lemmas 2 to 5
will change, but the lemmas correspond to at most 3 components (and states) and
hence can be easily proved (even using techniques like model-checking).

We agree with Reviewer-D that liveness is an important non-trivial property
that has to be proven for a cache-coherence protocol. The example of
ParentSendReq is an example of a voluntary rule, and the policy to fire such a
rule does indeed affect the liveness. In our development, we haven't proved
full liveness (which includes the liveness of the processor), but we have
proved the following (stated informally):
a) Any request (or response) that a cache wants to send can eventually be sent
(either to its parent or to its children)
b) Under the assumption that ParentSendReq rule (sending downgrade request from
a parent to its children) will fire eventually, whenever the parent has an
upgrade request from a child, every request will eventually get a response.

Thus, the only extraneous condition we impose is that a parent should "decide"
to send a downgrade request to (some of) its children whenever it sees an
upgrade request from one of its children. Eventually, such a request will be
sent via ParentSendReq rule.

We believe such a modular technique for verifying full-system correctness with
respect to SC will be useful for hardware designers/verification-engineers to
independently verify different components of the system. We do not model non-SC
systems (like ARM/POWER), which exhibit behaviors described by Reviewer-C
(involving shared store buffers, committing non-verified load values, etc).
Reviewer-C, please note that the idea of verification loads does indeed ensure
correctness, even if there were stores to that address in between the
speculative load and the verification load, as long as the final store read by
the verification load writes the same value as the speculative load. In fact,
our final theorem ensures that this is indeed the case, otherwise the
speculative system will not be an implementation of the SC reference system.

Reviewer-B please note that our SC model is exactly as specified by Lamport.
Also please note that by local buffers in Figure 2, we do not mean just store
buffers; it includes all local data accessible to the processor.

Thanks for notifying us of other related work (Hu et.al.'s "Decidable SC"
SPAA'03, Qadeer's "Verifying SC by model-checking" TPDS'03, Sorin's
"Verification-aware Architectures" PACT'07, "Lamport Clocks: Verifying
Directory Cache-Coherence Protocol" SPAA'98 and Moore's "ACL2 proof of cache
coherence" CAV'98), and for pointing the other bugs in our related work
section.

Reviewer-B's concerns:
- Ideal processor and memory: as described in that paragraph - memory responds
  instantaneously and processor executes atomically without speculation
- Bluespec served as the model: The motivation for LTS comes from using
  Bluespec to describe hardware, which is similar to our LTS formalization.
- Bluespec's competitive performance can be seen from [Dave et al. 2005], and
  other papers using Bluespec designs.
- Peek-beneath abstraction boundaries: To prove the full-system, the
  individual components are treated as black-boxes, completely captured by their
  interface specification.
- Good-ideas from worlds of PL semantics: Using LTSes as a precise hardware
  specification.
- Infinite families of hardware design ... resist reduction to finite-state
  models: reduction is applied to 'infinite families'
- Figure 2: logical view of system as 3-components: It is more-or-less accepted
  in the architecture community.
- Refining vs Equivalent: Throughout the paper, we are showing only refinement
  in one direction, though you are correct that in this particular example they
  are equivalent.
- Finer-grained architectural details: as mentioned in that paragraph (like
  register-renaming)
- Main-memory still exposes atomic loads and stores: Most architectures (including
  POWER and ARM) can be modeled using a memory that responds to loads and stores
  instantaneously, though the response (or request) can sit in a buffer between
  the memory and the processor, leading to relaxed memory models.

### 2014-12-15 Rebuttal.txt
stuff

@@
 
-We have submitted the source code of our Coq development as a suplement to our
-paper in the submission website. It can also be obtained from the public repository
-http://github.com/vmurali/SeqConsistency.
+We have submitted the source code of our Coq development as a supplement to our
+paper in the submission website (we didn't give any link to preserve
+anonymity).
 
-In terms of code-reuse, we have structured the proofs modularly as described in
-the paper. Thus, the proofs of the different components can be reused while
-changing the other components. For example, the proof of the cache-coherence
-protocol can be reused as is for verifying other (non-sequentially consistent)
-systems. Moreover, the proof of the cache-coherence is itself modular and can
-be adopted to other protocols (like MESI, MOSI); The proofs for Lemmas 2 to 5
-will change, but the lemmas correspond to at most 3 components (and states) and
-hence can be easily proved (even using techniques like model-checking).
+We have structured the proofs modularly as described in the paper. Thus, the
+proofs of the different components can be reused while changing the other
+components. For example, the proof of the cache-coherence protocol can be
+reused as is for verifying other (non-sequentially consistent) systems.
+Moreover, the proof of the cache-coherence is itself modular and can be adopted
+to other protocols (like MESI, MOSI); The proofs for Lemmas 2 to 5 will change,
+but the lemmas correspond to at most 3 components (and states) and hence can be
+easily proved (even using techniques like model-checking).
 
-We agree with Reviewer-D that liveness is an important non-trivial property
-that has to be proven for a cache-coherence protocol. The example of
-ParentSendReq is an example of a voluntary rule, and the policy to fire such a
-rule does indeed affect the liveness. In our development, we haven't proved
-full liveness (which includes the liveness of the processor), but we have
-proved the following (stated informally):
+We agree with Reviewer-D that proving liveness is important and non-trivial.
+The example of ParentSendReq is an example of a voluntary rule, and the policy
+to fire such a rule does indeed affect the liveness. In our development, we
+haven't proved full liveness of the system (which includes the liveness of the
+processor), but we have proved the following which specifies the liveness of
+the cache-coherence protocol:
 a) Any request (or response) that a cache wants to send can eventually be sent
 (either to its parent or to its children)
-b) Under the assumption that ParentSendReq rule (sending downgrade request from
-a parent to its children) will fire eventually, whenever the parent has an
-upgrade request from a child, every request will eventually get a response.
+b) Under the assumption that a parent will eventually "decide" to send a
+downgrade request to its children (using ParentSendReq rule) whenever it sees
+an upgrade request from a child (as opposed to ignoring such an upgrade request
+from a child forever), every request sent to either the parent or to a child
+will eventually get a response.
 
-Thus, the only extraneous condition we impose is that a parent should "decide"
-to send a downgrade request to (some of) its children whenever it sees an
-upgrade request from one of its children. Eventually, such a request will be
-sent via ParentSendReq rule.
+We chose "halting" for pedagogical reasons in the paper; in our actual Coq
+development, we use the more traditional trace-equivalence definition requiring
+inter-convertible states. It is worth noting that any safety property can be
+reduced to the problem of halting.
 
@@
 respect to SC will be useful for hardware designers/verification-engineers to
-independently verify different components of the system. We do not model non-SC
-systems (like ARM/POWER), which exhibit behaviors described by Reviewer-C
-(involving shared store buffers, committing non-verified load values, etc).
-Reviewer-C, please note that the idea of verification loads does indeed ensure
-correctness, even if there were stores to that address in between the
+independently verify different components of the system. We can't be sure that
+everything will generalize effortlessly to the trickiest relaxed memory models
+(like those exhibited by ARM/POWER as pointed by Reviewer-C because of no
+verification loads and/or shared store buffers), but it would have been hard to
+figure out the basic framework if we started on more complex examples first.
+We still think our example is quite complex and interesting, which Reviewer-D
+seems to agree with.
+
+Reviewer-C, please note that verification loads do indeed ensure correctness.
+This is true even if there were stores to that address in between the
 speculative load and the verification load, as long as the final store read by
-the verification load writes the same value as the speculative load. In fact,
-our final theorem ensures that this is indeed the case, otherwise the
-speculative system will not be an implementation of the SC reference system.
+the verification load writes the same value as the speculative load. Our final
+theorem ensures this; otherwise the speculative system will not be an
+implementation of the SC reference system.
 
@@
 Directory Cache-Coherence Protocol" SPAA'98 and Moore's "ACL2 proof of cache
-coherence" CAV'98), and for pointing the other bugs in our related work
-section.
+coherence" CAV'98), and to Reviewer-B for pointing the other bugs in our
+related work section.
 
@@
   specification.
-- Infinite families of hardware design ... resist reduction to finite-state
+- Infinite families of hardware design resist reduction to finite-state
   models: reduction is applied to 'infinite families'

### 2014-12-15 Rebuttal.txt
stuff

@@
 Reviewer-B's concerns:
-- Ideal processor and memory: as described in that paragraph - memory responds
+- Ideal processor and memory: As described in that paragraph - memory responds
   instantaneously and processor executes atomically without speculation
@@
   interface specification.
-- Good-ideas from worlds of PL semantics: Using LTSes as a precise hardware
+- Good-ideas from the world of PL semantics: Using LTSes as a precise hardware
   specification.
@@
   are equivalent.
-- Finer-grained architectural details: as mentioned in that paragraph (like
-  register-renaming)
+- Finer-grained architectural details: As mentioned in that paragraph (like
+  register-renaming).
 - Main-memory still exposes atomic loads and stores: Most architectures (including

### 2014-12-15 Rebuttal.txt
stuff

@@
 Directory Cache-Coherence Protocol" SPAA'98 and Moore's "ACL2 proof of cache
-coherence" CAV'98), and to Reviewer-B for pointing the other bugs in our
-related work section.
+coherence" CAV'98), and for pointing the other bugs in our related work
+section.
 

### 2014-12-15 Rebuttal.txt
stuff

@@
-Some of the questions raised by the first three reviewers are addressed in the
-comments of the Reviewer D.
+Some of the questions raised by the reviewers are addressed by Reviewer-D.
 
@@
 to fire such a rule does indeed affect the liveness. In our development, we
-haven't proved full liveness of the system (which includes the liveness of the
-processor), but we have proved the following which specifies the liveness of
-the cache-coherence protocol:
+haven't proved full-system liveness (which includes the liveness of the
+processor), but we have proved the liveness of the cache-coherence protocol, in
+particular, the following:
 a) Any request (or response) that a cache wants to send can eventually be sent
@@
 
-We chose "halting" for pedagogical reasons in the paper; in our actual Coq
-development, we use the more traditional trace-equivalence definition requiring
-inter-convertible states. It is worth noting that any safety property can be
-reduced to the problem of halting.
+We chose "halting" for pedagogical purposes; in our actual Coq-code we use the
+traditional trace-equivalence definition requiring inter-convertible states. It
+is worth noting that any safety property can be reduced to the problem of
+halting.
 
@@
 We still think our example is quite complex and interesting, which Reviewer-D
-seems to agree with.
+seems to agree with. Moreover, the performance benefits of a relaxed memory
+model over SC is being actively debated in the architecture community.
 
@@
 theorem ensures this; otherwise the speculative system will not be an
-implementation of the SC reference system.
+implementation of the reference SC.
 
-Reviewer-B please note that our SC model is exactly as specified by Lamport.
+Reviewer-B, please note that our SC model is exactly as specified by Lamport.
 Also please note that by local buffers in Figure 2, we do not mean just store

### 2014-12-15 Rebuttal.txt
stuff

@@
-Some of the questions raised by the reviewers are addressed by Reviewer-D.
+Some of the questions raised by the reviewers (including the justification of
+modularization advantages) are addressed by Reviewer-D.
 
 We have submitted the source code of our Coq development as a supplement to our
-paper in the submission website (we didn't give any link to preserve
-anonymity).
+paper in the submission website.
 

### 2014-12-15 Rebuttal.txt
A pass through rebuttal

@@
 paper in the submission website.
+
+In response to "points against" by Reviews B and C:
+We believe such a modular technique for verifying full-system correctness with
+respect to SC will be useful for hardware designers/verification-engineers to
+independently verify different components of the system. We can't be sure that
+everything will generalize effortlessly to the trickiest relaxed memory models
+(like those exhibited by ARM/POWER as pointed by Reviewer-C because of no
+verification loads and/or shared store buffers), but it would have been hard to
+figure out the basic framework if we started on more complex examples first.
+We still think our example is quite complex and interesting, which Reviewer-D
+seems to agree with, and our focus is on developing a general framework,
+not this particular specification or proof.
+
+Reviewer-C, please note that verification loads do indeed ensure SC correctness.
+This is true even if there were stores to that address in between the
+speculative load and the verification load, as long as the final store read by
+the verification load writes the same value as the speculative load. Our final
+theorem ensures this; otherwise the speculative system will not be an
+implementation of the reference SC.
+
+Reviewer-B, please note that our SC model is exactly as specified by Lamport
+(though it may be written in different notation).  Also please note that by local
+buffers in Figure 2, we do not mean just store buffers; it includes all local
+data accessible to the processor.
+
+>>>> QUESTION FOR MURALI: What in the reviews does the following paragraph respond to?
+     Is it just about the Coq proofs?  (If so, apologies for reordering)
+     To me it seems perilously close to just restating our paper abstract. <<<<<
 
@@
 
-We believe such a modular technique for verifying full-system correctness with
-respect to SC will be useful for hardware designers/verification-engineers to
-independently verify different components of the system. We can't be sure that
-everything will generalize effortlessly to the trickiest relaxed memory models
-(like those exhibited by ARM/POWER as pointed by Reviewer-C because of no
-verification loads and/or shared store buffers), but it would have been hard to
-figure out the basic framework if we started on more complex examples first.
-We still think our example is quite complex and interesting, which Reviewer-D
-seems to agree with. Moreover, the performance benefits of a relaxed memory
-model over SC is being actively debated in the architecture community.
-
-Reviewer-C, please note that verification loads do indeed ensure correctness.
-This is true even if there were stores to that address in between the
-speculative load and the verification load, as long as the final store read by
-the verification load writes the same value as the speculative load. Our final
-theorem ensures this; otherwise the speculative system will not be an
-implementation of the reference SC.
-
-Reviewer-B, please note that our SC model is exactly as specified by Lamport.
-Also please note that by local buffers in Figure 2, we do not mean just store
-buffers; it includes all local data accessible to the processor.
-
 Thanks for notifying us of other related work (Hu et.al.'s "Decidable SC"
@@
 Directory Cache-Coherence Protocol" SPAA'98 and Moore's "ACL2 proof of cache
-coherence" CAV'98), and for pointing the other bugs in our related work
+coherence" CAV'98), and for pointing out the other bugs in our related work
 section.
@@
   individual components are treated as black-boxes, completely captured by their
-  interface specification.
+  interface specification.  Peeking beneath boundaries would instead look inside
+  implementations.
 - Good-ideas from the world of PL semantics: Using LTSes as a precise hardware

### 2014-12-15 Rebuttal.txt
stuff

@@
 We still think our example is quite complex and interesting, which Reviewer-D
-seems to agree with, and our focus is on developing a general framework,
-not this particular specification or proof.
+seems to agree with.
 
@@
 
-Reviewer-B, please note that our SC model is exactly as specified by Lamport
-(though it may be written in different notation).  Also please note that by local
-buffers in Figure 2, we do not mean just store buffers; it includes all local
-data accessible to the processor.
-
->>>> QUESTION FOR MURALI: What in the reviews does the following paragraph respond to?
-     Is it just about the Coq proofs?  (If so, apologies for reordering)
-     To me it seems perilously close to just restating our paper abstract. <<<<<
-
-We have structured the proofs modularly as described in the paper. Thus, the
-proofs of the different components can be reused while changing the other
-components. For example, the proof of the cache-coherence protocol can be
-reused as is for verifying other (non-sequentially consistent) systems.
-Moreover, the proof of the cache-coherence is itself modular and can be adopted
-to other protocols (like MESI, MOSI); The proofs for Lemmas 2 to 5 will change,
-but the lemmas correspond to at most 3 components (and states) and hence can be
-easily proved (even using techniques like model-checking).
+Regarding Reviewer-D's question about reusability of our proofs and framework,
+the proof of the cache-coherence protocol can be reused as is for verifying
+other (relaxed memory) systems. Moreover, the proof of the cache-coherence is
+itself modular and can be adopted to other protocols (like MESI, MOSI); The
+proofs for Lemmas 2 to 5 will change, but the lemmas correspond to at most 3
+components (and states) and hence can be easily proved (even using techniques
+like model-checking).
 
@@
 halting.
+
+Reviewer-B, please note that our SC model is exactly as specified by Lamport
+(though it may be written in different notation).  Also please note that by local
+buffers in Figure 2, we do not mean just store buffers; it includes all local
+data accessible to the processor.
 

### 2014-12-15 Rebuttal.txt
removed old reviews, rebuttal

@@
 We still think our example is quite complex and interesting, which Reviewer-D
-seems to agree with.
+seems to agree with, and our focus is on developing a general framework,
+not this particular specification or proof.
 
@@
 
-Regarding Reviewer-D's question about reusability of our proofs and framework,
-the proof of the cache-coherence protocol can be reused as is for verifying
-other (relaxed memory) systems. Moreover, the proof of the cache-coherence is
-itself modular and can be adopted to other protocols (like MESI, MOSI); The
-proofs for Lemmas 2 to 5 will change, but the lemmas correspond to at most 3
-components (and states) and hence can be easily proved (even using techniques
-like model-checking).
+Reviewer-B, please note that our SC model is exactly as specified by Lamport
+(though it may be written in different notation).  Also please note that by local
+buffers in Figure 2, we do not mean just store buffers; it includes all local
+data accessible to the processor.
 
@@
 traditional trace-equivalence definition requiring inter-convertible states. It
-is worth noting that any safety property can be reduced to the problem of
-halting.
+is worth noting that any safety property can be reduced to the halting problem.
 
-Reviewer-B, please note that our SC model is exactly as specified by Lamport
-(though it may be written in different notation).  Also please note that by local
-buffers in Figure 2, we do not mean just store buffers; it includes all local
-data accessible to the processor.
+Thanks for notifying us of other related work (Hu's "Decidable SC" SPAA'03,
+Qadeer's "Verifying SC by model-checking" TPDS'03, Sorin's "Verification-aware
+Architectures" PACT'07, "Lamport Clocks: Verifying Directory Protocol" SPAA'98
+and Moore's "ACL2 cache coherence proof" CAV'98), and for pointing out other
+bugs in our related work section.
 
-Thanks for notifying us of other related work (Hu et.al.'s "Decidable SC"
-SPAA'03, Qadeer's "Verifying SC by model-checking" TPDS'03, Sorin's
-"Verification-aware Architectures" PACT'07, "Lamport Clocks: Verifying
-Directory Cache-Coherence Protocol" SPAA'98 and Moore's "ACL2 proof of cache
-coherence" CAV'98), and for pointing out the other bugs in our related work
-section.
-
+Beyond 500 words:
 Reviewer-B's concerns:
@@
   the memory and the processor, leading to relaxed memory models.
+
+Regarding Reviewer-D's question about reusability of our proofs and framework,
+the proof of the cache-coherence protocol can be reused as is for verifying
+other (relaxed memory) systems. Moreover, the proof of the cache-coherence is
+itself modular and can be adopted to other protocols (like MESI, MOSI); The
+proofs for Lemmas 2 to 5 will change, but the lemmas correspond to at most 3
+components (and states) and hence can be easily proved (even using techniques
+like model-checking).

### 2015-01-20 Rebuttal2.txt
stuff

1. Why waste time describing LTSes:
We introduced LTSes mainly to define the notion of refinement that we
are using, as well as to define the combination operators.

2. The user of a theorem-prover has to walk the "protocol space" through case analysis:
The theorem-prover has to walk the protocol space through case analysis, but
the proof-script automation in Coq significantly reduces the burden. One can
write proof-scripts that performs a proof-search, i.e. tries to apply
previously proven basic lemmas on each of the cases and leaves the user with
very few actual proof obligations that are complex or unique so as to not be
discharged using automated proof-search.

3. Invariants are the hardest part in parametric verification of a cache
coherence protocol, and we do not mention it:
We do mention the high level theorems that we use in the paper.  While we
restrict ourselves to the verification of the MSI protocol's implementation,
the proof is modular, and a large portion of the proof can be used, as is, for
verification of other protocols (like MESI, MOSI and MOESI), as we discuss in
the previous rebuttal. The theorem prover helps construct and prove the high
level theorems that make these protocols correct. These theorems depend on a
few basic properties pertaining that can be established easily, but were not
mentioned in the paper due to lack of space. These properties include a) state
changes occuring when and only when response messages are sent, b) absence of
response messages from the parent to a child and from the same child to the
parent simultaneously in the network, for the same address and c) FIFO
properties of the messages. These basic properties seem to be the foundation
over which the other cache coherence protocols lie -- they eliminate the corner
cases which make the proof difficult. For example, since responses from a child
to its parent and from the parent to the same child for the same address cannot
be present simultaneously, an upgrade response from the parent cannot cross a
downgrade response from the child, which would have eventually led to the
upgrade of the child and downgrade of the parent violating Lemma 3. As pointed
out by Reviewer-E, the theorems and these properties serve as a method to
derive a correct protocol (the different protocols are different system
instances which obey these properties).

4. Pipelining protocol engines and Bluespec:
Though the semantics of the atomic transitions in our paper execute one at a
time, different independent transitions can be made to occur in parallel in the
actual implementation without changing the semantics; Bluespec already does
this optimization. One can take this further, where each atomic transition can
be spread across multiple hardware clock cycles, while ensuring that the
overall atomic transaction semantics are obeyed. Bluespec currently does not
implement this, though there has been work in this area [Karczmarek et.al.].

5. Nirav's point: I don't see that Reviewer-E's scathing concern fall
into the core of what we're addressing; our focus on modularity is to
allow verification to be (mostly) carried over to different
microarchitectural variants, which is a bare minimum if one wants
architects to have any lateral freedom to design a working
implementation with a low-cost path to verification. I don't see why
there's any fundamental reason that "scalable" approaches he's pointed to
can't be fit into our methodology directly? I've only glanced at the
papers, but it seems like it should be so.
--- Note sure how to condense it and incorporate -Murali

Reviewer-F:
1. Evidence of proofs being machine-checked:
We have submitted the source code of our Coq development as a supplement to our
paper in the submission website.

2. Figure 3, Tags in Figures 5 & 6:
Both these points are valid and we could indicate these points explicitly.

3. Figure 6: The third case:
Reviewer-F's assertion is not true. The "upd" function merely determines if the
"real" state of the system is as dictated by the state (s,pc) or if it should
be modified to take into account the responses from the memory system. If the
processor is not waiting for the memory system response, then the state is
exactly (s,pc). We made an optimization in the fourth case where we make the
real state H whenever we know that the next instruction we decode is H; instead
the state can be left as (s,pc) in the fourth case.

4. Figure 8: Tags for non-speculative loads:
Our invariant is that there can be only one outstanding non-speculative load,
though there can be several outstanding speculative loads along with the
outstanding non-speculative load. Thus, anything without a tag is the
non-speculative load and there is only one of them, and anything with a tag is
a speculative load.

Reviewer-E's related work summary:

FM methodology: The methodology used for proving that the memory system
realizes cache coherence is not new. A theorem-proving approach to
verifying memory systems has been tacit in many publications. Researchers
preferred model checking, as opposed to pushing through theorem-proving
based proofs largely because of the daunting complexity of realistic
memory system protocols. One can find several approaches that address
specific aspects of memory system verification in, say, Wim Hesselink's
work : http://www.informatik.uni-trier.de/~ley/pers/hd/h/Hesselink:Wim_H=.

@@@ Murali response
Lots of PVS-based proofs for concurrent algorithms, mutual exclusion, etc...
(Needs to be read further, but it's just a bunch of mechanized proofs for
a lot of concurrent algorithms)
***

There has of course been recent trends to simplify protocols thru the DeNoVo
project:
http://scholar.google.com/citations?view_op=view_citation&hl=en&user=ufGYoL0AAAAJ&citation_for_view=ufGYoL0AAAAJ:8k81kl-MbHgC
The authors have not cited such related pieces of work and justified the
simplicity of their design.

@@@ Murali response
A new non-conventional cache coherence protocol that produces sequentially
consistent results for data-rate free programs. There's no formal proof that
this protocol is correct...
(Needs to be read further, but doesn't seem relevant)
***

The authors claim that they are achieving their proof for an arbitrary number
of processors. They do not mention anything about the invariants involved
in their parametric proofs. This leads me to suspect that these invariants
are pretty straightforward.  If one harks back to work addressing real
protocols, the Chou, Mannava, Park work
(http://scholar.google.com/citations?view_op=view_citation&hl=en&user=ufGYoL0AAAAJ&citation_for_view=ufGYoL0AAAAJ:qjMakFHDy7sC) shows how exceedingly hard
it is to discover these invariants. In fact, the CMP approach is an
incremental CEGAR-style method for discovering these invariants. Stating these
invariants explicitly would be virtually impossible - even for FLASH-like
protocols.  In fact, a whole series of papers have tried to address invariant
discovery in this space : http://dl.acm.org/citation.cfm?id=1517434

@@@ Murali response
Chou, Mannava, Park [CMP-method] approach uses three paradigms: a) symmetry of
cache coherence protocols b) abstraction of several nodes which behaves as just
another node and c) counter-example guided abtraction refinement (CEGAR) for
refining the abstraction's properties. Using CEGAR, they get invariants for the
rest of the system while concentrating on a finite number of nodes (2).  Each
refinement will not restrict the behavior, but will help towards proving a
safety property. Specifically, let P be an invariant for the abstract
transition system, and we should prove that it satisfies safety property X. We
start out with a P that doesn't satisfy X. Then we add constraints on P, while
permitting the same transitions. Eventually P will satisfy X. The manual
portion of this procedure is in strengthening P.
***

@@@ Murali response
The second paper tries to formulate better invariants in the first place using
the notion of "flows". It says what other messages can interfere when the other
message is being processed...
(Needs to be read further -- I don't understand "flows" clearly; can't say
anything more before I undestand that properly)
***

For hierarchical protocols, this work has been extended in e.g.,
http://link.springer.com/article/10.1007%2Fs10703-010-0092-y

@@@ Murali response
Same approach as [CMP-method] to hierarchical protocols.
***

### 2015-01-20 RelatedWork.tex
stuff

@@
 Cachet~\cite{StoyShenArvind:Proofs}, another single-level cache coherence
-protocol, was verified to be obey the CRF memory model~\cite{Shen:CRF}, and
+protocol, was verified to obey the CRF memory model~\cite{Shen:CRF}, and
 parts of the proof were implemented using PVS.

### 2015-01-20 Rebuttal2.txt
Arvind's changes

@@
 1. Why waste time describing LTSes:
-We introduced LTSes mainly to define the notion of refinement that we
-are using, as well as to define the combination operators.
+We introduced LTSes mainly to define the notion of refinement that we are
+using, as well as to define the operators to combine LTSes.
 
-2. The user of a theorem-prover has to walk the "protocol space" through case analysis:
+2. The user of a theorem-prover has to walk the "protocol space" through case
+analysis:
 The theorem-prover has to walk the protocol space through case analysis, but
 the proof-script automation in Coq significantly reduces the burden. One can
-write proof-scripts that performs a proof-search, i.e. tries to apply
-previously proven basic lemmas on each of the cases and leaves the user with
-very few actual proof obligations that are complex or unique so as to not be
-discharged using automated proof-search.
+write proof-scripts that perform a proof-search, i.e. try to apply previously
+proven basic lemmas on each of the cases and leave the user with very few
+actual proof obligations that are complex or unique so as to not be discharged
+using automated proof-search.
 
@@
 level theorems that make these protocols correct. These theorems depend on a
-few basic properties pertaining that can be established easily, but were not
-mentioned in the paper due to lack of space. These properties include a) state
-changes occuring when and only when response messages are sent, b) absence of
-response messages from the parent to a child and from the same child to the
-parent simultaneously in the network, for the same address and c) FIFO
-properties of the messages. These basic properties seem to be the foundation
-over which the other cache coherence protocols lie -- they eliminate the corner
-cases which make the proof difficult. For example, since responses from a child
-to its parent and from the parent to the same child for the same address cannot
-be present simultaneously, an upgrade response from the parent cannot cross a
-downgrade response from the child, which would have eventually led to the
-upgrade of the child and downgrade of the parent violating Lemma 3. As pointed
-out by Reviewer-E, the theorems and these properties serve as a method to
-derive a correct protocol (the different protocols are different system
-instances which obey these properties).
+few basic properties that can be established easily, but were not mentioned in
+the paper due to lack of space. These properties include a) state changes occur
+when and only when response messages are sent, b) the absence of response
+messages from the parent to a child and from the same child to the parent
+simultaneously in the network, for the same address and c) FIFO properties of
+the responses. These basic properties seem to be the foundation over which the
+other cache coherence protocols are built -- they eliminate the corner cases
+simplifying the proof. For example, since responses from a child to its parent
+and from the parent to the same child for the same address cannot be present
+simultaneously, an upgrade response from the parent cannot cross a downgrade
+response from the child, which would have eventually led to the upgrade of the
+child and downgrade of the parent violating Lemma 3. As pointed out by
+Reviewer-E, the theorems and these properties serve as a method to derive a
+correct protocol (the different protocols are different system instances which
+obey these properties).
 
@@
 time, different independent transitions can be made to occur in parallel in the
-actual implementation without changing the semantics; Bluespec already does
-this optimization. One can take this further, where each atomic transition can
-be spread across multiple hardware clock cycles, while ensuring that the
-overall atomic transaction semantics are obeyed. Bluespec currently does not
-implement this, though there has been work in this area [Karczmarek et.al.].
+actual implementation without changing the semantics; Bluespec compiler excels
+at this optimization.  One can take this further, where each atomic transition
+can be spread across multiple hardware clock cycles, while ensuring that the
+overall atomic transaction semantics are obeyed. Bluespec implementation
+currently does not include this optimization, though detailed algorithms to
+accomplish this have been published have been published [Karczmarek, MEMOCODE'14].
 
@@
 2. Figure 3, Tags in Figures 5 & 6:
-Both these points are valid and we could indicate these points explicitly.
+Both these points are valid and we should have indicated them explicitly.
 
@@
 FM methodology: The methodology used for proving that the memory system
-realizes cache coherence is not new. A theorem-proving approach to
-verifying memory systems has been tacit in many publications. Researchers
-preferred model checking, as opposed to pushing through theorem-proving
-based proofs largely because of the daunting complexity of realistic
-memory system protocols. One can find several approaches that address
-specific aspects of memory system verification in, say, Wim Hesselink's
-work : http://www.informatik.uni-trier.de/~ley/pers/hd/h/Hesselink:Wim_H=.
+realizes cache coherence is not new. A theorem-proving approach to verifying
+memory systems has been tacit in many publications. Researchers preferred model
+checking, as opposed to pushing through theorem-proving based proofs largely
+because of the daunting complexity of realistic memory system protocols. One
+can find several approaches that address specific aspects of memory system
+verification in, say, Wim Hesselink's work :
+http://www.informatik.uni-trier.de/~ley/pers/hd/h/Hesselink:Wim_H=.
 
@@
 The authors claim that they are achieving their proof for an arbitrary number
-of processors. They do not mention anything about the invariants involved
-in their parametric proofs. This leads me to suspect that these invariants
-are pretty straightforward.  If one harks back to work addressing real
-protocols, the Chou, Mannava, Park work
-(http://scholar.google.com/citations?view_op=view_citation&hl=en&user=ufGYoL0AAAAJ&citation_for_view=ufGYoL0AAAAJ:qjMakFHDy7sC) shows how exceedingly hard
-it is to discover these invariants. In fact, the CMP approach is an
-incremental CEGAR-style method for discovering these invariants. Stating these
-invariants explicitly would be virtually impossible - even for FLASH-like
-protocols.  In fact, a whole series of papers have tried to address invariant
-discovery in this space : http://dl.acm.org/citation.cfm?id=1517434
+of processors. They do not mention anything about the invariants involved in
+their parametric proofs. This leads me to suspect that these invariants are
+pretty straightforward.  If one harks back to work addressing real protocols,
+the Chou, Mannava, Park work
+(http://scholar.google.com/citations?view_op=view_citation&hl=en&user=ufGYoL0AAAAJ&citation_for_view=ufGYoL0AAAAJ:qjMakFHDy7sC)
+shows how exceedingly hard it is to discover these invariants. In fact, the CMP
+approach is an incremental CEGAR-style method for discovering these invariants.
+Stating these invariants explicitly would be virtually impossible - even for
+FLASH-like protocols.  In fact, a whole series of papers have tried to address
+invariant discovery in this space : http://dl.acm.org/citation.cfm?id=1517434
 
@@
 (Needs to be read further -- I don't understand "flows" clearly; can't say
-anything more before I undestand that properly)
+anything more before I understand that properly)
 ***

### 2015-01-20 Rebuttal2.txt
Arvind's changes
Small changes to PLDI rebuttal #2

@@
 using, as well as to define the operators to combine LTSes.
+
+adamc suggests: This description here sounds like the wrong level of detail.
+If we respond to the point, IMO it should be just to say that we wanted to review standard material for the benefit of readers unfamiliar with it; the point wasn't to present novel concepts.
 
@@
 the responses. These basic properties seem to be the foundation over which the
-other cache coherence protocols are built -- they eliminate the corner cases
+other cache coherence protocols are built -- they eliminate the corner cases,
 simplifying the proof. For example, since responses from a child to its parent
@@
 response from the child, which would have eventually led to the upgrade of the
-child and downgrade of the parent violating Lemma 3. As pointed out by
+child and downgrade of the parent, violating Lemma 3. As pointed out by
 Reviewer-E, the theorems and these properties serve as a method to derive a
@@
 currently does not include this optimization, though detailed algorithms to
-accomplish this have been published have been published [Karczmarek, MEMOCODE'14].
+accomplish this have been published [Karczmarek, MEMOCODE'14].
 
@@
 memory systems has been tacit in many publications. Researchers preferred model
-checking, as opposed to pushing through theorem-proving based proofs largely
+checking, as opposed to pushing through theorem-proving based proofs, largely
 because of the daunting complexity of realistic memory system protocols. One
@@
 @@@ Murali response
-Same approach as [CMP-method] to hierarchical protocols.
+Same approach as [CMP-method] to hierarchical protocols. It models only 2 level hierarchies!
 ***

### 2015-01-20 Rebuttal2.txt
Revise Rebuttal points (more work needed on 3a).

@@
-1. Why waste time describing LTSes:
-We introduced LTSes mainly to define the notion of refinement that we are
-using, as well as to define the operators to combine LTSes.
+Reviewer-E
 
-adamc suggests: This description here sounds like the wrong level of detail.
-If we respond to the point, IMO it should be just to say that we wanted to review standard material for the benefit of readers unfamiliar with it; the point wasn't to present novel concepts.
+1. On Describing LTSes:
 
-2. The user of a theorem-prover has to walk the "protocol space" through case
-analysis:
-The theorem-prover has to walk the protocol space through case analysis, but
-the proof-script automation in Coq significantly reduces the burden. One can
-write proof-scripts that perform a proof-search, i.e. try to apply previously
-proven basic lemmas on each of the cases and leave the user with very few
-actual proof obligations that are complex or unique so as to not be discharged
-using automated proof-search.
+We include the LTS description for the benefit of readers unfamiliar with it;
+the point wasn't to present novel concepts.
 
-3. Invariants are the hardest part in parametric verification of a cache
+2. Modularity
+
+ One of the major benefits of our approach is that it unifies the
+ modularization of the design and its verification task which enables
+ verification to be done in tandem with mircoarchitectural design as noted by
+ Reviewer-E.  Further, this allows us to require only incremental verification
+ effort for an incremental microarchitectural change, enabling scaling of the
+ verification task.
+
+3. Model Checking vs. Theorem Proving
+
+a. Finding invariants is fundamentally hard. In [PAPER],
+
+Finding invariants is fundamentally hard. We are equivalent in complexity but
+  because of our modularization we have a free source of abstraction from the
+  designer.
+
+   - In comparision, Quote paper. Invariants were dervied manually, took one
+  month, and relied on a manual abstraction of the system space based on
+  symettry. Our work is similar save that we do not require an abstraction.
+
+  In comparision to the <PAPER>, our invariants of similar complexity. In
+  general, we believe that such invariants relatively simple but subtle and
+  generally intuitively understood by the designers.
+
+b. On Case-Analysis:
+
+Much of the requisite protocol case analysis is done automatically via
+proof-script automation in Coq. In our experience, cases where proof-script
+automation fail required an additional invariant; we strongly believe that
+manually providing a similiar invariant are necessary in a model checking
+approach.
+
+c. Benefits of TP-based parameterization: It should also be noted that their
+  work while hierarchical is only 2-level while ours applies to arbitrary
+  levels of hierarchy.
+
+4. Pipelining protocol engines and Bluespec:
+
+While atomic transitions in our paper have one-at-a-time semantics, the
+Bluespec compiler generates hardware which executes independant transitions
+concurrently in a clock cycle. The Bluespec compiler excels at this
+optimization. Atomic transition can also be spread across multiple clock
+cycles, while maintaining atomic transaction semantics [Karczmarek,
+MEMOCODE'14].
+
+Reviewer-F:
+
+1. Evidence of proofs being machine-checked:
+
+We submitted the source code of our Coq development as a supplement to our
+paper in the submission website.
+
+2. Figure 3, tags in Figures 5 & 6:
+Both these points are valid and we should have indicated them explicitly.
+
+3. Figure 6: Clarification of the third case: The "upd" function merely
+determines if the "real" state of the system is as dictated by (s,pc) or if it
+should be modified to take into account memory responses. If the processor
+is not waiting for a memory response, then the state is exactly (s,pc). We made
+an optimization in the fourth case where we make the real state H whenever we
+know that the next instruction we decode is H; we could have it as (s,pc)
+instead.
+
+4. Figure 8: Tags for non-speculative loads: Our invariant is that there can be
+only one outstanding non-speculative load, though there can be several
+outstanding speculative loads concurrently. Anything without a tag is the
+current non-speculative load, and anything with a tag is a speculative load.
+
+========================
+
+4. Invariants are the hardest part in parametric verification of a cache
 coherence protocol, and we do not mention it:
@@
 
-4. Pipelining protocol engines and Bluespec:
-Though the semantics of the atomic transitions in our paper execute one at a
-time, different independent transitions can be made to occur in parallel in the
-actual implementation without changing the semantics; Bluespec compiler excels
-at this optimization.  One can take this further, where each atomic transition
-can be spread across multiple hardware clock cycles, while ensuring that the
-overall atomic transaction semantics are obeyed. Bluespec implementation
-currently does not include this optimization, though detailed algorithms to
-accomplish this have been published [Karczmarek, MEMOCODE'14].
-
-5. Nirav's point: I don't see that Reviewer-E's scathing concern fall
-into the core of what we're addressing; our focus on modularity is to
-allow verification to be (mostly) carried over to different
-microarchitectural variants, which is a bare minimum if one wants
-architects to have any lateral freedom to design a working
-implementation with a low-cost path to verification. I don't see why
-there's any fundamental reason that "scalable" approaches he's pointed to
-can't be fit into our methodology directly? I've only glanced at the
-papers, but it seems like it should be so.
---- Note sure how to condense it and incorporate -Murali
-
-Reviewer-F:
-1. Evidence of proofs being machine-checked:
-We have submitted the source code of our Coq development as a supplement to our
-paper in the submission website.
-
-2. Figure 3, Tags in Figures 5 & 6:
-Both these points are valid and we should have indicated them explicitly.
-
-3. Figure 6: The third case:
-Reviewer-F's assertion is not true. The "upd" function merely determines if the
-"real" state of the system is as dictated by the state (s,pc) or if it should
-be modified to take into account the responses from the memory system. If the
-processor is not waiting for the memory system response, then the state is
-exactly (s,pc). We made an optimization in the fourth case where we make the
-real state H whenever we know that the next instruction we decode is H; instead
-the state can be left as (s,pc) in the fourth case.
-
-4. Figure 8: Tags for non-speculative loads:
-Our invariant is that there can be only one outstanding non-speculative load,
-though there can be several outstanding speculative loads along with the
-outstanding non-speculative load. Thus, anything without a tag is the
-non-speculative load and there is only one of them, and anything with a tag is
-a speculative load.
+=====================================================================================================
+= Our Notes
+=====================================================================================================
 

### 2015-01-21 Rebuttal2.txt
Clean version

@@
-Reviewer-E
+One of the major benefits of our approach is that it unifies the modularization
+of the design and its verification task which enables verification to be done
+in tandem with mirco-architectural design (as noted by Reviewer-E). Further,
+this allows us to require only incremental verification effort for an
+incremental microarchitectural change, enabling scaling of the verification
+task.
 
-1. On Describing LTSes:
+Our approach vs. Model-Checking-based:
 
+a) Deriving Invariants:
+Finding invariants is fundamentally hard. The CMP-method (referred by
+Reviewer-E) uses symmetry to abstract the system, and then inserts invariants
+(called non-interference-lemmas) manually to eliminate counter-examples that
+violate the required property (the flow-based methodology [1] specifies these
+non-interference-lemmas in the form of properties of action sequences that
+takes place for any high-level transaction.) All these invariants are
+relatively simple but very subtle and are well understood by the designers. For
+example, for the FLASH protocol, one of the non-interference-lemma (Lemma-5 in
+[2]) says that whenever a cache is in E state, then the data it has is
+up-to-date. This is very similar to our Lemma-1. In general, we believe that
+the level of complexity of the lemmas in the two approaches are similar.
+
+The verification of the hierarchical protocol in [3] is restricted to 2-levels
+whereas our work applies to arbitrary levels of hierarchy. We are not sure if
+it's a fundamental restriction on the CMP-method, since the kind of symmetry
+that CMP-method exploits in a one or two-level system is not present in a
+hierarchical system.
+
+b) Case-Analysis:
+Much of the requisite protocol case analysis is done automatically via
+proof-script automation in Coq. In our experience, cases where proof-script
+automation fail required an additional invariant; we believe that manually
+providing a similiar invariant is necessary even in a model-checking approach.
+
+Reviewer-F:
+
+Figure-3, no tags in Figures-5,6:
+Both these points are valid and we should have indicated them explicitly.
+
+Figure-6, third case:
+The "upd" function merely determines if the "real" state of the system is as
+dictated by (s,pc) or if it should be modified to take into account memory
+responses. If the processor is not waiting for a memory response, then the
+state is exactly (s,pc). We made an optimization in the fourth case where we
+make the real state H whenever we know that the next instruction we decode is
+H; we could have it as (s,pc) instead.
+
+Figure-8: tags for non-speculative loads:
+Our invariant is that there can be only one outstanding non-speculative load,
+though there can be several outstanding speculative loads concurrently.
+Anything without a tag is the current non-speculative load, and anything with a
+tag is a speculative load.
+
+Reviewer-E:
+
+Pipelining protocol-engines:
+While atomic transitions in our paper have one-at-a-time semantics, the
+Bluespec compiler generates hardware which executes independant transitions
+concurrently in a clock cycle (as the semantics aren't violated); the compiler
+excels at this optimization. Atomic transition can also be spread across
+multiple clock cycles, while maintaining the semantics [4].
+
+Evidence of machine-checked proofs:
+We submitted our Coq source code as a supplement to our paper in the submission
+website.
+
+On Describing LTSes:
 We include the LTS description for the benefit of readers unfamiliar with it;
@@
 
-2. Modularity
-
- One of the major benefits of our approach is that it unifies the
- modularization of the design and its verification task which enables
- verification to be done in tandem with mircoarchitectural design as noted by
- Reviewer-E.  Further, this allows us to require only incremental verification
- effort for an incremental microarchitectural change, enabling scaling of the
- verification task.
-
-3. Model Checking vs. Theorem Proving
-
-a. Finding invariants is fundamentally hard. In [PAPER],
-
-Finding invariants is fundamentally hard. We are equivalent in complexity but
-  because of our modularization we have a free source of abstraction from the
-  designer.
-
-   - In comparision, Quote paper. Invariants were dervied manually, took one
-  month, and relied on a manual abstraction of the system space based on
-  symettry. Our work is similar save that we do not require an abstraction.
-
-  In comparision to the <PAPER>, our invariants of similar complexity. In
-  general, we believe that such invariants relatively simple but subtle and
-  generally intuitively understood by the designers.
-
-b. On Case-Analysis:
-
-Much of the requisite protocol case analysis is done automatically via
-proof-script automation in Coq. In our experience, cases where proof-script
-automation fail required an additional invariant; we strongly believe that
-manually providing a similiar invariant are necessary in a model checking
-approach.
-
-c. Benefits of TP-based parameterization: It should also be noted that their
-  work while hierarchical is only 2-level while ours applies to arbitrary
-  levels of hierarchy.
-
-4. Pipelining protocol engines and Bluespec:
-
-While atomic transitions in our paper have one-at-a-time semantics, the
-Bluespec compiler generates hardware which executes independant transitions
-concurrently in a clock cycle. The Bluespec compiler excels at this
-optimization. Atomic transition can also be spread across multiple clock
-cycles, while maintaining atomic transaction semantics [Karczmarek,
-MEMOCODE'14].
-
-Reviewer-F:
-
-1. Evidence of proofs being machine-checked:
-
-We submitted the source code of our Coq development as a supplement to our
-paper in the submission website.
-
-2. Figure 3, tags in Figures 5 & 6:
-Both these points are valid and we should have indicated them explicitly.
-
-3. Figure 6: Clarification of the third case: The "upd" function merely
-determines if the "real" state of the system is as dictated by (s,pc) or if it
-should be modified to take into account memory responses. If the processor
-is not waiting for a memory response, then the state is exactly (s,pc). We made
-an optimization in the fourth case where we make the real state H whenever we
-know that the next instruction we decode is H; we could have it as (s,pc)
-instead.
-
-4. Figure 8: Tags for non-speculative loads: Our invariant is that there can be
-only one outstanding non-speculative load, though there can be several
-outstanding speculative loads concurrently. Anything without a tag is the
-current non-speculative load, and anything with a tag is a speculative load.
-
-========================
-
-4. Invariants are the hardest part in parametric verification of a cache
-coherence protocol, and we do not mention it:
-We do mention the high level theorems that we use in the paper.  While we
-restrict ourselves to the verification of the MSI protocol's implementation,
-the proof is modular, and a large portion of the proof can be used, as is, for
-verification of other protocols (like MESI, MOSI and MOESI), as we discuss in
-the previous rebuttal. The theorem prover helps construct and prove the high
-level theorems that make these protocols correct. These theorems depend on a
-few basic properties that can be established easily, but were not mentioned in
-the paper due to lack of space. These properties include a) state changes occur
-when and only when response messages are sent, b) the absence of response
-messages from the parent to a child and from the same child to the parent
-simultaneously in the network, for the same address and c) FIFO properties of
-the responses. These basic properties seem to be the foundation over which the
-other cache coherence protocols are built -- they eliminate the corner cases,
-simplifying the proof. For example, since responses from a child to its parent
-and from the parent to the same child for the same address cannot be present
-simultaneously, an upgrade response from the parent cannot cross a downgrade
-response from the child, which would have eventually led to the upgrade of the
-child and downgrade of the parent, violating Lemma 3. As pointed out by
-Reviewer-E, the theorems and these properties serve as a method to derive a
-correct protocol (the different protocols are different system instances which
-obey these properties).
-
-=====================================================================================================
-= Our Notes
-=====================================================================================================
-
-Reviewer-E's related work summary:
-
-FM methodology: The methodology used for proving that the memory system
-realizes cache coherence is not new. A theorem-proving approach to verifying
-memory systems has been tacit in many publications. Researchers preferred model
-checking, as opposed to pushing through theorem-proving based proofs, largely
-because of the daunting complexity of realistic memory system protocols. One
-can find several approaches that address specific aspects of memory system
-verification in, say, Wim Hesselink's work :
-http://www.informatik.uni-trier.de/~ley/pers/hd/h/Hesselink:Wim_H=.
-
-@@@ Murali response
-Lots of PVS-based proofs for concurrent algorithms, mutual exclusion, etc...
-(Needs to be read further, but it's just a bunch of mechanized proofs for
-a lot of concurrent algorithms)
-***
-
-There has of course been recent trends to simplify protocols thru the DeNoVo
-project:
-http://scholar.google.com/citations?view_op=view_citation&hl=en&user=ufGYoL0AAAAJ&citation_for_view=ufGYoL0AAAAJ:8k81kl-MbHgC
-The authors have not cited such related pieces of work and justified the
-simplicity of their design.
-
-@@@ Murali response
-A new non-conventional cache coherence protocol that produces sequentially
-consistent results for data-rate free programs. There's no formal proof that
-this protocol is correct...
-(Needs to be read further, but doesn't seem relevant)
-***
-
-The authors claim that they are achieving their proof for an arbitrary number
-of processors. They do not mention anything about the invariants involved in
-their parametric proofs. This leads me to suspect that these invariants are
-pretty straightforward.  If one harks back to work addressing real protocols,
-the Chou, Mannava, Park work
-(http://scholar.google.com/citations?view_op=view_citation&hl=en&user=ufGYoL0AAAAJ&citation_for_view=ufGYoL0AAAAJ:qjMakFHDy7sC)
-shows how exceedingly hard it is to discover these invariants. In fact, the CMP
-approach is an incremental CEGAR-style method for discovering these invariants.
-Stating these invariants explicitly would be virtually impossible - even for
-FLASH-like protocols.  In fact, a whole series of papers have tried to address
-invariant discovery in this space : http://dl.acm.org/citation.cfm?id=1517434
-
-@@@ Murali response
-Chou, Mannava, Park [CMP-method] approach uses three paradigms: a) symmetry of
-cache coherence protocols b) abstraction of several nodes which behaves as just
-another node and c) counter-example guided abtraction refinement (CEGAR) for
-refining the abstraction's properties. Using CEGAR, they get invariants for the
-rest of the system while concentrating on a finite number of nodes (2).  Each
-refinement will not restrict the behavior, but will help towards proving a
-safety property. Specifically, let P be an invariant for the abstract
-transition system, and we should prove that it satisfies safety property X. We
-start out with a P that doesn't satisfy X. Then we add constraints on P, while
-permitting the same transitions. Eventually P will satisfy X. The manual
-portion of this procedure is in strengthening P.
-***
-
-@@@ Murali response
-The second paper tries to formulate better invariants in the first place using
-the notion of "flows". It says what other messages can interfere when the other
-message is being processed...
-(Needs to be read further -- I don't understand "flows" clearly; can't say
-anything more before I understand that properly)
-***
-
-For hierarchical protocols, this work has been extended in e.g.,
-http://link.springer.com/article/10.1007%2Fs10703-010-0092-y
-
-@@@ Murali response
-Same approach as [CMP-method] to hierarchical protocols. It models only 2 level hierarchies!
-***
+References:
+[1] Parameterized Verification using Message Flows: An Industrial Experience,
+John O'Leary, Murali Talupur, Mark R. Tuttle, FMCAD'09
+[2] A Simple Method for Parameterized Verification of Cache Coherence Protocols,
+Ching-Tsun Chou, Phanindra K. Mannava, Seungjoon Park, FMCAD'04
+[3] Efficient methods for formally verifying safety properties of hierarchical
+cache coherence protocol, Xiaofang Chen, Yu Yang, Ganesh Gopalakrishnan,
+Ching-Tsun Chou, FMSD'10
+[4] A new synthesis procedure for atomic rules containing multi-cycle function
+blocks, Michal Karczmarek, Arvind, Muralidaran Vijayaraghavan, MEMOCODE'14

### 2015-01-21 Rebuttal2.txt
Clean version
stuff

@@
 violate the required property (the flow-based methodology [1] specifies these
-non-interference-lemmas in the form of properties of action sequences that
+non-interference-lemmas in the form of properties of traces that
 takes place for any high-level transaction.) All these invariants are
@@
 b) Case-Analysis:
-Much of the requisite protocol case analysis is done automatically via
-proof-script automation in Coq. In our experience, cases where proof-script
-automation fail required an additional invariant; we believe that manually
-providing a similiar invariant is necessary even in a model-checking approach.
+Much of the requisite case-analysis is done automatically via proof-script
+automation in Coq. In our experience, cases where proof-script automation fail
+required an additional invariant; we believe that manually providing a similiar
+invariant is necessary even in a model-checking approach.
 

### 2015-01-21 Rebuttal2.txt
stuff

@@
 
+Beyond 500 words:
 On Describing LTSes:

### 2015-01-21 Rebuttal2.txt
Arvind's changes

--- 2015-01-21 Rebuttal2.txt
@@
 a) Deriving Invariants:
-Finding invariants is fundamentally hard. The CMP-method (referred by
-Reviewer-E) uses symmetry to abstract the system, and then inserts invariants
-(called non-interference-lemmas) manually to eliminate counter-examples that
-violate the required property (the flow-based methodology [1] specifies these
-non-interference-lemmas in the form of properties of traces that
-takes place for any high-level transaction.) All these invariants are
-relatively simple but very subtle and are well understood by the designers. For
-example, for the FLASH protocol, one of the non-interference-lemma (Lemma-5 in
-[2]) says that whenever a cache is in E state, then the data it has is
-up-to-date. This is very similar to our Lemma-1. In general, we believe that
-the level of complexity of the lemmas in the two approaches are similar.
+Finding invariants is fundamentally hard. The CMP-method referred by Reviewer-E
+uses symmetry to abstract the system, and then inserts invariants, called
+non-interference-lemmas, manually to eliminate counter-examples that violate
+the required property (the flow-based methodology [1] specifies invariants in
+the form of properties of traces that take place for any high-level
+transaction.) All these invariants are relatively simple but very subtle in
+terms of their efficacy. For example, for the FLASH protocol, one of the
+non-interference-lemma (Lemma-5 in [2]) says that whenever a cache is in E
+state, then the data it has is up-to-date. This is very similar to our Lemma-1.
+In general, we believe that the level of complexity of the lemmas in the two
+approaches are similar. The invariants are intuitively understood at the time
+of design of each module, making the task of formalizing them during the
+module's verification manageable.
 
@@
 it's a fundamental restriction on the CMP-method, since the kind of symmetry
-that CMP-method exploits in a one or two-level system is not present in a
+that CMP-method exploits in a one/two-level system is not present in a
 hierarchical system.
@@
 
+Beyond 500 words:
 Evidence of machine-checked proofs:

### 2015-01-21 Rebuttal2.txt
last set of changes

--- 2015-01-21 Rebuttal2.txt
@@
 of the design and its verification task which enables verification to be done
-in tandem with mirco-architectural design (as noted by Reviewer-E). Further,
+in tandem with micro-architectural design (as noted by Reviewer-E). Further,
 this allows us to require only incremental verification effort for an
-incremental microarchitectural change, enabling scaling of the verification
+incremental micro-architectural change, enabling scaling of the verification
 task.
@@
 a) Deriving Invariants:
-Finding invariants is fundamentally hard. The CMP-method (referred by
-Reviewer-E) uses symmetry to abstract the system, and then inserts invariants
-(called non-interference-lemmas) manually to eliminate counter-examples that
-violate the required property (the flow-based methodology [1] specifies these
-non-interference-lemmas in the form of properties of traces that
-takes place for any high-level transaction.) All these invariants are
-relatively simple but very subtle and are well understood by the designers. For
-example, for the FLASH protocol, one of the non-interference-lemma (Lemma-5 in
-[2]) says that whenever a cache is in E state, then the data it has is
-up-to-date. This is very similar to our Lemma-1. In general, we believe that
-the level of complexity of the lemmas in the two approaches are similar.
+Finding invariants is fundamentally hard. The CMP-method (Reviewer-E) uses
+symmetry to abstract the system, and then inserts "non-interference-lemmas"
+manually to eliminate counter-examples that violate the required property.
+Flow-based methodology [1] gives yet another way of manually specifying
+invariants. In general, we believe that the level of complexity of the lemmas
+in the two approaches are similar.  For example, for the FLASH protocol, one of
+the non-interference-lemma (Lemma-5 in [2]) says that whenever a cache is in
+E-state, then its data is up-to-date which is similar to our Lemma-1. The
+invariants are subtle but are intuitively understood at the time of design of
+each module, making the task of formalizing them during the module's
+verification straightforward.
 
@@
 automation in Coq. In our experience, cases where proof-script automation fail
-required an additional invariant; we believe that manually providing a similiar
+required an additional invariant; we believe that manually providing a similar
 invariant is necessary even in a model-checking approach.
@@
 
-Figure-6, third case:
+Figure-6, third-case:
 The "upd" function merely determines if the "real" state of the system is as
@@
 Figure-8: tags for non-speculative loads:
-Our invariant is that there can be only one outstanding non-speculative load,
-though there can be several outstanding speculative loads concurrently.
-Anything without a tag is the current non-speculative load, and anything with a
-tag is a speculative load.
+Our invariant is that there can be only one outstanding non-speculative load
+(ie. loads without tags), though there can be several outstanding speculative
+loads (ie. loads with tags) concurrently.
 
@@
 While atomic transitions in our paper have one-at-a-time semantics, the
-Bluespec compiler generates hardware which executes independant transitions
+Bluespec compiler generates hardware which executes independent transitions
 concurrently in a clock cycle (as the semantics aren't violated); the compiler
@@
 
+Beyond 500 words:
 References:

### 2015-01-21 Rebuttal2.txt
last set of changes
A light pass over PLDI response #2

@@
 One of the major benefits of our approach is that it unifies the modularization
-of the design and its verification task which enables verification to be done
+of the design and its verification task, which enables verification to be done
 in tandem with micro-architectural design (as noted by Reviewer-E). Further,
@@
 task.
+
+Reviewer-E:
 
@@
 invariants. In general, we believe that the level of complexity of the lemmas
-in the two approaches are similar.  For example, for the FLASH protocol, one of
-the non-interference-lemma (Lemma-5 in [2]) says that whenever a cache is in
-E-state, then its data is up-to-date which is similar to our Lemma-1. The
+between those approaches and ours are similar.  For example, for the FLASH protocol, one of
+the non-interference-lemmas (Lemma-5 in [2]) says that whenever a cache is in
+E-state, then its data is up-to-date, which is similar to our Lemma-1. The
 invariants are subtle but are intuitively understood at the time of design of
@@
 Much of the requisite case-analysis is done automatically via proof-script
-automation in Coq. In our experience, cases where proof-script automation fail
+automation in Coq. In our experience, cases where proof-script automation fails
 required an additional invariant; we believe that manually providing a similar
@@
 
+Pipelining protocol-engines:
+While atomic transitions in our paper have one-at-a-time semantics, the
+Bluespec compiler generates hardware which executes independent transitions
+concurrently in a clock cycle (as the semantics aren't violated); the compiler
+excels at this optimization. Atomic transitions can also be spread across
+multiple clock cycles, while maintaining the semantics [4].
+
+On Describing LTSes:
+We include the LTS description for the benefit of readers unfamiliar with it;
+the point wasn't to present novel concepts.
+
 Reviewer-F:
+
+Evidence of machine-checked proofs:
+We submitted our Coq source code as a supplement to our paper in the submission
+website.
 
@@
 Our invariant is that there can be only one outstanding non-speculative load
-(ie. loads without tags), though there can be several outstanding speculative
-loads (ie. loads with tags) concurrently.
-
-Reviewer-E:
-
-Pipelining protocol-engines:
-While atomic transitions in our paper have one-at-a-time semantics, the
-Bluespec compiler generates hardware which executes independent transitions
-concurrently in a clock cycle (as the semantics aren't violated); the compiler
-excels at this optimization. Atomic transition can also be spread across
-multiple clock cycles, while maintaining the semantics [4].
-
-Evidence of machine-checked proofs:
-We submitted our Coq source code as a supplement to our paper in the submission
-website.
-
-On Describing LTSes:
-We include the LTS description for the benefit of readers unfamiliar with it;
-the point wasn't to present novel concepts.
+(i.e. loads without tags), though there can be several outstanding speculative
+loads (i.e. loads with tags) concurrently.
 

### 2015-01-29 Rebuttal2.txt

@@
 a) Deriving Invariants:
-Finding invariants is fundamentally hard. The CMP-method (Reviewer-E) uses
-symmetry to abstract the system, and then inserts "non-interference-lemmas"
-manually to eliminate counter-examples that violate the required property.
-Flow-based methodology [1] gives yet another way of manually specifying
-invariants. In general, we believe that the level of complexity of the lemmas
-between those approaches and ours are similar.  For example, for the FLASH protocol, one of
-the non-interference-lemmas (Lemma-5 in [2]) says that whenever a cache is in
+Finding invariants is fundamentally hard. The CMP-method [1] uses symmetry to
+abstract the system, and then inserts "non-interference-lemmas" manually to
+eliminate counter-examples that violate the required property.  Flow-based
+methodology [2] gives yet another way of manually specifying invariants. In
+general, we believe that the level of complexity of the lemmas between those
+approaches and ours are similar. For example, for the FLASH protocol, one of
+the non-interference-lemmas (Lemma-5 in [3]) says that whenever a cache is in
 E-state, then its data is up-to-date, which is similar to our Lemma-1. The
@@
 
-The verification of the hierarchical protocol in [3] is restricted to 2-levels
+The verification of the hierarchical protocol in [4] is restricted to 2-levels
 whereas our work applies to arbitrary levels of hierarchy. We are not sure if
@@
 excels at this optimization. Atomic transitions can also be spread across
-multiple clock cycles, while maintaining the semantics [4].
+multiple clock cycles, while maintaining the semantics [5].
 
@@
 Figure-8: tags for non-speculative loads:
-Our invariant is that there can be only one outstanding non-speculative load
+Our invariant ensures there can be only one outstanding non-speculative load
 (i.e. loads without tags), though there can be several outstanding speculative
@@
 References:
-[1] Parameterized Verification using Message Flows: An Industrial Experience,
+[1] A simple method for parameterized verification of cache coherence protocols,
+Ching-Tsun Chou, Phanindra K Mannava, Seungjoon Park, FMCAD'04
+[2] Parameterized Verification using Message Flows: An Industrial Experience,
 John O'Leary, Murali Talupur, Mark R. Tuttle, FMCAD'09
-[2] A Simple Method for Parameterized Verification of Cache Coherence Protocols,
+[3] A Simple Method for Parameterized Verification of Cache Coherence Protocols,
 Ching-Tsun Chou, Phanindra K. Mannava, Seungjoon Park, FMCAD'04
-[3] Efficient methods for formally verifying safety properties of hierarchical
+[4] Efficient methods for formally verifying safety properties of hierarchical
 cache coherence protocol, Xiaofang Chen, Yu Yang, Ganesh Gopalakrishnan,
 Ching-Tsun Chou, FMSD'10
-[4] A new synthesis procedure for atomic rules containing multi-cycle function
+[5] A new synthesis procedure for atomic rules containing multi-cycle function
 blocks, Michal Karczmarek, Arvind, Muralidaran Vijayaraghavan, MEMOCODE'14

### 2016-03-28 Rebuttal2Murali.txt
stuff

1. Why waste time describing LTSes:
We introduced LTSes mainly to define the notion of refinement that we are
using, as well as to define the operators to combine LTSes.

2. The user of a theorem-prover has to walk the "protocol space" through case
analysis:
The theorem-prover has to walk the protocol space through case analysis, but
the proof-script automation in Coq significantly reduces the burden. One can
write proof-scripts that performs a proof-search, i.e. tries to apply
previously proven basic lemmas on each of the cases and leaves the user with
very few actual proof obligations that are complex or unique so as to not be
discharged using automated proof-search.

3. Invariants are the hardest part in parametric verification of a cache
coherence protocol, and we do not mention it:
We do mention the high level theorems that we use in the paper.  While we
restrict ourselves to the verification of the MSI protocol's implementation,
the proof is modular, and a large portion of the proof can be used, as is, for
verification of other protocols (like MESI, MOSI and MOESI), as we discuss in
the previous rebuttal. The theorem prover helps construct and prove the high
level theorems that make these protocols correct. These theorems depend on a
few basic properties that can be established easily, but were not mentioned in
the paper due to lack of space. These properties include a) state changes
occuring when and only when response messages are sent, b) absence of response
messages from the parent to a child and from the same child to the parent
simultaneously in the network, for the same address and c) FIFO properties of
the messages. These basic properties seem to be the foundation over which the
other cache coherence protocols lie -- they eliminate the corner cases which
make the proof difficult.  For example, since responses from a child to its
parent and from the parent to the same child for the same address cannot be
present simultaneously, an upgrade response from the parent cannot cross a
downgrade response from the child, which would have eventually led to the
upgrade of the child and downgrade of the parent violating Lemma 3. As pointed
out by Reviewer-E, the theorems and these properties serve as a method to
derive a correct protocol (the different protocols are different system
instances which obey these properties).

4. Pipelining protocol engines and Bluespec:
Though the semantics of the atomic transitions in our paper execute one at a
time, different independent transitions can be made to occur in parallel in the
actual implementation without changing the semantics; Bluespec already does
this optimization. One can take this further, where each atomic transition can
be spread across multiple hardware clock cycles, while ensuring that the
overall atomic transaction semantics are obeyed. Bluespec currently does not
implement this, though there has been work in this area [Karczmarek et.al.].

5. Nirav's point: I don't see that Reviewer-E's scathing concern fall
into the core of what we're addressing; our focus on modularity is to
allow verification to be (mostly) carried over to different
microarchitectural variants, which is a bare minimum if one wants
architects to have any lateral freedom to design a working
implementation with a low-cost path to verification. I don't see why
there's any fundamental reason that "scalable" approaches he's pointed to
can't be fit into our methodology directly? I've only glanced at the
papers, but it seems like it should be so.
--- Note sure how to condense it and incorporate -Murali

Reviewer-F:
1. Evidence of proofs being machine-checked:
We have submitted the source code of our Coq development as a supplement to our
paper in the submission website.

2. Figure 3, Tags in Figures 5 & 6:
Both these points are valid and we could indicate these points explicitly.

3. Figure 6: The third case:
Reviewer-F's assertion is not true. The "upd" function merely determines if the
"real" state of the system is as dictated by the state (s,pc) or if it should
be modified to take into account the responses from the memory system. If the
processor is not waiting for the memory system response, then the state is
exactly (s,pc). We made an optimization in the fourth case where we make the
real state H whenever we know that the next instruction we decode is H; instead
the state can be left as (s,pc) in the fourth case.

4. Figure 8: Tags for non-speculative loads:
Our invariant is that there can be only one outstanding non-speculative load,
though there can be several outstanding speculative loads along with the
outstanding non-speculative load. Thus, anything without a tag is the
non-speculative load and there is only one of them, and anything with a tag is
a speculative load.

Reviewer-E's related work summary:

FM methodology: The methodology used for proving that the memory system
realizes cache coherence is not new. A theorem-proving approach to verifying
memory systems has been tacit in many publications. Researchers preferred model
checking, as opposed to pushing through theorem-proving based proofs largely
because of the daunting complexity of realistic memory system protocols. One
can find several approaches that address specific aspects of memory system
verification in, say, Wim Hesselink's work :
http://www.informatik.uni-trier.de/~ley/pers/hd/h/Hesselink:Wim_H=.

@@@ Murali response
Lots of PVS-based proofs for concurrent algorithms, mutual exclusion, etc...
(Needs to be read further, but it's just a bunch of mechanized proofs for
a lot of concurrent algorithms)
***

There has of course been recent trends to simplify protocols thru the DeNoVo
project:
http://scholar.google.com/citations?view_op=view_citation&hl=en&user=ufGYoL0AAAAJ&citation_for_view=ufGYoL0AAAAJ:8k81kl-MbHgC
The authors have not cited such related pieces of work and justified the
simplicity of their design.

@@@ Murali response
A new non-conventional cache coherence protocol that produces sequentially
consistent results for data-rate free programs. There's no formal proof that
this protocol is correct...
(Needs to be read further, but doesn't seem relevant)
***

The authors claim that they are achieving their proof for an arbitrary number
of processors. They do not mention anything about the invariants involved in
their parametric proofs. This leads me to suspect that these invariants are
pretty straightforward.  If one harks back to work addressing real protocols,
the Chou, Mannava, Park work
(http://scholar.google.com/citations?view_op=view_citation&hl=en&user=ufGYoL0AAAAJ&citation_for_view=ufGYoL0AAAAJ:qjMakFHDy7sC)
shows how exceedingly hard it is to discover these invariants. In fact, the CMP
approach is an incremental CEGAR-style method for discovering these invariants.
Stating these invariants explicitly would be virtually impossible - even for
FLASH-like protocols.  In fact, a whole series of papers have tried to address
invariant discovery in this space : http://dl.acm.org/citation.cfm?id=1517434

@@@ Murali response
Chou, Mannava, Park [CMP-method] approach uses three paradigms: a) symmetry of
cache coherence protocols b) abstraction of several nodes which behaves as just
another node and c) counter-example guided abtraction refinement (CEGAR) for
refining the abstraction's properties. Using CEGAR, they get invariants for the
rest of the system while concentrating on a finite number of nodes (2).  Each
refinement will not restrict the behavior, but will help towards proving a
safety property. Specifically, let P be an invariant for the abstract
transition system, and we should prove that it satisfies safety property X. We
start out with a P that doesn't satisfy X. Then we add constraints on P, while
permitting the same transitions. Eventually P will satisfy X. The manual
portion of this procedure is in strengthening P.
***

@@@ Murali response
The second paper tries to formulate better invariants in the first place using
the notion of "flows". It says what other messages can interfere when the other
message is being processed...
(Needs to be read further -- I don't understand "flows" clearly; can't say
anything more before I understand that properly)
***

For hierarchical protocols, this work has been extended in e.g.,
http://link.springer.com/article/10.1007%2Fs10703-010-0092-y

@@@ Murali response
Same approach as [CMP-method] to hierarchical protocols.
***
